### 2017-01-20 article/article_v0.01.tex
Article draft is now available

\documentclass[titlepage,11pt,twoside]{article}

\usepackage[dvips]{graphicx}

\usepackage[myheadings]{fullpage}
\usepackage{pmetrika}
\usepackage{pmbib}
\usepackage{color}
\usepackage{mathtools}
\usepackage{amssymb}

%\usepackage{submit}

\newcommand{\bfU}{\mbox{\boldmath$\mathsf{U}$}}
\newcommand{\bfu}{\mbox{\boldmath$\mathsf{u}$}}
\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
\newcommand{\RR}{\mathbb{R}}
\DeclareMathOperator*{\V}{V}
\newcommand{\argmax}{\text{argmax}}

%\begin{figure}[h]
%\centerline{\includegraphics{figure03.eps}}
%\caption{Projection of item discrimination vectors onto $V_{\theta_T}$ hyperplance for a six item three-dimensional approximate sample structure.}
%\end{figure}

%\raggedbottom
\flushbottom

%\firstpage{1}
%\setcounter{lastpage}{999}
\setcounter{secnumdepth}{3}

\begin{document}

\linespacing{1}

\title{Something that is not PCADSC}

\author{Anne H. Petersen, Bo Bang, Karl Markussen}

\affil{University of Copenhagen}

\vspace{\fill}

\linespacing{1}

%\RepeatTitle{Psychometrics: From Practice to Theory and Back}

\begin{center}\vskip3pt

\vspace{32pt}

Abstract\vskip3pt

\end{center}

\begin{abstract}
abstract...
\begin{keywords}
keywords... 
\end{keywords}
\end{abstract}

\vspace{\fill}\newpage

\section{Introduction}
\label{sec:introduction}
The origin stories of datasets have largely changed over time. While data was often gathered with a specific scientific inquiry in mind in the past, a lot of data is accumulated without such a specific purpose nowadays. This is both due to the vast amounts of data registrations online and due to a trend towards more open source research in a lot of different scientific disciplines. Especially the latter phenomenon poses new challenges with respect to data quality assessment. When data is collected and made public without a specific end-point in mind, how do we ensure that e.g. differences in measurement instruments do not cause the data to be effectively divided into subsets that are simply not comparable?

A lot of sophisticated methods are available for answering this type of question when we are willing to assume a statistical model of some sort. But when these models are taken away, a remarkable void of methods is left behind. What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables. We propose a new method for this task, namely Principal Component Analysis-based Data Structure Comparison (PCADSC). This method employs the principal component decomposition of the data matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences. \hl{Mention R package.}

This manuscript is structured as follows: First, in Section \ref{sec:stateoftheart}, we present the data structure comparison problem in more detail and discuss what statistical methods are already available for solving similar challenges. Next, in Section \ref{sec:pcadscintro}, we move on to a description of the PCADSC procedure, including a brief introduction to principal component analysis (PCA) in general. In Section \ref{sec:dataexample}, we present a worked data example using the open source, online available PISA data (\hl{ref}), which is an example of a dataset where multiple data collection methods \hl{Eller måske lande?} have been employed.

\section{Something about state of the art}
\label{sec:stateoftheart}
\subsection{\hl{More detailed description of the type of problem we wish to address}}
\hl{
\begin{itemize}
\item Two subsets of a dataset, i.e. to datasets with the same variables, but different observations
\item Wish to compare structures without specifying a model or even any variables of interest
\item The most central example is the question of whether the two subsets can readily be combined in a (unknown) data analysis, or if the subset-inducing variable actually implies heterogeneity across the subset division
	\begin{itemize}
		\item Examples: Large scale open source datasets such as the PISA data and ESS (European Social Survey) data and ...(?). In these datasets, the data producers are very far away from the majority of the data analysts. Therefore, problem-specific recommendations about potential instrument-induced challenges in the datasets are not available for the data analysts. How can data producers ensure that this will not be an issue, at least not related to known data gathering differences?
		\item Other examples?
		\item Perhaps a description of what happens if we are to combine the two subsets of the datasets without taking a e.g. an instrument-effect into account. When will it cause problems (maybe: causal graph style?)?
		\item Mention somewhere: We want a solution that is largely independent of the sizes of the two subsets of data. Thereby, a lot of methods that compare each subset to the full dataset in some sense are excluded.
	\end{itemize}	
\end{itemize}
}

\subsection{\hl{Describe existing methods used to solve similar questions or parts of the question we are addressing}}
\hl{
\begin{itemize}
\item The simplest case: variable-by-variable tests in distributional differences
	\begin{itemize}
		\item Simple, but scales poorly
		\item Only relates to marginal differences and not to the interplay between variables
	\end{itemize}
\item Karl's papers?
\item Anne's papers: IRT-based methods for surveys
	\begin{itemize}
		\item Moves beyond the marginal approach, but needs a model pre-specified
		\item Thus, it is not a general data structure comparison method, but rather a fitted-model comparison method. It addresses the interplay between the model and the data, not the data alone. This is fundamentally a different (though related) question.
	\end{itemize}
\end{itemize}
}

\section{PCADSC - description of the method}
\label{sec:pcadscintro}
\hl{Description from Anne's master's thesis. Rewrite.}

As mentioned above, the purpose of PCADSC is comparing overall data structures in two or more subsets of a dataset. But before we can get further into describing this procedure, we must first define what exactly is meant by "overall structure". One such definition is the structure of the covariance matrix of the dataset. If we assume all variables in the dataset to be jointly normal with zero means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables. This gives it a very nice interpretation as a measure of the overall structure. If we do not accept the normality assumption, pairwise correlations and variable variances are still interesting quantities that say something about the interrelations between the variables, at least in terms of linear relationships. All in all, the empirical covariance matrix is a reasonable place to start looking for differences in "overall data structures".

Though the idea sounds appealing, it is quite difficult to assess similarity of matrices, and moreover, this becomes increasinly difficult for large numbers of variables and thus high dimensional covariance matrices. There is simply too much information to consider at once.

However, by clever use of linear algebra, we can construct a decomposition of the covariance matrix that makes it easier to gain an overview of the data. We propose a new method based on principal component analysis that seems to be able to identify differences in datasets based on intuitive, visual inspections. We refer to this method as principal component analysis-based data structure comparison (PCADSC) and we present the procedure below. But first, we give a minimal introduction to principal component analysis in general with reference to Koch (2014).

\subsection{Principal component analysis - a very brief introduction}
Principal component analysis (PCA) allows us to decompose a data matrix consisting of $n$ observations of $d$ variables into $d$ orthogonal vectors of length $d$. This is in itself not very impressive nor useful. But PCA does not result in just any basis of the data matrix. The basis chosen using PCA has the very favorable property that each vector is chosen such that they maximize the accumulated explained variance. These vectors are usually denoted \textit{principal component scores} or simply \textit{principal components} and they are constructed as linear combinations of the variables in the dataset (or, equivalently, as projections of the data matrix). The coefficients of each variable in these projections are denoted \textit{loadings}. Let $S \in \RR^{d \times d}$ denote the empirical covariance matrix and let $X \in \RR^{n \times d}$ denote our observed data matrix. $d$ is the number of variables while $n$ is the number of observations. The deconstruction procedure then goes as follows:
\begin{enumerate}
\item The vector of loadings corresponding to the first principal component, $\eta_1 \in \RR^d$, is chosen such that the component explains as much variance as possible. More precisely, for any unit vector $u \in \RR^d$ we find that $\V(u^T X^T)$ is maximized when $u = \eta_1$. Let $\lambda_1 = \V(\eta_1^T X^T)$. 
\item The second vector of loadings is then chosen as $\eta_2 = \argmax_{u: \, u \perp \eta_1} \V(u^T X^T)$ and we again define $\lambda_2 := \V(\eta_2 X)$.
\item The third vector of loadings is $\eta_3 = \argmax_{u: \, u \perp \{\eta_1, \eta_2\}} \V(u^T X^T)$ and $\lambda_3 :=  \V(\eta_3^T X^T)$ \\
\vdots 
\item[\refstepcounter{enumi} $d$.] The $d$th vector of loadings is $\eta_d = \argmax_{u: \, u \perp \{eta_1, ..., \eta_{d-1}\}} \V(u^T X^T)$ and $\lambda_d = V(\eta_d^T X^T)$.
\end{enumerate}
The $\eta_i$s and $\lambda_i$s are determined uniquely (up to scaling) and they are given as the eigenvectors and eigenvalues of $S$, respectively. Moreover, we find that $X = \sum_{i=1}^d \eta_j \eta_j^T X$, so the procedure is really just a clever choice of rotation of the data, such that it is split up into vectors that are ordered by their variances, or more precisely, their variance contributions. We define the contribution to the total variance of the $k$th principal component as $\frac{\lambda_k}{tr(S)}$ and please note that, due to the properties of eigenvalue decomposition, $tr(S) = \sum_{i=1}^d \lambda_i$, which is the sum of the variances of the variables in the dataset.

 In this introduction, we have assumed full rank of the covariance matrix $S$ for simplicity, but it should be noted that this assumption is in no way needed for the results to hold.

\subsection{PCA-based data structure comparison}
Above, we promised a method for intuitive, visual inspection of data structure similarities, but as of now, all intuition might have been lost in technicalities. The main point we want to emphasize from PCA is that we can interpret the eigenvector/eigenvalue-pairs of the covariance matrix as tools for obtaining a different representation of the dataset. Specifically, we can interpret the PCA loadings (eigenvectors) as weights that determine the relative influence of each variable in each component. If two different data sets with the same variables, but different samples of observations, then have similar loading patterns, the two data sets will agree on which variables explain the most variance and also \textit{how} this variance is explained.  Looking at the PCA loadings and the cumulative explained variance thus provides a straightforward non-parametric graphical approach for assessing similarity between two datasets. \\

Our proposal of a PCADSC method consist of three steps. These steps should be performed separately for each of the two (or more) datasets that we wish to compare. Note that the datasets must have the same variables, but different sample sizes are allowed. The three steps are:
\begin{enumerate}
\item \hl{Standardize. Full data or subsets? And also: Something about what to do if not all variables are numerical.}
\item Compute the PCA loadings and the variance contributions of each principal component.
\item For each principal component, standardize the loadings, i.e. scale them such that they sum to one.
\item Produce a plot consisting of a bar for each principal component, decorated with the cumulative variance contribution corresponding to this component. The bar should be of length one and colored according to the variables loading the component.
\end{enumerate}
The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.  \hl{Refer to example/show plot.}

\section{Data example stuff}
\label{sec:dataexample}
\hl{
\begin{itemize}
\item PISA 
\end{itemize}
}

\section{Discussion}
\label{sec:discussion}
\hl{
\begin{itemize}
\item Generalizing the results to non-numeric variables?
\item Generalizing the results to covariance matrices that are not of full rank?
\item ?
\end{itemize}
}

\section{Concluding Remarks}
\label{sec:conclusion}

\vspace{\fill}\clearpage

\begin{thebibliography}

\bibitem Ackerman, T.A. (1992). \hl{EXAMPLE POST.} A didactic explanation of item bias, item impact, and item validity from a multidimensional perspective. \textit{Journal of Educational Measurement}, \textit{29}, 67--91.

\end{thebibliography}
\vspace{\fill}

%\vfill\eject
\end{document}

### 2017-01-20 R/hairplot.R
Jeg har lavet en ny funktion til visuel sammenligning af to loadingsmatricer. Den kan afprøves på to datasæt via koden
hairplot(data1,data2)

@@ -1,0 +1,32 @@
+# Hairplot
+hairplot <- function(data1,data2,standardize=TRUE,make.plot=TRUE,arrow.len=0.05) {
+  # stadardize
+  data1 <- MuMIn::stdize(as.matrix(data1),scale=standardize)
+  data2 <- MuMIn::stdize(as.matrix(data2),scale=standardize)
+  
+  # eigen decomposition
+  d <-  ncol(data1)
+  n1 <- nrow(data1)
+  n2 <- nrow(data2)
+  eigen1 <- eigen(1/n1*t(data1)%*%data1)
+  eigen2 <- eigen(1/n2*t(data2)%*%data2)
+  
+  # find angles
+  angles <- matrix(0,d,d)
+  for (i in 1:d) for (j in 1:d) angles[i,j] <- asin(max(0,min(1,abs(sum(eigen1$vectors[,i]*eigen2$vectors[,j])))))
+  
+  # make plot
+  if (make.plot) {
+    plot(c(0.5,d+0.5),c(0.5,d+0.5),type="n",xlab="dataset 1: PC's",ylab="dataset 2: PC's",axes=FALSE)
+    axis(1,1:d)
+    axis(2,1:d)
+    pi <- 2*asin(1)
+    for (i in 1:d) for (j in 1:d) {
+      len <- sqrt(eigen1$values[i]/eigen1$values[1]*eigen2$values[j]/eigen2$values[1])
+      arrows(i,j,i+len*cos(angles[i,j]/2),j+len*sin(angles[i,j]/2),length=arrow.len)
+    }
+  }
+  
+  # return
+  return(angles)
+}

### 2017-01-20 R/hairplot.R
Added @export. I guess more Roxygen stuff needs to be added later.

@@ -1,20 +1,21 @@
 # Hairplot
+#' @export
 hairplot <- function(data1,data2,standardize=TRUE,make.plot=TRUE,arrow.len=0.05) {
   # stadardize
   data1 <- MuMIn::stdize(as.matrix(data1),scale=standardize)
   data2 <- MuMIn::stdize(as.matrix(data2),scale=standardize)
-  
+
   # eigen decomposition
   d <-  ncol(data1)
   n1 <- nrow(data1)
   n2 <- nrow(data2)
   eigen1 <- eigen(1/n1*t(data1)%*%data1)
   eigen2 <- eigen(1/n2*t(data2)%*%data2)
-  
+
   # find angles
   angles <- matrix(0,d,d)
   for (i in 1:d) for (j in 1:d) angles[i,j] <- asin(max(0,min(1,abs(sum(eigen1$vectors[,i]*eigen2$vectors[,j])))))
-  
+
   # make plot
   if (make.plot) {
     plot(c(0.5,d+0.5),c(0.5,d+0.5),type="n",xlab="dataset 1: PC's",ylab="dataset 2: PC's",axes=FALSE)
@@ -26,7 +27,7 @@ hairplot <- function(data1,data2,standardize=TRUE,make.plot=TRUE,arrow.len=0.05)
       arrows(i,j,i+len*cos(angles[i,j]/2),j+len*sin(angles[i,j]/2),length=arrow.len)
     }
   }
-  
+
   # return
   return(angles)
 }

### 2017-01-23 R/hairplot.R
Added @examples to hairplot code

@@ -1,4 +1,16 @@
-# Hairplot
+#' Hairplot
+#'
+#' @examples
+#' #load iris data
+#'   data(iris)
+#'
+#' #Define grouping variable, grouping the observations by whether their species is
+#' #Setosa or not
+#'  iris$group <- "setosa"
+#'  iris$group[iris$Species != "setosa"] <- "non-setosa"
+#'
+#' #Make hairplot
+#'  hairplot(iris[iris$group == "setosa", 1:4], iris[iris$group == "non-setosa", 1:4])
 #' @export
 hairplot <- function(data1,data2,standardize=TRUE,make.plot=TRUE,arrow.len=0.05) {
   # stadardize
@@ -31,3 +43,5 @@ hairplot <- function(data1,data2,standardize=TRUE,make.plot=TRUE,arrow.len=0.05)
   # return
   return(angles)
 }
+
+

### 2017-01-25 article/article_v0.01.tex
Update article_v0.01.tex

@@
 \label{sec:introduction}
-The origin stories of datasets have largely changed over time.
-While data was often gathered with a specific scientific inquiry in mind in the past, a lot of data is accumulated without such a specific purpose nowadays.
-This is both due to the vast amounts of data registrations online and due to a trend towards more open source research in a lot of different scientific disciplines.
-Especially the latter phenomenon poses new challenges with respect to data quality assessment.
-When data is collected and made public without a specific end-point in mind, how do we ensure that e.g.
-differences in measurement instruments do not cause the data to be effectively divided into subsets that are simply not comparable?
+The origin stories of datasets have changed over time.
+While data, in the past, data was collected with a specific scientific inquiry in mind, today, a lot of data is accumulated without a specific purpose.
+This is due to vast amounts of data being registered online and due to a trend towards more open source research.
+The latter phenomenon, in particular poses new challenges wrt.\ data quality assessment.
+When data are collected and made public without a specific end-point in mind, how do we ensure that e.g.\ differences in measurement instruments do not cause the data to be effectively divided into subsets that are simply not comparable?
 
-A lot of sophisticated methods are available for answering this type of question when we are willing to assume a statistical model of some sort.
-But when these models are taken away, a remarkable void of methods is left behind.
+Sophisticated methods for addressing this question are available when we are willing to assume a statistical model, but when these models are taken away, a remarkable void of methods is left behind.
 What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.

### 2017-02-19 article/article_v0.01.tex
Minor language stuff

@@
 The origin stories of datasets have changed over time.
-While data, in the past, data was collected with a specific scientific inquiry in mind, today, a lot of data is accumulated without a specific purpose.
+While in the past, data was often collected with a specific scientific inquiry in mind, today, a lot of data is accumulated without such a specific purpose.
 This is due to vast amounts of data being registered online and due to a trend towards more open source research.
-The latter phenomenon, in particular poses new challenges wrt.\ data quality assessment.
+The latter phenomenon in particular poses new challenges wrt.
+data quality assessment.
 When data are collected and made public without a specific end-point in mind, how do we ensure that e.g.\ differences in measurement instruments do not cause the data to be effectively divided into subsets that are simply not comparable?

### 2017-02-24 article/article_v0.01.tex
Typo 'eta' til '\eta'

@@
 \vdots 
-\item[\refstepcounter{enumi} $d$.] The $d$th vector of loadings is $\eta_d = \argmax_{u: \, u \perp \{eta_1, ..., \eta_{d-1}\}} \V(u^T X^T)$ and $\lambda_d = V(\eta_d^T X^T)$.
+\item[\refstepcounter{enumi} $d$.] The $d$th vector of loadings is $\eta_d = \argmax_{u: \, u \perp \{\eta_1, ..., \eta_{d-1}\}} \V(u^T X^T)$ and $\lambda_d = V(\eta_d^T X^T)$.
 \end{enumerate}

### 2017-02-25 article/article_v0.01.tex
Ny beskrivelse af PCA

--- 2017-02-19 article/article_v0.01.tex
@@
 \subsection{Principal component analysis - a very brief introduction}
-Principal component analysis (PCA) allows us to decompose a data matrix consisting of $n$ observations of $d$ variables into $d$ orthogonal vectors of length $d$.
+
+{\color{red}
+
+\emph{Principal component analysis} (PCA) may be interpreted in terms of the \emph{rank-q-reconstruction error}.
+To describe this consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, and let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ be their average.
+Suppose that we want to describe each of the observations by $q$ numbers instead of the original $d$ numbers.
+The rank-q-reconstruction error is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
+\begin{equation*}
+\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
+\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
+\end{equation*} 
+PCA ensures the existence of the subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error.
+Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of the empirical covariance $S = \frac{1}{n} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$.
+Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ of $S$ in the columns.
+The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a sign whenever the eigenvalues are different.
+If some of the eigenvalue are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
+In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero.
+It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
+\begin{equation*}
+\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
+\end{equation*}
+and equals $\sum_{i=q+1}^d \lambda_i$.
+In the terminology of PCA the eigenvectors $\eta_j \in \RR^d$ are called the \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}.
+The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called the \emph{scores}.
+The $j$th loading can also found as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores
+\begin{align*}
+\eta_j &= \argmax_{u \in \RR^d\colon: u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
+\lambda_j &= \frac{1}{n} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2
+\end{align*}
+
+It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
+This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and associated variation components $\lambda_j$ describe the structure of the dataset simultaneously for all approximating dimension $q$.
+This implies that the loadings and variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension $q$.
+}
+
+%Principal component analysis (PCA) allows us to decompose a data matrix consisting of $n$ observations of $d$ variables into $d$ orthogonal vectors of length $d$.
 This is in itself not very impressive nor useful.
@@
 The deconstruction procedure then goes as follows:
-\begin{enumerate}
-\item The vector of loadings corresponding to the first principal component, $\eta_1 \in \RR^d$, is chosen such that the component explains as much variance as possible.
+%\begin{enumerate}
+%\item The vector of loadings corresponding to the first principal component, $\eta_1 \in \RR^d$, is chosen such that the component explains as much variance as possible.
 More precisely, for any unit vector $u \in \RR^d$ we find that $\V(u^T X^T)$ is maximized when $u = \eta_1$.
 Let $\lambda_1 = \V(\eta_1^T X^T)$.
-\item The second vector of loadings is then chosen as $\eta_2 = \argmax_{u: \, u \perp \eta_1} \V(u^T X^T)$ and we again define $\lambda_2 := \V(\eta_2 X)$.
-\item The third vector of loadings is $\eta_3 = \argmax_{u: \, u \perp \{\eta_1, \eta_2\}} \V(u^T X^T)$ and $\lambda_3 :=  \V(\eta_3^T X^T)$ \\
-\vdots 
-\item[\refstepcounter{enumi} $d$.] The $d$th vector of loadings is $\eta_d = \argmax_{u: \, u \perp \{eta_1, ..., \eta_{d-1}\}} \V(u^T X^T)$ and $\lambda_d = V(\eta_d^T X^T)$.
-\end{enumerate}
-The $\eta_i$s and $\lambda_i$s are determined uniquely (up to scaling) and they are given as the eigenvectors and eigenvalues of $S$, respectively.
+%\item The second vector of loadings is then chosen as $\eta_2 = \argmax_{u: \, u \perp \eta_1} \V(u^T X^T)$ and we again define $\lambda_2 := \V(\eta_2 X)$.
+%\item The third vector of loadings is $\eta_3 = \argmax_{u: \, u \perp \{\eta_1, \eta_2\}} \V(u^T X^T)$ and $\lambda_3 :=  \V(\eta_3^T X^T)$ \\
+%\vdots 
+%\item[\refstepcounter{enumi} $d$.] The $d$th vector of loadings is $\eta_d = \argmax_{u: \, u \perp \{\eta_1, ..., \eta_{d-1}\}} \V(u^T X^T)$ and $\lambda_d = V(\eta_d^T X^T)$.
+%\end{enumerate}
+%The $\eta_i$s and $\lambda_i$s are determined uniquely (up to scaling) and they are given as the eigenvectors and eigenvalues of $S$, respectively.
 Moreover, we find that $X = \sum_{i=1}^d \eta_j \eta_j^T X$, so the procedure is really just a clever choice of rotation of the data, such that it is split up into vectors that are ordered by their variances, or more precisely, their variance contributions.
 We define the contribution to the total variance of the $k$th principal component as $\frac{\lambda_k}{tr(S)}$ and please note that, due to the properties of eigenvalue decomposition, $tr(S) = \sum_{i=1}^d \lambda_i$, which is the sum of the variances of the variables in the dataset.
-
- In this introduction, we have assumed full rank of the covariance matrix $S$ for simplicity, but it should be noted that this assumption is in no way needed for the results to hold.
+%
+% In this introduction, we have assumed full rank of the covariance matrix $S$ for simplicity, but it should be noted that this assumption is in no way needed for the results to hold.
 

### 2017-02-25 R/hairplot.R
Implemented Wally plot for PCADSC + minor changes

@@ -1,5 +1,13 @@
 #' Hairplot
 #'
+#' @description ????
+#'
+#' @param data1 ???
+#' @param data2 ???
+#' @param standardize ???
+#' @param make.plot ????
+#' @param arrow.len ???
+#'
 #' @examples
 #' #load iris data
 #'   data(iris)
@@ -11,6 +19,8 @@
 #'
 #' #Make hairplot
 #'  hairplot(iris[iris$group == "setosa", 1:4], iris[iris$group == "non-setosa", 1:4])
+#'
+#' @importFrom graphics axis arrows
 #' @export
 hairplot <- function(data1,data2,standardize=TRUE,make.plot=TRUE,arrow.len=0.05) {
   # stadardize

### 2017-02-25 R/makePCADSC.R

@@ -49,6 +49,12 @@ makePCADSC <- function(data, splitBy, var=NULL, covCO=NULL) {
   #define var
   if (is.null(var)) var <- setdiff(names(data), splitBy)
 
+  #TO DO:
+  #-  check if any variables are essentially empty. This will cause an error
+  #   as we then divide by zero when standardizing
+  #-  Make sure it deals with factor splitBy varibales correctly
+
+
   #If data is tibble, data.table or matrix, convert it to data.frame
   #If data is neither, throw error
   if (any(class(data) %in% c("data.table", "tbl", "tbl_df", "matrix"))) {

### 2017-02-25 R/qplot.R

@@ -16,6 +16,9 @@ setGeneric("plot")
 #' If \code{NULL} (the default), the original levels of the splitting variable
 #' are used.
 #'
+#' @param hideCumVar If \code{TRUE} (the default), cummulated explained variance
+#' percentages are not shown for each component.
+#'
 #' @importFrom methods setMethod
 #' @importFrom ggplot2 qplot ggplot aes_string geom_bar coord_flip scale_x_reverse
 #'             scale_y_continuous geom_text xlab ylab theme facet_wrap
@@ -25,7 +28,8 @@ setGeneric("plot")
 #'@include PCADSC.R
 #'@export
 setMethod("qplot","PCADSC",
-          function(x, varLabels=NULL, covCO=NULL, splitLabels=NULL) {
+          function(x, varLabels=NULL, covCO=NULL, splitLabels=NULL,
+                   hideCumVar = FALSE) {
 
   pcadscObj <- x
   splitLevels <- pcadscObj@splitLevels
@@ -61,20 +65,25 @@ setMethod("qplot","PCADSC",
   #browser()
     #note: aes_string is used for devtools::check() to stop complaining
     #about undocumented variables. It is not necessary at all.
-  ggplot(pcaFrame, aes_string(x="comp", y="loading", fill="var")) +
+  p <- ggplot(pcaFrame, aes_string(x="comp", y="loading", fill="var")) +
     geom_bar(stat="identity") +
     coord_flip() +
     scale_x_reverse(breaks = 1:nComp) +
-    scale_y_continuous(limits=c(0, 1.4),
-                       breaks=c(0, 0.25, 0.5, 0.75, 1)) +
-    geom_text(aes_string(label="cvc"), y=1.2, cex=4, na.rm=T) +
     xlab("Principal component") +
     ylab("Standardized loading") +
-    theme(legend.position="bottom") +
     facet_wrap(~ group, ncol=2,
                labeller=as_labeller(facetLabels)) +
     scale_fill_discrete(name=NULL, labels=varLabels) +
-    theme_bw()
+    theme_bw() +
+    theme(legend.position="bottom")
+  if (hideCumVar) {
+    return(p + scale_y_continuous(breaks=c(0, 0.25, 0.5, 0.75, 1)))
+  } else {
+    p +
+      scale_y_continuous(limits=c(0, 1.4),
+                         breaks=c(0, 0.25, 0.5, 0.75, 1)) +
[…]

### 2017-02-25 R/wallyPlot.R

@@ -1,0 +1,40 @@
+#' @title Wally plot for a PCADSC object
+#'
+#' @description Creates a Wally plot for a \code{\link{PCADSC}} object,
+#'  i.e. a grid of plots where the grouping variable for the PCADSC procedure
+#'  has been replaced by random grouping in all but one (the top-left) plot.
+#'  Thereby, the plot can be used to access the variability of PCADSC results.
+#'
+#' @param PCADSCobj A PCADSC object, as created by \code{\link{makePCADSC}}.
+#' @param data The dataset used to generate the PCADSC object.
+#' @param nrow The number of rows of plots in the outputted Wally plot.
+#' @param ncol The number of columns of plo0ts in the outputted Wally plot.
+#' @param covCO A cummulated variance cut-off limit. This parameter controls how
+#' many components are illustrated in each of the PCADSC plots.
+#' @param hideCumVar If \code{TRUE} (the default), cummulated explained variance
+#' percentages are not shown for each component.
+#'
+#' @importFrom ggplot2 theme xlab ylab element_blank
+#' @export
+wallyPCADSC <- function(PCADSCobj, data, nrow, ncol, covCO = 1, hideCumVar = TRUE) {
+  vars <- PCADSCobj@varNames
+  nObs1 <- PCADSCobj@nObs1
+
+  nTotal <- nrow * ncol
+  plots <- list()
+  plots[[1]] <- qplot(PCADSCobj, covCO = covCO, hideCumVar = hideCumVar) +
+    theme(legend.position = "none") + xlab("") + ylab("") +
+    theme(axis.text = element_blank(), axis.ticks = element_blank())
+  n <- nrow(data)
+
+  for (i in 2:nTotal) {
+    grp1 <- sample(1:n, nObs1)
+    data$grp <- "Group 2"
+    data$grp[grp1] <- "Group 1"
+    plots[[i]] <- qplot(makePCADSC(data, splitBy = "grp", var = vars),
+                      covCO = covCO, hideCumVar = hideCumVar) +
+      theme(legend.position = "none") + xlab("") + ylab("") +
+      theme(axis.text = element_blank(), axis.ticks = element_blank())
+  }
+  do.call("grid.arrange", c(plots, ncol = ncol, nrow = nrow))
+}

### 2017-02-25 article/article_v0.01.tex

@@
 \label{sec:dataexample}
-\hl{
-\begin{itemize}
-\item PISA 
-\end{itemize}
-}
+We will now turn to a concrete data example in order to illustrate the possibilities of each of the methods presented above \hl{more than just PCADSC?}.
+We use data from the 2012 version of the European Social Survey (ESS) project, a very large dataset that is available online at \hl{link?}.
+As the name suggests, the data comes from a survey that was conducted primarily in Europe aiming to collect information about the social conditions of the citizens \hl{ref?}.
+As with all international (or simply multi-center) studies, one might be concerned about whether or not the data from different countries can readily be combined.
+This is the question we will address using PCADSC in the current section.
+
+All computations and figures presented in this section where created using our R package PCADSC, which is available at \hl{... github? CRAN?}.
+
+\subsection{Data}
+\begin{figure}
+\center
+\includegraphics[scale=1]{surveyTable.png}
+\caption{\hl{Make a new table here}}
+\label{surveyTable}
+\end{figure}
+The ESS 2012 data contains a total of 626 variables collected from 54673 citizens in 29 countries.
+Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
+These 35 items can be collected into 6 distinct scales (\hl{REF: ESS6 Topline Results Series 5}), as illustrated in Figure \ref{surveyTable}.
+We represent each of these scales by a single variable, which is calculated as the average score within the items related to that variable (\hl{ref to someone that says that is sensible?}).
+For simplicity, we use only complete cases for this construction.
+
+In the following, we only compare two countries, namely Denmark and Bulgaria, which have $n_{BG} = 1798$ and $n_{DK} = 1498$ complete cases in the variables of interest, respectively.
+For these two countries, the ESS authors highlight differences in the relationship between the psychological well-being scales on an nation-aggregated level (\hl{REF: ESS6 Topline Results Series 5}) which might reflect large, cultural differences.
+Simply put, we do not expect happiness and psychological well-being to be the same phenomena in Bulgaria and Denmark \hl{more here about why, about the cultures?}.
+Therefore, a successful method for data comparisons should be able to detect these differences.
+
+\subsection{PCADSC}
+\begin{figure}
+\center
+\includegraphics[scale = 0.8]{essPCADSC.pdf}
+\caption{\hl{something}}
+\label{plotESSPCADSC}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[scale = 0.8]{essWallyPCADSC.pdf}
+\caption{\hl{something}}
+\label{plotESSPCADSCWally}
+\end{figure}
+
+Figure \ref{plotESSPCADSC} illustrates the results of conducting PCADSC on the psychological well-being data from the ESS.
+While the first principal component, which is responsible for explaining 50-60 \% of the variance in the data, is very similar, we see quite large differences in the remaining components.
+In the second component, we see that the two countries disagree in the relative importance of the scales \textit{Community wellbeing} and \textit{Vitality}.
+In the third and fourth components, general disagreement  is found, thus implying that components representing almost half of the variability in the data are not very similar across the two countries.
+This implies that the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, just as the ESS authors suggested.
+But did we really illustrate a data structure difference due to country differences or did we just illustrate the variability of the results of the PCA method?
+In order to investigate this further, we refer to a so-called \textit{Wally plot} (\hl{ref: Claus Ekstrøm}).
+In this plot, we compare the results of PCADSC conducted with grouping by country with several random, but similar grouping variables.
+Specifically, we produce 7 PCADSC plots where the country variable was replaced by a randomly generated variable that divides the observations into two groups of the same sizes as the country samples.
+The results are illustrated in Figure \ref{plotESSPCADSCWally}.
+Here, we see that the differences in the second component are not matched in any of the randomly grouped PCADSC runs.
+In fact, the 7 runs are remarkably similar, thereby illustrating that PCADSC seems to be very robust with respect to random groupings: The signal in the data is not blurred by the random subdivisions.
+When it comes to the differences in the third component for the two groups, we find much larger variability in the 7 random runs.
+\hl{more comments here}...
 

### 2017-02-26 article/article_v0.01.tex
More work on data example stufff, final commit before meeting

@@
 \usepackage{amssymb}
+\usepackage{hyperref}
 
@@
 \newcommand{\argmax}{\text{argmax}}
+\newcommand{\R}[1]{\texttt{#1}}
 
@@
 We will now turn to a concrete data example in order to illustrate the possibilities of each of the methods presented above \hl{more than just PCADSC?}.
-We use data from the 2012 version of the European Social Survey (ESS) project, a very large dataset that is available online at \hl{link?}.
-As the name suggests, the data comes from a survey that was conducted primarily in Europe aiming to collect information about the social conditions of the citizens \hl{ref?}.
-As with all international (or simply multi-center) studies, one might be concerned about whether or not the data from different countries can readily be combined.
+We use data from the 2012 version of the European Social Survey (ESS) project, a very large dataset that is freely available online at \url{www.europeansocialsurvey.org}.
+As the name suggests, the data comes from a survey that was conducted primarily in Europe aiming to collect information about the social conditions of the citizens \hl{reference?}.
+As with all international (or, simply, multi-center) studies, one might be concerned about whether or not the data from different countries can readily be combined.
 This is the question we will address using PCADSC in the current section.
 
-All computations and figures presented in this section where created using our R package PCADSC, which is available at \hl{... github? CRAN?}.
+All computations and figures presented in this section where created using our \R{R} package \R{PCADSC}, which is available online at \url{www.github.com/AnnePetersen1/PCADSC} \hl{maybe do a CRAN submission instead?}.
 
@@
 Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
-These 35 items can be collected into 6 distinct scales (\hl{REF: ESS6 Topline Results Series 5}), as illustrated in Figure \ref{surveyTable}.
+These 35 items can be divided into 6 distinct scales (\hl{REF: ESS6 Topline Results Series 5}), as illustrated in Figure \ref{surveyTable}.
 We represent each of these scales by a single variable, which is calculated as the average score within the items related to that variable (\hl{ref to someone that says that is sensible?}).
-For simplicity, we use only complete cases for this construction.
+For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below.
 
 In the following, we only compare two countries, namely Denmark and Bulgaria, which have $n_{BG} = 1798$ and $n_{DK} = 1498$ complete cases in the variables of interest, respectively.
-For these two countries, the ESS authors highlight differences in the relationship between the psychological well-being scales on an nation-aggregated level (\hl{REF: ESS6 Topline Results Series 5}) which might reflect large, cultural differences.
+For these two countries, the ESS authors \hl{different term?} particularly highlight differences in the relationship between the psychological well-being scales, at least on an nation-aggregated level (\hl{REF: ESS6 Topline Results Series 5}).
+This might be the result of large, cultural and socio-economic differences between the two countries.
 Simply put, we do not expect happiness and psychological well-being to be the same phenomena in Bulgaria and Denmark \hl{more here about why, about the cultures?}.
-Therefore, a successful method for data comparisons should be able to detect these differences.
+Therefore, a successful method for data comparisons should be able to detect these differences by looking at the differences in the interplay between the 6 scales of psychological well-being.
 
@@
 Figure \ref{plotESSPCADSC} illustrates the results of conducting PCADSC on the psychological well-being data from the ESS.
-While the first principal component, which is responsible for explaining 50-60 \% of the variance in the data, is very similar, we see quite large differences in the remaining components.
+While the first principal component, which is responsible for explaining 50-60 \% of the variance in the data, is very similar for the two countries, we see quite large differences in the remaining components.
 In the second component, we see that the two countries disagree in the relative importance of the scales \textit{Community wellbeing} and \textit{Vitality}.
-In the third and fourth components, general disagreement  is found, thus implying that components representing almost half of the variability in the data are not very similar across the two countries.
-This implies that the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, just as the ESS authors suggested.
-But did we really illustrate a data structure difference due to country differences or did we just illustrate the variability of the results of the PCA method?
-In order to investigate this further, we refer to a so-called \textit{Wally plot} (\hl{ref: Claus Ekstrøm}).
+In the third and fourth components, general disagreement  is found.
+All in all, components 2-4, representing almost half of the variability in the data, are not very similar across the two countries.
+Moreover, the two subsets of the data also differ with respect to how much variance is explained by each component, and the difference is particularly big for the first component.
+This component has approximately 15 \%  more explanatory power in the Bulgarian subsample than it does in the Danish.
+
+In summary, we find that the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics.
+\\
+
+But did we really illustrate a data structure difference due to country differences or did we just illustrate the variability of the results of the PCADSC method?
+In order to investigate this further, we look at a so-called \textit{Wally plot} (\hl{ref: Claus Ekstrøm}).
 In this plot, we compare the results of PCADSC conducted with grouping by country with several random, but similar grouping variables.
@@
 The results are illustrated in Figure \ref{plotESSPCADSCWally}.
-Here, we see that the differences in the second component are not matched in any of the randomly grouped PCADSC runs.
+Here, we see that the differences in the second component from the original PCADSC results are not matched in any of the randomly grouped PCADSC runs.
 In fact, the 7 runs are remarkably similar, thereby illustrating that PCADSC seems to be very robust with respect to random groupings: The signal in the data is not blurred by the random subdivisions.
 When it comes to the differences in the third component for the two groups, we find much larger variability in the 7 random runs.
-\hl{more comments here}...
+\hl{more comments here...
+Wait until we are sure exactly what we think about the results and what other PCA-based methods, we will do before/after.
+Particularly, how do we deal with eigen value differences?}
 

### 2017-02-26 ESS/ess_notes.R

@@ -1,0 +1,95 @@
+#Load data
+data <- read.csv("P:/PCADSC/data/ESS/ESS6.csv",
+                 na.strings = c(".", ".c", ".d", ".b"))
+
+useVars <- c("cntry", "stflife", "happy", "fltsd", "fltdpr", "enjlf", "wrhpp", "fltanx",
+             "fltpcfl", "dclvlf", "lchshcp", "accdng", "tmimdng", "tmabdng",
+             "tmendng","dngval", "sedirlf", "optftr", "lotsgot", "pstvms", "flrms",
+             "wrbknrm", "deaimpp", "flteeff", "slprl", "cldgng", "enrglot", "ppltrst",
+             "pplfair", "pplhlp", "pplahlp", "flclpla", "inprdsc", "flapppl", "rehlppl",
+             "fltlnl")
+
+anData <- na.omit(data[data$cntry %in% c("DK", "BG"), useVars])
+anData$cntry <- as.character(anData$cntry)
+
+sMat <- list()
+sMat$var <- names(anData)[-1]
+sMat$scale <- c(rep("Evaluative wellbeing", 2),
+               rep("Emotional wellbeing", 6),
+               rep("Functioning", 14),
+               rep("Vitality", 4),
+               rep("Community wellbeing", 5),
+               rep("Supportive relationships", 4))
+
+
+#Evaluative wellbeing
+sapply(anData[, sMat$var[sMat$scale == "Evaluative wellbeing"]], range) #same number of levels
+anData$EvaluativeWellbeing <- rowMeans(anData[, sMat$var[sMat$scale == "Evaluative wellbeing"]])
+
+#Emotional wellbeing
+sapply(anData[, sMat$var[sMat$scale == "Emotional wellbeing"]], range) #same number of levels
+for (v in sMat$var[sMat$scale == "Emotional wellbeing"])  {
+  anData[, v] <- anData[, v] - 1
+}
+for (v in c("fltsd", "fltdpr", "fltanx")) {
+  anData[, v] <- 3 - anData[, v]
+}
+anData$EmotionalWellbeing <- rowMeans(anData[, sMat$var[sMat$scale == "Emotional wellbeing"]])
+
+#Functioning
+sapply(anData[, sMat$var[sMat$scale == "Functioning"]], range)
+for (v in setdiff(sMat$var[sMat$scale == "Functioning"],
+                  c("tmimdng", "tmabdng", "tmendng", "sedirlf", "deaimpp")))  {
[…]

### 2017-02-27 article/article_v0.01.tex
Suggestion for new title + correction of typos

@@
 
-\title{Something that is not PCADSC}
+%\title{Something that is not PCADSC}
+\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
 
@@
 \end{equation*} 
-PCA ensures the existence of the subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error.
-Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of the empirical covariance $S = \frac{1}{n} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$.
+PCA ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error.
+Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of the empirical covariance $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$.
 Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ of $S$ in the columns.
@@
 and equals $\sum_{i=q+1}^d \lambda_i$.
-In the terminology of PCA the eigenvectors $\eta_j \in \RR^d$ are called the \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}.
-The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called the \emph{scores}.
+In the terminology of PCA the eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}.
+The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}.
 The $j$th loading can also found as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores
 \begin{align*}
-\eta_j &= \argmax_{u \in \RR^d\colon: u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
-\lambda_j &= \frac{1}{n} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2
+\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
+\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2
 \end{align*}

### 2017-02-27 R/cumeigen.R
cumeigen() for cumulated eigenvalues has been added.
Presently I can't build the package. But perhaps the "commit" will solve this!?

@@ -1,0 +1,128 @@
+##Author: Bo Markussen##
+#' Cumulative eigenvalues
+#'
+#' @title Cumulative eigenvalue plot
+#'
+#' @description Compares eigenvalues by cumulative Bland-Altman type plots. Kolmogorov-Smirnov and Cramer-von Mises tests evaluated by permutation tests
+#'
+#' @param data A dataset, either a \code{data.frame} or a \code{matrix} with variables
+#' in columns and observations in rows. WHAT TO DO ABOUT DATA.TABLES AND ALSO TIBBLES. BOTH WILL BE
+#' PROBLEMATIC AS OF NOW.
+#'
+#' @param splitBy A grouping variable with two levels defining the two groups within the
+#' dataset whose data structures we wish to compare. If \code{splitBy} has more than
+#' two levels, only the first two levels are used.
+#'
+#' @param var The variable names in \code{data} to include in the PCADSC. If \code{NULL}
+#' (the default), all variables except for \code{splitBy} are used.
+#'
+#' @param make.plot Should a plot be generated?
+#'
+#' @param B Number of simulations
+#'
+#' @examples
+#' #load iris data
+#' data(iris)
+#'
+#' #Define grouping variable, grouping the observations by whether their species is
+#' #Setosa or not
+#' iris$group <- "setosa"
+#' iris$group[iris$Species != "setosa"] <- "non-setosa"
+#'
+#' #make a PCADSC object, splitting the data by "group"
+#' cumeigen(iris, "group",var=setdiff(names(iris), c("group", "Species")))
+#'
+#' @export
+# cumulated differences
+cumeigen <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE) {
+  #define var
+  if (is.null(var)) var <- setdiff(names(data), splitBy)
+
+  #TO DO:
+  #-  check if any variables are essentially empty. This will cause an error
+  #   as we then divide by zero when standardizing
+  #-  Make sure it deals with factor splitBy varibales correctly
+
+  #If data is tibble, data.table or matrix, convert it to matrix
+  #If data is neither, throw error
[…]

### 2017-02-27 supplementary_code/cumulated_eigenvalues.R

@@ -1,7 +1,29 @@
 ##Author: Bo Markussen##
-
+#' Cumulative eigenvalues
+#'
+#' @description Compares eigenvalues by cumulative Bland-Altman type plots. Kolmogorov-Smirnov and Cramer-von Mises tests evaluated by permutation tests
+#'
+#' @param data1 First dataset
+#' @param data2 Second dataset
+#' @param standardize Should variables be standardized?
+#' @param make.plot Should a plot be generated?
+#' @param B Number of simulations
+#'
+#' @examples
+#' #load iris data
+#'   data(iris)
+#'
+#' #Define grouping variable, grouping the observations by whether their species is
+#' #Setosa or not
+#'  iris$group <- "setosa"
+#'  iris$group[iris$Species != "setosa"] <- "non-setosa"
+#'
+#' #Make cumulative eigenvalues
+#'  cumeigen(iris[iris$group == "setosa", 1:4], iris[iris$group == "non-setosa", 1:4])
+#'
+#' @export
 # cumulated differences
-cumdiff <- function(data1,data2,standardize=TRUE,make.plot=FALSE) {
+cumeigen <- function(data1,data2,standardize=TRUE,make.plot=FALSE,B=1000) {
   # standardize
   data1 <- MuMIn::stdize(as.matrix(data1),scale=standardize)
   data2 <- MuMIn::stdize(as.matrix(data2),scale=standardize)
@@ -12,21 +34,22 @@ cumdiff <- function(data1,data2,standardize=TRUE,make.plot=FALSE) {
   n2 <- nrow(data2)
   n12 <- nrow(data12)
   d12 <- ncol(data12)
-  eigen1 <- eigen(1/n1*t(data1)%*%data1) #n1-1 here?
-  eigen2 <- eigen(1/n2*t(data2)%*%data2) #n2-1 here?
-  eigen12 <- eigen(1/n12*t(data12)%*%data12)
+  eigen1 <- eigen(1/(n1-1)*t(data1)%*%data1)
+  eigen2 <- eigen(1/(n2-1)*t(data2)%*%data2)
+  eigen12 <- eigen(1/(n12-1)*t(data12)%*%data12)
 
   # cumulated differences:
   # 1) start from smallest eigenvalues
[…]

### 2017-02-27 R/cumeigen.R
Rette 'collate' fejl i DESCRIPTION. Og et par typos.

@@ -1,6 +1,3 @@
-##Author: Bo Markussen##
-#' Cumulative eigenvalues
-#'
 #' @title Cumulative eigenvalue plot
 #'
 #' @description Compares eigenvalues by cumulative Bland-Altman type plots. Kolmogorov-Smirnov and Cramer-von Mises tests evaluated by permutation tests
@@ -32,9 +29,9 @@
 #' #make a PCADSC object, splitting the data by "group"
 #' cumeigen(iris, "group",var=setdiff(names(iris), c("group", "Species")))
 #'
+#' @importFrom graphics polygon matlines lines
 #' @export
-# cumulated differences
-cumeigen <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE) {
+cumeigen <- function(data,splitBy,var=NULL,B=1000,make.plot=FALSE) {
   #define var
   if (is.null(var)) var <- setdiff(names(data), splitBy)
 
@@ -99,7 +96,7 @@ cumeigen <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE) {
   for (i in 1:B) {
     ii <- sample(1:n,n1)
     data1 <- as.matrix(stddata(data[ii,var]))
-    data2 <- as.matrix(stddata(data[setdiff(1:n,ii),var]))
+    data2 <- as.matrix(stddata(data[-ii,var]))
     y1 <- eigen(1/(n1-1)*t(data1)%*%data1,only.values=TRUE)$values
     y2 <- eigen(1/(n2-1)*t(data2)%*%data2,only.values=TRUE)$values
     y.sim[,i]  <- c(0,cumsum(y1-y2))

### 2017-02-28 R/hairplot.R
Replaced std-function from MuMIn with our std-function from makePCADSC.R

@@ -20,12 +20,12 @@
 #' #Make hairplot
 #'  hairplot(iris[iris$group == "setosa", 1:4], iris[iris$group == "non-setosa", 1:4])
 #'
-#' @importFrom graphics axis arrows
+#' @importFrom graphics axis arrows plot
 #' @export
 hairplot <- function(data1,data2,standardize=TRUE,make.plot=TRUE,arrow.len=0.05) {
   # stadardize
-  data1 <- MuMIn::stdize(as.matrix(data1),scale=standardize)
-  data2 <- MuMIn::stdize(as.matrix(data2),scale=standardize)
+  data1 <- as.matrix(stdData(data1)) #stdize(as.matrix(data1),scale=standardize)
+  data2 <- as.matrix(stdData(data2)) #stdize(as.matrix(data2),scale=standardize)
 
   # eigen decomposition
   d <-  ncol(data1)

### 2017-02-28 R/cumeigen.R
Added missing @importFroms

@@ -29,7 +29,8 @@
 #' #make a PCADSC object, splitting the data by "group"
 #' cumeigen(iris, "group",var=setdiff(names(iris), c("group", "Species")))
 #'
-#' @importFrom graphics polygon matlines lines
+#' @importFrom graphics polygon matlines lines legend
+#' @importFrom stats quantile
 #' @export
 cumeigen <- function(data,splitBy,var=NULL,B=1000,make.plot=FALSE) {
   #define var

### 2017-02-28 article/article_v0.01.tex
Definition of hairplot has been changed. This has been described in the article, and implemented in the PCADSC package.
Beside this other editorial changes have been made.

@@
 \newcommand{\R}[1]{\texttt{#1}}
+\newcommand{\acos}{\text{arccos}}
 
@@
 %\title{Something that is not PCADSC}
-\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
+%\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
+\title{Explorative data structure comparison with application to \ldots}
 
@@
 
-Though the idea sounds appealing, it is quite difficult to assess similarity of matrices, and moreover, this becomes increasinly difficult for large numbers of variables and thus high dimensional covariance matrices.
+Though the idea sounds appealing, it is quite difficult to assess similarity of matrices, and moreover, this becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices.
 There is simply too much information to consider at once.
@@
 Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of the empirical covariance $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$.
-Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ of $S$ in the columns.
-The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a sign whenever the eigenvalues are different.
-If some of the eigenvalue are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
+Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ in the columns.
+The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different.
+If some of the eigenvalues are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
 In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero.
@@
 \end{equation*}
-and equals $\sum_{i=q+1}^d \lambda_i$.
+and equals $\sum_{j=q+1}^d \lambda_j$.
 In the terminology of PCA the eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}.
@@
 \eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
-\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2
+\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
 \end{align*}
@@
 Looking at the PCA loadings and the cumulative explained variance thus provides a straightforward non-parametric graphical approach for assessing similarity between two datasets.
-\\
+
+{\color{red}
+We propose the following three diagnostic plots:
+\begin{description}
+\item[CumVarAgreement:] Cumulated Bland-Altman plot for the variation decomposition.
+\item[Hairplot:] Plot of the loadings for one of the datasets in the coordinate system of the other dataset.
+\item[Pancakeplot:] This is the ``classical'' PCADSC.
+\end{description}
+
+Below we describe the details of these plots.
+Consider two different datasets in the same $d$ variables.
+Let $\lambda_{k1} \ge \dotsm \ge \lambda_{kd}$ and $\eta_{k1},\dotsc,\eta_{kd}$ be the variation components and the loadings for dataset $k=1,2$.
+
+\medskip
+
+\textbf{CumVarAgreement:} To be done.
+
+\bigskip
+
+\textbf{Hairplot:} Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
+The empirical covariance matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
+\begin{equation*}
+S_1 = \sum_{k=1}^d \lambda_{1j} \eta_{1k} \eta_{1k}^\top
+= \lambda_{\max} \sum_{k=1}^d 
+\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg) 
+\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top.
+\end{equation*}
+We propose to visualize this decomposition by plotting arrows with length $l_{kj}$ and angle $\theta_{kj}/2$ anticlockwise from the diagonal in the $j$th row and the $k$th column of a $d \times d$ grid display, where
+\begin{align*}
+l_{kj} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k} \eta_{2j}^\top \rvert, &
+\theta_{kj} &= \acos(\lvert \eta_{1k} \eta_{2j}^\top \rvert).
+\end{align*}
+Using the same grid display, but interchanging the first and the second dataset as well as the rows and the columns, and using the angles clockwise, similar arrows are plotted in a different color.
+
+In our implementation we have used blue for the first and the second set of arrows, respectively.
+In this plot the blue arrows in the $k$th column visualize the decomposition of the $k$th principal components for the first dataset in the coordinate system of the second dataset.
+The scaling of the length by $\lambda_{\max}$ is made so that the longest arrow has unit length, and the absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors.
+This indeterminacy also implies the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$.
+We can visualize both the decomposition of $S_1$ and $S_2$ in the same plot by divide the angles by two and using anticlockwise and clockwise shifts from the diagonal.
+Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset.
+
+If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as the off-diagonal arrows would have zero length.
+Differences in the variation components are visualized as differences in the lengths of the arrows, also in the diagonal.
+And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal as well as arrows of non vanishing length in the off-diagonal.
+
+\bigskip
+
+Remarks: Presently permutation tests based on the Kolmogorov-Smirnov and the Cram\'er-von Mises test statistics has been implemented for CumVarAgreement.
+This should also be possible for the Hairplot based on the off-diagonal components.
+For the Pancakeplot a Wallyplot has been implemented (however, in Claus' Wally plot the position of the ``observed dataset'' is also random).
+
+\bigskip
+
+}
 
@@
 \includegraphics[scale = 0.8]{essWallyPCADSC.pdf}
-\caption{\hl{something}}
+\caption{\hl{something.}}
 \label{plotESSPCADSCWally}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[scale = 0.8]{essHairplot.png}
+\caption{The blue arrows show the principal components of the first dataset decomposed in the coordinate system of the principal components of the second dataset. And the red arrows the reverse.}
+\label{plotESSHairplot}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[scale = 0.8]{essCumVarAgreement.png}
+\caption{\hl{something.}}
+\label{plotCumVarAgreement}
 \end{figure}
@@
 This component has approximately 15 \%  more explanatory power in the Bulgarian subsample than it does in the Danish.
+
+{\color{red} 
+Figure \ref{plotESSHairplot} shows the hairplot.
+The blue arrows visualize the decomposition of the principal components for the first dataset in the coordinate system of the second dataset.
+We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3.
+The red arrows visualize the decomposition of the principal components for the second dataset in the coordinate system of the first dataset.
+We see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3.
+For PC1, PC5 and PC6 the main difference is in the size of the variation component.
+}
 

### 2017-02-28 R/cumeigen.R

@@ -31,7 +31,7 @@
 #'
 #' @importFrom graphics polygon matlines lines
 #' @export
-cumeigen <- function(data,splitBy,var=NULL,B=1000,make.plot=FALSE) {
+cumeigen <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE) {
   #define var
   if (is.null(var)) var <- setdiff(names(data), splitBy)
 

### 2017-02-28 R/hairplot.R

@@ -1,38 +1,88 @@
-#' Hairplot
+#' @title Hairplot
 #'
-#' @description ????
+#' @description Compares eigenvalues and eigenvectors form two datasets. Kolmogorov-Smirnov and Cramer-von Mises tests evaluated by permutation tests might be implemented later.
 #'
-#' @param data1 ???
-#' @param data2 ???
-#' @param standardize ???
-#' @param make.plot ????
-#' @param arrow.len ???
+#' @param data A dataset, either a \code{data.frame} or a \code{matrix} with variables
+#' in columns and observations in rows. WHAT TO DO ABOUT DATA.TABLES AND ALSO TIBBLES. BOTH WILL BE
+#' PROBLEMATIC AS OF NOW.
+#'
+#' @param splitBy A grouping variable with two levels defining the two groups within the
+#' dataset whose data structures we wish to compare. If \code{splitBy} has more than
+#' two levels, only the first two levels are used.
+#'
+#' @param var The variable names in \code{data} to include in the PCADSC. If \code{NULL}
+#' (the default), all variables except for \code{splitBy} are used.
+#'
+#' @param make.plot Should a plot be generated?
+#'
+#' @param B Number of simulations
+#'
+#' @param arrow.len Length of arrow on plot
 #'
 #' @examples
 #' #load iris data
-#'   data(iris)
+#' data(iris)
 #'
 #' #Define grouping variable, grouping the observations by whether their species is
 #' #Setosa or not
-#'  iris$group <- "setosa"
-#'  iris$group[iris$Species != "setosa"] <- "non-setosa"
+#' iris$group <- "setosa"
+#' iris$group[iris$Species != "setosa"] <- "non-setosa"
 #'
-#' #Make hairplot
-#'  hairplot(iris[iris$group == "setosa", 1:4], iris[iris$group == "non-setosa", 1:4])
+#' #make a PCADSC object, splitting the data by "group"
+#' hairplot(iris, "group",var=setdiff(names(iris), c("group", "Species")))
 #'
 #' @importFrom graphics axis arrows
 #' @export
-hairplot <- function(data1,data2,standardize=TRUE,make.plot=TRUE,arrow.len=0.05) {
-  # stadardize
-  data1 <- MuMIn::stdize(as.matrix(data1),scale=standardize)
[…]

### 2017-02-28 ESS/ess_notes.R
Redid structure a bit + implemented new annotation option for pancake plots and parameter for choosing which components to plot

@@ -98,7 +98,7 @@ ggsave(plot = p2, file = "./article/essWallyPCADSC.pdf",
 
 
 #Do cumeigen stuff:
-PCADSC:::cumeigen(anData, "cntry",  c("EvaluativeWellbeing",
+cumeigen(anData, "cntry",  c("EvaluativeWellbeing",
                                      "EmotionalWellbeing",
                                      "Functioning",
                                      "Vitality",

### 2017-02-28 R/loadComp.R

@@ -30,17 +30,19 @@ loadComp <- function(data, vars = NULL, varCO=NULL) {
   pcvc <- pcvc/sum(pcvc)
   cpcvc <- cvc <- cumsum(pcvc)
   cvc <- paste(round(cvc*100, 2), "%")
+  varpct <- paste(round(pcvc*100, 2), "%")
 
   #combine
   pxx <- reshape2::melt(px)
-  pxx$cvc_raw <- pxx$cvc <- pxx$pcvc <- rep(NA, nrow(pxx))
+  pxx$varpct_raw <- pxx$cvc_raw <- pxx$cvc <- pxx$varpct <- rep(NA, nrow(pxx))
   for (i in 1:n) {
     thisComp <- which(pxx$Var2==i)
     pxx$cvc[thisComp[1]] <- cvc[i]
-    pxx$pcvc[thisComp] <- pcvc[i]
+    pxx$varpct[thisComp[1]] <- varpct[i]
     pxx$cvc_raw[thisComp] <- cpcvc[i]
+    pxx$varpct_raw[thisComp] <- pcvc[i]
   }
 
-  colnames(pxx) <- c("var", "comp", "loading", "pcvc", "cvc", "cvc_raw")
+  colnames(pxx) <- c("var", "comp", "loading", "varpct", "cvc", "cvc_raw", "varpct_raw")
   list(loadings=pxx, nObs=p$n.obs)
 }

### 2017-02-28 R/pancakePlot.R

@@ -1,0 +1,111 @@
+#' @title ??
+#'
+#' @description ??
+#' @param x ??
+#' @param varLabels A vector of character string labels for the variables used in
+#' \code{pcadscObj}. If non-null, these labels appear in the plot instead of the
+#' variable names. Note that they must be listed in the same order as the variables
+#' in \code{pcadscObj} and this order can be expected by calling
+#' \code{pcadscObj$varNames}. If \code{NULL} (the default), these variable
+#' names are used.
+#'
+#' @param covCO Not implemented, something like a variance contribution cut-off, maybe?
+#'
+#' @param splitLabels Labels for the two categories of the splitting variable used
+#' to create the PCADSC object, \code{pcadscObj}, given as a named list (see examples).
+#' These labels will appear in the headers of the two PCADSC plots.
+#' If \code{NULL} (the default), the original levels of the splitting variable
+#' are used.
+#'
+#' @param varAnnotation If \code{"cum"} (the default), cummulated explained variance
+#' percentages are annotated to the right of the bars for each component. If \code{"raw"},
+#' the non-cummulated percentages of explained variance are added instead. If \code{NULL},
+#' no annotation is added. Note that \code{"cum"} is only allowed if all components
+#' are plottet (see \code{useComp}).
+#'
+#' @param useComps A vector of integers with the numbers of the principal component
+#' that should be included in the plot.
+#'
+#' @importFrom methods setMethod
+#' @importFrom ggplot2 qplot ggplot aes_string geom_bar coord_flip scale_x_reverse
+#'             scale_y_continuous geom_text xlab ylab theme facet_wrap
+#'             as_labeller scale_fill_discrete theme_bw
+#'
+#'@export
+pancakePlot <- function(x, varLabels=NULL, covCO=NULL, splitLabels=NULL,
+                        varAnnotation = "cum", useComps = NULL) {
+  pcadscObj <- x
+  splitLevels <- pcadscObj@splitLevels
+  nCat1 <- pcadscObj@nObs1
+  nCat2 <- pcadscObj@nObs2
+  splitBy <- pcadscObj@splitBy
+  pcaFrame <- pcadscObj@pcaFrame
+
[…]

### 2017-02-28 R/print.R

@@ -15,8 +15,8 @@ setMethod("print", "PCADSC", function(x, ...) {
   splitLevels <- pcadscObj@splitLevels
   nObs1 <- pcadscObj@nObs1
   nObs2 <- pcadscObj@nObs2
-  cvc1 <- cumsum(unique(frame$pcvc[frame$group == splitLevels[1]]))
-  cvc2 <- cumsum(unique(frame$pcvc[frame$group == splitLevels[2]]))
+  cvc1 <- cumsum(unique(frame$varpct_raw[frame$group == splitLevels[1]]))
+  cvc2 <- cumsum(unique(frame$varpct_raw[frame$group == splitLevels[2]]))
   nImportant1 <- min(which(cvc1 > 0.95))
   nImportant2 <- min(which(cvc2 > 0.95))
   name1 <- paste("\"", splitLevels[1], "\"", sep="")

### 2017-02-28 R/qplot.R

@@ -1,3 +1,4 @@
+#'@importFrom ggplot2 qplot
 setGeneric("qplot")
 setGeneric("plot")
 
@@ -26,66 +27,9 @@ setGeneric("plot")
 #'
 #'@describeIn PCADSC Plot a PCADSC object
 #'@include PCADSC.R
+#'@include pancakePlot.R
 #'@export
-setMethod("qplot","PCADSC",
-          function(x, varLabels=NULL, covCO=NULL, splitLabels=NULL,
-                   hideCumVar = FALSE) {
-
-  pcadscObj <- x
-  splitLevels <- pcadscObj@splitLevels
-  nCat1 <- pcadscObj@nObs1
-  nCat2 <- pcadscObj@nObs2
-  splitBy <- pcadscObj@splitBy
-  pcaFrame <- pcadscObj@pcaFrame
-
-  if (is.null(covCO)) covCO <- 1
-  if (covCO != 1) {
-    maxUseComp <- min(pcaFrame$comp[pcaFrame$cvc_raw > covCO])
-    pcaFrame <- pcaFrame[pcaFrame$comp <= maxUseComp, ]
-  }
-  nComp <- max(pcaFrame$comp)
-
-  if (is.null(varLabels)) {
-    varLabels <- pcadscObj@varNames
-  }
-  if (is.null(splitLabels)) {
-    splitLabels <- splitLevels
-  } else {
-    sl1 <- splitLabels[[which(names(splitLabels)==splitLevels[1])]]
-    sl2 <- splitLabels[[which(names(splitLabels)==splitLevels[2])]]
-    splitLabels <- c(sl1, sl2)
-  }
-
-
-
-  facetLabels <- c(paste(splitLabels[1], ", n = ", nCat1, sep=""),
-                   paste(splitLabels[2], ", n = ", nCat2, sep=""))
-  names(facetLabels) <- splitLevels
-
-  #browser()
-    #note: aes_string is used for devtools::check() to stop complaining
-    #about undocumented variables. It is not necessary at all.
-  p <- ggplot(pcaFrame, aes_string(x="comp", y="loading", fill="var")) +
-    geom_bar(stat="identity") +
-    coord_flip() +
-    scale_x_reverse(breaks = 1:nComp) +
-    xlab("Principal component") +
-    ylab("Standardized loading") +
-    facet_wrap(~ group, ncol=2,
-               labeller=as_labeller(facetLabels)) +
[…]

### 2017-02-28 R/wallyPancake.R

@@ -11,18 +11,20 @@
 #' @param ncol The number of columns of plo0ts in the outputted Wally plot.
 #' @param covCO A cummulated variance cut-off limit. This parameter controls how
 #' many components are illustrated in each of the PCADSC plots.
-#' @param hideCumVar If \code{TRUE} (the default), cummulated explained variance
-#' percentages are not shown for each component.
+#' @param varAnnotation ...
+#' @param useComps ...
 #'
 #' @importFrom ggplot2 theme xlab ylab element_blank
 #' @export
-wallyPCADSC <- function(PCADSCobj, data, nrow, ncol, covCO = 1, hideCumVar = TRUE) {
+wallyPancake <- function(PCADSCobj, data, nrow, ncol, covCO = 1,
+                         varAnnotation = "cum", useComps = NULL) {
   vars <- PCADSCobj@varNames
   nObs1 <- PCADSCobj@nObs1
 
   nTotal <- nrow * ncol
   plots <- list()
-  plots[[1]] <- qplot(PCADSCobj, covCO = covCO, hideCumVar = hideCumVar) +
+  plots[[1]] <- qplot(PCADSCobj, covCO = covCO, varAnnotation = varAnnotation,
+                      useComps = useComps) +
     theme(legend.position = "none") + xlab("") + ylab("") +
     theme(axis.text = element_blank(), axis.ticks = element_blank())
   n <- nrow(data)
@@ -32,7 +34,8 @@ wallyPCADSC <- function(PCADSCobj, data, nrow, ncol, covCO = 1, hideCumVar = TRU
     data$grp <- "Group 2"
     data$grp[grp1] <- "Group 1"
     plots[[i]] <- qplot(makePCADSC(data, splitBy = "grp", var = vars),
-                      covCO = covCO, hideCumVar = hideCumVar) +
+                      covCO = covCO, varAnnotation = varAnnotation,
+                      useComps = useComps) +
       theme(legend.position = "none") + xlab("") + ylab("") +
       theme(axis.text = element_blank(), axis.ticks = element_blank())
   }

### 2017-02-28 ESS/ess_notes.R
Added hairplot

@@ -79,7 +79,7 @@ anData$SupportiveRelationships <- rowMeans(anData[, sMat$var[sMat$scale ==
 library(PCADSC)
 a <- makePCADSC(anData, "cntry",  c("EvaluativeWellbeing",
                                "EmotionalWellbeing",
-                               "Functioning",
+                               "Functioning,"
                                "Vitality",
                                "CommunityWellbeing",
                                "SupportiveRelationships"))
@@ -103,4 +103,20 @@ cumeigen(anData, "cntry",  c("EvaluativeWellbeing",
                                      "Functioning",
                                      "Vitality",
                                      "CommunityWellbeing",
-                                     "SupportiveRelationships"))
+                                     "SupportiveRelationships"), make.plot = T)
+#Do hairplot:
+hairplot(anData, "cntry",  c("EvaluativeWellbeing",
+                             "EmotionalWellbeing",
+                             "Functioning",
+                             "Vitality",
+                             "CommunityWellbeing",
+                             "SupportiveRelationships"))
+
+#Data overview
+library(dataMaid)
+clean(anData, useVars = c("cntry", "EvaluativeWellbeing",
+                                     "EmotionalWellbeing",
+                                     "Functioning",
+                                     "Vitality",
+                                     "CommunityWellbeing",
+                                     "SupportiveRelationships"), replace = TRUE)

### 2017-03-01 article/article_v0.01.tex
opdateret manuskript
opdateret manuskript

@@
 \usepackage{hyperref}
+\usepackage{natbib}
 
@@
 
-\author{Anne H. Petersen, Bo Bang, Karl Markussen}
+\author{Anne H. Petersen, Bo Markussen, Karl Bang Christensen}
 
@@
 \label{sec:introduction}
-The origin stories of datasets have changed over time.
-While in the past, data was often collected with a specific scientific inquiry in mind, today, a lot of data is accumulated without such a specific purpose.
-This is due to vast amounts of data being registered online and due to a trend towards more open source research.
+
+Classical statistical methodology is aimed at analyzing data from designed experiments and historically statistical analyses have been done by researchers who knew the design and origin story of the data set well.
+The origin stories of data sets have changed over time and today a lot of data is accumulated without specific purpose.
+This is due to vast amounts of data being registered online and to a trend towards more open source research.
 The latter phenomenon in particular poses new challenges wrt.
 data quality assessment.
-When data are collected and made public without a specific end-point in mind, how do we ensure that e.g.\ differences in measurement instruments do not cause the data to be effectively divided into subsets that are simply not comparable?
+When data are collected and made public without a specific end-point in mind, how do we ensure that differences in, say, choice measurement instruments, mode of administration, or sampling frame do not cause the data to be effectively divided into subsets that are simply not comparable?
+
+Surveys often use mixed modes of administration, e.g.
+mail and telephone, and while this can improve response rates, the mode of administration can affect results \cite{Brambilla1987,McHorney1994} and differences in response behavior can lead to biased results.
+Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful.
+
+The rapid growth of web surveys, due to low cost, timeliness, and other factors, generate large data sources that lack a sampling frame of the general population.
+However, it can be problematic to combine online panels (pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions) \cite{Liu2016}.
 
@@
 
-\subsection{Principal component analysis - a very brief introduction}
+\subsection{Principal component analysis}
 
-{\color{red}
+Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$ denote the empirical covariance.
 
 \emph{Principal component analysis} (PCA) may be interpreted in terms of the \emph{rank-q-reconstruction error}.
-To describe this consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, and let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ be their average.
 Suppose that we want to describe each of the observations by $q$ numbers instead of the original $d$ numbers.
@@
 PCA ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error.
-Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of the empirical covariance $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$.
+Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of $S$.
 Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ in the columns.
@@
 and equals $\sum_{j=q+1}^d \lambda_j$.
-In the terminology of PCA the eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}.
+The eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}.
 The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}.
@@
 \end{align*}
-
 It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
-This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and associated variation components $\lambda_j$ describe the structure of the dataset simultaneously for all approximating dimension $q$.
-This implies that the loadings and variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension $q$.
-}
+This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$.
+This implies that the loadings and variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension $q$.
 
@@
 \vspace{\fill}\clearpage
+\newpage
+\bibliographystyle{apa}
+\bibliography{bib}
 
-\begin{thebibliography}
-
-\bibitem Ackerman, T.A.
-(1992).
-\hl{EXAMPLE POST.} A didactic explanation of item bias, item impact, and item validity from a multidimensional perspective.
-\textit{Journal of Educational Measurement}, \textit{29}, 67--91.
-
-\end{thebibliography}
-\vspace{\fill}
-
-%\vfill\eject
 \end{document}

### 2017-03-02 article/article_v0.01.tex
Worked on the "Method" section in the article. Work isn't yet completed.

@@
 
-Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$ denote the empirical covariance.
-
-\emph{Principal component analysis} (PCA) may be interpreted in terms of the \emph{rank-q-reconstruction error}.
-Suppose that we want to describe each of the observations by $q$ numbers instead of the original $d$ numbers.
-The rank-q-reconstruction error is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
+Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance.
+Suppose that we want to describe the observations by $q$ numbers instead of the original $d$ numbers.
+The associated \emph{rank-q-reconstruction error} is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
 \begin{equation*}
@@
 \end{equation*}
-PCA ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error.
+\emph{Principal component analysis} (PCA) ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error.
 Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of $S$.
@@
 Above, we promised a method for intuitive, visual inspection of data structure similarities, but as of now, all intuition might have been lost in technicalities.
-The main point we want to emphasize from PCA is that we can interpret the eigenvector/eigenvalue-pairs of the covariance matrix as tools for obtaining a different representation of the dataset.
-Specifically, we can interpret the PCA loadings (eigenvectors) as weights that determine the relative influence of each variable in each component.
-If two different data sets with the same variables, but different samples of observations, then have similar loading patterns, the two data sets will agree on which variables explain the most variance and also \textit{how} this variance is explained.
+The main point we want to emphasize from PCA is that whereas the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables.
+If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both data situations.
+This can be the case while the two sets of scores could be arbitrarily different, which e.g.\ could happen if the two datasets were taken from two different populations of subjects.
+On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variables are used differently in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.
+
+%Specifically, we can interpret the loadings as weights that determine the relative influence of each variable in each principal component.
+
+%the two data sets will agree on which variables explain the most variance and also how this variance is explained.
 Looking at the PCA loadings and the cumulative explained variance thus provides a straightforward non-parametric graphical approach for assessing similarity between two datasets.
 
-{\color{red}
-We propose the following three diagnostic plots:
-\begin{description}
-\item[CumVarAgreement:] Cumulated Bland-Altman plot for the variation decomposition.
-\item[Hairplot:] Plot of the loadings for one of the datasets in the coordinate system of the other dataset.
-\item[Pancakeplot:] This is the ``classical'' PCADSC.
-\end{description}
-
-Below we describe the details of these plots.
-Consider two different datasets in the same $d$ variables.
-Let $\lambda_{k1} \ge \dotsm \ge \lambda_{kd}$ and $\eta_{k1},\dotsc,\eta_{kd}$ be the variation components and the loadings for dataset $k=1,2$.
+In this paper we propose three diagnostic plots for comparing the loading patterns in two different datasets.
+These three plots will be referred to as the \emph{CumEigenPlot}, the \emph{HairPlot} and the \emph{PancakePlot}.
+In order to describe these plots we consider two different datasets with $n_1$ and $n_2$ observations, respectively, but in the same $d$ variables.
+We assume that the variables have been standardized separately within the two datasets to have zero mean and unit standard deviation \hl{(justification for this choice should be further elaborated.
+And also: Something about what to do if not all variables are numerical.)}.
+The empirical covariance matrices for the two datasets are called $S_1$ and $S_2$, respectively.
+Due to the standardization these are the same as the empirical correlation matrices.
+Finally, let $\lambda_{i1} \ge \dotsm \ge \lambda_{id}$ and $\eta_{i1},\dotsc,\eta_{id}$ be the variation components and the loadings for dataset $i=1,2$.
 
@@
 
-\textbf{CumVarAgreement:} To be done.
+Description of the \emph{CumEigenPlot}: \hl{This is a cumulated Bland-Altman plot for the variation components. Details will be given later decomposition.}
 
-\bigskip
+\medskip
 
-\textbf{Hairplot:} Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
+Description of the \emph{HairPlot:} Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
 The empirical covariance matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
@@
 \Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
-\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top.
+\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
 \end{equation*}
-We propose to visualize this decomposition by plotting arrows with length $l_{kj}$ and angle $\theta_{kj}/2$ anticlockwise from the diagonal in the $j$th row and the $k$th column of a $d \times d$ grid display, where
+and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset.
+We propose to visualize these two decompositions in a $d \times d$ grid display.
+In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell.
+The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ anticlockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal.
+To facilitate the following description we will refer to the arrows drawn anticlockwise as the blue arrows, and the arrows drawn clockwise as the red arrows.
+The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
 \begin{align*}
-l_{kj} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k} \eta_{2j}^\top \rvert, &
-\theta_{kj} &= \acos(\lvert \eta_{1k} \eta_{2j}^\top \rvert).
+\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k} \eta_{2j}^\top \rvert, &
+\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j} \eta_{1k}^\top \rvert, &
+\theta_{jk} &= \acos(\lvert \eta_{1k} \eta_{2j}^\top \rvert).
 \end{align*}
-Using the same grid display, but interchanging the first and the second dataset as well as the rows and the columns, and using the angles clockwise, similar arrows are plotted in a different color.
 
-In our implementation we have used blue for the first and the second set of arrows, respectively.
-In this plot the blue arrows in the $k$th column visualize the decomposition of the $k$th principal components for the first dataset in the coordinate system of the second dataset.
-The scaling of the length by $\lambda_{\max}$ is made so that the longest arrow has unit length, and the absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors.
-This indeterminacy also implies the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$.
-We can visualize both the decomposition of $S_1$ and $S_2$ in the same plot by divide the angles by two and using anticlockwise and clockwise shifts from the diagonal.
+The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors.
+This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using anticlockwise and clockwise shifts from the diagonal.
+Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.
+
+In the \emph{HairPlot} the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal components for the first dataset in the coordinate system of the second dataset.
 Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset.
+If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
+Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
+And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells as well as arrows of non vanishing length in the off-diagonal cells.
 
-If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as the off-diagonal arrows would have zero length.
-Differences in the variation components are visualized as differences in the lengths of the arrows, also in the diagonal.
-And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal as well as arrows of non vanishing length in the off-diagonal.
+%Using the same grid display, but interchanging the first and the second dataset as well as the rows and the columns, and using the angles clockwise, similar arrows are plotted in a different color.
+
+%In our implementation we have used blue for the first and the second set of arrows, respectively.
 
@@
 
-}
-
-Our proposal of a PCADSC method consist of three steps.
+Description of the \emph{PancakePlot}: Our proposal of a PCADSC method consist of three steps.
 These steps should be performed separately for each of the two (or more) datasets that we wish to compare.

### 2017-03-02 article/article_v0.01.tex
Emphasized difference between correlation and "covariance of standardized variables".

@@
 
-In this paper we propose three diagnostic plots for comparing the loading patterns in two different datasets.
-These three plots will be referred to as the \emph{CumEigenPlot}, the \emph{HairPlot} and the \emph{PancakePlot}.
-In order to describe these plots we consider two different datasets with $n_1$ and $n_2$ observations, respectively, but in the same $d$ variables.
-We assume that the variables have been standardized separately within the two datasets to have zero mean and unit standard deviation \hl{(justification for this choice should be further elaborated.
-And also: Something about what to do if not all variables are numerical.)}.
-The empirical covariance matrices for the two datasets are called $S_1$ and $S_2$, respectively.
-Due to the standardization these are the same as the empirical correlation matrices.
-Finally, let $\lambda_{i1} \ge \dotsm \ge \lambda_{id}$ and $\eta_{i1},\dotsc,\eta_{id}$ be the variation components and the loadings for dataset $i=1,2$.
+In this paper we propose three diagnostic plots, referred to as the \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot}, for comparing the loading patterns in two datasets.
+In order to describe these plots we consider two different datasets in the same $d$ variables and with $n_1$ and $n_2$ observations, respectively.
+For dataset $i=1,2$ let $S_i \in \RR^{d \times d}$ be the empirical correlation matrix, and let $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0 $ and $\eta_{i1},\dotsc,\eta_{id} \in \RR^{d}$ be the corresponding variation components and loadings.
+The correlation matrices correspond to the covariance matrices for the variables after these have been standardized to unit standard deviation separately within the two datasets.
+
+\hl{To do: justification for using correlation matrices should be further elaborated.
+And also: Something about what to do if not all variables are numerical.
+Using correlation matrices is not the same as assuming standadized variables; there is a difference in the permutations.}
 

### 2017-03-04 article/article_v0.01.tex
Worked on description of CumEigenPlot. Not yet completed.

@@
 In order to describe these plots we consider two different datasets in the same $d$ variables and with $n_1$ and $n_2$ observations, respectively.
-For dataset $i=1,2$ let $S_i \in \RR^{d \times d}$ be the empirical correlation matrix, and let $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0 $ and $\eta_{i1},\dotsc,\eta_{id} \in \RR^{d}$ be the corresponding variation components and loadings.
+For dataset $i=1,2$ let $S_i \in \RR^{d \times d}$ be the empirical correlation matrix \hl{(find a way to introduce the empirial correlation matrix instead)}, and let $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0 $ and $\eta_{i1},\dotsc,\eta_{id} \in \RR^{d}$ be the corresponding variation components and loadings.
 The correlation matrices correspond to the covariance matrices for the variables after these have been standardized to unit standard deviation separately within the two datasets.
+Similarly, let $S \in \RR^{d \times d}$ be the empirical correlation matrix for the combined dataset with $n_1+n_2$ observations, and let $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ and $\eta_1,\dotsc,\eta_d \in \RR^d$ be the corresponding variation components and loadings.
 
@@
 
-Description of the \emph{CumEigenPlot}: \hl{This is a cumulated Bland-Altman plot for the variation components. Details will be given later decomposition.}
+Description of the \emph{CumEigenPlot}: This plot compares the variation components.
+If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
+And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
+In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets we plot a piecewise linear curve connecting the points
+\begin{align*}
+(0,0), &&
+(\lambda_1,\lambda_{11}-\lambda_{12}), &&
+(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{12}-\lambda_{22}), &&
+\ldots, &&
+\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg).
+\end{align*}
+This may be seen as a cumulated Bland-Altman plot for the variation components (\hl{reference to cumulated residuals and to Bland-Altman}).
+Note that since we are using correlation matrices the last point will always be $(d,0)$.
+Thus, this curve will begin and end at the x-axis.
+And the larger excursions it makes away from the x-axis the less alike the cumulated variation components for the two datasets are.
+
+Whether the excursions in the observed curve are large or within the range of sample variation can be quantified by a permutation test.
+The idea is that we randomly reallocate the $n_1+n_2$ observations in the combined dataset to two datasets with $n_1$ and $n_2$ observations, respectively, and then repeat the procedure described above.
+The random reallocation by construction ensures that the two resampled datasets are alike except their sample size and sampling variation.
+In the \emph{CumEigenPlot} we make 1000 independent random reallocations, and plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95pct coverage intervals.
+This provides a graphical validation tool.
+P-values for the null hypothesis that the variation components are the same are easily provided for any appropriate test statistic by the resampling procedure as well.
+We have implemented the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
+\begin{align*}
+\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
+\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)^2.
+\end{align*}
 
@@
 Description of the \emph{HairPlot:} Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
-The empirical covariance matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
+The empirical correlation matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
 \begin{equation*}

### 2017-03-04 article/article_v0.01.tex
Worked on description of CumEigenPlot. Not yet completed.

@@
 
-Description of the \emph{CumEigenPlot}: This plot compares the variation components.
+Description of the \emph{CumEigenPlot}: This plot only visualizes the variation components.
 If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
@@
 In the \emph{CumEigenPlot} we make 1000 independent random reallocations, and plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95pct coverage intervals.
-This provides a graphical validation tool.
-P-values for the null hypothesis that the variation components are the same are easily provided for any appropriate test statistic by the resampling procedure as well.
-We have implemented the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
-\begin{align*}
-\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
-\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)^2.
-\end{align*}
+
+\hl{TO DO: Describe KS and CvM-tests.}
 
@@
 Description of the \emph{HairPlot:} Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
-The empirical correlation matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
+The empirical covariance matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
 \begin{equation*}

### 2017-03-05 article/article_v0.01.tex
Still working on Methods.

--- 2017-03-04 article/article_v0.01.tex
@@
 
-However, by clever use of linear algebra, we can construct a decomposition of the covariance matrix that makes it easier to gain an overview of the data.
+However, by 
+%clever 
+use of linear algebra, we can construct a decomposition of the covariance matrix that makes it easier to gain an overview of the data.
 We propose a new method based on principal component analysis that seems to be able to identify differences in datasets based on intuitive, visual inspections.
@@
 
-In this paper we propose three diagnostic plots, referred to as the \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot}, for comparing the loading patterns in two datasets.
-In order to describe these plots we consider two different datasets in the same $d$ variables and with $n_1$ and $n_2$ observations, respectively.
+In this paper we propose three diagnostic plots for comparing the loading patterns in two datasets $x_{11},\dotsc,x_{1 n_1} \in \RR^d$ and $x_{21},\dotsc,x_{2 n_2} \in \RR^d$ in the same $d$ variables.
+The construction of these plots, which are called \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot}, proceeds in the following steps:
+\begin{enumerate}
+\item Standardized the observations $x_{i1},\dotsc,x_{i n_i}$ to have zero mean and unit standard deviation separately for each of the $d$ variables within each datasets.
+Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized observations, and form the principal component analysis
+\begin{equation*}
+S_i = \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$.} 
+\end{equation*}
+\item Combine the standardized datasets into a joint dataset $\tilde{x}_1,\dotsc,\tilde{x}_n \in \RR^d$ with $n=n_1+n_2$ observations, and form the principal component analysis
+\begin{equation*}
+S = \frac{1}{n-2} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top.
+\end{equation*}
+\item Make the diagnostic plot of interest. Details of these plots will be described below.
+\item To investigate the sample variation randomly reallocate the combined dataset formed in step 2 to two datasets of $n_1$ and $n_2$ observations, respectively, and redo step 1 and 3.
+This is done several times, e.g.\ 10000 times, and compared to the plot for the original allocation.
+\end{enumerate}
+
+Before describing the details of the  \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot} we make some general remarks.
+The covariance matrix for the standardized variables is the same as the correlation matrix for the original variables.
+The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1.
+This also holds for the reallocated datasets, and in particular we have $\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$.
+The random reallocations in step 4 provide resamples under the null hypothesis that the two sample populations have the same structure in the $d$ variables, except that the two populations are allowed to have the own standard deviations for each of the variables.
+This exception is induced by the initial step 1, which performs a standardization within the two datasets before they are combined in step 2.
+These resamples can be used for a visualization of the sample variation, and if a test statistic can be defined in step 3, then a p-value for the null hypothesis may be computed.
+Such p-values are know as permutation tests for Goodness-of-Fit (\hl{reference?}).
+
+%In order to describe these plots we consider two different datasets in the same $d$ variables and with $n_1$ and $n_2$ observations, respectively.
 For dataset $i=1,2$ let $S_i \in \RR^{d \times d}$ be the empirical correlation matrix \hl{(find a way to introduce the empirial correlation matrix instead)}, and let $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0 $ and $\eta_{i1},\dotsc,\eta_{id} \in \RR^{d}$ be the corresponding variation components and loadings.
@@
 
-\hl{To do: justification for using correlation matrices should be further elaborated.
-And also: Something about what to do if not all variables are numerical.
-Using correlation matrices is not the same as assuming standadized variables; there is a difference in the permutations.}
+\hl{To do: justification for using correlation matrices should be further elaborated. And also: Something about what to do if not all variables are numerical. 
+%Using correlation matrices is not the same as assuming standadized variables; there is a difference in the permutations.
+}
 
@@
 \begin{equation*}
-S_1 = \sum_{k=1}^d \lambda_{1j} \eta_{1k} \eta_{1k}^\top
+S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
 = \lambda_{\max} \sum_{k=1}^d

### 2017-03-05 article/article_v0.01.tex
Having had big fight with GitHub. Hope it is solved now. But lets see.

@@
 In this paper we propose three diagnostic plots for comparing the loading patterns in two datasets $x_{11},\dotsc,x_{1 n_1} \in \RR^d$ and $x_{21},\dotsc,x_{2 n_2} \in \RR^d$ in the same $d$ variables.
-The construction of these plots, which are called \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot}, proceeds in the following steps:
+The construction of these plots, which we call \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot}, proceeds in the following steps:
 \begin{enumerate}
@@
 \end{equation*}
-\item Make the diagnostic plot of interest. Details of these plots will be described below.
+\item Make the diagnostic plot of interest. Details of the plots will be described below.
 \item To investigate the sample variation randomly reallocate the combined dataset formed in step 2 to two datasets of $n_1$ and $n_2$ observations, respectively, and redo step 1 and 3.
@@
 Description of the \emph{CumEigenPlot}: This plot compares the variation components. 
-If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
+%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
 And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
@@
 (\lambda_1,\lambda_{11}-\lambda_{12}), &&
-(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{12}-\lambda_{22}), &&
+(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
 \ldots, &&
@@
 This may be seen as a cumulated Bland-Altman plot for the variation components (\hl{reference to cumulated residuals and to Bland-Altman}).
-Note that since we are using correlation matrices the last point will always be $(d,0)$.
+Note that due to the standardization the last point will be equal to $(d,0)$.
 Thus, this curve will begin and end at the x-axis.
@@
 
-Whether the excursions in the observed curve are large or within the range of sample variation can be quantified by a permutation test.
-The idea is that we randomly reallocate the $n_1+n_2$ observations in the combined dataset to two datasets with $n_1$ and $n_2$ observations, respectively, and then repeat the procedure described above.
-The random reallocation by construction ensures that the two resampled datasets are alike except their sample size and sampling variation.
-In the \emph{CumEigenPlot} we make 1000 independent random reallocations, and plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95pct coverage intervals.
-This provides a graphical validation tool.
-P-values for the null hypothesis that the variation components are the same are easily provided for any appropriate test statistic by the resampling procedure as well.
-We have implemented the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
+In step 3 we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics given by
 \begin{align*}
@@
 \end{align*}
+From the random reallocations in step 4 we hence get p-values for the Kolmogorov-Smirnov and the Cram\'er-von Mises Goodness-of-Fit tests.
+Step 4 is also used to visualize the sampling variation in a display, where we plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95pct coverage intervals.
+If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicate differences between the two datasets.
+
+%Whether the excursions in the observed curve are large or within the range of sample variation can be quantified by a permutation test.
+The idea is that we randomly reallocate the $n_1+n_2$ observations in the combined dataset to two datasets with $n_1$ and $n_2$ observations, respectively, and then repeat the procedure described above.
+The random reallocation by construction ensures that the two resampled datasets are alike except their sample size and sampling variation.
+In the \emph{CumEigenPlot} we make 1000 independent random reallocations, and plot  P-values for the null hypothesis that the variation components are the same are easily provided for any appropriate test statistic by the resampling procedure as well.
 

### 2017-03-06 R/cumeigen.R
Translated cumeigen plot into ggplot2

@@ -31,6 +31,9 @@
 #'
 #' @importFrom graphics polygon matlines lines legend
 #' @importFrom stats quantile
+#' @importFrom ggplot2 aes geom_polygon geom_line scale_x_continuous
+#' scale_y_continuous theme_bw theme element_blank xlab ylab geom_line
+#' scale_linetype_manual annotate unit
 #' @export
 cumeigen <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE) {
   #define var
@@ -109,18 +112,63 @@ cumeigen <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE) {
   KS.pvalue <- mean(KS.sim >= KS.obs)
   CvM.pvalue <- mean(CvM.sim >= CvM.obs)
 
-  # make plot
-  if (make.plot) {
-    y.min <- apply(y.sim,1,quantile,probs=0.025)
-    y.max <- apply(y.sim,1,quantile,probs=0.975)
-    plot(c(x[1],x[d+1]),c(1.01*min(c(y.obs,as.vector(y.sim))),1.01*max(c(y.obs,as.vector(y.sim)))),type="n",
-         xlab="Cumulated joint eigenvalues",ylab="Cumulated difference of eigenvalues")
-    polygon(c(x,rev(x)),c(y.min,rev(y.max)),col="aliceblue",border=NA)
-    matlines(x,y.sim[,1:min(20,B)],col="gray")
-    lines(x,y.obs,lwd=2)
-    legend("topleft",paste(c("KS: p=","CvM: p="),round(c(KS.pvalue,CvM.pvalue),3),sep=""))
+  if (!make.plot) {
+    # return
+    return(list(KS.pvalue=KS.pvalue,CvM.pvalue=CvM.pvalue))
   }
 
+  # make plot
+ # if (make.plot) {
+    y.min <- apply(y.sim,1,quantile,probs=0.025)
+    y.max <- apply(y.sim,1,quantile,probs=0.975)
+    nSims <- min(20, B)
+    y.sim.small <- y.sim[, 1:nSims]
+    pFsim <- data.frame(x = rep(x, ncol(y.sim.small)),
+                     y = c(y.sim.small),
+                     type = c(rep("sim", ncol(y.sim.small)*nrow(y.sim.small))),
+                     run = c(rep(1:ncol(y.sim.small), each = length(x))))
[…]

### 2017-03-06 article/article_v0.01.tex

@@
 \end{equation*}
-\item Combine the standardized datasets into a joint dataset $\tilde{x}_1,\dotsc,\tilde{x}_n \in \RR^d$ with $n=n_1+n_2$ observations, and form the principal component analysis
+\item Combine the standardized datasets into a joint dataset $\tilde{x}_1,\dotsc,\tilde{x}_n \in \RR^d$ with $n=n_1+n_2$ observations, standardize this dataset, and form the principal component analysis
 \begin{equation*}
@@
 \item Make the diagnostic plot of interest. Details of the plots will be described below.
-\item To investigate the sample variation randomly reallocate the combined dataset formed in step 2 to two datasets of $n_1$ and $n_2$ observations, respectively, and redo step 1 and 3.
+\item To investigate the sample variation randomly reallocate the combined and standardized dataset formed in step 2 to two datasets of $n_1$ and $n_2$ observations, respectively, and redo step 1 and 3.
 This is done several times, e.g.\ 10000 times, and compared to the plot for the original allocation.
@@
 \end{align*}
-From the random reallocations in step 4 we hence get p-values for the Kolmogorov-Smirnov and the Cram\'er-von Mises Goodness-of-Fit tests.
+From the random reallocations in step 4 we get p-values for the Kolmogorov-Smirnov and the Cram\'er-von Mises Goodness-of-Fit tests.
 Step 4 is also used to visualize the sampling variation in a display, where we plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95pct coverage intervals.
@@
 
-\medskip
+\bigskip
 
-Description of the \emph{HairPlot:} Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
+Description of the \emph{HairPlot:} This plot simultaneously compares the variation components and the loadings.
+Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
 The empirical correlation matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
@@
 And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells as well as arrows of non vanishing length in the off-diagonal cells.
+The name \emph{HairPlot} arise by seeing the arrows as \emph{hairs}.
+Using this interpretation we may say that a more ``hairy'' plot indicates less agreement between the two loading patterns.
 
-%Using the same grid display, but interchanging the first and the second dataset as well as the rows and the columns, and using the angles clockwise, similar arrows are plotted in a different color.
-
-%In our implementation we have used blue for the first and the second set of arrows, respectively.
+Along with the visual comparison we have also implemented numerical tests in step 3.
+Separately for the off-diagonal grid cells we may use the lengths $\mu_{jk}$ and $\nu_{jk}$ as test statistics, and for the on-diagonal grid cells we use the distance between the scaled loadings as test statistics, that is
+\begin{equation*}
+\chi_k = \lambda_{\max}^{-1} \lvert \lambda_{1k} \eta_{1k} - \lambda_{2k} \eta_{2k} \rvert, \quad \text{for $k=1,\dotsc,d$.}
+\end{equation*}
+In total this gives $2d^2 - d$ tests, which should be corrected for the family wise type I error using a methods that doesn't assume independence among the tests.
+We also construct overall Kolmogorov-Smirnov and Cram\'er-von Mises type test statistics by
+\begin{align*}
+\text{KS}_1 &= \max_{j,k=1,\dotsc,d} (1_{j \neq k} \mu_{jk} + 1_{j=k} \chi_k), &
+\text{CvM}_1 &= \sum_{j=1}^d \sum_{k=1}^d (1_{j \neq k} \mu_{jk} + 1_{j=k} \chi_k)^2, \\
+\text{KS}_2 &= \max_{j,k=1,\dotsc,d} (1_{j \neq k} \nu_{jk} + 1_{j=k} \chi_k), &
+\text{CvM}_2 &= \sum_{j=1}^d \sum_{k=1}^d (1_{j \neq k} \nu_{jk} + 1_{j=k} \chi_k)^2.
+\end{align*}
 
@@
 
-Remarks: Presently permutation tests based on the Kolmogorov-Smirnov and the Cram\'er-von Mises test statistics has been implemented for CumVarAgreement.
-This should also be possible for the Hairplot based on the off-diagonal components.
-For the Pancakeplot a Wallyplot has been implemented (however, in Claus' Wally plot the position of the ``observed dataset'' is also random).
+Description of the \emph{PancakePlot}: This plot compares the loadings.
 
-\bigskip
-
-Description of the \emph{PancakePlot}: Our proposal of a PCADSC method consist of three steps.
+Our proposal of a PCADSC method consist of three steps.
 These steps should be performed separately for each of the two (or more) datasets that we wish to compare.

### 2017-03-06 article/article_v0.01.tex
Decription of the plots completed in the article. This is a first draft.

@@
 Description of the \emph{PancakePlot}: This plot compares the loadings.
+The \emph{PancakePlot} consists of two parallel vertical barplots, one for each of the datasets, with one bar for each of the $d$ loading vector $\eta_{ij} \in \RR^d$.
+In the bars the relative contributions of the $d$ variables are visualized by separate colors for the different variables.
+Due to the indeterminacy of the sign all the signs are removed from the coefficients in the loadings, and the relative contribution is defined as the resulting coefficients divided by their total sum.
+The bars are ordered according to the variation components, and we also insert the cumulative variation components next to the barplots.
+The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.
+\hl{Refer to example/show plot.}
 
-Our proposal of a PCADSC method consist of three steps.
+We not implemented a test statistic for the \emph{PancakePlot}, but in order to visualize the sample variation the observed plot can be shown together with resampled plots generated in step 4.
+
+%Our proposal of a PCADSC method consist of three steps.
 These steps should be performed separately for each of the two (or more) datasets that we wish to compare.
@@
 The three steps are:
-\begin{enumerate}
-\item \hl{Standardize. Full data or subsets? And also: Something about what to do if not all variables are numerical.}
-\item Compute the PCA loadings and the variance contributions of each principal component.
-\item For each principal component, standardize the loadings, i.e. scale them such that they sum to one.
-\item Produce a plot consisting of a bar for each principal component, decorated with the cumulative variance contribution corresponding to this component.
+%\begin{enumerate}
+%\item \hl{Standardize. Full data or subsets? And also: Something about what to do if not all variables are numerical.}
+%\item Compute the PCA loadings and the variance contributions of each principal component.
+%\item For each principal component, standardize the loadings, i.e. scale them such that they sum to one.
+%\item Produce a plot consisting of a bar for each principal component, decorated with the cumulative variance contribution corresponding to this component.
 The bar should be of length one and colored according to the variables loading the component.
-\end{enumerate}
-The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.
-\hl{Refer to example/show plot.}
+%\end{enumerate}
 

### 2017-03-06 R/cumeigen.R
Translated hairplot into ggplot2 syntax + added missing dependencies for cumeigen

@@ -31,7 +31,7 @@
 #'
 #' @importFrom graphics polygon matlines lines legend
 #' @importFrom stats quantile
-#' @importFrom ggplot2 aes geom_polygon geom_line scale_x_continuous
+#' @importFrom ggplot2 ggplot aes_string geom_polygon geom_line scale_x_continuous
 #' scale_y_continuous theme_bw theme element_blank xlab ylab geom_line
 #' scale_linetype_manual annotate unit
 #' @export
@@ -132,10 +132,10 @@ cumeigen <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE) {
     yBreaks <- round(c(0 + yMaxVal * c(1/3, 2/3, 3/3), 0,
                        0 - yMaxVal * c(1/3, 2/3, 3/3)),1)
 
-    ggplot(pFsim, aes(x = x, y = y)) +
+    ggplot(pFsim, aes_string(x = "x", y = "y")) +
       annotate(geom = "polygon", x = c(x, rev(x)), y = c(y.min, rev(y.max)),
                fill = "aliceblue") +
-     geom_line(aes(group = run, linetype = factor(run)), col = "grey", size = 0) +
+     geom_line(aes_string(group = "run", linetype = factor("run")), col = "grey", size = 0) +
       scale_x_continuous(limits = c(x[1],x[d+1]), breaks = 0:length(x)) +
       scale_y_continuous(limits = c(yMinVal, yMaxVal),
                          breaks = yBreaks) +

### 2017-03-06 R/hairplot.R

@@ -32,6 +32,9 @@
 #' hairplot(iris, "group",var=setdiff(names(iris), c("group", "Species")))
 #'
 #' @importFrom graphics axis arrows
+#' @importFrom ggplot2 ggplot aes_string scale_x_continuous scale_y_continuous
+#' theme_bw theme element_blank xlab ylab geom_segment unit scale_color_manual
+#' @importFrom grid arrow
 #' @export
 hairplot <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE,arrow.len=0.05) {
   #define var
@@ -86,27 +89,62 @@ hairplot <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE,arrow.len=0.05)
 
   # find angles
   angles <- matrix(0,d,d)
-  for (i in 1:d) for (j in 1:d) angles[i,j] <- asin(max(0,min(1,abs(sum(eigen1$vectors[,i]*eigen2$vectors[,j])))))
-
-  # make plot
-  if (make.plot) {
-    plot(c(0.5,d+0.5),c(0.5,d+0.5),type="n",xlab="dataset 1: PC's",ylab="dataset 2: PC's",axes=FALSE)
-    axis(1,1:d)
-    axis(2,1:d)
-    pi <- 2*asin(1)
-    max.eigen <- max(c(eigen1$values[1],eigen2$values[1]))
-    for (i in 1:d) for (j in 1:d) {
-      # plot PC1's in coordinate system of PC2
-      len <- sqrt(eigen1$values[i]/max.eigen)*abs(sum(eigen1$vectors[,i]*eigen2$vectors[,j]))
-      arrows(i,j,i+len*cos((pi-angles[i,j])/2),j+len*sin((pi-angles[i,j])/2),length=arrow.len,col="blue")
-    }
-    for (i in 1:d) for (j in 1:d) {
-      # plot PC2's in coordinate system of PC1
-      len <- sqrt(eigen2$values[j]/max.eigen)*abs(sum(eigen1$vectors[,i]*eigen2$vectors[,j]))
-      arrows(i,j,i+len*cos(angles[i,j]/2),j+len*sin(angles[i,j]/2),length=arrow.len,col="red")
-    }
+  for (i in 1:d) for (j in 1:d) angles[i,j] <- asin(max(0,min(1,abs(sum(eigen1$vectors[,i]*
+                                                                          eigen2$vectors[,j])))))
+  # Return angles if no plot is required
+  if (!make.plot) {
[…]

### 2017-03-08 R/hairplot.R
HairPlot: Fixed bug in data.frame for color-code

@@ -114,7 +114,7 @@ hairplot <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE,arrow.len=0.05)
                             rep(1:d, d) + c(len2)*cos(c(angles)/2)),
                    yend = c(rep(1:d, each = d) + c(len1)*sin((pi-c(angles))/2),
                             rep(1:d, each = d) + c(len2)*sin(c(angles)/2)),
-                   type = rep(c("1st", "2nd"), each = 2*d))
+                   type = rep(c("1st", "2nd"), each = d*d))
 
   ggplot(pF[, , drop = F], aes_string(x = "x", y = "y", col = "type",
                                       xend = "xend", yend = "yend")) +

### 2017-03-08 article/article_v0.01.tex
Comments/minor stuff for sections 3.2 - 3.2.1

@@
 \usepackage{natbib}
+%\usepackage[backend=biber,natbib=true, style=authoryear-comp]{biblatex}
 
@@
 
-Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance.
+Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average\hl{(s?)} and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance \hl{Is $\bar{x} \in \RR^d$ with $\bar{x} = (\bar{x_1}, ..., \bar{x_d})^T$?}.
 Suppose that we want to describe the observations by $q$ numbers instead of the original $d$ numbers.
@@
 If some of the eigenvalues are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
-In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero.
+In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero \hl{Anne: What about covariance matrices that are not of full rank (e.g.
+because of colinearity)?
+Won't we find eigenvalue multiplicities $> 1$ here as well?}.
 It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
@@
 The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}.
-The $j$th loading can also found as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores
+The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores:
 \begin{align*}
@@
 \end{align*}
-It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
+\hl{Anne: Should we maybe assume standardized $x$s here also?
+I feel like it is part of the PCA algorithm and subtracting the mean in the above seems off for that reason.} It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
 This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$.
@@
 
-\subsection{PCA-based data structure comparison}
+\subsection{PCA-based data structure comparisons}
 Above, we promised a method for intuitive, visual inspection of data structure similarities, but as of now, all intuition might have been lost in technicalities.
@@
 Before describing the details of the  \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot} we make some general remarks.
-The covariance matrix for the standardized variables is the same as the correlation matrix for the original variables.
+First, it should be noted that PCA is sensitive to scaling, as the procedure deconstructs the covariance matrix in components according to the most explained variance.
+This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data.
+Therefore, the variables should always be scaled prior to performing PCA.
+Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices.
+\hl{Do we still need more about this topic?
+Then we can add something like:
+\begin{itemize}
+\item Following the greedy interpretation of PCA, we know that the first eigenvector is chosen as
+$$\eta_1 = \argmax_{u \in \RR^d} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2  = \text{argmax}_{u \in \RR^d} \V(u^T X^T) $$
+\item Now, imagine that $\V(x_1) >> \V(x_j)$ for all $j \in 2, ..., d$.
+Note that $(1, 0, ..., 0) X^T = x_i$.
+Then, clearly a vector $u$ that only picks out the first variable from the data matrix $X$, namely $u^T = (1, 0, ..., 0)$ will allow us to only look at the variance of $x_1$, which is very large.
+This has nothing to do with the interplay between variables, as the phenomenon will disappear if we scale $x_1$
+\item A more formal explanation could come from this kind of intuition, if needed.
+\end{itemize}
+}
 The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1.
@@
 This exception is induced by the initial step 1, which performs a standardization within the two datasets before they are combined in step 2.
-These resamples can be used for a visualization of the sample variation, and if a test statistic can be defined in step 3, then a p-value for the null hypothesis may be computed.
-Such p-values are know as permutation tests for Goodness-of-Fit (\hl{reference?}).
+These resamples can be used for a visualization of the sample variation, and if a test statistic can be defined in step 3, then a $p$-value for the null hypothesis may be computed.
+Such $p$-values are know as permutation tests for Goodness-of-Fit (\hl{reference?}).
 
@@
 
-\hl{To do: justification for using correlation matrices should be further elaborated. And also: Something about what to do if not all variables are numerical. 
+\hl{To do: Something about what to do if not all variables are numerical. Anne: Maybe leave this for the discussion?
 %Using correlation matrices is not the same as assuming standadized variables; there is a difference in the permutations.
@@
 
-Description of the \emph{CumEigenPlot}: This plot compares the variation components. 
+\subsubsection{Description of the \emph{CumEigenPlot}}
+ This plot compares the variation components. 
 %If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
@@
 And the larger excursions it makes away from the x-axis the less alike the cumulated variation components for the two datasets are.
+\hl{Is this an informative x-axis?
+Should we maybe consider scaling it?}
 
@@
 \end{align*}
-From the random reallocations in step 4 we get p-values for the Kolmogorov-Smirnov and the Cram\'er-von Mises Goodness-of-Fit tests.
-Step 4 is also used to visualize the sampling variation in a display, where we plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95pct coverage intervals.
-If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicate differences between the two datasets.
+\hl{Maybe more here on what the null hypothesis of these tests is?} 
+From the random reallocations in step 4 we get $p$-values for the Kolmogorov-Smirnov and the Cram\'er-von Mises Goodness-of-Fit tests.
+Step 4 is also used to visualize the sampling variation in a display, where we plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95 \% coverage intervals.
+If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicates differences between the two datasets.
 
@@
 
-Description of the \emph{HairPlot:} This plot simultaneously compares the variation components and the loadings.
+\subsubsection{Description of the \emph{HairPlot}} 
+This plot simultaneously compares the variation components and the loadings.
 Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
@@
 
-\bigskip
+%\bigskip
 
-Description of the \emph{PancakePlot}: This plot compares the loadings.
+\subsubsection{Description of the \emph{PancakePlot}}
+This plot compares the loadings.
 The \emph{PancakePlot} consists of two parallel vertical barplots, one for each of the datasets, with one bar for each of the $d$ loading vector $\eta_{ij} \in \RR^d$.

### 2017-03-08 ESS/ess_notes.R

@@ -8,7 +8,7 @@ useVars <- c("cntry", "stflife", "happy", "fltsd", "fltdpr", "enjlf", "wrhpp", "
              "wrbknrm", "deaimpp", "flteeff", "slprl", "cldgng", "enrglot", "ppltrst",
              "pplfair", "pplhlp", "pplahlp", "flclpla", "inprdsc", "flapppl", "rehlppl",
              "fltlnl")
-
+#
 #anData <- na.omit(data[data$cntry %in% c("DK", "BG"), useVars])
 anData <- na.omit(data[data$cntry %in% c("DK", "SE"), useVars])
 
@@ -77,21 +77,23 @@ anData$SupportiveRelationships <- rowMeans(anData[, sMat$var[sMat$scale ==
 
 #Do PCADSC
 library(PCADSC)
+library(ggplot2)
+library(gridExtra)
 a <- makePCADSC(anData, "cntry",  c("EvaluativeWellbeing",
                                "EmotionalWellbeing",
-                               "Functioning,"
+                               "Functioning",
                                "Vitality",
                                "CommunityWellbeing",
                                "SupportiveRelationships"))
 
 
-p1 <- qplot(a)
+p1 <- pancakePlot(a)
 
 ggsave(plot = p1, file = "./article/essPCADSC.pdf",
        width = 8, height = 4)
 
 set.seed(1234)
-p2 <- wallyPCADSC(a, anData, nrow = 4, ncol = 2)
+p2 <- wallyPancake(a, anData, nrow = 4, ncol = 2)
 
 ggsave(plot = p2, file = "./article/essWallyPCADSC.pdf",
        width = 8, height = 10)

### 2017-03-15 article/article_v0.01.tex
Slides + added comments and minor corrections to the rest of section 3.

@@
 %\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
-\title{Explorative data structure comparison with application to \ldots}
+\title{Exploratory data structure comparisons using Principal Component Analysis}
 
@@
 \hl{Anne: Should we maybe assume standardized $x$s here also?
-I feel like it is part of the PCA algorithm and subtracting the mean in the above seems off for that reason.} It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
+I feel like it is part of the PCA algorithm and subtracting the mean in the above seems off for that reason.
+Also: Something is wrong with the dimensions here.
+$u$ is $d \times 1$, $x_i$ is $d \times 1$(?) - how do we get $\eta_j$ to also be $d \times 1$?} It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
 This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$.
@@
 \item Following the greedy interpretation of PCA, we know that the first eigenvector is chosen as
-$$\eta_1 = \argmax_{u \in \RR^d} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2  = \text{argmax}_{u \in \RR^d} \V(u^T X^T) $$
+$$\eta_1 = \argmax_{u \in \RR^d} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2  = \text{argmax}_{u \in \RR^d} \tilde{\V}(u^T X^T) $$
 \item Now, imagine that $\V(x_1) >> \V(x_j)$ for all $j \in 2, ..., d$.
@@
 \end{align*}
-\hl{Maybe more here on what the null hypothesis of these tests is?} 
+\hl{Maybe more here on what the null hypothesis and range of these test statistics are?} 
 From the random reallocations in step 4 we get $p$-values for the Kolmogorov-Smirnov and the Cram\'er-von Mises Goodness-of-Fit tests.
@@
 The empirical correlation matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
+%\begin{equation*}
+%S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
+%= \lambda_{\max} \sum_{k=1}^d
+%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
+%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
+%\end{equation*}
 \begin{equation*}
@@
 = \lambda_{\max} \sum_{k=1}^d
-\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
-\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
+\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
+\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)^\top,
 \end{equation*}
-and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset.
+and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset \hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?}.
 We propose to visualize these two decompositions in a $d \times d$ grid display.
@@
 \begin{align*}
-\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k} \eta_{2j}^\top \rvert, &
-\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j} \eta_{1k}^\top \rvert, &
-\theta_{jk} &= \acos(\lvert \eta_{1k} \eta_{2j}^\top \rvert).
+\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
+\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
+\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
 \end{align*}
-
+\hl{Maybe emphasize inner product/angle-interpretation here?}
 The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors.
@@
 
-In the \emph{HairPlot} the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal components for the first dataset in the coordinate system of the second dataset.
+In the \emph{HairPlot} the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component for the first dataset in the coordinate system of the second dataset.
 Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset.
-If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
+\hl{Can we expand on this in a non-technical way?} If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
 Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
@@
 
-Along with the visual comparison we have also implemented numerical tests in step 3.
+Along with the visual comparison we have also implemented numerical tests in step 3 \hl{Anne: Is a test really necessary here?
+Isn't the point of this plot to point us towards the most interesting components, not to quantify exactly how much they differ?
+I'm unsure as to whether a significant difference will necessarily be empirically/clinically relevant and likewise, if a non-significant difference will always be one we can disregard of...}.
 Separately for the off-diagonal grid cells we may use the lengths $\mu_{jk}$ and $\nu_{jk}$ as test statistics, and for the on-diagonal grid cells we use the distance between the scaled loadings as test statistics, that is
@@
 The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.
-\hl{Refer to example/show plot.}
+\hl{Refer to example/show plot.
+Also, from Anne to Anne: A bit further elaboration here, especially mention the loadings as weights interpretation}
 
@@
 %\end{enumerate}
+
+\hl{Mention somewere how we propose for people to use the plots: First CE, then hair, then pancake focusing in hairy components.}
 

### 2017-03-15 ESS/ess_notes.R

@@ -74,12 +74,78 @@ anData$fltlnl <- 10 - ((anData$fltlnl - 1) * (10/3))
 anData$SupportiveRelationships <- rowMeans(anData[, sMat$var[sMat$scale ==
                                                                "Supportive relationships"]])
 
+data_SEDK <- anData
+
+anData <- na.omit(data[data$cntry %in% c("DK", "BG"), useVars])
+
+anData$cntry <- as.character(anData$cntry)
+
+sMat <- list()
+sMat$var <- names(anData)[-1]
+sMat$scale <- c(rep("Evaluative wellbeing", 2),
+                rep("Emotional wellbeing", 6),
+                rep("Functioning", 14),
+                rep("Vitality", 4),
+                rep("Community wellbeing", 5),
+                rep("Supportive relationships", 4))
+
+
+#Evaluative wellbeing
+sapply(anData[, sMat$var[sMat$scale == "Evaluative wellbeing"]], range) #same number of levels
+anData$EvaluativeWellbeing <- rowMeans(anData[, sMat$var[sMat$scale == "Evaluative wellbeing"]])
+
+#Emotional wellbeing
+sapply(anData[, sMat$var[sMat$scale == "Emotional wellbeing"]], range) #same number of levels
+for (v in sMat$var[sMat$scale == "Emotional wellbeing"])  {
+  anData[, v] <- anData[, v] - 1
+}
+for (v in c("fltsd", "fltdpr", "fltanx")) {
+  anData[, v] <- 3 - anData[, v]
+}
+anData$EmotionalWellbeing <- rowMeans(anData[, sMat$var[sMat$scale == "Emotional wellbeing"]])
+
+#Functioning
+sapply(anData[, sMat$var[sMat$scale == "Functioning"]], range)
+for (v in setdiff(sMat$var[sMat$scale == "Functioning"],
+                  c("tmimdng", "tmabdng", "tmendng", "sedirlf", "deaimpp")))  {
+  anData[, v] <- (anData[, v]-1)*2.5
+}
+for (v in c("dclvlf", "accdng", "dngval", "optftr", "lotsgot", "pstvms")) {
+  anData[, v] <- 10 - anData[, v]
+}
+anData$Functioning <- rowMeans(anData[, sMat$var[sMat$scale == "Functioning"]])
+
+#Vitality
+sapply(anData[, sMat$var[sMat$scale == "Vitality"]], range)
[…]

### 2017-03-19 article/article_v0.9.tex
Redid section 4 + minor general editing. Saved new version that we should use from now on.

--- 2017-03-15 article/article_v0.01.tex
@@
 \usepackage{natbib}
+\usepackage{multirow}
 %\usepackage[backend=biber,natbib=true, style=authoryear-comp]{biblatex}
@@
 %\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
-\title{Exploratory data structure comparisons using Principal Component Analysis}
+\title{Exploratory data structure comparisons: Three new tools based on Principal Component Analysis}
 
@@
 
-Classical statistical methodology is aimed at analyzing data from designed experiments and historically statistical analyses have been done by researchers who knew the design and origin story of the data set well.
-The origin stories of data sets have changed over time and today a lot of data is accumulated without specific purpose.
+Classical statistical methodology is aimed at analyzing data from designed experiments and historically, statistical analyses have been done by researchers who knew the design and origin story of the data set well.
+However, the origin stories of data sets have changed over time and today a lot of data is accumulated without a specific purpose in mind.
 This is due to vast amounts of data being registered online and to a trend towards more open source research.
@@
 What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
-We propose a new method for this task, namely Principal Component Analysis-based Data Structure Comparison (PCADSC).
-This method employs the principal component decomposition of the data matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
-\hl{Mention R package.}
+We propose a suite of three new tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC).
+These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
+Software for using these tools is available in an \texttt{R}-package.
+\hl{how to refer to our package?}
 
 This manuscript is structured as follows: First, in Section \ref{sec:stateoftheart}, we present the data structure comparison problem in more detail and discuss what statistical methods are already available for solving similar challenges.
-Next, in Section \ref{sec:pcadscintro}, we move on to a description of the PCADSC procedure, including a brief introduction to principal component analysis (PCA) in general.
-In Section \ref{sec:dataexample}, we present a worked data example using the open source, online available PISA data (\hl{ref}), which is an example of a dataset where multiple data collection methods \hl{Eller måske lande?} have been employed.
+Next, in Section \ref{sec:pcadscintro}, we move on to a description of the PCADSC procedures, including a brief introduction to principal component analysis (PCA) in general.
+In Section \ref{sec:dataexample}, we present a worked data example using open source, online available data on psychological well-being in three European countries.
+More specifically, we compare data from Denmark, which has been repeatedly been rewarded "happiest country in the world", with data from Bulgaria and Sweden, respectively, to investigate whether or not psychological well-being - and thus happiness - is really a concept that is universal beyond country borders, or if such rankings of happiness are inherently meaningless.
+At last, we discuss limitations and merits of the PCADSC tools in Section \ref{sec:discussion}.
 
@@
 
-\section{PCADSC - description of the method}
+\section{PCA-based tools for data structure comparisons}
 \label{sec:pcadscintro}
-\hl{Description from Anne's master's thesis. Rewrite.}
-
 As mentioned above, the purpose of PCADSC is comparing overall data structures in two or more subsets of a dataset.
-But before we can get further into describing this procedure, we must first define what exactly is meant by "overall structure".
+But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context.
 One such definition is the structure of the covariance matrix of the dataset.
@@
 This gives it a very nice interpretation as a measure of the overall structure.
-If we do not accept the normality assumption, pairwise correlations and variable variances are still interesting quantities that say something about the interrelations between the variables, at least in terms of linear relationships.
-All in all, the empirical covariance matrix is a reasonable place to start looking for differences in "overall data structures".
+If we do not accept the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data.
+All in all, the empirical covariance matrix is a reasonable place to start looking for differences in \textit{overall data structures}.
 
-Though the idea sounds appealing, it is quite difficult to assess similarity of matrices, and moreover, this becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices.
+Though the idea sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices.
 There is simply too much information to consider at once.
-
-However, by 
-%clever 
-use of linear algebra, we can construct a decomposition of the covariance matrix that makes it easier to gain an overview of the data.
-We propose a new method based on principal component analysis that seems to be able to identify differences in datasets based on intuitive, visual inspections.
-We refer to this method as principal component analysis-based data structure comparison (PCADSC) and we present the procedure below.
-But first, we give a minimal introduction to principal component analysis in general with reference to Koch (2014).
+However, by use of linear algebra, we can decompose and recompose the covariance matrix, such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance.
+In particular, we find a new representation of the covariance matrix that makes it to gain an overview of the most interesting aspects of the data.
+On such decomposition strategy is using \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.
 
 \subsection{Principal component analysis}
-
 Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average\hl{(s?)} and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance \hl{Is $\bar{x} \in \RR^d$ with $\bar{x} = (\bar{x_1}, ..., \bar{x_d})^T$?}.
@@
 In this paper we propose three diagnostic plots for comparing the loading patterns in two datasets $x_{11},\dotsc,x_{1 n_1} \in \RR^d$ and $x_{21},\dotsc,x_{2 n_2} \in \RR^d$ in the same $d$ variables.
-The construction of these plots, which we call \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot}, proceeds in the following steps:
+The construction of these plots, which we call \textit{CE plot}, \textit{angle plot} and \textit{chroma plot}, proceeds in the following steps:
 \begin{enumerate}
@@
 
-Before describing the details of the  \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot} we make some general remarks.
+Before describing the details of the  \emph{CE plot}, the \emph{angle plot} and \emph{chroma plot}, we make some general remarks.
 First, it should be noted that PCA is sensitive to scaling, as the procedure deconstructs the covariance matrix in components according to the most explained variance.
@@
 
-\subsubsection{Description of the \emph{CumEigenPlot}}
- This plot compares the variation components. 
+\subsubsection{The cumulative eigenvalue plot}
+ The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the covariance matrix. 
 %If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
@@
 \end{align*}
-This may be seen as a cumulated Bland-Altman plot for the variation components (\hl{reference to cumulated residuals and to Bland-Altman}).
+This may be seen as a cumulative Bland-Altman plot for the variation components (\hl{reference to cumulative residuals and to Bland-Altman}).
 Note that due to the standardization the last point will be equal to $(d,0)$.
 Thus, this curve will begin and end at the x-axis.
-And the larger excursions it makes away from the x-axis the less alike the cumulated variation components for the two datasets are.
+And the larger excursions it makes away from the x-axis the less alike the cumulative variation components for the two datasets are.
 \hl{Is this an informative x-axis?
@@
 
-\subsubsection{Description of the \emph{HairPlot}} 
+\subsubsection{The angle plot} 
 This plot simultaneously compares the variation components and the loadings.
@@
 
-In the \emph{HairPlot} the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component for the first dataset in the coordinate system of the second dataset.
+In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component for the first dataset in the coordinate system of the second dataset.
 Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset.
@@
 And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells as well as arrows of non vanishing length in the off-diagonal cells.
-The name \emph{HairPlot} arise by seeing the arrows as \emph{hairs}.
-Using this interpretation we may say that a more ``hairy'' plot indicates less agreement between the two loading patterns.
 
-Along with the visual comparison we have also implemented numerical tests in step 3 \hl{Anne: Is a test really necessary here?
-Isn't the point of this plot to point us towards the most interesting components, not to quantify exactly how much they differ?
-I'm unsure as to whether a significant difference will necessarily be empirically/clinically relevant and likewise, if a non-significant difference will always be one we can disregard of...}.
-Separately for the off-diagonal grid cells we may use the lengths $\mu_{jk}$ and $\nu_{jk}$ as test statistics, and for the on-diagonal grid cells we use the distance between the scaled loadings as test statistics, that is
-\begin{equation*}
-\chi_k = \lambda_{\max}^{-1} \lvert \lambda_{1k} \eta_{1k} - \lambda_{2k} \eta_{2k} \rvert, \quad \text{for $k=1,\dotsc,d$.}
-\end{equation*}
-In total this gives $2d^2 - d$ tests, which should be corrected for the family wise type I error using a methods that doesn't assume independence among the tests.
-We also construct overall Kolmogorov-Smirnov and Cram\'er-von Mises type test statistics by
-\begin{align*}
-\text{KS}_1 &= \max_{j,k=1,\dotsc,d} (1_{j \neq k} \mu_{jk} + 1_{j=k} \chi_k), &
-\text{CvM}_1 &= \sum_{j=1}^d \sum_{k=1}^d (1_{j \neq k} \mu_{jk} + 1_{j=k} \chi_k)^2, \\
-\text{KS}_2 &= \max_{j,k=1,\dotsc,d} (1_{j \neq k} \nu_{jk} + 1_{j=k} \chi_k), &
-\text{CvM}_2 &= \sum_{j=1}^d \sum_{k=1}^d (1_{j \neq k} \nu_{jk} + 1_{j=k} \chi_k)^2.
-\end{align*}
-
-%\bigskip
-
-\subsubsection{Description of the \emph{PancakePlot}}
+\subsubsection{The chroma plot}
 This plot compares the loadings.
-The \emph{PancakePlot} consists of two parallel vertical barplots, one for each of the datasets, with one bar for each of the $d$ loading vector $\eta_{ij} \in \RR^d$.
+The \emph{chroma plot} consists of two parallel vertical barplots, one for each of the datasets, with one bar for each of the $d$ loading vector $\eta_{ij} \in \RR^d$.
 In the bars the relative contributions of the $d$ variables are visualized by separate colors for the different variables.
@@
 
-We not implemented a test statistic for the \emph{PancakePlot}, but in order to visualize the sample variation the observed plot can be shown together with resampled plots generated in step 4.
+We not implemented a test statistic for the \emph{chroma plot}, but in order to visualize the sample variation the observed plot can be shown together with resampled plots generated in step 4.
 
@@
 
-\hl{Mention somewere how we propose for people to use the plots: First CE, then hair, then pancake focusing in hairy components.}
+\hl{Mention somewhere how we propose for people to use the plots: First CE, then angle, then chroma focusing on interesting components.}
 
-\section{Data example stuff}
+\section{European differences in psychological well-being: A data example}
 \label{sec:dataexample}
-We will now turn to a concrete data example in order to illustrate the possibilities of each of the methods presented above \hl{more than just PCADSC?}.
-We use data from the 2012 version of the European Social Survey (ESS) project, a very large dataset that is freely available online at \url{www.europeansocialsurvey.org}.
+We will now turn to a concrete data example in order to illustrate the capabilities of the methods presented above.
+We use data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
+This investigation is motivated by an increasingly popular new tendency to publish miscellaneous rankings of countries in fields as different as educational quality (\hl{ref: PISA stuff}), \hl{XXX} and citizen happiness (\hl{ref OECD?}).
+From a methodological point of view, such international rankings are very concerning, as they rely on the fundamental assumption  that the measured concepts are inherently the same across countries.
+The PCADSC tools qualify as a suite of methods for exploring the validity of this assumption empirically.
+
+In the rankings of happiness, Denmark has repeatedly been awarded with the title of "happiest country" (\hl{ref?}.
+However, little critical discussion has been aimed at whether or not this title is in any way meaningful.
+Does there exist such a thing as a universal, internationally valid concept of happiness?
+Or do different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socioeconomic settings?
+This is in fact a question concerning comparability of data structures.
+If two countries differ e.g.
+in terms of how social networks are typically build and structured, with one emphasizing family relations and the other mostly focusing on other social relations, having a weak family connection does not have the same implications in the first country as it does in the second one.
+More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being.
+The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures.
+And therefore, comparing the two countries in these measures is not a meaningful endeavor.
+
+In this section we will use the PCADSC tools to unveil such differences between Denmark and Bulgaria.
+These two countries have previously been highlighted to be very different in terms of what defines happiness (\hl{REF: ESS6 Topline Results Series 5}).
+This might be the result of large, cultural and socioeconomic differences between the two countries.
+Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.
+
+We also compare the Danish data with Swedish data in order to illustrate that the PCADSC tools actually do have some discriminatory power.
+Denmark and Sweden are both Scandinavian countries and are often deemed very similar in terms of culture and history.
+Therefore, we expect fundamental concepts such as psychological well-being to similar across these two countries.
+
+%As Denmark has repeatedly been awarded with the title of "the happiest country", we have included this country
+
+%Specifically, we take on a critical perspective on
+
+%, a very large dataset that is freely available online at \url{www.europeansocialsurvey.org}.
 As the name suggests, the data comes from a survey that was conducted primarily in Europe aiming to collect information about the social conditions of the citizens \hl{reference?}.
@@
 \subsection{Data}
+%\begin{figure}
+%\center
+%\includegraphics[scale=1]{surveyTable.png}
+%\caption{\hl{Make a new table here}}
+%\label{surveyTable}
+%\end{figure}
+
+%\begin{table}[t]
+%\centering
+%\begin{tabular}{cl}
+%Wellbeing dimension & ESS item \\
+%\multirow{2}{*}{Evaluative wellbeing} & X \\ & Y \\ 
+%\end{tabular}
+%\end{table}
+
+\begin{table}[t]
+\centering
+\begin{tabular}{lccccccccc}
+  \hline
+  & \multicolumn{3}{c}{Denmark} & \multicolumn{3}{c}{Bulgaria} & \multicolumn{3}{c}{Sweden} \\
+ & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \\ 
+  \hline
+ Evaluative wellbeing & 8.00 & 8.75 & 9.50 & 3.50 & 5.00 & 7.00 & 7.00 & 8.00 & 9.00 \\ 
+ Emotional wellbeing & 7.22 & 8.33 & 8.89 & 5.00 & 6.67 & 7.78 & 6.67 & 7.78 & 8.89 \\ 
+Functioning & 6.93 & 7.57 & 8.21 & 5.50 & 6.68 & 7.68 & 6.39 & 7.04 & 7.68 \\ 
+Vitality & 6.67 & 7.50 & 8.33 & 5.83 & 7.50 & 8.33 & 6.67 & 7.50 & 9.17 \\ 
+ Community wellbeing & 5.83 & 6.77 & 7.57 & 3.70 & 4.67 & 5.70 & 5.66 & 6.57 & 7.37 \\ 
+Supportive relationships & 7.42 & 8.25 & 8.92 & 6.17 & 7.25 & 8.08 & 7.42 & 8.25 & 8.75 \\ 
+   \hline
+\end{tabular}
+\caption{The 1st quartile, the median and the third quartile of the distrubutions of each of the six dimensions of psychological well-being, stratified by country.
+Note that the scales are constructed such that they all run from 0-10.}
+\label{tableDistr}
+\end{table}
+
+The ESS 2012 data contains a total of 626 variables collected from 54673 citizens in 29 countries.
+Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
+These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}.
+More details on these scales is to be found in (\hl{ref: ESS6 Topline Results Series 5}).
+We represent each of these scales by a single variable, which is calculated as the average score within the items related to that variable (\hl{ref to someone that says that is sensible?}) and scaled such that it takes a value between 0 and 10.
+For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below.
+This gives us $n_{DK} = 1498$ observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$.
+Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
+
+\subsection{Comparing Denmark and Bulgaria}
 \begin{figure}
 \center
-\includegraphics[scale=1]{surveyTable.png}
-\caption{\hl{Make a new table here}}
-\label{surveyTable}
-\end{figure}
-The ESS 2012 data contains a total of 626 variables collected from 54673 citizens in 29 countries.
-Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
-These 35 items can be divided into 6 distinct scales (\hl{REF: ESS6 Topline Results Series 5}), as illustrated in Figure \ref{surveyTable}.
-We represent each of these scales by a single variable, which is calculated as the average score within the items related to that variable (\hl{ref to someone that says that is sensible?}).
-For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below.
-
-In the following, we only compare two countries, namely Denmark and Bulgaria, which have $n_{BG} = 1798$ and $n_{DK} = 1498$ complete cases in the variables of interest, respectively.
-For these two countries, the ESS authors \hl{different term?} particularly highlight differences in the relationship between the psychological well-being scales, at least on an nation-aggregated level (\hl{REF: ESS6 Topline Results Series 5}).
-This might be the result of large, cultural and socio-economic differences between the two countries.
-Simply put, we do not expect happiness and psychological well-being to be the same phenomena in Bulgaria and Denmark \hl{more here about why, about the cultures?}.
-Therefore, a successful method for data comparisons should be able to detect these differences by looking at the differences in the interplay between the 6 scales of psychological well-being.
-
-\subsection{PCADSC}
-\begin{figure}
-\center
-\includegraphics[scale = 0.8]{essPCADSC.pdf}
-\caption{\hl{something}}
-\label{plotESSPCADSC}
+\includegraphics[scale = 0.7]{essDKBGce.pdf}
+\includegraphics[scale = 0.7]{essDKBGhair.pdf}
+\caption{The CE plot (top) and the angle plot (bottom) resulting from comparing Bulgarian and Danish data on psychosocial well-being.
+Dataset 1 refers to the Bulgarian subsample, while Dataset 2 is the Danish data.
+The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the assumption of no difference in data structures.
+In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
+\label{plotBG.cehair}
 \end{figure}
@@
 \center
-\includegraphics[scale = 0.8]{essWallyPCADSC.pdf}
-\caption{\hl{something.}}
-\label{plotESSPCADSCWally}
+\includegraphics[scale = 0.7]{essDKBGpancake234.pdf}
+\caption{\hl{something}}
+\label{plotBG.pancake}
 \end{figure}
 
-\begin{figure}
-\center
-\includegraphics[scale = 0.8]{essHairplot.png}
-\caption{The blue arrows show the principal components of the first dataset decomposed in the coordinate system of the principal components of the second dataset. And the red arrows the reverse.}
-\label{plotESSHairplot}
-\end{figure}
+%\begin{figure}
+%\center
+%\includegraphics[scale = 0.7]{essDKBGWallyPCADSC234.pdf}
+%\caption{\hl{something.}}
+%\label{plotBG.wally}
+%\end{figure}
 
-\begin{figure}
-\center
-\includegraphics[scale = 0.8]{essCumVarAgreement.png}
-\caption{\hl{something.}}
-\label{plotCumVarAgreement}
-\end{figure}
+Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales.
+The CE plot show a remarkable degree of lacking comparability: The cumulative difference in the eigenvalue by far exceeds what could come about randomly if there really was no difference in the data structure.
+This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero.
 
-Figure \ref{plotESSPCADSC} illustrates the results of conducting PCADSC on the psychological well-being data from the ESS.
-While the first principal component, which is responsible for explaining 50-60 \% of the variance in the data, is very similar for the two countries, we see quite large differences in the remaining components.
+Moving on to the angle plot, we find that the differences are primarily to be found in the second, third and fourth principal components (PCs): The blue arrows visualize the decomposition of the principal components for the first dataset in the coordinate system of the second dataset.
+We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3.
+The red arrows visualize the decomposition of the principal components for the second dataset in the coordinate system of the first dataset.
+Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3.
+Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.
+
+ The chroma blot in Figure \ref{plotBG.pancake} allows us to look closer into these components.
+Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish.
+In the Danish data, on the other hand, we find that \textit{Vitality} and \textit{Emotional well-being} seem to play bigger roles, as they appear with larger loadings in more high-ranking components in this sample, relative to the Bulgarian.
+
+All in all, we find that psychological well-being does not seem to be the same concept in Bulgaria as in Denmark.
+The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalue) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle and chroma plots).
+In Bulgaria, interpersonal features seem to be more informative of psychological wellbeing, whereas in Denmark, individual characteristic play a relatively bigger role.
+Thus, the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics.
+\\
+
+\hl{I have dropped the Wally-plots for now, as I'm not really sure they are that useful.}
+
+% While the first principal component, which is responsible for explaining 50-60 \% of the variance in the data, is very similar for the two countries, we see quite large differences in the remaining components.
 In the second component, we see that the two countries disagree in the relative importance of the scales \textit{Community wellbeing} and \textit{Vitality}.
@@
 
-{\color{red}
-Figure \ref{plotESSHairplot} shows the hairplot.
-The blue arrows visualize the decomposition of the principal components for the first dataset in the coordinate system of the second dataset.
-We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3.
-The red arrows visualize the decomposition of the principal components for the second dataset in the coordinate system of the first dataset.
-We see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3.
+%{\color{red}
+%Figure \ref{plotESSHairplot} shows the hairplot.
+The blue arrows visualize the decomposition of the principal components for the %first dataset in the coordinate system of the second dataset.
+We see that PC2 also loads on PC3, that PC3 also loads on PC4, and %that PC4 also loads on PC2 and PC3.
+The red arrows visualize the decomposition of the principal components for the second dataset %in the coordinate system of the first dataset.
+We see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that %PC4 also loads on PC3.
 For PC1, PC5 and PC6 the main difference is in the size of the variation component.
-}
+%}
 
-In summary, we find that the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics.
-\\
-
-But did we really illustrate a data structure difference due to country differences or did we just illustrate the variability of the results of the PCADSC method?
+%But did we really illustrate a data structure difference due to country differences or did we just illustrate the variability of the results of the PCADSC method?
 In order to investigate this further, we look at a so-called \textit{Wally plot} (\hl{ref: Claus Ekstrøm}).
@@
 
+\subsection{Comparing Denmark and Sweden}
+\begin{figure}
+\center
+\includegraphics[scale = 0.7]{essDKSEce.pdf}
+\includegraphics[scale = 0.7]{essDKSEhair.pdf}
+\caption{A CE (top) and a hair (bottom) plot for comparing the Danish (Dataset 1) and the Swedish (Dataset 2) psychological well-being data.
+The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
+\label{plotSE.cehair}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[scale = 0.7]{essDKSEpancake.pdf}
+\caption{A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples.
+Note that the bars for each component is annotated with its cumulative variance score, that is, how much variance can be explained by having information of this and the preceding components.}
+\label{plotSE.pancake}
+\end{figure}
+
+%\begin{figure}
+%\center
+%\includegraphics[scale = 0.7]{essDKSEWallyPCADSC.pdf}
+%\caption{\hl{something.}}
+%\label{plotSE.wally}
+%\end{figure}
+
+We now turn to the comparison of Denmark and Sweden in terms of psychological well-being.
+Figure \ref{plotSE.cehair} show the CE and angle plots for these two countries.
+In the CE plot, we now find the cumulative eigenvalue curve to be just within the acceptance region of the null-hypothesis.
+This is also reflected by the two tests, which now produce $p$-values of $p_{KS} = 0.14$ and $p_{CvM} = 0.09$, respectively, thus accepting the null-hypothesis at the typical 5\% level, but not with overwhelming evidence.
+
+The angle plot from Figure \ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent.
+This implies that if one already has e.g.
+the information held in the first PC from the Danish data, this information is in itself mostly sufficient to describe the first PC of the Swedish data.
+
+Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80 \% of the variance in both datasets), and slight, but increasing, differences in the remaining three components.
+We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25 \% of the variance in both datasets.
+In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g.
+conclude using Table \ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular, the unhappy people in Denmark (represented by the 1st quartiles) are generally a lot happier in Denmark than they are in Sweden.
+A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.
+
 \section{Discussion}
@@
 \item Generalizing the results to non-numeric variables?
+	\begin{itemize}
+		\item Interpretation for binary variables?
+		\item Any meaningful way to allow for nominal, categorical variables?
+		\item ... ?
+	\end{itemize}
 \item Generalizing the results to covariance matrices that are not of full rank?
-\item ?
+\item Investigate sensitivity towards the sample sizes $n_1$ and $n_2$
+\item Limitations: What sorts of problems can never be found using PCADSC? 
+\begin{itemize}
+\item Differences in scaling, as we standardize all variables
+\item More?
+\end{itemize}
 \end{itemize}
@@
 
-\section{Concluding Remarks}
-\label{sec:conclusion}
+%\section{Concluding Remarks}
+%\label{sec:conclusion}
 

### 2017-03-19 ESS/ess_notes.R

@@ -37,7 +37,7 @@ for (v in sMat$var[sMat$scale == "Emotional wellbeing"])  {
 for (v in c("fltsd", "fltdpr", "fltanx")) {
   anData[, v] <- 3 - anData[, v]
 }
-anData$EmotionalWellbeing <- rowMeans(anData[, sMat$var[sMat$scale == "Emotional wellbeing"]])
+anData$EmotionalWellbeing <- (10/3)*rowMeans(anData[, sMat$var[sMat$scale == "Emotional wellbeing"]])
 
 #Functioning
 sapply(anData[, sMat$var[sMat$scale == "Functioning"]], range)
@@ -58,9 +58,9 @@ for (v in sMat$var[sMat$scale == "Vitality"])  {
 for (v in c("flteeff", "slprl", "cldgng")) {
   anData[, v] <- 3 - anData[, v]
 }
-anData$Vitality <- rowMeans(anData[, sMat$var[sMat$scale == "Vitality"]])
+anData$Vitality <- (10/3)*rowMeans(anData[, sMat$var[sMat$scale == "Vitality"]])
 
-#Community vellbeing
+#Community wellbeing
 sapply(anData[, sMat$var[sMat$scale == "Community wellbeing"]], range)
 anData$pplahlp <- (10/6)*anData$pplahlp
 anData$flclpla <- 10 - ((10/4)*(anData$flclpla - 1))
@@ -102,7 +102,7 @@ for (v in sMat$var[sMat$scale == "Emotional wellbeing"])  {
 for (v in c("fltsd", "fltdpr", "fltanx")) {
   anData[, v] <- 3 - anData[, v]
 }
-anData$EmotionalWellbeing <- rowMeans(anData[, sMat$var[sMat$scale == "Emotional wellbeing"]])
+anData$EmotionalWellbeing <- (10/3)*rowMeans(anData[, sMat$var[sMat$scale == "Emotional wellbeing"]])
 
 #Functioning
 sapply(anData[, sMat$var[sMat$scale == "Functioning"]], range)
@@ -123,9 +123,9 @@ for (v in sMat$var[sMat$scale == "Vitality"])  {
 for (v in c("flteeff", "slprl", "cldgng")) {
   anData[, v] <- 3 - anData[, v]
 }
-anData$Vitality <- rowMeans(anData[, sMat$var[sMat$scale == "Vitality"]])
[…]

### 2017-03-23 article/article_v0.9.tex
Edited section about PCADSC tools (now section 2)

@@
 This is due to vast amounts of data being registered online and to a trend towards more open source research.
-The latter phenomenon in particular poses new challenges wrt.
-data quality assessment.
+The latter phenomenon in particular poses new challenges wrt.\ data quality assessment.
 When data are collected and made public without a specific end-point in mind, how do we ensure that differences in, say, choice measurement instruments, mode of administration, or sampling frame do not cause the data to be effectively divided into subsets that are simply not comparable?
+Three examples of this are
 
-Surveys often use mixed modes of administration, e.g.
+\begin{itemize}
+\item Surveys often use mixed modes of administration, e.g.
 mail and telephone, and while this can improve response rates, the mode of administration can affect results \cite{Brambilla1987,McHorney1994} and differences in response behavior can lead to biased results.
 Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful.
+\item The rapid growth of web surveys, due to low cost, timeliness, and other factors, generate large data sources that lack a sampling frame of the general population.
+However, it can be problematic to combine online panels (pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions) \cite{Liu2016}.
+\item Large scale open source data sets such as the PISA data and ESS (European Social Survey) data.
+In these data sets, data analysts are far away from the data producers and problem-specific recommendations about potential instrument-induced challenges in the data sets are not available for the data analysts.
+\end{itemize}
 
-The rapid growth of web surveys, due to low cost, timeliness, and other factors, generate large data sources that lack a sampling frame of the general population.
-However, it can be problematic to combine online panels (pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions) \cite{Liu2016}.
+\hl{Maybe something here about what PCA is usually used for, e.g. dimension reduction, exploratory data analysis, ...}
+
+Assume that we have two data sets with the same variables, but different observations represented as a data set with
+a subset-inducing variable.
+Assume next that we wish to compare structures without specifying a model, or even a variable of interest.
+The central question is whether the two data sets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity across the subset division.
 
 Sophisticated methods for addressing this question are available when we are willing to assume a statistical model, but when these models are taken away, a remarkable void of methods is left behind.
-What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
+What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a data set without assuming neither directional nor hierarchical relationships between the variables.
+Using parametric models is not a general data structure comparison method, but rather a fitted-model comparison method.
+It addresses the interplay between the model and the data, not the data alone.
+Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and not to the interplay between variables.
+Entry-by-entry comparison of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase.
+For these methods significant $p$-values are likely to abound as the sample size increases.
+Parametric models using, e.g., latent variable models moves beyond the marginal approach, but need a pre-specified model.
+
 We propose a suite of three new tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC).
@@
 Software for using these tools is available in an \texttt{R}-package.
-\hl{how to refer to our package?}
+\hl{how to refer to our package?} This yields a solution that is largely independent of the sizes of the two subsets of data.
+Thereby, a lot of methods that compare each subset to the full data set in some sense are excluded.
 
-This manuscript is structured as follows: First, in Section \ref{sec:stateoftheart}, we present the data structure comparison problem in more detail and discuss what statistical methods are already available for solving similar challenges.
-Next, in Section \ref{sec:pcadscintro}, we move on to a description of the PCADSC procedures, including a brief introduction to principal component analysis (PCA) in general.
+This manuscript is structured as follows: First, in Section Section \ref{sec:pcadscintro}, we describe the PCADSC procedures, including a brief introduction to principal component analysis (PCA) in general.
 In Section \ref{sec:dataexample}, we present a worked data example using open source, online available data on psychological well-being in three European countries.
-More specifically, we compare data from Denmark, which has been repeatedly been rewarded "happiest country in the world", with data from Bulgaria and Sweden, respectively, to investigate whether or not psychological well-being - and thus happiness - is really a concept that is universal beyond country borders, or if such rankings of happiness are inherently meaningless.
-At last, we discuss limitations and merits of the PCADSC tools in Section \ref{sec:discussion}.
-
-\section{Something about state of the art}
-\label{sec:stateoftheart}
-\subsection{\hl{More detailed description of the type of problem we wish to address}}
-\hl{
-\begin{itemize}
-\item Two subsets of a dataset, i.e. to datasets with the same variables, but different observations
-\item Wish to compare structures without specifying a model or even any variables of interest
-\item The most central example is the question of whether the two subsets can readily be combined in a (unknown) data analysis, or if the subset-inducing variable actually implies heterogeneity across the subset division
-	\begin{itemize}
-		\item Examples: Large scale open source datasets such as the PISA data and ESS (European Social Survey) data and ...(?).
-In these datasets, the data producers are very far away from the majority of the data analysts.
-Therefore, problem-specific recommendations about potential instrument-induced challenges in the datasets are not available for the data analysts.
-How can data producers ensure that this will not be an issue, at least not related to known data gathering differences?
-		\item Other examples?
-		\item Perhaps a description of what happens if we are to combine the two subsets of the datasets without taking a e.g.
-an instrument-effect into account.
-When will it cause problems (maybe: causal graph style?)?
-		\item Mention somewhere: We want a solution that is largely independent of the sizes of the two subsets of data.
-Thereby, a lot of methods that compare each subset to the full dataset in some sense are excluded.
-	\end{itemize}
-\end{itemize}
-}
-
-\subsection{\hl{Describe existing methods used to solve similar questions or parts of the question we are addressing}}
-\hl{
-\begin{itemize}
-\item The simplest case: variable-by-variable tests in distributional differences
-	\begin{itemize}
-		\item Simple, but scales poorly
-		\item Only relates to marginal differences and not to the interplay between variables
-	\end{itemize}
-\item Karl's papers?
-\item Anne's papers: IRT-based methods for surveys
-	\begin{itemize}
-		\item Moves beyond the marginal approach, but needs a model pre-specified
-		\item Thus, it is not a general data structure comparison method, but rather a fitted-model comparison method.
-It addresses the interplay between the model and the data, not the data alone.
-This is fundamentally a different (though related) question.
-	\end{itemize}
-\end{itemize}
-}
+More specifically, we compare data from Denmark, which has repeatedly been rewarded "happiest country in the world", with data from Bulgaria and Sweden, respectively, to investigate whether or not psychological well-being - and thus happiness - is really a concept that is universal beyond country borders, or if such rankings of happiness are inherently meaningless.
+At last, we discuss limitations and merits of the PCADSC tools in Section \ref{sec.Discussion}.
 
@@
 
-Though the idea sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices.
+A naive approach for data structure comparisons might therefore be to compute the empirical covariance matrices on each of the two data subsets and simply compare these matrices.
+Though the idea perhaps sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices.
 There is simply too much information to consider at once.
-However, by use of linear algebra, we can decompose and recompose the covariance matrix, such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance.
-In particular, we find a new representation of the covariance matrix that makes it to gain an overview of the most interesting aspects of the data.
+However, by use of linear algebra, we can decompose and recompose the covariance matrix such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance.
+In particular, we find a new representation of the covariance matrix that makes it possible to gain an overview of the most interesting aspects of the data.
 On such decomposition strategy is using \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.
 
-\subsection{Principal component analysis}
-Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average\hl{(s?)} and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance \hl{Is $\bar{x} \in \RR^d$ with $\bar{x} = (\bar{x_1}, ..., \bar{x_d})^T$?}.
-Suppose that we want to describe the observations by $q$ numbers instead of the original $d$ numbers.
-The associated \emph{rank-q-reconstruction error} is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
+\subsection{A brief introduction to Principal component analysis}
+Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = (\bar{x}_1, ..., \bar{x}_d)^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$.
+We assume in the following that all variables of $X$ have a numerical interpretation, e.g.
+by being continuous or ordinal and categorical.
+For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost.
+More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the orignial $d$-dimensional one.
+Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
 \begin{equation*}
@@
 \end{equation*}
-\emph{Principal component analysis} (PCA) ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error.
-Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of $S$.
-Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ in the columns.
-The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different.
-If some of the eigenvalues are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
-In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero \hl{Anne: What about covariance matrices that are not of full rank (e.g.
-because of colinearity)?
-Won't we find eigenvalue multiplicities $> 1$ here as well?}.
-It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
+and the theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error.
+\hl{Reference here?
+Maybe to like original Pearson article?}
+
+More specifically, the rank-q-reconstruction error is attained when we choose
 \begin{equation*}
@@
 \end{equation*}
-and equals $\sum_{j=q+1}^d \lambda_j$.
-The eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}.
-The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}.
-The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores:
+where $\eta_1, ..., \eta_q$ are the $q$ first eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, ..., \lambda_d$.
+In the PCA framework, we refer to the eigenvectors as \textit{loadings}.
+The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
+$$\text{trace}(S) = \sum_{j=1}^d \hat{V}(X_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j$$
+where $X_j \in \RR^n$ denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
+This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
+And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different, the representation hereby obtained is a valid object of inference.
+If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_0$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
+
+The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
 \begin{align*}
@@
 \end{align*}
-\hl{Anne: Should we maybe assume standardized $x$s here also?
-I feel like it is part of the PCA algorithm and subtracting the mean in the above seems off for that reason.
-Also: Something is wrong with the dimensions here.
-$u$ is $d \times 1$, $x_i$ is $d \times 1$(?) - how do we get $\eta_j$ to also be $d \times 1$?} It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
+where the initial subspace is defined as $\hat{K}_0 = \{0\}$.
+It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
 This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$.
-This implies that the loadings and variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension $q$.
+This implies that the loadings and the variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension, $q$, a priori.
 
-%Principal component analysis (PCA) allows us to decompose a data matrix consisting of $n$ observations of $d$ variables into $d$ orthogonal vectors of length $d$.
-This is in itself not very impressive nor useful.
-But PCA does not result in just any basis of the data matrix.
-The basis chosen using PCA has the very favorable property that each vector is chosen such that they maximize the accumulated explained variance.
-These vectors are usually denoted \textit{principal component scores} or simply \textit{principal components} and they are constructed as linear combinations of the variables in the dataset (or, equivalently, as projections of the data matrix).
-The coefficients of each variable in these projections are denoted \textit{loadings}.
-Let $S \in \RR^{d \times d}$ denote the empirical covariance matrix and let $X \in \RR^{n \times d}$ denote our observed data matrix.
-$d$ is the number of variables while $n$ is the number of observations.
-The deconstruction procedure then goes as follows:
-%\begin{enumerate}
-%\item The vector of loadings corresponding to the first principal component, $\eta_1 \in \RR^d$, is chosen such that the component explains as much variance as possible.
-More precisely, for any unit vector $u \in \RR^d$ we find that $\V(u^T X^T)$ is maximized when $u = \eta_1$.
-Let $\lambda_1 = \V(\eta_1^T X^T)$.
-%\item The second vector of loadings is then chosen as $\eta_2 = \argmax_{u: \, u \perp \eta_1} \V(u^T X^T)$ and we again define $\lambda_2 := \V(\eta_2 X)$.
-%\item The third vector of loadings is $\eta_3 = \argmax_{u: \, u \perp \{\eta_1, \eta_2\}} \V(u^T X^T)$ and $\lambda_3 :=  \V(\eta_3^T X^T)$ \\
-%\vdots
-%\item[\refstepcounter{enumi} $d$.] The $d$th vector of loadings is $\eta_d = \argmax_{u: \, u \perp \{\eta_1, ..., \eta_{d-1}\}} \V(u^T X^T)$ and $\lambda_d = V(\eta_d^T X^T)$.
-%\end{enumerate}
-%The $\eta_i$s and $\lambda_i$s are determined uniquely (up to scaling) and they are given as the eigenvectors and eigenvalues of $S$, respectively.
-Moreover, we find that $X = \sum_{i=1}^d \eta_j \eta_j^T X$, so the procedure is really just a clever choice of rotation of the data, such that it is split up into vectors that are ordered by their variances, or more precisely, their variance contributions.
-We define the contribution to the total variance of the $k$th principal component as $\frac{\lambda_k}{tr(S)}$ and please note that, due to the properties of eigenvalue decomposition, $tr(S) = \sum_{i=1}^d \lambda_i$, which is the sum of the variances of the variables in the dataset.
-%
-% In this introduction, we have assumed full rank of the covariance matrix $S$ for simplicity, but it should be noted that this assumption is in no way needed for the results to hold.
-
-\subsection{PCA-based data structure comparisons}
-Above, we promised a method for intuitive, visual inspection of data structure similarities, but as of now, all intuition might have been lost in technicalities.
-The main point we want to emphasize from PCA is that whereas the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables.
-If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both data situations.
+\subsection{Using PCA for data structure comparisons}
+All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful quantities to compare across such different datasets.
+Usually, when performing PCA with other purposes in mind, the main interest lies in the \textit{scores}, i.e.
+the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings.
+But where the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables.
+If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both datasets.
 This can be the case while the two sets of scores could be arbitrarily different, which e.g.\ could happen if the two datasets were taken from two different populations of subjects.
@@
 
-%Specifically, we can interpret the loadings as weights that determine the relative influence of each variable in each principal component.
-
-%the two data sets will agree on which variables explain the most variance and also how this variance is explained.
-Looking at the PCA loadings and the cumulative explained variance thus provides a straightforward non-parametric graphical approach for assessing similarity between two datasets.
-
-In this paper we propose three diagnostic plots for comparing the loading patterns in two datasets $x_{11},\dotsc,x_{1 n_1} \in \RR^d$ and $x_{21},\dotsc,x_{2 n_2} \in \RR^d$ in the same $d$ variables.
-The construction of these plots, which we call \textit{CE plot}, \textit{angle plot} and \textit{chroma plot}, proceeds in the following steps:
+The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables.
+We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$.
+For each of these two datasets, we complete the following steps:
 \begin{enumerate}
-\item Standardized the observations $x_{i1},\dotsc,x_{i n_i}$ to have zero mean and unit standard deviation separately for each of the $d$ variables within each datasets.
-Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized observations, and form the principal component analysis
+\item Standardized each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized datasets. 
+\item Form the principal component analysis
 \begin{equation*}
@@
 \end{equation*}
-\item Combine the standardized datasets into a joint dataset $\tilde{x}_1,\dotsc,\tilde{x}_n \in \RR^d$ with $n=n_1+n_2$ observations, standardize this dataset, and form the principal component analysis
-\begin{equation*}
-S = \frac{1}{n-2} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top.
-\end{equation*}
-\item Make the diagnostic plot of interest. Details of the plots will be described below.
-\item To investigate the sample variation randomly reallocate the combined and standardized dataset formed in step 2 to two datasets of $n_1$ and $n_2$ observations, respectively, and redo step 1 and 3.
-This is done several times, e.g.\ 10000 times, and compared to the plot for the original allocation.
+thereby obtaining loadings $\eta_{ik}$ and variance components $\lambda_{ik}$ for $i \in \{1, 2\}$ and $k \in 1, ..., d$.
 \end{enumerate}
+The hereby obtains PCA decompositions of the covariance matrices can then be compared.
+We present three diagnostics plots that are designed to shine a light on different types and nuances of data structure differences.
+These three plots are
 
-Before describing the details of the  \emph{CE plot}, the \emph{angle plot} and \emph{chroma plot}, we make some general remarks.
-First, it should be noted that PCA is sensitive to scaling, as the procedure deconstructs the covariance matrix in components according to the most explained variance.
+\begin{description}
+\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA.
+The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
+\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e.
+loading and variance component pairs) that are most similar and most different across the two datasets
+\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
+\end{description}
+For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.
+
+But before diving deeper into the details of the \emph{CE plot}, the \emph{angle plot} and the \emph{chroma plot}, a general remark about standardization in PCA is in place.
+We would like to emphasize that PCA is sensitive to scaling, as the procedure deconstructs the covariance matrix in components according to the most explained variance.
 This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data.
@@
 Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices.
-\hl{Do we still need more about this topic?
-Then we can add something like:
-\begin{itemize}
-\item Following the greedy interpretation of PCA, we know that the first eigenvector is chosen as
-$$\eta_1 = \argmax_{u \in \RR^d} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2  = \text{argmax}_{u \in \RR^d} \tilde{\V}(u^T X^T) $$
-\item Now, imagine that $\V(x_1) >> \V(x_j)$ for all $j \in 2, ..., d$.
-Note that $(1, 0, ..., 0) X^T = x_i$.
-Then, clearly a vector $u$ that only picks out the first variable from the data matrix $X$, namely $u^T = (1, 0, ..., 0)$ will allow us to only look at the variance of $x_1$, which is very large.
-This has nothing to do with the interplay between variables, as the phenomenon will disappear if we scale $x_1$
-\item A more formal explanation could come from this kind of intuition, if needed.
-\end{itemize}
-}
-The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1.
-This also holds for the reallocated datasets, and in particular we have $\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$.
-The random reallocations in step 4 provide resamples under the null hypothesis that the two sample populations have the same structure in the $d$ variables, except that the two populations are allowed to have the own standard deviations for each of the variables.
-This exception is induced by the initial step 1, which performs a standardization within the two datasets before they are combined in step 2.
-These resamples can be used for a visualization of the sample variation, and if a test statistic can be defined in step 3, then a $p$-value for the null hypothesis may be computed.
-Such $p$-values are know as permutation tests for Goodness-of-Fit (\hl{reference?}).
-
-%In order to describe these plots we consider two different datasets in the same $d$ variables and with $n_1$ and $n_2$ observations, respectively.
-For dataset $i=1,2$ let $S_i \in \RR^{d \times d}$ be the empirical correlation matrix \hl{(find a way to introduce the empirial correlation matrix instead)}, and let $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0 $ and $\eta_{i1},\dotsc,\eta_{id} \in \RR^{d}$ be the corresponding variation components and loadings.
-The correlation matrices correspond to the covariance matrices for the variables after these have been standardized to unit standard deviation separately within the two datasets.
-Similarly, let $S \in \RR^{d \times d}$ be the empirical correlation matrix for the combined dataset with $n_1+n_2$ observations, and let $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ and $\eta_1,\dotsc,\eta_d \in \RR^d$ be the corresponding variation components and loadings.
-
-\hl{To do: Something about what to do if not all variables are numerical. Anne: Maybe leave this for the discussion?
-%Using correlation matrices is not the same as assuming standadized variables; there is a difference in the permutations.
-}
-
-\medskip
+The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that $\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$.
 
@@
 And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
-In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets we plot a piecewise linear curve connecting the points
+In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
 \begin{align*}
@@
 This may be seen as a cumulative Bland-Altman plot for the variation components (\hl{reference to cumulative residuals and to Bland-Altman}).
-Note that due to the standardization the last point will be equal to $(d,0)$.
+Note that due to the standardization, the last point will always be equal to $(d,0)$.
 Thus, this curve will begin and end at the x-axis.
-And the larger excursions it makes away from the x-axis the less alike the cumulative variation components for the two datasets are.
-\hl{Is this an informative x-axis?
-Should we maybe consider scaling it?}
+And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are.
+Moreover, a positive cumulative differences implies that dataset 1 holds more information in the first components than dataset 2 does.
 
-In step 3 we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics given by
+In order to test whether these cumulative differences are statistical artifacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
 \begin{align*}
@@
 \end{align*}
-\hl{Maybe more here on what the null hypothesis and range of these test statistics are?} 
-From the random reallocations in step 4 we get $p$-values for the Kolmogorov-Smirnov and the Cram\'er-von Mises Goodness-of-Fit tests.
-Step 4 is also used to visualize the sampling variation in a display, where we plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95 \% coverage intervals.
+We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and standardized datasets into two new datasets of $n_1$ and $n_2$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics.
+This should be done a large (e.g.
+10000) number of times.
+Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to even larger test statistics than the one we found for the original datasets.
+
+The permutation test results are also used to visualize the uncertainty of the CE curve in the plots.
+In the CE plots shown in the following section, we plot the observed curve together with 20 of the resampled curves, as well as a shaded region visualizing pointwise 95 \% coverage intervals.
 If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicates differences between the two datasets.
@@
 In the \emph{CumEigenPlot} we make 1000 independent random reallocations, and plot  P-values for the null hypothesis that the variation components are the same are easily provided for any appropriate test statistic by the resampling procedure as well.
-
-\bigskip
 
@@
 In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell.
-The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ anticlockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal.
-To facilitate the following description we will refer to the arrows drawn anticlockwise as the blue arrows, and the arrows drawn clockwise as the red arrows.
+The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal.
+To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows.
 The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
@@
 \end{align*}
-\hl{Maybe emphasize inner product/angle-interpretation here?}
+Note that for two $d$-dimensional, unit length vectors $a$ and $b$, $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the sample correlation.
+Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance component contributions.
 The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors.
-This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using anticlockwise and clockwise shifts from the diagonal.
+This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal.
 Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.
@@
 Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
-And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells as well as arrows of non vanishing length in the off-diagonal cells.
+And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.
 
 \subsubsection{The chroma plot}
-This plot compares the loadings.
-The \emph{chroma plot} consists of two parallel vertical barplots, one for each of the datasets, with one bar for each of the $d$ loading vector $\eta_{ij} \in \RR^d$.
-In the bars the relative contributions of the $d$ variables are visualized by separate colors for the different variables.
-Due to the indeterminacy of the sign all the signs are removed from the coefficients in the loadings, and the relative contribution is defined as the resulting coefficients divided by their total sum.
-The bars are ordered according to the variation components, and we also insert the cumulative variation components next to the barplots.
+This plot compares the loadings of the two datasets.
+The chroma plot consists of two panels, one for each dataset, made up of colored bars.
+These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions.
+More specifically, when illustrating the $i$th principal component, we plot a vertical bar of length one that has been divided into $d$ segments of different colors and the wideness of the $j$th such segment is given by
+$$\omega_{ij} = \frac{|(\eta_i)_j|}{\sum_{k=1}^d |(\eta_i)_k|}$$
+where $(\eta_i)_j$ denotes the $j$th entry of $\eta_i$.
+Due to the indeterminacy of the sign, all the signs are removed from the coefficients in the loadings.
+The bars are ordered according to the variation components and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance component contributions,
+$$\tilde\sigma^c_i = \frac{\sum_{j = 1}^i \lambda_i}{\sum_{k=1}^d \lambda_k} = \frac{\sum_{j = 1}^i \lambda_i}{d}$$
+Especially when $d$ is large, we recommend plotting only a select set of interesting principal components, e.g.
+as identified by use of the angle plot.
+In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_i = \frac{\lambda_i}{d}$.
+
 The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.
-\hl{Refer to example/show plot.
-Also, from Anne to Anne: A bit further elaboration here, especially mention the loadings as weights interpretation}
-
-We not implemented a test statistic for the \emph{chroma plot}, but in order to visualize the sample variation the observed plot can be shown together with resampled plots generated in step 4.
+For each component, the loadings describe how influential the different variables are on that component.
+Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such that \textit{"variable $x$ is generally more influential in subset 1 than it is in subset 2"}, thereby helping us to understand where and why the data structure differences are found.
 
@@
 %\end{enumerate}
-
-\hl{Mention somewhere how we propose for people to use the plots: First CE, then angle, then chroma focusing on interesting components.}
 
@@
 \section{Discussion}
-\label{sec:discussion}
+\label{sec.Discussion}
+When deciding whether to combine two data sets for analysis, the issue of heterogeneity across data sets must be addressed.
+Simple methods suffer drawbacks and will likely scale poorly.
+Parametric models using, e.g., latent variable models moves beyond the marginal approach, but need a pre-specified model.
+
+New tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if the two two data sets can be combine for analysis were proposed and discussed in the paper.
+They employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences yielding a solution that is largely independent of the sizes  the two data sets.
+
 \hl{
-\begin{itemize}
-\item Generalizing the results to non-numeric variables?
-	\begin{itemize}
-		\item Interpretation for binary variables?
-		\item Any meaningful way to allow for nominal, categorical variables?
-		\item ... ?
-	\end{itemize}
-\item Generalizing the results to covariance matrices that are not of full rank?
-\item Investigate sensitivity towards the sample sizes $n_1$ and $n_2$
-\item Limitations: What sorts of problems can never be found using PCADSC? 
-\begin{itemize}
-\item Differences in scaling, as we standardize all variables
-\item More?
-\end{itemize}
-\end{itemize}
+The methodology is quite general and in principle any plot that can be made can be included ('di-scree plot') ... [rephrase!!]
 }
+
+Further topics need to be addressed.
+These include generalizing the methods to be able to address (i) binary, ordinal or even nominal categorical variables, (ii) covariance matrices that are not of full rank.
+
+More evaluation of the performance is also needed. Investigating sensitivity towards the sample sizes $n_1$ and $n_2$
+
+The limitations of the PCADSC procedures should also be sudied in more detail.
+It seems unlikely that the procedure would be able to disclose differences in scaling, since all variables are standardized in the procedure.
+This type of heterogeneity should thus be adjusted for in later analyses of the combined data set.
 

### 2017-03-23 article/article_v0.9.tex
Edited section about PCADSC tools (now section 2)

@@
 The bars are ordered according to the variation components and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance component contributions,
-$$\tilde\sigma^c_i = \frac{\sum_{j = 1}^i \lambda_i}{\sum_{k=1}^d \lambda_k} = \frac{\sum_{j = 1}^i \lambda_i}{d}$$
+$$\sigma_c = \frac{\sum_{j = 1}^i \lambda_i}{\sum_{k=1}^d \lambda_k} = \frac{\sum_{j = 1}^i \lambda_i}{d}$$
 Especially when $d$ is large, we recommend plotting only a select set of interesting principal components, e.g.
 as identified by use of the angle plot.
-In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_i = \frac{\lambda_i}{d}$.
+In this scenario, the annotations should rather be the non-cumulative variance contributions, $\frac{\lambda_i}{d}$.
 

### 2017-03-26 article/article_v0.9.tex
Edited section 3 (data example)

--- 2017-03-23 article/article_v0.9.tex
@@
 \usepackage{multirow}
+%\usepackage{colortbl}
 %\usepackage[backend=biber,natbib=true, style=authoryear-comp]{biblatex}
+\usepackage[table]{xcolor}
+\definecolor{lightgray}{gray}{0.9}
 
@@
 \DeclareMathOperator*{\V}{V}
-\newcommand{\argmax}{\text{argmax}}
+\DeclareMathOperator*{\argmax}{argmax}
+%\newcommand{\argmax}{\text{argmax}}
 \newcommand{\R}[1]{\texttt{#1}}
@@
 %\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
-\title{Exploratory data structure comparisons: Three new tools based on Principal Component Analysis}
+\title{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}
 
@@
 
-%\RepeatTitle{Psychometrics: From Practice to Theory and Back}
+%\RepeatTitle{Exploratory data structure comparisons: Three new tools based on Principal Component Analysis}
 
@@
 \begin{keywords}
-keywords...
+Principal component analysis, multi-center studies, data structure comparisons \hl{Is this how you do keywords? Karl? Bo?}
 \end{keywords}
@@
 \hl{Maybe something here about what PCA is usually used for, e.g. dimension reduction, exploratory data analysis, ...}
+
+\hl{Maybe broaden the raison d'etre of PCADSC a bit: It is also a useful tool for doing initial, exploratory data analysis and it helps us to identify potential problems or challenges early in the study design development, before concrete model choices have been made, thereby aiding a more economic use of resources and a less ad-hoc approach to the types of problems that PCADSC can find.}
 
@@
 We present three diagnostics plots that are designed to shine a light on different types and nuances of data structure differences.
-These three plots are
+These three plots are:
 
@@
 Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices.
-The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that $\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$.
+The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
+$$\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$$
+which will simplify some expressions below.
 
@@
 \ldots, &&
-\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg).
+\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
 \end{align*}
@@
 We use data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
-This investigation is motivated by an increasingly popular new tendency to publish miscellaneous rankings of countries in fields as different as educational quality (\hl{ref: PISA stuff}), \hl{XXX} and citizen happiness (\hl{ref OECD?}).
+This investigation is motivated by an increasingly popular new tendency to publish miscellaneous rankings of countries in fields as different as educational quality (e.g.
+the PISA tests) and citizen happiness (e.g.
+the UN \textit{World Happiness Report} project).
 From a methodological point of view, such international rankings are very concerning, as they rely on the fundamental assumption  that the measured concepts are inherently the same across countries.
@@
 
-In the rankings of happiness, Denmark has repeatedly been awarded with the title of "happiest country" (\hl{ref?}.
-However, little critical discussion has been aimed at whether or not this title is in any way meaningful.
-Does there exist such a thing as a universal, internationally valid concept of happiness?
-Or do different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socioeconomic settings?
+For international comparisons of educational systems, the PISA tests have repeatedly been criticized for not being meaningful objects for international comparisons, especially due to problems with differential item functions (\cite{Kankaras2014}; \cite{Kreiner2014}) and translation problems (\cite{Asil2016}).
+\hl{Til Karl: Måske kender du nogle flere studier her, vi bør nævne?}
+
+In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though \cite{Veenhoven2012} presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability and \cite{Lolle2016} shows highly potent translation issues for the term \textit{happiness}.
+The main question is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
+Or do different aspects of psychological well-being or happiness simply not have the same relative meaning in different cultural and socioeconomic settings?
 This is in fact a question concerning comparability of data structures.
@@
 And therefore, comparing the two countries in these measures is not a meaningful endeavor.
+In this paper, we will focus on a single aspect of overall happiness, namely psychological well-being,
 
-In this section we will use the PCADSC tools to unveil such differences between Denmark and Bulgaria.
-These two countries have previously been highlighted to be very different in terms of what defines happiness (\hl{REF: ESS6 Topline Results Series 5}).
-This might be the result of large, cultural and socioeconomic differences between the two countries.
+In this section, we use the PCADSC tools to unveil international differences in one aspect of happiness, namely psychological well-being.
+Our starting point is Denmark, a small, northern European country that has repeatedly been awarded with the title of "happiest country in the world"  by the \textit{World Happiness Report}, most recently in 2016 (\cite{WHR2016}), and we wish to investigate if this title is really meaningful at all.
+In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria.
+Though both countries are European and thus not geographically nor culturally as far apart as some other countries might be, these two countries have previously been highlighted to be very different in terms of what defines happiness (\cite{ESStopline5} {\hl{Is this the correct way to refer to a technical report?}).
+Moreover,intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated (\cite{Rodriguez2014}), with a much less strong relationship between the two in Northern- compared to other European countries.
+In particular, interpersonal relations should play a less important role in Denmark, compared to Bulgaria.
 Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.
@@
 
-%As Denmark has repeatedly been awarded with the title of "the happiest country", we have included this country
-
-%Specifically, we take on a critical perspective on
-
-%, a very large dataset that is freely available online at \url{www.europeansocialsurvey.org}.
-As the name suggests, the data comes from a survey that was conducted primarily in Europe aiming to collect information about the social conditions of the citizens \hl{reference?}.
-As with all international (or, simply, multi-center) studies, one might be concerned about whether or not the data from different countries can readily be combined.
-This is the question we will address using PCADSC in the current section.
-
-All computations and figures presented in this section where created using our \R{R} package \R{PCADSC}, which is available online at \url{www.github.com/AnnePetersen1/PCADSC} \hl{maybe do a CRAN submission instead?}.
+All computations and figures presented in this section where created using our \R{R} package \R{PCADSC}, which is available online at \url{www.github.com/AnnePetersen1/PCADSC}.
 
 \subsection{Data}
-%\begin{figure}
-%\center
-%\includegraphics[scale=1]{surveyTable.png}
-%\caption{\hl{Make a new table here}}
-%\label{surveyTable}
-%\end{figure}
-
-%\begin{table}[t]
-%\centering
-%\begin{tabular}{cl}
-%Wellbeing dimension & ESS item \\
-%\multirow{2}{*}{Evaluative wellbeing} & X \\ & Y \\
-%\end{tabular}
-%\end{table}
 
@@
 
-The ESS 2012 data contains a total of 626 variables collected from 54673 citizens in 29 countries.
+The ESS 2012 data contains a total of 626 variables collected from 54673 citizens of 29 countries.
 Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
 These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}.
-More details on these scales is to be found in (\hl{ref: ESS6 Topline Results Series 5}).
-We represent each of these scales by a single variable, which is calculated as the average score within the items related to that variable (\hl{ref to someone that says that is sensible?}) and scaled such that it takes a value between 0 and 10.
+More details on these scales can be found in (\cite{ESStopline5}) and the relationship between questionnaire items and scales is summarized in Table \ref{table:items} in the appendices.
+We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10.
 For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below.
-This gives us $n_{DK} = 1498$ observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$.
+This gives us $n_{DK} = 1498$ observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations.
 Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
@@
 \includegraphics[scale = 0.7]{essDKBGpancake234.pdf}
-\caption{\hl{something}}
+\caption{A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data.
+The component-bars are annotated with their relative variance contributions (denoted $\tilde\sigma_i$ in the above).}
 \label{plotBG.pancake}
@@
 Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales.
-The CE plot show a remarkable degree of lacking comparability: The cumulative difference in the eigenvalue by far exceeds what could come about randomly if there really was no difference in the data structure.
+The CE plot show a remarkable degree of lacking comparability: The cumulative differences in the eigenvalues by far exceed what could come about randomly if there really were no difference in the data structures.
 This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero.
 
-Moving on to the angle plot, we find that the differences are primarily to be found in the second, third and fourth principal components (PCs): The blue arrows visualize the decomposition of the principal components for the first dataset in the coordinate system of the second dataset.
+Moving on to the angle plot, we find that the differences are primarily to be found in the second, third and fourth principal components (PCs): The blue arrows visualize the decomposition of the principal components for the Bulgarian dataset in the coordinate system of the Danish dataset.
 We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3.
-The red arrows visualize the decomposition of the principal components for the second dataset in the coordinate system of the first dataset.
+The red arrows visualize the decomposition of the principal components for the Danish dataset in the coordinate system of the Bulgarian dataset.
 Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3.
@@
 
-All in all, we find that psychological well-being does not seem to be the same concept in Bulgaria as in Denmark.
-The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalue) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle and chroma plots).
-In Bulgaria, interpersonal features seem to be more informative of psychological wellbeing, whereas in Denmark, individual characteristic play a relatively bigger role.
-Thus, the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics.
-\\
-
-\hl{I have dropped the Wally-plots for now, as I'm not really sure they are that useful.}
+All in all, we find that psychological well-being does not seem to be the same concept in Bulgaria and Denmark.
+The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalues) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle- and chroma plots).
+In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, which corresponds with previous findings.
+Thus, the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics (\cite{ESStopline5}).
+Moreover, the two countries cannot be ranked in terms of which country is "the most happy", at least not by referring to psychological well-being dimensions such as those encountered here.
 
@@
 \caption{A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples.
-Note that the bars for each component is annotated with its cumulative variance score, that is, how much variance can be explained by having information of this and the preceding components.}
+Note that the bars for each component is annotated with its cumulative variance score (denoted $\tilde\sigma^c_i$ in the above), that is, how much variance can be explained by having information of this and the preceding components.}
 \label{plotSE.pancake}
@@
 We now turn to the comparison of Denmark and Sweden in terms of psychological well-being.
-Figure \ref{plotSE.cehair} show the CE and angle plots for these two countries.
+Figure \ref{plotSE.cehair} shows the CE- and angle plots for these two countries.
 In the CE plot, we now find the cumulative eigenvalue curve to be just within the acceptance region of the null-hypothesis.
-This is also reflected by the two tests, which now produce $p$-values of $p_{KS} = 0.14$ and $p_{CvM} = 0.09$, respectively, thus accepting the null-hypothesis at the typical 5\% level, but not with overwhelming evidence.
+This is also reflected by the two tests, which now produce $p$-values of $p_\text{KS} = 0.14$ and $p_\text{CvM} = 0.09$, respectively, thus accepting the null-hypothesis at the typical 5\% level, but not with overwhelming evidence.
 
@@
 In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g.
-conclude using Table \ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular, the unhappy people in Denmark (represented by the 1st quartiles) are generally a lot happier in Denmark than they are in Sweden.
+conclude using Table \ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally a lot happier than the least happy people in Sweden.
 A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.
@@
 
+\appendix
+
+\section{Supplementary tables}
+\begin{table}[h]
+\centering
+\small
+\rowcolors{1}{}{lightgray}
+\begin{tabular}{ll}
+\hline
+Scale & Items \\
+\hline
+\rowcolor{lightgray} & How satisfied with life as a whole \\
+\rowcolor{lightgray}\multirow{-2}{*}{Evaluative wellbeing }& How happy are you\\
+
+\rowcolor{white} & Felt sad, how often in the past week\\
+\rowcolor{white} & Felt depressed, how often in the past week\\
+\rowcolor{white} & Enjoyed life, how often in the past week\\
+\rowcolor{white} & Were happy, how often in the past week\\
+\rowcolor{white} & Felt anxious, how often in the past week\\
+\rowcolor{white}\multirow{-6}{*}{Emotional wellbeing} & Felt calm and peaceful, how often in the past week\\
+
+\rowcolor{lightgray} & Free to decide how to live my life\\
+\rowcolor{lightgray} & Little chance to show how capable I am\\
+\rowcolor{lightgray} & Feel accomplishment from what I do\\
+\rowcolor{lightgray} & Interested in what you are doing\\
+\rowcolor{lightgray} & Absorbed in what you are doing\\
+\rowcolor{lightgray} & Enthusiastic about what you are doing\\
+\rowcolor{lightgray} & Feel what I do in life is valuable and worthwhile\\
+\rowcolor{lightgray} & Have a sense of direction\\
+\rowcolor{lightgray} & Always optimistic about my future\\
+\rowcolor{lightgray} & There are lots of things I feel I am good at\\
+\rowcolor{lightgray} & In general feel very positive about myself\\
+\rowcolor{lightgray} & At times feel as if I am a failure\\
+\rowcolor{lightgray} & When things go wrong in my life it takes a long time to get back to normal\\
+\rowcolor{lightgray}\multirow{-14}{*}{Functioning} & Deal with important problems\\
+
+\rowcolor{white} & Felt everything did an effort, how often in the past week\\
+\rowcolor{white} & Sleep was restless, how often in the past week\\
+\rowcolor{white} & Could not get going, how often in the past week\\
+\rowcolor{white}\multirow{-4}{*}{Vitality} & Had a lot of energy, how often in the past week\\
+
+\rowcolor{lightgray} &  Most people can be trusted\\
+\rowcolor{lightgray} &  People try to take advantage\\
+\rowcolor{lightgray} &  Most of the time people are helpful\\
+\rowcolor{lightgray} &  Feel people in local area help one another\\
+\rowcolor{lightgray}\multirow{-2}{*}{Community wellbeing}& Feel close to the people in local area\\
+
+\rowcolor{white} & How many with whom you can discuss intimate matters\\
+\rowcolor{white} & Feel appreciated by those you are close to\\
+\rowcolor{white} & Receive help and support\\
+\rowcolor{white}\multirow{-4}{*}{Supportive relationships} & Felt lonely, how often in the past week\\
+\hline
+\end{tabular}
+\caption{Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
+Note that before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
+\label{table:items}
+\end{table}
+
 \end{document}

### 2017-03-30 article/article_v0.9.tex
Karl's changes

@@
 Assume next that we wish to compare structures without specifying a model, or even a variable of interest.
-The central question is whether the two data sets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity across the subset division.
+The central question is whether the two data sets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be taken into account.
 
@@
 What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a data set without assuming neither directional nor hierarchical relationships between the variables.
-Using parametric models is not a general data structure comparison method, but rather a fitted-model comparison method.
+The use of parametric models does not constitute general data structure comparison method, but rather a fitted-model comparison method.
 It addresses the interplay between the model and the data, not the data alone.
@@
 These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
-Software for using these tools is available in an \texttt{R}-package.
-\hl{how to refer to our package?} This yields a solution that is largely independent of the sizes of the two subsets of data.
-Thereby, a lot of methods that compare each subset to the full data set in some sense are excluded.
+This yields a solution that is largely independent of the sizes of the two subsets of data.
+The proposed tools are implemented in the statistical software \texttt{R} (\cite{R}) statistical software in our package, \texttt{PCADSC}, which is available at \hl{?
+How to refer to a package on Github?}.
 
-This manuscript is structured as follows: First, in Section Section \ref{sec:pcadscintro}, we describe the PCADSC procedures, including a brief introduction to principal component analysis (PCA) in general.
-In Section \ref{sec:dataexample}, we present a worked data example using open source, online available data on psychological well-being in three European countries.
-More specifically, we compare data from Denmark, which has repeatedly been rewarded "happiest country in the world", with data from Bulgaria and Sweden, respectively, to investigate whether or not psychological well-being - and thus happiness - is really a concept that is universal beyond country borders, or if such rankings of happiness are inherently meaningless.
-At last, we discuss limitations and merits of the PCADSC tools in Section \ref{sec.Discussion}.
+This manuscript is describes the procedures, including a brief introduction to principal component analysis (PCA) in general and presents a worked data example using open source, online available data on psychological well-being in three European countries.
+More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries.
 
@@
 \label{sec:pcadscintro}
-As mentioned above, the purpose of PCADSC is comparing overall data structures in two or more subsets of a dataset.
+As mentioned above, the purpose of PCADSC is to compare overall data structures in two subsets of a data set.
 But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context.
@@
 These three plots are:
-
 \begin{description}

### 2017-04-02 R/CEPlot.R
Redid entire structure. Changed S4 to S3, reorganized everything to be focused around PCADSC objects. Renamed functins in accordance with article.  Old code is available in /supplementary_code/s4version

@@ -1,0 +1,93 @@
+#' @title Cumulative eigenvalue plot
+#'
+#' @description Compares eigenvalues by cumulative Bland-Altman type plots.
+#' Kolmogorov-Smirnov and Cramer-von Mises tests evaluated by permutation tests
+#'
+#' @param x ...
+#'
+#' @param nDraw ...
+#'
+#' @examples
+#' #load iris data
+#' data(iris)
+#'
+#' #Define grouping variable, grouping the observations by whether their species is
+#' #Setosa or not
+#' iris$group <- "setosa"
+#' iris$group[iris$Species != "setosa"] <- "non-setosa"
+#'
+#' #make a PCADSC object, splitting the data by "group"
+#' irisPCADSC <- PCADSC(iris, "group", setdiff(names(iris), c("group", "Species")))
+#'
+#' #make a CE plot
+#' CEPlot(irisPCADSC)
+#'
+#' @importFrom stats quantile
+#' @importFrom ggplot2 ggplot aes_string geom_polygon geom_line scale_x_continuous
+#' scale_y_continuous theme_bw theme element_blank xlab ylab geom_line
+#' scale_linetype_manual annotate unit
+#' @export
+CEPlot <- function(x, nDraw = NULL) {
+
+  #Check whether x has a valid class
+  objName <- deparse(substitute(x))
+  if ("PCADSC" %in% class(x)) {
+    if (!is.null(x$CEInfo)) {
+      obj <- x$CEInfo
+    } else {
+      stop(paste(objName, "does not contain any CE information.",
+                 "Please call doCE() on", objName, "before making a CEPlot."))
+    }
+  } else if ("CEInfo" %in% class(x)) {
+    obj <- x
+  } else {
+    stop(paste(objName, "must be of class PCADSC or CEInfo."))
+  }
+
+  #Unpack CEInfo object
+  xVals <- obj$xVals
+  y.obs <- obj$y.obs
+  y.sim <- obj$y.sim
+  KS.obs <- obj$KS.obs
+  KS.pvalue <- obj$KS.pvalue
+  CvM.obs <- obj$CvM.obs
+  CvM.pvalue <- obj$CvM.pvalue
+  B <- obj$B
+  d <- obj$d
+
+  #Make CE plot
+  if(is.null(nDraw)) nDraw <- min(B, 20)
+  y.min <- apply(y.sim, 1, quantile, probs=0.025)
[…]

### 2017-04-02 R/PCADSC.R

@@ -1,40 +1,149 @@
-#' A PCADSC object
+#' @title Compute the elements used for PCADSC
+#' @description Calculate standardized PCA loadings and accumulated variance contributions
+#' separately on two parts of a dataset in order to perform non-parametric data structure
+#' comparisons.
 #'
-#' Blabla
+#' @param data A dataset, either a \code{data.frame} or a \code{matrix} with variables
+#' in columns and observations in rows. WHAT TO DO ABOUT DATA.TABLES AND ALSO TIBBLES. BOTH WILL BE
+#' PROBLEMATIC AS OF NOW.
 #'
-#' @slot pcaFrame ...
-#' @slot splitBy ...
-#' @slot splitLevels ...
-#' @slot varNames ...
-#' @slot n1 ...
-#' @slot n2 ...
-#' @slot nObs1 ...
-#' @slot nObs2 ...
+#' @param splitBy A grouping variable with two levels defining the two groups within the
+#' dataset whose data structures we wish to compare. If \code{splitBy} has more than
+#' two levels, only the first two levels are used.
+#'
+#' @param vars The variable names in \code{data} to include in the PCADSC. If \code{NULL}
+#' (the default), all variables except for \code{splitBy} are used.
+#'
+#' @param doCE ...
+#'
+#' @param doAngle ...
+#'
+#' @param doChroma ...
+#'
+#' @param B ...
+#'
+#' @return A list including the following objects: ...
+#'
+#' @details Blablabla.
+#' Note that \code{\link[tibble]{tibble}}s and \code{\link[data.table]{data.table}}s are
+#' accepted as input, but they are instantly converted to \code{\link{data.frame}}s. Future
+#' releases might include specific implementation for these data representations.
 #'
 #' @examples
+#'
 #' #load iris data
-#'   data(iris)
+#' data(iris)
 #'
 #' #Define grouping variable, grouping the observations by whether their species is
 #' #Setosa or not
-#'  iris$group <- "setosa"
-#'  iris$group[iris$Species != "setosa"] <- "non-setosa"
+#' iris$group <- "setosa"
+#' iris$group[iris$Species != "setosa"] <- "non-setosa"
 #'
 #' #make a PCADSC object, splitting the data by "group"
-#'  irisPCADSC <- makePCADSC(iris, "group",
-#'                         var=setdiff(names(iris), c("group", "Species")))
[…]

### 2017-04-02 R/anglePlot.R

@@ -1,0 +1,62 @@
+#' @title Angle plot
+#'
+#' @description Compares eigenvalues and eigenvectors form two datasets. Kolmogorov-Smirnov and Cramer-von Mises tests evaluated by permutation tests might be implemented later.
+#'
+#' @param x ...
+#'
+#' @examples
+#' #load iris data
+#' data(iris)
+#'
+#' #Define grouping variable, grouping the observations by whether their species is
+#' #Setosa or not
+#' iris$group <- "setosa"
+#' iris$group[iris$Species != "setosa"] <- "non-setosa"
+#'
+#' #make a PCADSC object, splitting the data by "group"
+#' irisPCADSC <- PCADSC(iris, "group", setdiff(names(iris), c("group", "Species")))
+#'
+#' #make an angle plot
+#' anglePlot(irisPCADSC)
+#'
+#' @importFrom ggplot2 ggplot aes_string scale_x_continuous scale_y_continuous
+#' theme_bw theme element_blank xlab ylab geom_segment unit scale_color_manual
+#' @importFrom grid arrow
+#' @export
+anglePlot <- function(x) {
+
+  #graphical settings
+  arrow.len <- 0.05
+
+  #Check whether x has a valid class
+  objName <- deparse(substitute(x))
+  if ("PCADSC" %in% class(x)) {
+    if (!is.null(x$angleInfo)) {
+      obj <- x$angleInfo
+    } else {
+      stop(paste(objName, "does not contain any angle information.",
+                 "Please call doAngle() on", objName, "before making an anglePlot."))
+    }
+  } else if ("angleInfo" %in% class(x)) {
+    obj <- x
+  } else {
+    stop(paste(objName, "must be of class PCADSC or angleInfo."))
+  }
+
+  #Unpack angleInfo object
+  aF <- obj$aF
+  d <- obj$d
+  splitLevels <- obj$splitLevels
+
+  #Make anglePlot
+  ggplot(aF, aes_string(x = "x", y = "y", col = "type",
+                        xend = "xend", yend = "yend")) +
+    scale_x_continuous(limits = c(0.5,d+0.5), breaks = 1:d) +
+    scale_y_continuous(limits = c(0.5,d+0.5), breaks = 1:d) +
+    theme_bw() +
+    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+    xlab(paste("PCs for", splitLevels[1])) +
[…]

### 2017-04-02 R/chromaPlot.R

@@ -1,7 +1,10 @@
-#' @title ??
+#' @title Chroma plot
 #'
 #' @description ??
-#' @param x ??
+#'
+#' @param x Either a \code{PCADSC} object or a \code{chromaInfo} object, as produced
+#' by \code{\link{PCADSC}} and \code{\link{doChroma}}.
+#'
 #' @param varLabels A vector of character string labels for the variables used in
 #' \code{pcadscObj}. If non-null, these labels appear in the plot instead of the
 #' variable names. Note that they must be listed in the same order as the variables
@@ -9,7 +12,7 @@
 #' \code{pcadscObj$varNames}. If \code{NULL} (the default), these variable
 #' names are used.
 #'
-#' @param covCO Not implemented, something like a variance contribution cut-off, maybe?
+#' @param covCO Variance contribution cut-off
 #'
 #' @param splitLabels Labels for the two categories of the splitting variable used
 #' to create the PCADSC object, \code{pcadscObj}, given as a named list (see examples).
@@ -26,20 +29,51 @@
 #' @param useComps A vector of integers with the numbers of the principal component
 #' that should be included in the plot.
 #'
+#' @examples
+#' #load iris data
+#' data(iris)
+#'
+#' #Define grouping variable, grouping the observations by whether their species is
+#' #Setosa or not
+#' iris$group <- "setosa"
+#' iris$group[iris$Species != "setosa"] <- "non-setosa"
+#'
+#' #make a PCADSC object, splitting the data by "group"
+#' irisPCADSC <- PCADSC(iris, "group", setdiff(names(iris), c("group", "Species")))
+#'
+#' #make a chroma plot
+#' chromaPlot(irisPCADSC)
+#'
+#'
 #' @importFrom methods setMethod
 #' @importFrom ggplot2 qplot ggplot aes_string geom_bar coord_flip scale_x_reverse
 #'             scale_y_continuous geom_text xlab ylab theme facet_wrap
 #'             as_labeller scale_fill_discrete theme_bw
 #'
 #'@export
-pancakePlot <- function(x, varLabels=NULL, covCO=NULL, splitLabels=NULL,
+chromaPlot <- function(x, varLabels=NULL, covCO=NULL, splitLabels=NULL,
                         varAnnotation = "cum", useComps = NULL) {
[…]

### 2017-04-02 R/doAngle.R

@@ -1,0 +1,61 @@
+#' @title Compute angle information
+#'
+#' @description ...
+#'
+#' @param x ...
+#'
+#'
+#' @export
+doAngle <- function(x) {
+  UseMethod("doAngle")
+}
+
+#x: pcaRes
+#' @export
+doAngle.pcaRes <- function(x) {
+  load1 <- x$load1
+  load2 <- x$load2
+  eigen1 <- x$eigen1
+  eigen2 <- x$eigen2
+  n1 <- x$n1
+  n2 <- x$n2
+  d <- x$d
+  splitBy <- x$splitBy
+  splitLevels <- x$splitLevels
+  vars <- x$vars
+
+  #make angleFrame for plotting
+
+  # find angles
+  angles <- matrix(0, d, d)
+  for (i in 1:d) for (j in 1:d) angles[i,j] <- asin(max(0,min(1,abs(sum(load1[,i]*load2[,j])))))
+
+  #Calculate vector lengths
+  max.eigen <- max(c(eigen1[1],eigen2[1]))
+  len1 <- len2 <- matrix(NA, d, d)
+
+  for (i in 1:d) for (j in 1:d) {
+    len1[i,j] <- sqrt(eigen1[i]/max.eigen)*abs(sum(load1[,i]*load2[,j]))
+    len2[i,j] <- sqrt(eigen2[j]/max.eigen)*abs(sum(load1[,i]*load2[,j]))
+  }
+
+  #Store in data.frame ready for plotting
+  aF <- data.frame(x = rep(1:d, 2*d),
+                   y = rep(rep(1:d, each = d),2),
+                   xend = c(rep(1:d, d) + c(len1)*cos((pi-c(angles))/2),
+                            rep(1:d, d) + c(len2)*cos(c(angles)/2)),
+                   yend = c(rep(1:d, each = d) + c(len1)*sin((pi-c(angles))/2),
+                            rep(1:d, each = d) + c(len2)*sin(c(angles)/2)),
+                   type = rep(c("1st", "2nd"), each = d*d))
+
+  #pack and return output
+  out <- list(aF = aF, splitLevels = splitLevels, d = d)
+  class(out) <- "chromaInfo"
+  out
+}
+
+#' @export
+doAngle.PCADSC <- function(x) {
+  x$angleInfo <- doAngle(x$pcaRes)
+  x
+}

### 2017-04-02 R/doCE.R

@@ -1,0 +1,87 @@
+#' @title Compute cumulative eigenvalue information
+#'
+#' @description ...
+#'
+#' @param x ...
+#'
+#' @param ... ...
+#'
+#' @export
+doCE <- function(x, ...) {
+  UseMethod("doCE")
+}
+
+#x: pcaRes
+#' @export
+doCE.pcaRes <- function(x, data, B, ...) {
+
+  #check whether B is positive
+  if (B < 1) {
+    stop("B must be positive.")
+  }
+
+  #Unpack x
+  n1 <- x$n1
+  nBoth <- x$nBoth
+  d <- x$d
+  vars <- x$vars
+
+  #Calculate cumulative sums for the observed data
+  ceResObs <- calcCum(x)
+  xVals <- ceResObs$x
+  y.obs <- ceResObs$y
+  KS.obs <- ceResObs$KS
+  CvM.obs <- ceResObs$CvM
+
+  #Calculate cumulative sums for the randomly partitioned data
+  y.sim   <- matrix(0, d+1, B)
+  KS.sim  <- rep(0, B)
+  CvM.sim <- rep(0, B)
+
+  for (i in 1:B) {
+    splitVar <- rep("2", nBoth)
+    ii <- sample(1:nBoth, n1)
+    splitVar[ii] <- "1"
+    data$splitVar <- factor(splitVar)
+    thisCEres <- calcCum(doPCA(data, "splitVar", c("1", "2"), vars, doBoth = FALSE),
+                         x = xVals)
+    y.sim[,i]  <- thisCEres$y
+    KS.sim[i]  <- thisCEres$KS
+    CvM.sim[i] <- thisCEres$CvM
+  }
+
+  #Calculate p-values
+  KS.pvalue <- mean(KS.sim >= KS.obs)
+  CvM.pvalue <- mean(CvM.sim >= CvM.obs)
+
+  #Pack and return output
+  out <- list(d = d, B = B, xVals = xVals, y.obs = y.obs, y.sim = y.sim,
+              KS.obs = KS.obs, KS.pvalue = KS.pvalue,
+              CvM.obs = CvM.obs, CvM.pvalue = CvM.pvalue)
+  class(out) <- "CEInfo"
+  out
+}
+
+#' @export
+doCE.PCADSC <- function(x, ...) {
+  x$CEInfo <- doCE(x$pcaRes, x$data, x$B)
+  x
+}
+
+
[…]

### 2017-04-02 R/doChroma.R

@@ -1,0 +1,85 @@
+#' @title Compute chroma information
+#'
+#' @description Compute standardized PCA loadings and cummulative variance contributions
+#' BLABLA
+#'
+#' @param x ...
+#'
+#' @export
+doChroma <- function(x) {
+  UseMethod("doChroma")
+}
+
+#x: pcaRes
+#' @export
+doChroma.pcaRes <- function(x) {
+  load1 <- x$load1
+  load2 <- x$load2
+  eigen1 <- x$eigen1
+  eigen2 <- x$eigen2
+  n1 <- x$n1
+  n2 <- x$n2
+  d <- x$d
+  splitBy <- x$splitBy
+  splitLevels <- x$splitLevels
+  vars <- x$vars
+
+  #make chromaFrame for plotting
+  cF1 <- chromaFrame(load1, eigen1, d, vars)
+  cF2 <- chromaFrame(load2, eigen2, d, vars)
+  cF <- as.data.frame(rbind(cF1, cF2))
+  cF$group <- rep(splitLevels, each = d^2)
+
+  #pack and return output
+  out <- list(splitLevels = splitLevels, n1 = n1, n2 = n2, splitBy = splitBy,
+              cF = cF, vars = vars)
+  class(out) <- "chromaInfo"
+  out
+}
+
+#' @export
+doChroma.PCADSC <- function(x) {
+  x$chromaInfo <- doChroma(x$pcaRes)
+  x
+}
+
+
+
+
+################Not exported below##################################################
+
+#Compute standardized PCA loadings and cummulative variance contributions,
+#pack it up for plotting in a data.frame
+#' @importFrom stats na.omit princomp
+#' @importFrom reshape2 melt
+chromaFrame <- function(loadings, eigenvalues, d, vars) {
+ #p <- pca
+ # p <- stats::princomp(sapply(stats::na.omit(data[, vars]), as.numeric))
+  px <- round(matrix(c(loadings), d, dimnames=list(vars, 1:d)), 4)
+
+  for (i in 1:d) { #standardize
+    px[, i] <- abs(px[, i]/sum(abs(px[, i])))
+  }
+
+  #cummulative variance contributions and
+  #principal component variance constribution
+  pcvc <- eigenvalues
+  pcvc <- pcvc/sum(pcvc)
[…]

### 2017-04-02 R/qplot.R

@@ -1,36 +1,0 @@
-#'@importFrom ggplot2 qplot
-setGeneric("qplot")
-setGeneric("plot")
-
-#' @param varLabels A vector of character string labels for the variables used in
-#' \code{pcadscObj}. If non-null, these labels appear in the plot instead of the
-#' variable names. Note that they must be listed in the same order as the variables
-#' in \code{pcadscObj} and this order can be expected by calling
-#' \code{pcadscObj$varNames}. If \code{NULL} (the default), these variable
-#' names are used.
-#'
-#' @param covCO Not implemented, something like a variance contribution cut-off, maybe?
-#'
-#' @param splitLabels Labels for the two categories of the splitting variable used
-#' to create the PCADSC object, \code{pcadscObj}, given as a named list (see examples).
-#' These labels will appear in the headers of the two PCADSC plots.
-#' If \code{NULL} (the default), the original levels of the splitting variable
-#' are used.
-#'
-#' @param hideCumVar If \code{TRUE} (the default), cummulated explained variance
-#' percentages are not shown for each component.
-#'
-#' @importFrom methods setMethod
-#' @importFrom ggplot2 qplot ggplot aes_string geom_bar coord_flip scale_x_reverse
-#'             scale_y_continuous geom_text xlab ylab theme facet_wrap
-#'             as_labeller scale_fill_discrete theme_bw
-#'
-#'@describeIn PCADSC Plot a PCADSC object
-#'@include PCADSC.R
-#'@include pancakePlot.R
-#'@export
-setMethod("qplot","PCADSC", pancakePlot)
-
-
-#'@rdname qplot,PCADSC-method
-setMethod("plot", "PCADSC", function(x, y = NULL, ...) qplot(x, ...))

### 2017-04-02 supplementary_code/s4version/PCADSC_old.R

@@ -1,0 +1,40 @@
+# #' A PCADSC object
+# #'
+# #' Blabla
+# #'
+# #' @slot pcaFrame ...
+# #' @slot splitBy ...
+# #' @slot splitLevels ...
+# #' @slot varNames ...
+# #' @slot n1 ...
+# #' @slot n2 ...
+# #' @slot nObs1 ...
+# #' @slot nObs2 ...
+# #'
+# #' @examples
+# #' #load iris data
+# #'   data(iris)
+# #'
+# #' #Define grouping variable, grouping the observations by whether their species is
+# #' #Setosa or not
+# #'  iris$group <- "setosa"
+# #'  iris$group[iris$Species != "setosa"] <- "non-setosa"
+# #'
+# #' #make a PCADSC object, splitting the data by "group"
+# #'  irisPCADSC <- makePCADSC(iris, "group",
+# #'                         var=setdiff(names(iris), c("group", "Species")))
+# #' #plot it
+# #'  qplot(irisPCADSC)
+# #'
+# #' #print it
+# #'  print(irisPCADSC)
+# #'
+# #'
+# #'
+# #' @importFrom methods setClass new
+# #' @export
+# PCADSC <- setClass("PCADSC",
+#                    slots = list(pcaFrame = "data.frame", splitBy = "character",
+#                                 splitLevels = "character", varNames = "character",
+#                                n1 = "numeric", n2 = "numeric", nObs1 = "numeric",
+#                                 nObs2 = "numeric"))

### 2017-04-02 supplementary_code/s4version/anglePlot.R

@@ -1,4 +1,4 @@
-#' @title Hairplot
+#' @title Angle plot
 #'
 #' @description Compares eigenvalues and eigenvectors form two datasets. Kolmogorov-Smirnov and Cramer-von Mises tests evaluated by permutation tests might be implemented later.
 #'
@@ -29,14 +29,14 @@
 #' iris$group[iris$Species != "setosa"] <- "non-setosa"
 #'
 #' #make a PCADSC object, splitting the data by "group"
-#' hairplot(iris, "group",var=setdiff(names(iris), c("group", "Species")))
+#' anglePlot(iris, "group",var=setdiff(names(iris), c("group", "Species")))
 #'
 #' @importFrom graphics axis arrows
 #' @importFrom ggplot2 ggplot aes_string scale_x_continuous scale_y_continuous
 #' theme_bw theme element_blank xlab ylab geom_segment unit scale_color_manual
 #' @importFrom grid arrow
 #' @export
-hairplot <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE,arrow.len=0.05) {
+anglePlot <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE,arrow.len=0.05) {
   #define var
   if (is.null(var)) var <- setdiff(names(data), splitBy)
 

### 2017-04-02 supplementary_code/s4version/cePlot.R

@@ -27,7 +27,7 @@
 #' iris$group[iris$Species != "setosa"] <- "non-setosa"
 #'
 #' #make a PCADSC object, splitting the data by "group"
-#' cumeigen(iris, "group",var=setdiff(names(iris), c("group", "Species")))
+#' cePlot(iris, "group",var=setdiff(names(iris), c("group", "Species")))
 #'
 #' @importFrom graphics polygon matlines lines legend
 #' @importFrom stats quantile
@@ -35,7 +35,7 @@
 #' scale_y_continuous theme_bw theme element_blank xlab ylab geom_line
 #' scale_linetype_manual annotate unit
 #' @export
-cumeigen <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE) {
+cePlot <- function(data,splitBy,var=NULL,B=1000,make.plot=TRUE) {
   #define var
   if (is.null(var)) var <- setdiff(names(data), splitBy)
 

### 2017-04-02 supplementary_code/s4version/loadComp.R

@@ -7,13 +7,13 @@
 #' @importFrom stats na.omit princomp
 #' @importFrom reshape2 melt
 loadComp <- function(data, vars = NULL, varCO=NULL) {
-  if (is.null(varCO)) {
-    varCO <- 1
-  }
+ # if (is.null(varCO)) {
+#    varCO <- 1
+#  }
 
-  if (is.null(vars)) {
-    vars <- names(data)
-  }
+#  if (is.null(vars)) {
+#    vars <- names(data)
+#  }
 
   n <- length(vars)
   p <- stats::princomp(sapply(stats::na.omit(data[, vars]), as.numeric))

### 2017-04-02 supplementary_code/s4version/qplot.R

@@ -1,0 +1,36 @@
+# #'@importFrom ggplot2 qplot
+#setGeneric("qplot")
+#setGeneric("plot")
+
+# #' @param varLabels A vector of character string labels for the variables used in
+# #' \code{pcadscObj}. If non-null, these labels appear in the plot instead of the
+# #' variable names. Note that they must be listed in the same order as the variables
+# #' in \code{pcadscObj} and this order can be expected by calling
+# #' \code{pcadscObj$varNames}. If \code{NULL} (the default), these variable
+# #' names are used.
+# #'
+# #' @param covCO Not implemented, something like a variance contribution cut-off, maybe?
+# #'
+# #' @param splitLabels Labels for the two categories of the splitting variable used
+# #' to create the PCADSC object, \code{pcadscObj}, given as a named list (see examples).
+# #' These labels will appear in the headers of the two PCADSC plots.
+# #' If \code{NULL} (the default), the original levels of the splitting variable
+# #' are used.
+# #'
+# #' @param hideCumVar If \code{TRUE} (the default), cummulated explained variance
+# #' percentages are not shown for each component.
+# #'
+# #' @importFrom methods setMethod
+# #' @importFrom ggplot2 qplot ggplot aes_string geom_bar coord_flip scale_x_reverse
+# #'             scale_y_continuous geom_text xlab ylab theme facet_wrap
+# #'             as_labeller scale_fill_discrete theme_bw
+# #'
+# #'@describeIn PCADSC Plot a PCADSC object
+# #'@include PCADSC.R
+# #'@include pancakePlot.R
+# #'@export
+# setMethod("qplot","PCADSC", pancakePlot)
+
+
+# #'@rdname qplot,PCADSC-method
+# setMethod("plot", "PCADSC", function(x, y = NULL, ...) qplot(x, ...))

### 2017-04-02 supplementary_code/s4version/wallyPlot.R

@@ -16,7 +16,7 @@
 #'
 #' @importFrom ggplot2 theme xlab ylab element_blank
 #' @export
-wallyPancake <- function(PCADSCobj, data, nrow, ncol, covCO = 1,
+wallyPlot <- function(PCADSCobj, data, nrow, ncol, covCO = 1,
                          varAnnotation = "cum", useComps = NULL) {
   vars <- PCADSCobj@varNames
   nObs1 <- PCADSCobj@nObs1

### 2017-04-10 article/article_v0.9.tex
Jeg er gået igang med at indskrive mine små ændringer

@@
 \begin{keywords}
-Principal component analysis, multi-center studies, data structure comparisons \hl{Is this how you do keywords? Karl? Bo?}
+multi-center studies, data heterogeneity, data structure comparisons, principal component analysis (PCA), goodness of fit, European Social Survey.
 \end{keywords}
@@
 \item Surveys often use mixed modes of administration, e.g.
-mail and telephone, and while this can improve response rates, the mode of administration can affect results \cite{Brambilla1987,McHorney1994} and differences in response behavior can lead to biased results.
+mail and telephone, and while this can improve response rates, the mode of administration can affect results \citep{Brambilla1987,McHorney1994}, and differences in response behaviour can lead to biased results.
 Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful.
 \item The rapid growth of web surveys, due to low cost, timeliness, and other factors, generate large data sources that lack a sampling frame of the general population.
-However, it can be problematic to combine online panels (pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions) \cite{Liu2016}.
+However, it can be problematic to combine online panels (pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions), see \cite{Liu2016}.
 \item Large scale open source data sets such as the PISA data and ESS (European Social Survey) data.
@@
 
-\hl{Maybe something here about what PCA is usually used for, e.g. dimension reduction, exploratory data analysis, ...}
-
 \hl{Maybe broaden the raison d'etre of PCADSC a bit: It is also a useful tool for doing initial, exploratory data analysis and it helps us to identify potential problems or challenges early in the study design development, before concrete model choices have been made, thereby aiding a more economic use of resources and a less ad-hoc approach to the types of problems that PCADSC can find.}
 
-Assume that we have two data sets with the same variables, but different observations represented as a data set with
-a subset-inducing variable.
+Assume that we have two data sets with the same variables, but different observations represented as a data set with a subset-inducing variable.
 Assume next that we wish to compare structures without specifying a model, or even a variable of interest.
@@
 
-Sophisticated methods for addressing this question are available when we are willing to assume a statistical model, but when these models are taken away, a remarkable void of methods is left behind.
+Sophisticated methods for addressing this question are available when we are willing to assume a statistical model \hl{Skal der referencer her?
+Nogen forslag?}, but when these models are taken away, a remarkable void of methods is left behind.
 What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a data set without assuming neither directional nor hierarchical relationships between the variables.
@@
 Entry-by-entry comparison of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase.
-For these methods significant $p$-values are likely to abound as the sample size increases.
 Parametric models using, e.g., latent variable models moves beyond the marginal approach, but need a pre-specified model.
+
+%For these methods significant $p$-values are likely to abound as the sample size increases. Bo: Jeg tror denne sætning bare kan udelades.
 
@@
 
-This manuscript is describes the procedures, including a brief introduction to principal component analysis (PCA) in general and presents a worked data example using open source, online available data on psychological well-being in three European countries.
+This manuscript describes the procedures, including a brief introduction to principal component analysis (PCA) in general and presents a worked data example using open source, online available data on psychological well-being in three European countries.
 More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries.
@@
 \end{equation*}
-and the theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error.
-\hl{Reference here?
-Maybe to like original Pearson article?}
+and the theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \citep{HastieEtAl2009}.
 

### 2017-04-10 article/article_v0.9.tex
Jeg er færdig med at tilføje mine rettelser i denne omgang. Der er flyttet lidt rundt på "standardization" argumentet. Diskussions afsnittet er stadigvæk ufærdigt.

@@
 One such definition is the structure of the covariance matrix of the dataset.
-If we assume all variables in the dataset to be jointly normal with zero means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables.
+If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables.
 This gives it a very nice interpretation as a measure of the overall structure.
@@
 
-A naive approach for data structure comparisons might therefore be to compute the empirical covariance matrices on each of the two data subsets and simply compare these matrices.
+A naive approach for data structure comparisons might therefore be to compute the empirical covariance matrices on each of the two data subsets and simply compare these matrices entry-by-entry.
 Though the idea perhaps sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices.
@@
 
+At this point a general remark about standardization is in place.
+PCA deconstructs the covariance matrix in components according to the most explained variance.
+This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data.
+Therefore, the variables often are standardized prior to performing PCA.
+Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices.
+The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation.
+In this paper we will adopt this approach, and in particular, we will allow for differences between scalings of the variables in the two data subsets.
+
+\hl{Bo: I næste afsnit bruges stadigvæk ``covariance''. Det er vel ok?}
+
 \subsection{A brief introduction to Principal component analysis}
@@
 For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost.
-More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the orignial $d$-dimensional one.
+More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the original $d$-dimensional one.
 Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
@@
 \end{equation*}
-and the theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \citep{HastieEtAl2009}.
+The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \citep{HastieEtAl2009}.
 
@@
 \end{equation*}
-where $\eta_1, ..., \eta_q$ are the $q$ first eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, ..., \lambda_d$.
+where $\eta_1, ..., \eta_q$ are the $q$ first eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$.
 In the PCA framework, we refer to the eigenvectors as \textit{loadings}.
 The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
-$$\text{trace}(S) = \sum_{j=1}^d \hat{V}(X_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j$$
-where $X_j \in \RR^n$ denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
+\begin{equation*}
+\text{trace}(S) = \sum_{j=1}^d \hat{V}(X_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
+\end{equation*}
+where $X_j \in \RR^n$ \hl{(Bo: $X_1$ og $X_2$ bruges i en anden betydning i afsnit 2.2.
+Omformuler?)} denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
 This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
 And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different, the representation hereby obtained is a valid object of inference.
-If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_0$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
+If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
 
@@
 The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables.
-We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$.
+We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the apparently same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$.
 For each of these two datasets, we complete the following steps:
@@
 \end{equation*}
-thereby obtaining loadings $\eta_{ik}$ and variance components $\lambda_{ik}$ for $i \in \{1, 2\}$ and $k \in 1, ..., d$.
+thereby obtaining loadings $\eta_{ik}$ and variance components $\lambda_{ik}$ for $i \in \{1, 2\}$ and $k \in 1,\dotsc,d$.
 \end{enumerate}
-The hereby obtains PCA decompositions of the covariance matrices can then be compared.
+The hereby obtains PCA decompositions of the correlation matrices can then be compared.
 We present three diagnostics plots that are designed to shine a light on different types and nuances of data structure differences.
@@
 
-But before diving deeper into the details of the \emph{CE plot}, the \emph{angle plot} and the \emph{chroma plot}, a general remark about standardization in PCA is in place.
-We would like to emphasize that PCA is sensitive to scaling, as the procedure deconstructs the covariance matrix in components according to the most explained variance.
-This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data.
-Therefore, the variables should always be scaled prior to performing PCA.
-Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices.
-The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
-$$\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$$
-which will simplify some expressions below.
+The standardization makes the variables comparable on the same scale (units of standard deviation), and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
+\begin{equation*}
+\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d.
+\end{equation*}
+This identity will simplify some expressions below.
 
 \subsubsection{The cumulative eigenvalue plot}
- The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the covariance matrix.
+The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix.
 %If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
@@
 \end{align*}
-This may be seen as a cumulative Bland-Altman plot for the variation components (\hl{reference to cumulative residuals and to Bland-Altman}).
+This may be seen as a cumulative Bland-Altman plot \citep{AltmanBland83,LinEtAl2002} for the variation components.
 Note that due to the standardization, the last point will always be equal to $(d,0)$.
@@
 
-In order to test whether these cumulative differences are statistical artifacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
+In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
 \begin{align*}
@@
 \end{align*}
-We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and standardized datasets into two new datasets of $n_1$ and $n_2$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics.
+We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_1$ and $n_2$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics.
 This should be done a large (e.g.
@@
 \end{equation*}
-and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset \hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?}.
+and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset \hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?
+Bo: Jeg har ikke noget bud på en sådan udvidelse lige nu og her --- skal vi ikke bare sige at det er ok i sin nuværende form?}.
 We propose to visualize these two decompositions in a $d \times d$ grid display.
@@
 These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions.
-More specifically, when illustrating the $i$th principal component, we plot a vertical bar of length one that has been divided into $d$ segments of different colors and the wideness of the $j$th such segment is given by
-$$\omega_{ij} = \frac{|(\eta_i)_j|}{\sum_{k=1}^d |(\eta_i)_k|}$$
-where $(\eta_i)_j$ denotes the $j$th entry of $\eta_i$.
+More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors and the wideness of the $j$th segments are given by
+\begin{align*}
+\omega_{1ij} &= \frac{\lvert\eta_{1ij}\rvert}{\sum_{k=1}^d \lvert\eta_{1ik}\rvert}, &
+\omega_{2ij} &= \frac{\lvert\eta_{2ij}\rvert}{\sum_{k=1}^d \lvert\eta_{2ik}\rvert},
+\end{align*}
+where $\eta_{1ij}$ and $\eta_{2ij}$ denotes the $j$th entry of $\eta_{1i}$ and $\eta_{2i}$, respectively.
 Due to the indeterminacy of the sign, all the signs are removed from the coefficients in the loadings.
 The bars are ordered according to the variation components and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance component contributions,
-$$\tilde\sigma^c_i = \frac{\sum_{j = 1}^i \lambda_i}{\sum_{k=1}^d \lambda_k} = \frac{\sum_{j = 1}^i \lambda_i}{d}$$
+\begin{align*}
+\tilde\sigma^c_{1i} &= \frac{\sum_{j=1}^i \lambda_{1j}}{\sum_{k=1}^d \lambda_{1k}} = \frac{\sum_{j=1}^i \lambda_{1j}}{d}, &
+\tilde\sigma^c_{2i} &= \frac{\sum_{j=1}^i \lambda_{2j}}{\sum_{k=1}^d \lambda_{2k}} = \frac{\sum_{j=1}^i \lambda_{2j}}{d}.
+\end{align*}
 Especially when $d$ is large, we recommend plotting only a select set of interesting principal components, e.g.
 as identified by use of the angle plot.
-In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_i = \frac{\lambda_i}{d}$.
+In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$.
 
@@
 For international comparisons of educational systems, the PISA tests have repeatedly been criticized for not being meaningful objects for international comparisons, especially due to problems with differential item functions (\cite{Kankaras2014}; \cite{Kreiner2014}) and translation problems (\cite{Asil2016}).
-\hl{Til Karl: Måske kender du nogle flere studier her, vi bør nævne?}
+\hl{Til Karl: Måske kender du nogle flere studier her, vi bør nævne?
+Bo: F.eks.
+\cite{ZwitserEtAl2017}?}
 
@@
 The main question is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
-Or do different aspects of psychological well-being or happiness simply not have the same relative meaning in different cultural and socioeconomic settings?
+Or do different aspects of psychological well-being or happiness simply not have the same relative meaning in different cultural and socio-economic settings?
 This is in fact a question concerning comparability of data structures.
@@
 The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures.
-And therefore, comparing the two countries in these measures is not a meaningful endeavor.
+And therefore, comparing the two countries in these measures is not a meaningful endeavour.
 In this paper, we will focus on a single aspect of overall happiness, namely psychological well-being,
@@
 In this section, we use the PCADSC tools to unveil international differences in one aspect of happiness, namely psychological well-being.
-Our starting point is Denmark, a small, northern European country that has repeatedly been awarded with the title of "happiest country in the world"  by the \textit{World Happiness Report}, most recently in 2016 (\cite{WHR2016}), and we wish to investigate if this title is really meaningful at all.
+Our starting point is Denmark, a small, northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 (\cite{WHR2016}), and we wish to investigate if this title is really meaningful at all.
 In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria.
-Though both countries are European and thus not geographically nor culturally as far apart as some other countries might be, these two countries have previously been highlighted to be very different in terms of what defines happiness (\cite{ESStopline5} {\hl{Is this the correct way to refer to a technical report?}).
+Though both countries are European and thus not geographically nor culturally as far apart as some other countries might be, these two countries have previously been highlighted to be very different in terms of what defines happiness (\cite{ESStopline5} {\hl{Is this the correct way to refer to a technical report?
+Bo: Ja, jeg synes det er fint.}).
 Moreover,intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated (\cite{Rodriguez2014}), with a much less strong relationship between the two in Northern- compared to other European countries.
@@
 In the CE plot, we now find the cumulative eigenvalue curve to be just within the acceptance region of the null-hypothesis.
-This is also reflected by the two tests, which now produce $p$-values of $p_\text{KS} = 0.14$ and $p_\text{CvM} = 0.09$, respectively, thus accepting the null-hypothesis at the typical 5\% level, but not with overwhelming evidence.
+This is also reflected by the two tests, which now produce $p$-values of $p_\text{KS} = 0.14$ and $p_\text{CvM} = 0.09$, respectively.
+Thus, it is not very unreasonable to assume  equal variance components in the two datasets.
+
+% but not with overwhelming evidence.
+% Bo: absence of proof \neq proof of absence.
 

### 2017-04-23 ESS/ess_notes.R
Updated figures for article

@@ -178,7 +178,9 @@ xtable(m, digits = 2)
 library(PCADSC)
 library(ggplot2)
 library(gridExtra)
-a1 <- makePCADSC(data_SEDK, "cntry",  c("EvaluativeWellbeing",
+
+set.seed(1234)
+a1 <- PCADSC(data_SEDK, "cntry",  c("EvaluativeWellbeing",
                                "EmotionalWellbeing",
                                "Functioning",
                                "Vitality",
@@ -186,12 +188,13 @@ a1 <- makePCADSC(data_SEDK, "cntry",  c("EvaluativeWellbeing",
                                "SupportiveRelationships"))
 
 
-p1 <- pancakePlot(a1)
+p1 <- chromaPlot(a1)
 
 ggsave(plot = p1, file = "./article/essDKSEpancake.pdf",
        width = 8, height = 5)
 
-a2 <- makePCADSC(data_BGDK, "cntry",  c("EvaluativeWellbeing",
+set.seed(123424)
+a2 <- PCADSC(data_BGDK, "cntry",  c("EvaluativeWellbeing",
                                        "EmotionalWellbeing",
                                        "Functioning",
                                        "Vitality",
@@ -199,50 +202,24 @@ a2 <- makePCADSC(data_BGDK, "cntry",  c("EvaluativeWellbeing",
                                        "SupportiveRelationships"))
 
 
-p2 <- pancakePlot(a2)
+p2 <- chromaPlot(a2)
 
 ggsave(plot = p2, file = "./article/essDKBGpancake.pdf",
        width = 8, height = 5)
 
-p2c <- pancakePlot(a2, useComps = 2:4)
+p2c <- chromaPlot(a2, useComps = 2:4)
 
 ggsave(plot = p2c, file = "./article/essDKBGpancake234.pdf",
        width = 8, height = 5)
 
 
-set.seed(1234)
-p3a <- wallyPancake(a1, data_SEDK, nrow = 3, ncol = 3, varAnnotation = NULL)
-p3b <- wallyPancake(a2, data_BGDK, nrow = 3, ncol = 3, varAnnotation = NULL)
-set.seed(12324)
-p3c <- wallyPancake(a2, data_BGDK, nrow = 3, ncol = 3,
-                    useComps = 2:4, varAnnotation = NULL)
-
-ggsave(plot = p3a, file = "./article/essDKSEWallyPCADSC.pdf",
-       width = 8, height = 10)
[…]

### 2017-04-23 article/article_v0.9.5.tex
First complete version of the article is done.

--- 2017-04-10 article/article_v0.9.tex
@@
 \documentclass[titlepage,11pt,twoside]{article}
-
 \usepackage[dvips]{graphicx}
@@
 \usepackage{pmetrika}
-\usepackage{pmbib}
+%\usepackage{pmbib}
+
 \usepackage{color}
@@
 %\usepackage{colortbl}
-%\usepackage[backend=biber,natbib=true, style=authoryear-comp]{biblatex}
 \usepackage[table]{xcolor}
 \definecolor{lightgray}{gray}{0.9}
+\usepackage[T1]{fontenc}
 
-%\usepackage{submit}
+\usepackage{caption,setspace}
+\captionsetup{font={footnotesize,stretch=1.5}}
+\renewcommand{\baselinestretch}{2}
 
@@
 \DeclareMathOperator*{\argmax}{argmax}
-%\newcommand{\argmax}{\text{argmax}}
 \newcommand{\R}[1]{\texttt{#1}}
@@
 
+\begin{titlepage}
 \linespacing{1}
 
-%\title{Something that is not PCADSC}
-%\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
 \title{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}
 
-\author{Anne H. Petersen, Bo Markussen, Karl Bang Christensen}
+\author{Anne H. Petersen}
+\affil{Department of Public Health, University of Copenhagen}
 
-\affil{University of Copenhagen}
+\author{Bo Markussen}
+\affil{Department of Mathematical Sciences, University of Copenhagen}
+
+\author{Karl Bang Christensen}
+\affil{Department of Public Health, University of Copenhagen}
 
@@
 
-\linespacing{1}
+\end{titlepage}\vspace*{24pt}
 
@@
 \begin{abstract}
-abstract...
+\linespacing{1}
+
+In empirical psychometric work, one often encounters data that is somehow divided into two distinct susbets.
+This might be due to multi-center sampling frames, or to variations in instruments, orders of questionnaire item, modes of administration etc.
+Often, one wishes to combine such data into one an conduct only a single analysis, but this is only meaningful if the two subsets are fundamentally similar in structures, that is, if they measure the same underlying concepts.
+However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc and require making full model assumptions.
+We present a suite of visual, non-parametric diagnostic tools for investigating differences in data structures for two datasets containing different observations of the same variables, refered to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC).
+This approach requires no directional nor hierarchical assumptions on the data.
+Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets and by each targeting a different aspect of the PCA information, the three plots presented here give a thorough and deep understanding of where differences occur and why it might be the case.
+
+The functionality of the PCADSC tools is presented by use of data from the European Social Survey on psycological well-being in three European countries, namely Denmark, Sweden and Bulgaria.
+Here, we find that the concepts of psychological well-being are not the same when comparing Denmark and Bulgaria, and thus a comparison of e.g.
+mean scores in different well-being aspects is not meaningful.
+However, when comparing the two Scandinavian countries, Denmark and Sweden, we find very similar data structures and thus very similar concepts of well-being.
+Therefore, inter-country comparisons are warranted for these two countries.
+
 \begin{keywords}
-multi-center studies, data heterogeneity, data structure comparisons, principal component analysis (PCA), goodness of fit, European Social Survey.
+multi-center studies, data heterogeneity, data structure comparisons, principal component analysis (PCA), European Social Survey.
 \end{keywords}
@@
 
-Classical statistical methodology is aimed at analyzing data from designed experiments and historically, statistical analyses have been done by researchers who knew the design and origin story of the data set well.
-However, the origin stories of data sets have changed over time and today a lot of data is accumulated without a specific purpose in mind.
-This is due to vast amounts of data being registered online and to a trend towards more open source research.
-The latter phenomenon in particular poses new challenges wrt.\ data quality assessment.
-When data are collected and made public without a specific end-point in mind, how do we ensure that differences in, say, choice measurement instruments, mode of administration, or sampling frame do not cause the data to be effectively divided into subsets that are simply not comparable?
-Three examples of this are
+Data comparability is a classical topic in psychometrics and, more generally, in statistics.
+Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability (and thus collapsibility) needs to be assesed empirically.
+This happens e.g.
+when data are collected across several centers (or even countries) or when different versions of an instrument, the mode of administration or the order of survey questionnaire items are applied.
+All of these practices are essential parts of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions.
+However, if we do not address the question of collapsibility among the data subsets arriving from e.g.
+different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.
 
-\begin{itemize}
-\item Surveys often use mixed modes of administration, e.g.
-mail and telephone, and while this can improve response rates, the mode of administration can affect results \citep{Brambilla1987,McHorney1994}, and differences in response behaviour can lead to biased results.
-Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful.
-\item The rapid growth of web surveys, due to low cost, timeliness, and other factors, generate large data sources that lack a sampling frame of the general population.
-However, it can be problematic to combine online panels (pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions), see \cite{Liu2016}.
-\item Large scale open source data sets such as the PISA data and ESS (European Social Survey) data.
-In these data sets, data analysts are far away from the data producers and problem-specific recommendations about potential instrument-induced challenges in the data sets are not available for the data analysts.
-\end{itemize}
+Sophisticated methods for addressing this question are available when we are willing to assume a statistical model.
+But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment very ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data.
+The use of (parametric) models does thus not constitute a valid general data structure comparison method, but rather a fitted-model comparison method.
+It addresses the interplay between the model and the data, not the data alone.
+However, useful tools for initial, exploratory investigations of data comparability data that help researchers identify potential problems or challenges early in the study design development are mostly absent.
+What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
+Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables.
+On the other hand, e.g.
+entry-by-entry comparisons of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase.
 
-\hl{Maybe broaden the raison d'etre of PCADSC a bit: It is also a useful tool for doing initial, exploratory data analysis and it helps us to identify potential problems or challenges early in the study design development, before concrete model choices have been made, thereby aiding a more economic use of resources and a less ad-hoc approach to the types of problems that PCADSC can find.}
+While the issue is not in any way new, its relative importance is increasing due to new developments in data availability and data collection.
+Classical statistical methods are generally aimed at analyzing data from designed experiments and historically, statistical analyses have been conducted by researchers who knew the design and the source of the dataset well.
+However, the origin stories of datasets have changed over time and today, a lot of data are accumulated without a specific purpose in mind, as data collection and sharing has become easier and more affordable through the development of new technologies.
+This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for all to use.
+However, large scale publicly available datasets, such as the PISA data and European Social Survey (ESS) data, are often used by data analysts that are far removed from the data producers and therefore, problem-specific recommendations about e.g.
+potential instrument-induced challenges in the datasets are not available for these data analysts.
+For this reason, a thorough investigation of data structure comparability becomes a crucial step in any meaningful data analysis when using such open-source data.
 
-Assume that we have two data sets with the same variables, but different observations represented as a data set with a subset-inducing variable.
-Assume next that we wish to compare structures without specifying a model, or even a variable of interest.
-The central question is whether the two data sets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be taken into account.
+The data structure comparison problem is however not limited to the context of new data sharing platforms.
+Even in single-population survey studies, a mixture of modes of administration is often used, e.g.
+mail and telephone, and while this can improve response rates, it can also be quite problematic \citep{Brambilla1987, McHorney1994}, and differences in response behavior can lead to biased results.
+Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful, so the comparability issue is not just a statistical puzzle.
+Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu \citeyearpar{Liu2016}, who warns against the combination of online panel data (i.e.
+pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions).
 
-Sophisticated methods for addressing this question are available when we are willing to assume a statistical model \hl{Skal der referencer her?
-Nogen forslag?}, but when these models are taken away, a remarkable void of methods is left behind.
-What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a data set without assuming neither directional nor hierarchical relationships between the variables.
-The use of parametric models does not constitute general data structure comparison method, but rather a fitted-model comparison method.
-It addresses the interplay between the model and the data, not the data alone.
-Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and not to the interplay between variables.
-Entry-by-entry comparison of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase.
-Parametric models using, e.g., latent variable models moves beyond the marginal approach, but need a pre-specified model.
-
-%For these methods significant $p$-values are likely to abound as the sample size increases. Bo: Jeg tror denne sætning bare kan udelades.
+The combarability issue that is present in all the examples presented so far can be summarized as follows: Assume that we have two datasets with the same variables, but different observations, or alternatively (and equivalently) a single dataset with a subset-inducing variable.
+Assume next that we wish to compare data structures without specifying a model, or even a variable of interest.
+The central question is then whether the two datasets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be somehow be dealt with, either in terms of modeling choices or simply by conducting several separate analyses.
 
@@
 These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
-This yields a solution that is largely independent of the sizes of the two subsets of data.
-The proposed tools are implemented in the statistical software \texttt{R} (\cite{R}) statistical software in our package, \texttt{PCADSC}, which is available at \hl{?
-How to refer to a package on Github?}.
+%This yields a solution that is largely independent of the sizes of the two subsets of data. Anne: Det ved vi da ikke om det gør? Er det ikke netop det, vi siger, at man burde teste som næste step?
+The proposed tools are implemented in the statistical software \texttt{R} in our package, \texttt{PCADSC} \citep{PCADSC}.
 
-This manuscript describes the procedures, including a brief introduction to principal component analysis (PCA) in general and presents a worked data example using open source, online available data on psychological well-being in three European countries.
-More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries.
+This manuscript describes the procedures, including a brief introduction to principal component analysis (PCA), and presents a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey.
+More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is completely ignored e.g.
+when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project.
 
@@
 \label{sec:pcadscintro}
-As mentioned above, the purpose of PCADSC is to compare overall data structures in two subsets of a data set.
+As mentioned above, the purpose of PCADSC is to compare overall data structures in two subsets of a dataset.
 But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context.
@@
 In particular, we find a new representation of the covariance matrix that makes it possible to gain an overview of the most interesting aspects of the data.
-On such decomposition strategy is using \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.
+One such decomposition strategy is known as \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.
 
@@
 PCA deconstructs the covariance matrix in components according to the most explained variance.
-This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data.
-Therefore, the variables often are standardized prior to performing PCA.
-Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices.
+This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will always be deemed highly influential, no matter the structure of the data.
+Therefore, the variables are typically standardized prior to performing PCA.
+Note that the covariance matrix of the standardized variables is the same as the correlation matrix of the original variables, so post-standardization PCA simply corresponds to performing data structure comparisons of the correlation matrices rather than the covariance matrices.
 The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation.
-In this paper we will adopt this approach, and in particular, we will allow for differences between scalings of the variables in the two data subsets.
+In this paper we will adopt this standardization policy, and in particular, we will allow for differences between scalings of the original, raw variables in the two data subsets.
 
-\hl{Bo: I næste afsnit bruges stadigvæk ``covariance''. Det er vel ok?}
+%\hl{Bo: I næste afsnit bruges stadigvæk ``covariance''. Det er vel ok?} Anne: Synes det fungerer fint.
 
-\subsection{A brief introduction to Principal component analysis}
+\subsection{A brief introduction to principal component analysis}
 Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = (\bar{x}_1, ..., \bar{x}_d)^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$.
@@
 The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \citep{HastieEtAl2009}.
-
 More specifically, the rank-q-reconstruction error is attained when we choose
@@
 \begin{equation*}
-\text{trace}(S) = \sum_{j=1}^d \hat{V}(X_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
+\text{trace}(S) = \sum_{j=1}^d \hat{V}(x_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
 \end{equation*}
-where $X_j \in \RR^n$ \hl{(Bo: $X_1$ og $X_2$ bruges i en anden betydning i afsnit 2.2.
-Omformuler?)} denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
+where $x_j \in \RR^n$ 
+%\hl{(Bo: $X_1$ og $X_2$ bruges i en anden betydning i afsnit 2.2. Omformuler?)} Anne: Hov, ja, hermed x_j så det passer med notationen ovenfor.
+denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
 This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
@@
 where the initial subspace is defined as $\hat{K}_0 = \{0\}$.
-It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
-This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$.
-This implies that the loadings and the variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension, $q$, a priori.
+It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
+This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$.
+This implies that the loadings and the variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.
 
 \subsection{Using PCA for data structure comparisons}
-All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful quantities to compare across such different datasets.
+All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets.
 Usually, when performing PCA with other purposes in mind, the main interest lies in the \textit{scores}, i.e.
@@
 If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both datasets.
-This can be the case while the two sets of scores could be arbitrarily different, which e.g.\ could happen if the two datasets were taken from two different populations of subjects.
-On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variables are used differently in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.
+This can be the case while the two sets of scores could be arbitrarily different, which could e.g.\ happen if the two datasets were taken from two different populations of subjects.
+On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.
 
@@
 \begin{enumerate}
-\item Standardized each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized datasets.
+\item Standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized datasets.
 \item Form the principal component analysis
@@
 \end{enumerate}
-The hereby obtains PCA decompositions of the correlation matrices can then be compared.
-We present three diagnostics plots that are designed to shine a light on different types and nuances of data structure differences.
+The hereby obtained PCA decompositions of the correlation matrices can then be compared.
+We present three diagnostics plots that are designed to shine a light on different types and levels of data structure differences.
 These three plots are:
@@
 
-The standardization makes the variables comparable on the same scale (units of standard deviation), and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
+Note that the standardization implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
 \begin{equation*}
@@
 Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
-The empirical correlation matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
+Then the empirical correlation matrix for the first dataset, $S_1$, has the following orthogonal decomposition in the coordinate system of the second dataset
 %\begin{equation*}
@@
 \end{equation*}
-and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset \hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?
+and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset %\hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?
 Bo: Jeg har ikke noget bud på en sådan udvidelse lige nu og her --- skal vi ikke bare sige at det er ok i sin nuværende form?}.
@@
 \end{align*}
-Note that for two $d$-dimensional, unit length vectors $a$ and $b$, $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the sample correlation.
+Note that for two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the sample correlation.
 Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance component contributions.
@@
 
-In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component for the first dataset in the coordinate system of the second dataset.
-Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset.
-\hl{Can we expand on this in a non-technical way?} If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
+In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the first dataset in the coordinate system of the second dataset.
+Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the second dataset in the coordinate system of the first dataset.
+%\hl{Can we expand on this in a non-technical way?
+- Karl: it seems that we cannot}
+If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
 Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
@@
 These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions.
-More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors and the wideness of the $j$th segments are given by
+More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors.
+The width of the $j$th colored segments are given by
 \begin{align*}
@@
 For each component, the loadings describe how influential the different variables are on that component.
-Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such that \textit{"variable $x$ is generally more influential in subset 1 than it is in subset 2"}, thereby helping us to understand where and why the data structure differences are found.
-
-%Our proposal of a PCADSC method consist of three steps.
-These steps should be performed separately for each of the two (or more) datasets that we wish to compare.
-Note that the datasets must have the same variables, but different sample sizes are allowed.
-The three steps are:
-%\begin{enumerate}
-%\item \hl{Standardize. Full data or subsets? And also: Something about what to do if not all variables are numerical.}
-%\item Compute the PCA loadings and the variance contributions of each principal component.
-%\item For each principal component, standardize the loadings, i.e. scale them such that they sum to one.
-%\item Produce a plot consisting of a bar for each principal component, decorated with the cumulative variance contribution corresponding to this component.
-The bar should be of length one and colored according to the variables loading the component.
-%\end{enumerate}
+Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such as \textit{"variable $x$ is generally more influential in subset 1 than it is in subset 2"}, thereby helping us to understand where and why potential data structure differences are found.
 
@@
 \label{sec:dataexample}
-We will now turn to a concrete data example in order to illustrate the capabilities of the methods presented above.
+We will now turn to a concrete data example in order to illustrate the capabilities of the tools presented above.
 We use data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
-This investigation is motivated by an increasingly popular new tendency to publish miscellaneous rankings of countries in fields as different as educational quality (e.g.
+The topic of our investigation is motivated by an increasingly popular tendency to publish rankings of countries in fields as different as educational quality (e.g.
 the PISA tests) and citizen happiness (e.g.
 the UN \textit{World Happiness Report} project).
-From a methodological point of view, such international rankings are very concerning, as they rely on the fundamental assumption  that the measured concepts are inherently the same across countries.
+From a methodological point of view, such international rankings are very problematic, as they rely on the fundamental assumption that the measured concepts are inherently the same across countries.
 The PCADSC tools qualify as a suite of methods for exploring the validity of this assumption empirically.
 
-For international comparisons of educational systems, the PISA tests have repeatedly been criticized for not being meaningful objects for international comparisons, especially due to problems with differential item functions (\cite{Kankaras2014}; \cite{Kreiner2014}) and translation problems (\cite{Asil2016}).
-\hl{Til Karl: Måske kender du nogle flere studier her, vi bør nævne?
-Bo: F.eks.
-\cite{ZwitserEtAl2017}?}
-
-In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though \cite{Veenhoven2012} presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability and \cite{Lolle2016} shows highly potent translation issues for the term \textit{happiness}.
-The main question is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
-Or do different aspects of psychological well-being or happiness simply not have the same relative meaning in different cultural and socio-economic settings?
+In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though \cite{Veenhoven2012} presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability, while \cite{Lolle2016} shows highly potent translation issues for the term \textit{happiness}.
+But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
+If not, the very idea of translation is meaningless, since different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socio-economic settings.
 This is in fact a question concerning comparability of data structures.
@@
 The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures.
-And therefore, comparing the two countries in these measures is not a meaningful endeavour.
-In this paper, we will focus on a single aspect of overall happiness, namely psychological well-being,
+And therefore, comparing the two countries in these measures is not a meaningful endeavor.
 
 In this section, we use the PCADSC tools to unveil international differences in one aspect of happiness, namely psychological well-being.
-Our starting point is Denmark, a small, northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 (\cite{WHR2016}), and we wish to investigate if this title is really meaningful at all.
+Our starting point is Denmark, a small, northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \citep{WHR2016}, and we wish to investigate if this title is really meaningful at all.
 In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria.
-Though both countries are European and thus not geographically nor culturally as far apart as some other countries might be, these two countries have previously been highlighted to be very different in terms of what defines happiness (\cite{ESStopline5} {\hl{Is this the correct way to refer to a technical report?
-Bo: Ja, jeg synes det er fint.}).
-Moreover,intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated (\cite{Rodriguez2014}), with a much less strong relationship between the two in Northern- compared to other European countries.
+Though both countries are European and thus not geographically nor culturally as far apart as some other countries might be, these two countries have previously been highlighted to be very different in terms of what defines happiness \citep{ESStopline5}.
+Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \citep{Rodriguez2014}, with a much less pronounced relationship between the two in Northern European countries, compared to other European countries.
 In particular, interpersonal relations should play a less important role in Denmark, compared to Bulgaria.
@@
 We also compare the Danish data with Swedish data in order to illustrate that the PCADSC tools actually do have some discriminatory power.
-Denmark and Sweden are both Scandinavian countries and are often deemed very similar in terms of culture and history.
-Therefore, we expect fundamental concepts such as psychological well-being to similar across these two countries.
+Denmark and Sweden are both Northern European countries and are often deemed very similar in terms of culture and history.
+Therefore, we expect fundamental concepts such as psychological well-being to be similar across these two countries.
 
-All computations and figures presented in this section where created using our \R{R} package \R{PCADSC}, which is available online at \url{www.github.com/AnnePetersen1/PCADSC}.
+All computations and figures presented in this section were created using our \R{R} package \R{PCADSC} \citep{PCADSC}.
 
@@
 These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}.
-More details on these scales can be found in (\cite{ESStopline5}) and the relationship between questionnaire items and scales is summarized in Table \ref{table:items} in the appendices.
+More details on these scales can be found in \citep{ESStopline5} and the relationship between questionnaire items and scales is summarized in Table \ref{table:items} in the appendices.
 We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10.
@@
 \begin{figure}
-\center
-\includegraphics[scale = 0.7]{essDKBGce.pdf}
-\includegraphics[scale = 0.7]{essDKBGhair.pdf}
+\centerline{\includegraphics[scale = 0.65]{essDKBGce.pdf}}
+\centerline{\includegraphics[scale = 0.65]{essDKBGhair.pdf}}
 \caption{The CE plot (top) and the angle plot (bottom) resulting from comparing Bulgarian and Danish data on psychosocial well-being.
-Dataset 1 refers to the Bulgarian subsample, while Dataset 2 is the Danish data.
-The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the assumption of no difference in data structures.
+The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
 In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
@@
 \begin{figure}
-\center
-\includegraphics[scale = 0.7]{essDKBGpancake234.pdf}
+\centerline{\includegraphics[scale = 0.7]{essDKBGpancake234.pdf}}
 \caption{A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data.
@@
 %\begin{figure}
-%\center
-%\includegraphics[scale = 0.7]{essDKBGWallyPCADSC234.pdf}
+%\centerline{\includegraphics[scale = 0.7]{essDKBGWallyPCADSC234.pdf}}
 %\caption{\hl{something.}}
@@
 
-Moving on to the angle plot, we find that the differences are primarily to be found in the second, third and fourth principal components (PCs): The blue arrows visualize the decomposition of the principal components for the Bulgarian dataset in the coordinate system of the Danish dataset.
+Moving on to the angle plot, we find that the differences are primarily to be found in the second, third and fourth principal components (PCs).
+The blue arrows visualize the decomposition of the principal components for the Bulgarian dataset in the coordinate system of the Danish dataset.
 We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3.
@@
 
-All in all, we find that psychological well-being does not seem to be the same concept in Bulgaria and Denmark.
+All in all, we find that psychological well-being does \textit{not} seem to be the same concept in Bulgaria and Denmark.
 The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalues) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle- and chroma plots).
-In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, which corresponds with previous findings.
+In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, corresponding well with the previous findings mentioned above.
 Thus, the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics (\cite{ESStopline5}).
@@
 \begin{figure}
-\center
-\includegraphics[scale = 0.7]{essDKSEce.pdf}
-\includegraphics[scale = 0.7]{essDKSEhair.pdf}
-\caption{A CE (top) and a hair (bottom) plot for comparing the Danish (Dataset 1) and the Swedish (Dataset 2) psychological well-being data.
+\centerline{\includegraphics[scale = 0.65]{essDKSEce.pdf}}
+\centerline{\includegraphics[scale = 0.65]{essDKSEhair.pdf}}
+\caption{A CE (top) and a hair (bottom) plot for comparing the Danish and the Swedish psychological well-being data.
 The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
@@
 \center
-\includegraphics[scale = 0.7]{essDKSEpancake.pdf}
+\centerline{\includegraphics[scale = 0.7]{essDKSEpancake.pdf}}
 \caption{A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples.
-Note that the bars for each component is annotated with its cumulative variance score (denoted $\tilde\sigma^c_i$ in the above), that is, how much variance can be explained by having information of this and the preceding components.}
+Note that for each component, the bar is annotated with its cumulative variance contribution (denoted $\tilde\sigma^c_i$ in the above), that is, how much variance can be explained by having information of this and the preceding components.}
 \label{plotSE.pancake}
@@
 In the CE plot, we now find the cumulative eigenvalue curve to be just within the acceptance region of the null-hypothesis.
-This is also reflected by the two tests, which now produce $p$-values of $p_\text{KS} = 0.14$ and $p_\text{CvM} = 0.09$, respectively.
-Thus, it is not very unreasonable to assume  equal variance components in the two datasets.
+This is also reflected by the two tests, which now produce $p$-values of $p_\text{KS} = 0.17$ and $p_\text{CvM} = 0.11$, respectively.
+Thus, it is not unreasonable to assume equal variance components in the two datasets.
 
@@
 
-The angle plot from Figure \ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent.
+The angle plot in Figure \ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent.
 This implies that if one already has e.g.
@@
 \label{sec.Discussion}
-When deciding whether to combine two data sets for analysis, the issue of heterogeneity across data sets must be addressed.
-Simple methods suffer drawbacks and will likely scale poorly.
-Parametric models using, e.g., latent variable models moves beyond the marginal approach, but need a pre-specified model.
+Three new tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if two datasets can be combined for analysis were proposed and discussed in this paper.
+They all employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
+% yielding a solution that is largely independent of the sizes  the two datasets.
+Anne: We don't know if that is true.
 
-New tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if the two two data sets can be combine for analysis were proposed and discussed in the paper.
-They employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences yielding a solution that is largely independent of the sizes  the two data sets.
+In an analysis of data from the 2012 version of the European Social Survey (ESS) project, we illustrated that the PCADSC tools help inform analyses of inter-country differences in psychological well-being and happiness.
+It should be noted that even though concerns about pooling data from different countries are not yet very prevalent when it comes to happiness rankings, the subject has been very well scrutinized in the field of international educational rankings.
+For instance, the PISA tests have repeatedly been criticized for not being meaningful objects of international comparisons, especially due to problems with differential item functions \citep{Kankaras2014,Kreiner2014,ZwitserEtAl2017} and translation problems \citep{Asil2016}.
+This highlights lack of data comparability as a empirically relevant and concerning issue that there is no way around: When conducting internal rankings, the issue of data structure heterogeneity should always be addressed.
 
-\hl{
-The methodology is quite general and in principle any plot that can be made can be included ('di-scree plot') ... [rephrase!!]
-}
+Though the PCADSC methods comprises three diagnostic plots as of now, the PCA decomposition can of course be utilized further to bring about new PCADSC plots that illustrate more aspects of data structure differences.
+For instance, the scree plot (as introduced by \cite{Cattell1966}), which plots the ordered eigenvalues against their component numbers, could easily be recreated as a PCADSC tool by creating scree plots for both data subsets in the same coordinate system, yielding a simple graphical comparison.
+The resampling approach used in the CE plot could then be adopted, thus annotating the scree plots with resampled curves.
+This general approach to model-fit evaluation using cumulative residuals, as inspired by Lin et al.
+\citeyearpar{LinEtAl2002}, is very helpful in combination with the permutation tests.
 
-Further topics need to be addressed.
-These include generalizing the methods to be able to address (i) binary, ordinal or even nominal categorical variables, (ii) covariance matrices that are not of full rank.
+The proposed methods also have limitations.
+First of all, as PCA is performed after standardization of the variables, the PCADSC methods cannot detect differences in mean values nor in marginal variances - this information is thrown away as the very first step.
+However, such differences are of a fundamentally different nature than those we have discussed so far.
+Differences in mean values are often the main interest of the analysis and should of course not be regarded as a data comparability issue.
+Differences in marginal variances, on the other hand, can pose some modeling challenges, but none that are not generally solvable by use of random effects models.
+We therefore do not consider this property of PCADSC to be a major drawback.
 
-More evaluation of the performance is also needed. Investigating sensitivity towards the sample sizes $n_1$ and $n_2$
+A more prominent limitation is the fact that the PCASDC methods are only valid for variables that have a numerical interpretation.
+In particular, this means that the methods can not be used for data with nominal categorical variables.
+This limitation is inherited from the PCA method and thus, moving beyond it will entail replacing the PCA framework with a more general one.
+Multiple correspondence analysis (MCA) is often suggested as a categorical extension of PCA \citep{Abdi2010} and including MCA in the PCADSC tools is thus a natural next step.
 
-The limitations of the PCADSC procedures should also be sudied in more detail.
-It seems unlikely that the procedure would be able to disclose differences in scaling, since all variables are standardized in the procedure.
-This type of heterogeneity should thus be adjusted for in later analyses of the combined data set.
+Moreover, though we have illustrated the discriminatory power of PCASDC in one data example, of course further evaluation of its performance is needed.  
+While the best feature of PCADSC is perhaps the intuitive, visual nature of the tools, it is also the biggest weakness.
+For statistical methods that result in a number (or a set of numbers), the standard way of evaluating performance is to report on simulation studies, but in the case of statistical methods that report graphics this is not feasible.
+Therefore, a systematic evaluation of the performance of the PCADSC methods is not very straight-forward to design: It really depends on whether or not users can detect actual differences in data structures by using the diagnostic plots.
+We suggest that the variability of the methods could be studied further using \textit{Wally plots} \citep{Ekstrøm2014}, which are learning tools developed for teaching students how to interpret residual plots.
+First in line as a topic for more thorough investigations of performance is the sensitivity towards the sample sizes of the two datasets, $n_1$ and $n_2$.
+
+All in all, PCADSC represents a first step towards addressing a known issue that has otherwise been met with unsatisfactory ad-hoc methods whose assumptions rest upon the very hypothesis they are testing.
+The PCADSC tools perform quite well in the data example provided here, yielding conclusions that are in accordance with well-established theories and they are simple to use and test further using the PCADSC \texttt{R} package.
 
@@
 
-\vspace{\fill}\clearpage
-\newpage
 \bibliographystyle{apa}
@@
 
-\section{Supplementary tables}
+%\newpage
+%\section{Supplementary tables}
 \begin{table}[h]
 \centering
-\small
+\scriptsize
 \rowcolors{1}{}{lightgray}
+\bgroup
+\def\arraystretch{1.5}
 \begin{tabular}{ll}
@@
 
-\rowcolor{white} & Felt everything did an effort, how often in the past week\\
+\rowcolor{white} & Felt everything did as effort, how often in the past week\\
 \rowcolor{white} & Sleep was restless, how often in the past week\\
@@
 \rowcolor{lightgray} &  Feel people in local area help one another\\
-\rowcolor{lightgray}\multirow{-2}{*}{Community wellbeing}& Feel close to the people in local area\\
+\rowcolor{lightgray}\multirow{-5}{*}{Community wellbeing}& Feel close to the people in local area\\
 
@@
 \end{tabular}
+\egroup
 \caption{Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.

### 2017-04-25 article/article_v0.9.5.tex
Rettelser og ændringer tilføjet. Om muligt har jeg udkommenteret tidliger ord eller sætninger via '%'. Således at disse er lettere at finde.

@@
 
-In empirical psychometric work, one often encounters data that is somehow divided into two distinct susbets.
+In empirical psychometric work, one often encounters data that is somehow divided into two distinct subsets.
 This might be due to multi-center sampling frames, or to variations in instruments, orders of questionnaire item, modes of administration etc.
-Often, one wishes to combine such data into one an conduct only a single analysis, but this is only meaningful if the two subsets are fundamentally similar in structures, that is, if they measure the same underlying concepts.
-However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc and require making full model assumptions.
-We present a suite of visual, non-parametric diagnostic tools for investigating differences in data structures for two datasets containing different observations of the same variables, refered to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC).
+Often, one wishes to combine such data into one an conduct only a single analysis, but this is only meaningful if the two subsets are similar in structures, that is, if they measure the same underlying concepts.
+However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc or require making full model assumptions.
+We present a suite of three visual, non-parametric diagnostic tools for investigating differences in data structures for two datasets containing different observations of the same variables, refered to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC).
 This approach requires no directional nor hierarchical assumptions on the data.
-Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets and by each targeting a different aspect of the PCA information, the three plots presented here give a thorough and deep understanding of where differences occur and why it might be the case.
+Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets, and by targeting different aspects of the PCA information, the three plots give a thorough and deep understanding of where differences occur and why it might be the case.
 
@@
 Data comparability is a classical topic in psychometrics and, more generally, in statistics.
-Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability (and thus collapsibility) needs to be assesed empirically.
+Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability (and thus collapsibility) needs to be assessed empirically.
 This happens e.g.
@@
 Sophisticated methods for addressing this question are available when we are willing to assume a statistical model.
-But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment very ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data.
-The use of (parametric) models does thus not constitute a valid general data structure comparison method, but rather a fitted-model comparison method.
+But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment %very
+ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a %valid 
+general data structure comparison method, but rather a fitted-model comparison method.
 It addresses the interplay between the model and the data, not the data alone.
@@
 However, the origin stories of datasets have changed over time and today, a lot of data are accumulated without a specific purpose in mind, as data collection and sharing has become easier and more affordable through the development of new technologies.
-This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for all to use.
+This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for %all
+everyone to use.
 However, large scale publicly available datasets, such as the PISA data and European Social Survey (ESS) data, are often used by data analysts that are far removed from the data producers and therefore, problem-specific recommendations about e.g.
-potential instrument-induced challenges in the datasets are not available for these data analysts.
+potential instrument-induced challenges in the datasets %are not
+may not be available for these data analysts.
 For this reason, a thorough investigation of data structure comparability becomes a crucial step in any meaningful data analysis when using such open-source data.
@@
 
-The combarability issue that is present in all the examples presented so far can be summarized as follows: Assume that we have two datasets with the same variables, but different observations, or alternatively (and equivalently) a single dataset with a subset-inducing variable.
+The comparability issue that is present in all the examples presented so far can be summarized as follows: Assume that we have two datasets with the same variables, but different observations, or alternatively (and equivalently) a single dataset with a subset-inducing variable.
 Assume next that we wish to compare data structures without specifying a model, or even a variable of interest.
@@
 More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is completely ignored e.g.
-when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project.
+when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project \citep{WHR2016}.
 
@@
 If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables.
-This gives it a very nice interpretation as a measure of the overall structure.
-If we do not accept the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data.
+%This gives it a very nice interpretation as a measure of the overall structure. If we do not accept 
+And without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data.
 All in all, the empirical covariance matrix is a reasonable place to start looking for differences in \textit{overall data structures}.
@@
 \begin{equation*}
-\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
+\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\},
 \end{equation*}
@@
 This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
-And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different, the representation hereby obtained is a valid object of inference.
+And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are %different,
+distinct, the representation hereby obtained is a valid object of inference.
 If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
@@
 \eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
-\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
+\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2,
 \end{align*}
@@
 We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the apparently same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$.
-For each of these two datasets, we complete the following steps:
+%
+Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For each of these %two
+three datasets, we complete the following steps:
 \begin{enumerate}
-\item Standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized datasets.
+\item Within each dataset standardize each of the variables to have mean zero and unit standard deviation.
+Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ and $\tilde{x}_j \in \RR^d$ for $j=1,\dotsc,n$ be the standardized datasets.
 \item Form the principal component analysis
-\begin{equation*}
-S_i = \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$.}
-\end{equation*}
-thereby obtaining loadings $\eta_{ik}$ and variance components $\lambda_{ik}$ for $i \in \{1, 2\}$ and $k \in 1,\dotsc,d$.
+\begin{align*}
+S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} &
+S &= \frac{1}{n-1} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top,
+\end{align*}
+thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda$ for $k=1,\dotsc,d$.
 \end{enumerate}
@@
 10000) number of times.
-Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to even larger test statistics than the one we found for the original datasets.
+Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to %even larger
+test statistics at least as large as %than 
+the one we found for the original datasets.
 
@@
 \centerline{\includegraphics[scale = 0.65]{essDKSEhair.pdf}}
-\caption{A CE (top) and a hair (bottom) plot for comparing the Danish and the Swedish psychological well-being data.
+\caption{A CE (top) and a angle (bottom) plot for comparing the Danish and the Swedish psychological well-being data.
 The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
@@
 The resampling approach used in the CE plot could then be adopted, thus annotating the scree plots with resampled curves.
-This general approach to model-fit evaluation using cumulative residuals, as inspired by Lin et al.
-\citeyearpar{LinEtAl2002}, is very helpful in combination with the permutation tests.
+This general approach to model-fit evaluation is very helpful in combination with permutation tests, %using cumulative residuals,
+as inspired by Lin et al. \citeyearpar{LinEtAl2002}.
+Similarly, the angle plot could also be annotated with p-values from a permutation test based on appropriate test statistics.
 

### 2017-04-25 article/article_v0.9.5.tex
Looked through Bo's comments and added minor extra stuff.

@@
 %
-Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For each of these %two
+Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For these %two
 three datasets, we complete the following steps:
 \begin{enumerate}
-\item Within each dataset standardize each of the variables to have mean zero and unit standard deviation.
+\item Within each dataset, standardize each of the variables to have mean zero and unit standard deviation.
 Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ and $\tilde{x}_j \in \RR^d$ for $j=1,\dotsc,n$ be the standardized datasets.
-\item Form the principal component analysis
+\item Form the principal component analyses
 \begin{align*}
@@
 \end{align*}
-thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda$ for $k=1,\dotsc,d$.
+thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda_k$ for $k=1,\dotsc,d$.
 \end{enumerate}
@@
 
-Though the PCADSC methods comprises three diagnostic plots as of now, the PCA decomposition can of course be utilized further to bring about new PCADSC plots that illustrate more aspects of data structure differences.
+Though the PCADSC toolbox comprises three diagnostic plots as of now, the PCA decomposition can of course be utilized further to bring about new PCADSC plots that illustrate more aspects of data structure differences.
 For instance, the scree plot (as introduced by \cite{Cattell1966}), which plots the ordered eigenvalues against their component numbers, could easily be recreated as a PCADSC tool by creating scree plots for both data subsets in the same coordinate system, yielding a simple graphical comparison.
@@
 This general approach to model-fit evaluation is very helpful in combination with permutation tests, %using cumulative residuals,
-as inspired by Lin et al. \citeyearpar{LinEtAl2002}.
-Similarly, the angle plot could also be annotated with p-values from a permutation test based on appropriate test statistics.
+as inspired by Lin et al. \citeyearpar{LinEtAl2002}. Similarly, the angle plot could also be annotated with p-values from a permutation test based on appropriate test statistics.
 

### 2017-04-25 article/article_v0.9.5.tex
Added (Anne's revision of) Karl's comments and changes.

@@
 
-Data comparability is a classical topic in psychometrics and, more generally, in statistics.
+Data comparability is a classical topic in statistics and a recurring one in psychometrics.
 Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability (and thus collapsibility) needs to be assessed empirically.
 This happens e.g.
-when data are collected across several centers (or even countries) or when different versions of an instrument, the mode of administration or the order of survey questionnaire items are applied.
+when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration or the order of survey questionnaire items are applied.
 All of these practices are essential parts of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions.
@@
 
-\section{European differences in psychological well-being: A data example}
+\section{Comparing psychological well-being: A data example}
 \label{sec:dataexample}
@@
 The topic of our investigation is motivated by an increasingly popular tendency to publish rankings of countries in fields as different as educational quality (e.g.
-the PISA tests) and citizen happiness (e.g.
+the PISA project) and citizen happiness (e.g.
 the UN \textit{World Happiness Report} project).
@@
 If two countries differ e.g.
-in terms of how social networks are typically build and structured, with one emphasizing family relations and the other mostly focusing on other social relations, having a weak family connection does not have the same implications in the first country as it does in the second one.
+in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one.
 More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being.
@@
 \centering
-\begin{tabular}{lccccccccc}
+%\begin{tabular}{lccccccccc}
+%  \hline
+%  & \multicolumn{3}{c}{Denmark} & \multicolumn{3}{c}{Bulgaria} & \multicolumn{3}{c}{Sweden} \\
+% & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \\
+%  \hline
+% Evaluative wellbeing & 8.00 & 8.75 & 9.50 & 3.50 & 5.00 & 7.00 & 7.00 & 8.00 & 9.00 \\
+% Emotional wellbeing & 7.22 & 8.33 & 8.89 & 5.00 & 6.67 & 7.78 & 6.67 & 7.78 & 8.89 \\
+%Functioning & 6.93 & 7.57 & 8.21 & 5.50 & 6.68 & 7.68 & 6.39 & 7.04 & 7.68 \\
+%Vitality & 6.67 & 7.50 & 8.33 & 5.83 & 7.50 & 8.33 & 6.67 & 7.50 & 9.17 \\
+% Community wellbeing & 5.83 & 6.77 & 7.57 & 3.70 & 4.67 & 5.70 & 5.66 & 6.57 & 7.37 \\
+%Supportive relationships & 7.42 & 8.25 & 8.92 & 6.17 & 7.25 & 8.08 & 7.42 & 8.25 & 8.75 \\
+%   \hline
+%\end{tabular}
+\begin{tabular}{lccc}
   \hline
-  & \multicolumn{3}{c}{Denmark} & \multicolumn{3}{c}{Bulgaria} & \multicolumn{3}{c}{Sweden} \\
- & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \\
+  & Denmark & Bulgaria & Sweden \\
   \hline
- Evaluative wellbeing & 8.00 & 8.75 & 9.50 & 3.50 & 5.00 & 7.00 & 7.00 & 8.00 & 9.00 \\
- Emotional wellbeing & 7.22 & 8.33 & 8.89 & 5.00 & 6.67 & 7.78 & 6.67 & 7.78 & 8.89 \\
-Functioning & 6.93 & 7.57 & 8.21 & 5.50 & 6.68 & 7.68 & 6.39 & 7.04 & 7.68 \\
-Vitality & 6.67 & 7.50 & 8.33 & 5.83 & 7.50 & 8.33 & 6.67 & 7.50 & 9.17 \\
- Community wellbeing & 5.83 & 6.77 & 7.57 & 3.70 & 4.67 & 5.70 & 5.66 & 6.57 & 7.37 \\
-Supportive relationships & 7.42 & 8.25 & 8.92 & 6.17 & 7.25 & 8.08 & 7.42 & 8.25 & 8.75 \\
+ Evaluative wellbeing & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
+ Emotional wellbeing & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
+Functioning & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
+Vitality & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
+ Community wellbeing & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
+Supportive relationships & 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
    \hline
 \end{tabular}
-\caption{The 1st quartile, the median and the third quartile of the distrubutions of each of the six dimensions of psychological well-being, stratified by country.
+
+\caption{Median values of each of the six dimensions of psychological well-being, stratified by country.
+25 and 75 percentiles are listed in parentheses.
 Note that the scales are constructed such that they all run from 0-10.}

### 2017-05-17 article/article_v0.9.5.tex
Added corrections according to Niels Keiding's feedback

@@
 This might be due to multi-center sampling frames, or to variations in instruments, orders of questionnaire item, modes of administration etc.
-Often, one wishes to combine such data into one an conduct only a single analysis, but this is only meaningful if the two subsets are similar in structures, that is, if they measure the same underlying concepts.
+Often, one wishes to combine such data into one and conduct only a single analysis, but this is only meaningful if the two subsets are similar in structures, that is, if they measure the same underlying concepts.
 However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc or require making full model assumptions.
@@
 It addresses the interplay between the model and the data, not the data alone.
-However, useful tools for initial, exploratory investigations of data comparability data that help researchers identify potential problems or challenges early in the study design development are mostly absent.
+However, useful tools for initial, exploratory investigations of data comparability that help researchers identify potential problems or challenges early in the study design development are mostly absent.
 What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
@@
 On the other hand, e.g.
-entry-by-entry comparisons of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase.
+entry-by-entry comparisons of the two empirical correlation matrices quickly become unmanageable as the number of variables increase.
 
@@
 One such definition is the structure of the covariance matrix of the dataset.
-If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables.
+If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the joint distribution of all the variables.
 %This gives it a very nice interpretation as a measure of the overall structure. If we do not accept 
@@
 The hereby obtained PCA decompositions of the correlation matrices can then be compared.
-We present three diagnostics plots that are designed to shine a light on different types and levels of data structure differences.
+We present three diagnostic plots that are designed to shine a light on different types and levels of data structure differences.
 These three plots are:
@@
 \subsubsection{The cumulative eigenvalue plot}
-The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix.
+The cumulative eigenvalue (CE) plot compares the variation components, i.e.
+the eigenvalues of the correlation matrix.
+These eigenvalues represent how much information is withheld in each component in terms of explained variance.
+Thus, by comparing cumulative sums of variance components, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative.
 %If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
@@
 (0,0), &&
-(\lambda_1,\lambda_{11}-\lambda_{12}), &&
+(\lambda_1,\lambda_{11}-\lambda_{21}), &&
 (\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
@@
 \end{align*}
-This may be seen as a cumulative Bland-Altman plot \citep{AltmanBland83,LinEtAl2002} for the variation components.
+%This may be seen as a cumulative Bland-Altman plot \citep{AltmanBland83,LinEtAl2002} for the variation components.
+%Anne: NK pointed out that this is somewhat confusing for people who are not familiar with the Bland-Altman plot, and as far as I can tell, what we do is quite different anyway, so I think no reference is really needed(?)
 Note that due to the standardization, the last point will always be equal to $(d,0)$.
@@
 And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are.
-Moreover, a positive cumulative differences implies that dataset 1 holds more information in the first components than dataset 2 does.
+Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does.
 
@@
 
-In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though \cite{Veenhoven2012} presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability, while \cite{Lolle2016} shows highly potent translation issues for the term \textit{happiness}.
+In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though \cite{Veenhoven2012} presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability, while \cite{Lolle2016} show highly potent translation issues for the term \textit{happiness}.
 But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
@@
 
-In this section, we use the PCADSC tools to unveil international differences in one aspect of happiness, namely psychological well-being.
-Our starting point is Denmark, a small, northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \citep{WHR2016}, and we wish to investigate if this title is really meaningful at all.
+In this section, we use the PCADSC tools to examine international differences in one aspect of happiness, namely psychological well-being.
+Our starting point is Denmark, a small, Northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \citep{WHR2016}, and we wish to investigate if this title is really meaningful at all.
 In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria.
-Though both countries are European and thus not geographically nor culturally as far apart as some other countries might be, these two countries have previously been highlighted to be very different in terms of what defines happiness \citep{ESStopline5}.
-Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \citep{Rodriguez2014}, with a much less pronounced relationship between the two in Northern European countries, compared to other European countries.
-In particular, interpersonal relations should play a less important role in Denmark, compared to Bulgaria.
+Though both countries are European and thus neither very far apart geographically nor culturally, these two countries have previously been highlighted to be very different in terms of what defines happiness \citep{ESStopline5}.
+Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \citep{Rodriguez2014}.
+When comparing Northern European countries to other European countries, a much less pronounced relationship between the two concepts is found.
+In particular, interpersonal relations play a less important role in Denmark, compared to Bulgaria.
 Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.
 
-We also compare the Danish data with Swedish data in order to illustrate that the PCADSC tools actually do have some discriminatory power.
+We also compare the Danish data with Swedish data in order to investigate if the PCADSC tools actually do hold discriminatory power.
 Denmark and Sweden are both Northern European countries and are often deemed very similar in terms of culture and history.
-Therefore, we expect fundamental concepts such as psychological well-being to be similar across these two countries.
+Therefore, we expect fundamental concepts, such as psychological well-being, to be similar across these two countries.
 
@@
 
-The ESS 2012 data contains a total of 626 variables collected from 54673 citizens of 29 countries.
+The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries.
 Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
@@
 For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below.
-This gives us $n_{DK} = 1498$ observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations.
+This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample.
+All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations.
 Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
@@
 \caption{A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data.
-The component-bars are annotated with their relative variance contributions (denoted $\tilde\sigma_i$ in the above).}
+The component-bars are annotated with their relative variance contributions (denoted $\tilde\sigma$ in the previous section).}
 \label{plotBG.pancake}
@@
 
-The chroma blot in Figure \ref{plotBG.pancake} allows us to look closer into these components.
+The chroma plot in Figure \ref{plotBG.pancake} allows us to look closer into these components.
 Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish.
@@
 In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, corresponding well with the previous findings mentioned above.
-Thus, the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics (\cite{ESStopline5}).
-Moreover, the two countries cannot be ranked in terms of which country is "the most happy", at least not by referring to psychological well-being dimensions such as those encountered here.
+Thus, the datasets are fundamentally different and we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics (\cite{ESStopline5}).
+Moreover, this also implies that two countries cannot be ranked in terms of which country is "the most happy", at least not by referring to psychological well-being dimensions such as those encountered here.
 
@@
 \caption{A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples.
-Note that for each component, the bar is annotated with its cumulative variance contribution (denoted $\tilde\sigma^c_i$ in the above), that is, how much variance can be explained by having information of this and the preceding components.}
+Note that for each component, the bar is annotated with its cumulative variance contribution (denoted $\tilde\sigma^c$ in the previous section), that is, how much variance can be explained by having information of this and the preceding components.}
 \label{plotSE.pancake}
@@
 
-Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80 \% of the variance in both datasets), and slight, but increasing, differences in the remaining three components.
-We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25 \% of the variance in both datasets.
+Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80\% of the variance in both datasets), and slight, but increasing, differences in the remaining three components.
+We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25\% of the variance in both datasets.
 In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g.
@@
 It should be noted that even though concerns about pooling data from different countries are not yet very prevalent when it comes to happiness rankings, the subject has been very well scrutinized in the field of international educational rankings.
-For instance, the PISA tests have repeatedly been criticized for not being meaningful objects of international comparisons, especially due to problems with differential item functions \citep{Kankaras2014,Kreiner2014,ZwitserEtAl2017} and translation problems \citep{Asil2016}.
-This highlights lack of data comparability as a empirically relevant and concerning issue that there is no way around: When conducting internal rankings, the issue of data structure heterogeneity should always be addressed.
+For instance, the PISA tests have repeatedly been criticized for not being meaningful objects of international comparisons, especially due to problems with differential item functioning \citep{Kankaras2014,Kreiner2014,ZwitserEtAl2017} and translation problems \citep{Asil2016}.
+This highlights lack of data comparability as an empirically relevant and concerning issue that there is no way around: When conducting internal rankings, the issue of data structure heterogeneity should always be addressed.
 
@@
 This general approach to model-fit evaluation is very helpful in combination with permutation tests, %using cumulative residuals,
-as inspired by Lin et al. \citeyearpar{LinEtAl2002}. Similarly, the angle plot could also be annotated with p-values from a permutation test based on appropriate test statistics.
+as inspired by Lin et al. \citeyearpar{LinEtAl2002}. Similarly, the angle plot could also be annotated with $p$-values from a permutation test based on appropriate test statistics.
 
 The proposed methods also have limitations.
-First of all, as PCA is performed after standardization of the variables, the PCADSC methods cannot detect differences in mean values nor in marginal variances - this information is thrown away as the very first step.
+First of all, as PCA is performed after standardization of the variables, the PCADSC methods cannot detect differences in neither mean values nor in marginal variances - this information is thrown away as the very first step.
 However, such differences are of a fundamentally different nature than those we have discussed so far.
@@
 A more prominent limitation is the fact that the PCASDC methods are only valid for variables that have a numerical interpretation.
-In particular, this means that the methods can not be used for data with nominal categorical variables.
+In particular, this means that the methods cannot be used for data with nominal categorical variables.
 This limitation is inherited from the PCA method and thus, moving beyond it will entail replacing the PCA framework with a more general one.

### 2017-05-17 ESS/ess_notes.R

@@ -8,7 +8,10 @@ useVars <- c("cntry", "stflife", "happy", "fltsd", "fltdpr", "enjlf", "wrhpp", "
              "wrbknrm", "deaimpp", "flteeff", "slprl", "cldgng", "enrglot", "ppltrst",
              "pplfair", "pplhlp", "pplahlp", "flclpla", "inprdsc", "flapppl", "rehlppl",
              "fltlnl")
-#
+
+nAllDK <- nrow(data[data$cntry == "DK", ])
+nAllBG <- nrow(data[data$cntry == "BG", ])
+nAllSE <- nrow(data[data$cntry == "SE", ])
 #anData <- na.omit(data[data$cntry %in% c("DK", "BG"), useVars])
 anData <- na.omit(data[data$cntry %in% c("DK", "SE"), useVars])
 
@@ -140,6 +143,15 @@ anData$SupportiveRelationships <- rowMeans(anData[, sMat$var[sMat$scale ==
                                                                "Supportive relationships"]])
 
 data_BGDK <- anData
+
+#Amount of missing information
+nMissDK <- nAllDK - sum(data_BGDK$cntry == "DK")
+nMissSE <- nAllSE - sum(data_SEDK$cntry == "SE")
+nMissBG <- nAllBG - sum(data_BGDK$cntry == "BG")
+pctMissDK <- nMissDK/nAllDK
+pctMissSE <- nMissSE/nAllSE
+pctMissBG <- nMissBG/nAllBG
+
 
 #Summarize data
 uVs <- c("EvaluativeWellbeing",

### 2017-06-18 ESS/ess_notes.R
New code for saving figure files

@@ -191,6 +191,30 @@ library(PCADSC)
 library(ggplot2)
 library(gridExtra)
 
+#DK/BG
+set.seed(123424)
+a2 <- PCADSC(data_BGDK, "cntry",  c("EvaluativeWellbeing",
+                                    "EmotionalWellbeing",
+                                    "Functioning",
+                                    "Vitality",
+                                    "CommunityWellbeing",
+                                    "SupportiveRelationships"))
+
+fig5_CE <- CEPlot(a2) +
+  scale_y_continuous(breaks = round(seq(-0.3, 0.5, 0.1),1), limits = c(-0.2, 0.5))
+fig5_angle <- anglePlot(a2)
+
+fig6_chroma <- chromaPlot(a2, useComps = 2:4) #figure 6
+
+
+ggsave(plot = grid.arrange(fig5_CE, fig5_angle, nrow = 2), file = "./article/PlosOne/Fig5.eps",
+       device = "eps", dpi = 400, width = 7, height = 8)
+
+ggsave(plot = fig6_chroma, file = "./article/PlosOne/Fig6.eps",
+       device = "eps", dpi = 400, width = 7, height = 4)
+
+
+#DK/SE
 set.seed(1234)
 a1 <- PCADSC(data_SEDK, "cntry",  c("EvaluativeWellbeing",
                                "EmotionalWellbeing",
@@ -198,60 +222,16 @@ a1 <- PCADSC(data_SEDK, "cntry",  c("EvaluativeWellbeing",
                                "Vitality",
                                "CommunityWellbeing",
                                "SupportiveRelationships"))
-
-
-p1 <- chromaPlot(a1)
-
-ggsave(plot = p1, file = "./article/essDKSEpancake.pdf",
-       width = 8, height = 5)
-
-set.seed(123424)
-a2 <- PCADSC(data_BGDK, "cntry",  c("EvaluativeWellbeing",
-                                       "EmotionalWellbeing",
-                                       "Functioning",
-                                       "Vitality",
-                                       "CommunityWellbeing",
-                                       "SupportiveRelationships"))
-
-
-p2 <- chromaPlot(a2)
-
-ggsave(plot = p2, file = "./article/essDKBGpancake.pdf",
-       width = 8, height = 5)
-
-p2c <- chromaPlot(a2, useComps = 2:4)
-
[…]

### 2017-06-18 ESS/sim_notes.R

@@ -37,29 +37,46 @@ po1 <- PCADSC(dF1, "group")
 po2 <- PCADSC(dF2, "group")
 
 CE1 <- CEPlot(po1) +
-  scale_y_continuous(breaks = round(seq(-0.6, 0.6, 0.1),1), limits = c(-0.5, 0.3))
+  scale_y_continuous(breaks = round(seq(-0.6, 0.6, 0.1),1), limits = c(-0.5, 0.3)) +
+  ggtitle("Dataset B") +  theme(plot.title = element_text(hjust = 0.5))
 
 CE2 <- CEPlot(po2) +
-  scale_y_continuous(breaks = round(seq(-0.6, 0.6, 0.1),1), limits = c(-0.3, 0.3))
+  scale_y_continuous(breaks = round(seq(-0.6, 0.6, 0.1),1), limits = c(-0.3, 0.3)) +
+  ggtitle("Dataset A") +  theme(plot.title = element_text(hjust = 0.5))
+
+angle1 <- anglePlot(po1) +
+  ggtitle("Dataset B") +  theme(plot.title = element_text(hjust = 0.5))
+angle2 <- anglePlot(po2) +
+  ggtitle("Dataset A") +  theme(plot.title = element_text(hjust = 0.5))
+
+chroma1 <- chromaPlot(po1) +
+  ggtitle("Dataset B") +  theme(plot.title = element_text(hjust = 0.5))
+chroma2 <- chromaPlot(po2) +
+  ggtitle("Dataset A") +  theme(plot.title = element_text(hjust = 0.5))
+
+#ggsave(plot = CE1, file = "./article/simCE1.pdf",
+ #      width = 8, height = 5)
+#ggsave(plot = CE2, file = "./article/simCE2.pdf",
+#       width = 8, height = 5)
+
+ggsave(plot = grid.arrange(CE2, CE1, nrow = 2), file = "./article/PlosOne/Fig2.eps",
+       device = "eps", dpi = 400, width = 7, height = 8)
 
 
-angle1 <- anglePlot(po1)
-angle2 <- anglePlot(po2)
+ggsave(plot = grid.arrange(angle2, angle1, nrow = 2), file = "./article/PlosOne/Fig3.eps",
+       device = "eps", dpi = 400, width = 7, height = 8)
 
-chroma1 <- chromaPlot(po1)
-chroma2 <- chromaPlot(po2)
+ggsave(plot = grid.arrange(chroma2, chroma1, nrow = 2), file = "./article/PlosOne/Fig4.eps",
[…]

### 2017-06-18 article/PlosOne/S1_appendix.tex
PlosOne style manuscript (including figures) done

\documentclass{article}
\usepackage[dvips]{graphicx}

\usepackage{fullpage}
\usepackage{color}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage[table]{xcolor}
\definecolor{lightgray}{gray}{0.9}
\usepackage[T1]{fontenc}

\usepackage{caption,setspace}
\captionsetup{font={footnotesize,stretch=1.5}}
%\renewcommand{\baselinestretch}{2}

\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes, positioning, calc}

\newcommand{\bfU}{\mbox{\boldmath$\mathsf{U}$}}
\newcommand{\bfu}{\mbox{\boldmath$\mathsf{u}$}}
\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
\newcommand{\RR}{\mathbb{R}}
\DeclareMathOperator*{\V}{V}
\DeclareMathOperator*{\argmax}{argmax}
\newcommand{\R}[1]{\texttt{#1}}
\newcommand{\acos}{\text{arccos}}

\begin{document}

\section*{S1 Appendix: Details about the simulated data}
This supplementary appendix describes the origin of the simulated data used for presenting the PCADSC plots. Dataset A consisted of 1000 independent simulated realizations from the $N(0, \Sigma_1)$-distribution, while dataset B consisted of 500 independent realizations from $N(0, \Sigma_1)$ and 500 independent realizations from $N(0, \Sigma_2)$. Dataset A was furthermore appended with a grouping variable, randomly dividing the observations into two groups. Dataset B also included a grouping variable, and this variable contained information about which of the two normal distributions each observation was simulated from. The covariance matrices, $\Sigma_1$ and $\Sigma_2$, were defined by

$$\Sigma_1 = \begin{pmatrix}
    1.0  &  0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
 0.0  &  1.0 & 0.0 & 0.0&  0.0 & 0.0 \\
 0.0   & 0.0 &  1.0 & 0.7 & 0.0 & 0.0 \\
 0.0 &  0.0 & 0.7 & 1.0 & 0.0 & 0.0 \\
0.0 &    0.0 &  0.0 & 0.0 & 1.0 & 0.4 \\
 0.0 & 0.0 &  0.0 & 0.0 & 0.4 & 1.0
\end{pmatrix}$$
and
$$
\Sigma_2 = \begin{pmatrix}
 1.0 & 0.2 & 0.1 & 0.0 & 0.0 & 0.0 \\
 0.2 & 1.0 & 0.1 & 0.0 & 0.0 & 0.0 \\
 0.1 & 0.1 & 1.0 & 0.0 & 0.0 & 0.0 \\
 0.0 & 0.0 & 0.0 & 1.0 & 0.3 & 0.1 \\
 0.0 & 0.0 & 0.0 & 0.3 & 1.0 & 0.2 \\
 0.0 & 0.0 & 0.0 & 0.1 & 0.2 & 1.0
\end{pmatrix},
$$
respectively.
\end{document}

### 2017-06-18 article/PlosOne/article_v1.1.tex
PlosOne style manuscript (including figures) done

--- 2017-05-17 article/article_v0.9.5.tex
@@
-\documentclass[titlepage,11pt,twoside]{article}
-\usepackage[dvips]{graphicx}
+% Template for PLoS
+% Version 3.4 January 2017
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that 
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
 
-\usepackage[myheadings]{fullpage}
-\usepackage{pmetrika}
-%\usepackage{pmbib}
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 
-\usepackage{color}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage{hyperref}
-\usepackage{natbib}
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8x]{inputenc}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Leave date blank
+\date{}
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\setlength{\headheight}{27.023pt}
+\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\sf PLOS}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%MY PREAMBLE
 \usepackage{multirow}
-%\usepackage{colortbl}
-\usepackage[table]{xcolor}
-\definecolor{lightgray}{gray}{0.9}
-\usepackage[T1]{fontenc}
-
-\usepackage{caption,setspace}
-\captionsetup{font={footnotesize,stretch=1.5}}
-\renewcommand{\baselinestretch}{2}
-
-\newcommand{\bfU}{\mbox{\boldmath$\mathsf{U}$}}
-\newcommand{\bfu}{\mbox{\boldmath$\mathsf{u}$}}
-\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
 \newcommand{\RR}{\mathbb{R}}
@@
 
-%\begin{figure}[h]
-%\centerline{\includegraphics{figure03.eps}}
-%\caption{Projection of item discrimination vectors onto $V_{\theta_T}$ hyperplance for a six item three-dimensional approximate sample structure.}
-%\end{figure}
+\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+  \renewenvironment{#1}%
+     {\linenomath\csname old#1\endcsname}%
+     {\csname oldend#1\endcsname\endlinenomath}}% 
+\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+  \patchAmsMathEnvironmentForLineno{#1}%
+  \patchAmsMathEnvironmentForLineno{#1*}}%
+\AtBeginDocument{%
+\patchBothAmsMathEnvironmentsForLineno{equation}%
+\patchBothAmsMathEnvironmentsForLineno{align}%
+\patchBothAmsMathEnvironmentsForLineno{flalign}%
+\patchBothAmsMathEnvironmentsForLineno{alignat}%
+\patchBothAmsMathEnvironmentsForLineno{gather}%
+\patchBothAmsMathEnvironmentsForLineno{multline}%
+}
 
-%\raggedbottom
-\flushbottom
+\renewcommand{\baselinestretch}{2}
 
-%\firstpage{1}
-%\setcounter{lastpage}{999}
-\setcounter{secnumdepth}{3}
+%% END MACROS SECTION
 
 \begin{document}
+\vspace*{0.2in}
 
-\begin{titlepage}
-\linespacing{1}
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Anne Helby Petersen\textsuperscript{1*\textcurrency},
+Bo Markussen\textsuperscript{2\textcurrency},
+Karl Bang Christensen\textsuperscript{1\textcurrency},
+\\
+\bigskip
+\textbf{1} Department of Public Health, University of Copenhagen, Copenhagen, Denmark
+\\
+\textbf{2} Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark
+\\
+\bigskip
 
-\title{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+% 
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+\textcurrency These authors contributed equally to this work.
 
-\author{Anne H. Petersen}
-\affil{Department of Public Health, University of Copenhagen}
+% Current address notes
+%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+% \textcurrency b Insert second current address 
+% \textcurrency c Insert third current address
 
-\author{Bo Markussen}
-\affil{Department of Mathematical Sciences, University of Copenhagen}
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* ahpe@sund.ku.dk
 
-\author{Karl Bang Christensen}
-\affil{Department of Public Health, University of Copenhagen}
-
-\vspace{\fill}
-
-\end{titlepage}\vspace*{24pt}
-
-%\RepeatTitle{Exploratory data structure comparisons: Three new tools based on Principal Component Analysis}
-
-\begin{center}\vskip3pt
-
-\vspace{32pt}
-
-Abstract\vskip3pt
-
-\end{center}
-
-\begin{abstract}
-\linespacing{1}
-
+\end{flushleft}
+% Please keep the abstract below 300 words CHECK
+\section*{Abstract}
 In empirical psychometric work, one often encounters data that is somehow divided into two distinct subsets.
@@
 Often, one wishes to combine such data into one and conduct only a single analysis, but this is only meaningful if the two subsets are similar in structures, that is, if they measure the same underlying concepts.
-However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc or require making full model assumptions.
+However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc and require making full model assumptions.
 We present a suite of three visual, non-parametric diagnostic tools for investigating differences in data structures for two datasets containing different observations of the same variables, refered to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC).
 This approach requires no directional nor hierarchical assumptions on the data.
-Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets, and by targeting different aspects of the PCA information, the three plots give a thorough and deep understanding of where differences occur and why it might be the case.
+Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets, and by targeting different aspects of the PCA information, the three plots give a thorough and deep understanding of where differences occur and why so.
 
@@
 
-\begin{keywords}
-multi-center studies, data heterogeneity, data structure comparisons, principal component analysis (PCA), European Social Survey.
-\end{keywords}
-\end{abstract}
+\linenumbers
 
-\vspace{\fill}\newpage
-
-\section{Introduction}
+\section*{Introduction}
 \label{sec:introduction}
@@
 Sophisticated methods for addressing this question are available when we are willing to assume a statistical model.
-But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment %very
-ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a %valid 
+But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data.
+The use of (parametric) models does thus not constitute a %valid
 general data structure comparison method, but rather a fitted-model comparison method.
@@
 However, the origin stories of datasets have changed over time and today, a lot of data are accumulated without a specific purpose in mind, as data collection and sharing has become easier and more affordable through the development of new technologies.
-This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for %all
-everyone to use.
+This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for everyone to use.
 However, large scale publicly available datasets, such as the PISA data and European Social Survey (ESS) data, are often used by data analysts that are far removed from the data producers and therefore, problem-specific recommendations about e.g.
-potential instrument-induced challenges in the datasets %are not
-may not be available for these data analysts.
+potential instrument-induced challenges in the datasets may not be available for these data analysts.
 For this reason, a thorough investigation of data structure comparability becomes a crucial step in any meaningful data analysis when using such open-source data.
@@
 Even in single-population survey studies, a mixture of modes of administration is often used, e.g.
-mail and telephone, and while this can improve response rates, it can also be quite problematic \citep{Brambilla1987, McHorney1994}, and differences in response behavior can lead to biased results.
-Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful, so the comparability issue is not just a statistical puzzle.
-Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu \citeyearpar{Liu2016}, who warns against the combination of online panel data (i.e.
-pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions).
+mail and telephone, and while this can improve response rates, it can also be quite problematic \cite{Brambilla1987, McHorney1994}, and differences in response behavior can lead to biased results.
+Powers, Mishra and Young report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful \cite{Powers2005}, so the comparability issue is not just a statistical puzzle.
+Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu, who warns against the combination of online panel data (i.e.
+pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions) \cite{Liu2016}.
 
@@
 These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
-%This yields a solution that is largely independent of the sizes of the two subsets of data. Anne: Det ved vi da ikke om det gør? Er det ikke netop det, vi siger, at man burde teste som næste step?
-The proposed tools are implemented in the statistical software \texttt{R} in our package, \texttt{PCADSC} \citep{PCADSC}.
+The proposed tools are implemented in the statistical software \texttt{R} in our package, \texttt{PCADSC} \cite{PCADSC}.
 
@@
 More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is completely ignored e.g.
-when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project \citep{WHR2016}.
+when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project \cite{WHR2016}.
 
-\section{PCA-based tools for data structure comparisons}
+\section*{PCA-based tools for data structure comparisons}
 \label{sec:pcadscintro}
@@
 
-%\hl{Bo: I næste afsnit bruges stadigvæk ``covariance''. Det er vel ok?} Anne: Synes det fungerer fint.
-
-\subsection{A brief introduction to principal component analysis}
-Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = (\bar{x}_1, ..., \bar{x}_d)^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$.
+\subsection*{A brief introduction to principal component analysis}
+\label{sec:introPCA}
+Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = {(\bar{x}_1, ..., \bar{x}_d)}^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) {(x_i-\bar{x})}^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$.
 We assume in the following that all variables of $X$ have a numerical interpretation, e.g.
@@
 \end{equation*}
-The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \citep{HastieEtAl2009}.
+The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \cite{HastieEtAl2009}.
 More specifically, the rank-q-reconstruction error is attained when we choose
@@
 \end{equation*}
-where $x_j \in \RR^n$ 
-%\hl{(Bo: $X_1$ og $X_2$ bruges i en anden betydning i afsnit 2.2. Omformuler?)} Anne: Hov, ja, hermed x_j så det passer med notationen ovenfor.
-denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
+where $x_j \in \RR^n$ denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
 This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
@@
 
-\subsection{Using PCA for data structure comparisons}
+\subsection*{Using PCA for data structure comparisons}
+
 All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets.
@@
 \begin{align*}
-S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} &
+S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} \\
 S &= \frac{1}{n-1} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top,
@@
 
-\subsubsection{The cumulative eigenvalue plot}
+While we present the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions.
+These plots are available in Figs~\ref{plot.simCE}, \ref{plot.simAngle} and \ref{plot.simChroma} for the CE-, angle- and chroma plots, respectively, and they are based on two simulated datasets:
+\begin{description}
+\item[Dataset A:] This dataset contains 1000 independent simulations from the same underlying model, namely \textit{model 1} from Fig~\ref{fig.simGraphs}.
+The observations are randomly divided into two groups.
+\item[Dataset B:] This dataset contains 500 independent simulations from \textit{model 1} from Fig~\ref{fig.simGraphs} and 500 independent simulations from \textit{model 2} in the same figure.
+The observations are naturally divided into groups corresponding to which model they were simulated from.
+\end{description} 
+Further details about the data simulations are available in the \nameref{appendix.simData}.
+In dataset A, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures.
+In dataset B, on the other hand, there are two fundamentally different, underlying datastructures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.
+
+\begin{figure}[!h]
+\caption{{\bf Graphical representations of the two models used for simulating example data for the PCADSC plots.} Square nodes represent observed variables, while rounded nodes represent latent variables.
+Arrows are used to illustrate causal links, and thus we e.g.
+find in \textit{Model 1} that the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable, namely $X_1$.
+Further details on the models and the simulated data are available in the \nameref{appendix.simData}, including covariance matrices for the two models.}
+\label{fig.simGraphs}
+\end{figure}
+
+\subsubsection*{The cumulative eigenvalue plot}
 The cumulative eigenvalue (CE) plot compares the variation components, i.e.
@@
 Thus, by comparing cumulative sums of variance components, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative.
-%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
-And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
 In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
 \begin{align*}
-(0,0), &&
-(\lambda_1,\lambda_{11}-\lambda_{21}), &&
-(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
-\ldots, &&
-\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
+&(0,0), \; 
+(\lambda_1,\lambda_{11}-\lambda_{21}), \;
+(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), \;
+\ldots, \\
+&\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
 \end{align*}
-%This may be seen as a cumulative Bland-Altman plot \citep{AltmanBland83,LinEtAl2002} for the variation components.
-%Anne: NK pointed out that this is somewhat confusing for people who are not familiar with the Bland-Altman plot, and as far as I can tell, what we do is quite different anyway, so I think no reference is really needed(?)
 Note that due to the standardization, the last point will always be equal to $(d,0)$.
@@
 Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does.
+Two examples of cumulative eigenvalue plots are available in Fig~\ref{plot.simCE}.
+
+\begin{figure}[!h]
+\caption{{\bf CE plots for the simulated datasets.} Dataset A, which includes no differences in data structures, is found in the top panel, and dataset B, where the observations stem from two different underlying models, is presented in the bottom panel.
+Both plots are annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
+The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.
+In the top panel, we see a cumulative eigenvalue curve that falls well within the confidence region, thereby suggesting no difference in eigenvalues (as expected).
+This also agrees with the large $p$-values and the fact that only a single underlying model was used for generating this data.
+In the bottom panel, the conclusion is reversed, which is also in accordance with the nature of the simulated data, which contains two distinct groups.}
+\label{plot.simCE}
+\end{figure}
 
@@
 \text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
-\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)^2.
+\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} {\bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)}^2.
 \end{align*}
@@
 
-%Whether the excursions in the observed curve are large or within the range of sample variation can be quantified by a permutation test.
-The idea is that we randomly reallocate the $n_1+n_2$ observations in the combined dataset to two datasets with $n_1$ and $n_2$ observations, respectively, and then repeat the procedure described above.
-The random reallocation by construction ensures that the two resampled datasets are alike except their sample size and sampling variation.
-In the \emph{CumEigenPlot} we make 1000 independent random reallocations, and plot  P-values for the null hypothesis that the variation components are the same are easily provided for any appropriate test statistic by the resampling procedure as well.
-
-\subsubsection{The angle plot}
+\subsubsection*{The angle plot}
 This plot simultaneously compares the variation components and the loadings.
@@
 Then the empirical correlation matrix for the first dataset, $S_1$, has the following orthogonal decomposition in the coordinate system of the second dataset
-%\begin{equation*}
-%S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
-%= \lambda_{\max} \sum_{k=1}^d
-%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
-%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
-%\end{equation*}
 \begin{equation*}
@@
 \Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
-\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)^\top,
+{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)}^\top,
 \end{equation*}
-and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset %\hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?
-Bo: Jeg har ikke noget bud på en sådan udvidelse lige nu og her --- skal vi ikke bare sige at det er ok i sin nuværende form?}.
+and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset.
 We propose to visualize these two decompositions in a $d \times d$ grid display.
@@
 Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.
+Fig~\ref{plot.simAngle} presents two examples of angle plots based on simulated data.
+
+\begin{figure}[!h]
+\caption{{\bf Angle plots for the simulated datasets.}Dataset A is used in the top panel and dataset B is used in the bottom panel.
+The blue show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.
+For dataset A, we find very short off-diagonal arrows, suggesting that the two PCA decompositions agree on the relative placement of the information in  the data among the 6 PCs.
+For dataset B, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}
+\label{plot.simAngle}
+\end{figure}
 
@@
 Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the second dataset in the coordinate system of the first dataset.
-%\hl{Can we expand on this in a non-technical way?
-- Karl: it seems that we cannot}
 If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
@@
 
-\subsubsection{The chroma plot}
+\subsubsection*{The chroma plot}
 This plot compares the loadings of the two datasets.
@@
 In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$.
+In Fig~ \ref{plot.simChroma}, two examples of chroma plots are available.
+
+\begin{figure}[!h]
+\caption{Chroma plots for the simulated datasets, with dataset A in the top and dataset B in the bottom.
+Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance ($\tilde\sigma_{ij}^c$).
+For dataset A, we see very similar visual patterns across the two panels, corresponding well with the fact that both groups are drawn from the same underlying model.
+For dataset B, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups.}
+\label{plot.simChroma}
+\end{figure}
 
@@
 
-\section{Comparing psychological well-being: A data example}
+\section*{Comparing psychological well-being: A data example}
 \label{sec:dataexample}
@@
 
-In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though \cite{Veenhoven2012} presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability, while \cite{Lolle2016} show highly potent translation issues for the term \textit{happiness}.
+In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though Veenhoven presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability \cite{Veenhoven2012}, while Lolle \& Andersen show highly potent translation issues for the term \textit{happiness} \cite{Lolle2016}.
 But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
@@
 In this section, we use the PCADSC tools to examine international differences in one aspect of happiness, namely psychological well-being.
-Our starting point is Denmark, a small, Northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \citep{WHR2016}, and we wish to investigate if this title is really meaningful at all.
+Our starting point is Denmark, a small, Northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \cite{WHR2016}, and we wish to investigate if this title is really meaningful at all.
 In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria.
-Though both countries are European and thus neither very far apart geographically nor culturally, these two countries have previously been highlighted to be very different in terms of what defines happiness \citep{ESStopline5}.
-Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \citep{Rodriguez2014}.
+Though both countries are European and thus neither very far apart geographically nor culturally, these two countries have previously been highlighted to be very different in terms of what defines happiness \cite{ESStopline5}.
+Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \cite{Rodriguez2014}.
 When comparing Northern European countries to other European countries, a much less pronounced relationship between the two concepts is found.
@@
 
-All computations and figures presented in this section were created using our \R{R} package \R{PCADSC} \citep{PCADSC}.
+All computations and figures presented in this section were created using our \R{R} package \R{PCADSC} \cite{PCADSC}.
 
-\subsection{Data}
+\subsection*{Data}
 
-\begin{table}[t]
+The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries.
+Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
+These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}.
+More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in the \nameref{table:items}.
+We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10.
+For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below.
+This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample.
+All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations.
+Table~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
+
+\begin{table}[!ht]
 \centering
-%\begin{tabular}{lccccccccc}
-%  \hline
-%  & \multicolumn{3}{c}{Denmark} & \multicolumn{3}{c}{Bulgaria} & \multicolumn{3}{c}{Sweden} \\
-% & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \\
-%  \hline
-% Evaluative wellbeing & 8.00 & 8.75 & 9.50 & 3.50 & 5.00 & 7.00 & 7.00 & 8.00 & 9.00 \\
-% Emotional wellbeing & 7.22 & 8.33 & 8.89 & 5.00 & 6.67 & 7.78 & 6.67 & 7.78 & 8.89 \\
-%Functioning & 6.93 & 7.57 & 8.21 & 5.50 & 6.68 & 7.68 & 6.39 & 7.04 & 7.68 \\
-%Vitality & 6.67 & 7.50 & 8.33 & 5.83 & 7.50 & 8.33 & 6.67 & 7.50 & 9.17 \\
-% Community wellbeing & 5.83 & 6.77 & 7.57 & 3.70 & 4.67 & 5.70 & 5.66 & 6.57 & 7.37 \\
-%Supportive relationships & 7.42 & 8.25 & 8.92 & 6.17 & 7.25 & 8.08 & 7.42 & 8.25 & 8.75 \\
-%   \hline
-%\end{tabular}
 \begin{tabular}{lccc}
@@
 \end{tabular}
-
 \caption{Median values of each of the six dimensions of psychological well-being, stratified by country.
@@
 
-The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries.
-Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
-These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}.
-More details on these scales can be found in \citep{ESStopline5} and the relationship between questionnaire items and scales is summarized in Table \ref{table:items} in the appendices.
-We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10.
-For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below.
-This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample.
-All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations.
-Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
+\subsection*{Comparing Denmark and Bulgaria}
+Fig~\ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales.
+The CE plot show a remarkable degree of lacking comparability: The cumulative differences in the eigenvalues by far exceed what could come about randomly if there really were no difference in the data structures.
+This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero.
 
-\subsection{Comparing Denmark and Bulgaria}
-\begin{figure}
-\centerline{\includegraphics[scale = 0.65]{essDKBGce.pdf}}
-\centerline{\includegraphics[scale = 0.65]{essDKBGhair.pdf}}
-\caption{The CE plot (top) and the angle plot (bottom) resulting from comparing Bulgarian and Danish data on psychosocial well-being.
+\begin{figure}[!h]
+\caption{{\bf CE- and angle plots comparing Danish and Bulgarian psychological well-being.} The CE plot (top) and the angle plot (bottom) resulting from comparing Bulgarian and Danish data on psychosocial well-being.
 The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
@@
 \end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[scale = 0.7]{essDKBGpancake234.pdf}}
-\caption{A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data.
-The component-bars are annotated with their relative variance contributions (denoted $\tilde\sigma$ in the previous section).}
-\label{plotBG.pancake}
-\end{figure}
-
-%\begin{figure}
-%\centerline{\includegraphics[scale = 0.7]{essDKBGWallyPCADSC234.pdf}}
-%\caption{\hl{something.}}
-%\label{plotBG.wally}
-%\end{figure}
-
-Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales.
-The CE plot show a remarkable degree of lacking comparability: The cumulative differences in the eigenvalues by far exceed what could come about randomly if there really were no difference in the data structures.
-This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero.
 
@@
 
-The chroma plot in Figure \ref{plotBG.pancake} allows us to look closer into these components.
+The chroma plot in Fig~\ref{plotBG.pancake} allows us to look closer into these components.
 Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish.
 In the Danish data, on the other hand, we find that \textit{Vitality} and \textit{Emotional well-being} seem to play bigger roles, as they appear with larger loadings in more high-ranking components in this sample, relative to the Bulgarian.
+
+\begin{figure}[!h]
+\caption{{\bf Chroma plot for comparing Danish and Bulgarian psychological well-being.} A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data.
+The component-bars are annotated with their relative variance contributions (denoted $\tilde\sigma$ in the previous section).}
+\label{plotBG.pancake}
+\end{figure}
 
@@
 In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, corresponding well with the previous findings mentioned above.
-Thus, the datasets are fundamentally different and we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics (\cite{ESStopline5}).
+Thus, the datasets are fundamentally different and we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics \cite{ESStopline5}.
 Moreover, this also implies that two countries cannot be ranked in terms of which country is "the most happy", at least not by referring to psychological well-being dimensions such as those encountered here.
 
-% While the first principal component, which is responsible for explaining 50-60 \% of the variance in the data, is very similar for the two countries, we see quite large differences in the remaining components.
-In the second component, we see that the two countries disagree in the relative importance of the scales \textit{Community wellbeing} and \textit{Vitality}.
-In the third and fourth components, general disagreement  is found.
-All in all, components 2-4, representing almost half of the variability in the data, are not very similar across the two countries.
-Moreover, the two subsets of the data also differ with respect to how much variance is explained by each component, and the difference is particularly big for the first component.
-This component has approximately 15 \%  more explanatory power in the Bulgarian subsample than it does in the Danish.
+\subsection*{Comparing Denmark and Sweden}
+We now turn to the comparison of Denmark and Sweden in terms of psychological well-being.
+Fig~\ref{plotSE.cehair} shows the CE- and angle plots for these two countries.
+In the CE plot, we now find the cumulative eigenvalue curve to be just within the acceptance region of the null-hypothesis.
+This is also reflected by the two tests, which now produce $p$-values of $p_\text{KS} = 0.17$ and $p_\text{CvM} = 0.11$, respectively.
+Thus, it is not unreasonable to assume equal variance components in the two datasets.
 
-%{\color{red}
-%Figure \ref{plotESSHairplot} shows the hairplot.
-The blue arrows visualize the decomposition of the principal components for the %first dataset in the coordinate system of the second dataset.
-We see that PC2 also loads on PC3, that PC3 also loads on PC4, and %that PC4 also loads on PC2 and PC3.
-The red arrows visualize the decomposition of the principal components for the second dataset %in the coordinate system of the first dataset.
-We see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that %PC4 also loads on PC3.
-For PC1, PC5 and PC6 the main difference is in the size of the variation component.
-%}
-
-%But did we really illustrate a data structure difference due to country differences or did we just illustrate the variability of the results of the PCADSC method?
-In order to investigate this further, we look at a so-called \textit{Wally plot} (\hl{ref: Claus Ekstrøm}).
-In this plot, we compare the results of PCADSC conducted with grouping by country with several random, but similar grouping variables.
-Specifically, we produce 7 PCADSC plots where the country variable was replaced by a randomly generated variable that divides the observations into two groups of the same sizes as the country samples.
-The results are illustrated in Figure \ref{plotESSPCADSCWally}.
-Here, we see that the differences in the second component from the original PCADSC results are not matched in any of the randomly grouped PCADSC runs.
-In fact, the 7 runs are remarkably similar, thereby illustrating that PCADSC seems to be very robust with respect to random groupings: The signal in the data is not blurred by the random subdivisions.
-When it comes to the differences in the third component for the two groups, we find much larger variability in the 7 random runs.
-\hl{more comments here...
-Wait until we are sure exactly what we think about the results and what other PCA-based methods, we will do before/after.
-Particularly, how do we deal with eigen value differences?}
-
-\subsection{Comparing Denmark and Sweden}
-\begin{figure}
-\centerline{\includegraphics[scale = 0.65]{essDKSEce.pdf}}
-\centerline{\includegraphics[scale = 0.65]{essDKSEhair.pdf}}
-\caption{A CE (top) and a angle (bottom) plot for comparing the Danish and the Swedish psychological well-being data.
+\begin{figure}[!h]
+\caption{{\bf CE- and angle plots comparing Danish and Swedish psychological well-being.} A CE (top) and a angle (bottom) plot for comparing the Danish and the Swedish psychological well-being data.
 The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
@@
 
-\begin{figure}
-\center
-\centerline{\includegraphics[scale = 0.7]{essDKSEpancake.pdf}}
-\caption{A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples.
+The angle plot in Fig~\ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent.
+This implies that if one already has e.g.
+the information held in the first PC from the Danish data, this information is in itself mostly sufficient to describe the first PC of the Swedish data.
+
+Looking at the chroma plot in Fig~\ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80\% of the variance in both datasets), and slight, but increasing, differences in the remaining three components.
+We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25\% of the variance in both datasets.
+In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g.
+conclude using Table~\ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally a lot happier than the least happy people in Sweden.
+A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.
+
+\begin{figure}[!h]
+\caption{{\bf Chroma plot comparing Danish and Bulgarian psychological well-being.} A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples.
 Note that for each component, the bar is annotated with its cumulative variance contribution (denoted $\tilde\sigma^c$ in the previous section), that is, how much variance can be explained by having information of this and the preceding components.}
@@
 
-%\begin{figure}
-%\center
-%\includegraphics[scale = 0.7]{essDKSEWallyPCADSC.pdf}
-%\caption{\hl{something.}}
-%\label{plotSE.wally}
-%\end{figure}
-
-We now turn to the comparison of Denmark and Sweden in terms of psychological well-being.
-Figure \ref{plotSE.cehair} shows the CE- and angle plots for these two countries.
-In the CE plot, we now find the cumulative eigenvalue curve to be just within the acceptance region of the null-hypothesis.
-This is also reflected by the two tests, which now produce $p$-values of $p_\text{KS} = 0.17$ and $p_\text{CvM} = 0.11$, respectively.
-Thus, it is not unreasonable to assume equal variance components in the two datasets.
-
-% but not with overwhelming evidence.
-% Bo: absence of proof \neq proof of absence.
-
-The angle plot in Figure \ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent.
-This implies that if one already has e.g.
-the information held in the first PC from the Danish data, this information is in itself mostly sufficient to describe the first PC of the Swedish data.
-
-Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80\% of the variance in both datasets), and slight, but increasing, differences in the remaining three components.
-We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25\% of the variance in both datasets.
-In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g.
-conclude using Table \ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally a lot happier than the least happy people in Sweden.
-A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.
-
-\section{Discussion}
+\section*{Discussion}
 \label{sec.Discussion}
@@
 They all employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
-% yielding a solution that is largely independent of the sizes  the two datasets.
-Anne: We don't know if that is true.
 
@@
 It should be noted that even though concerns about pooling data from different countries are not yet very prevalent when it comes to happiness rankings, the subject has been very well scrutinized in the field of international educational rankings.
-For instance, the PISA tests have repeatedly been criticized for not being meaningful objects of international comparisons, especially due to problems with differential item functioning \citep{Kankaras2014,Kreiner2014,ZwitserEtAl2017} and translation problems \citep{Asil2016}.
+For instance, the PISA tests have repeatedly been criticized for not being meaningful objects of international comparisons, especially due to problems with differential item functioning \cite{Kankaras2014,Kreiner2014,ZwitserEtAl2017} and translation problems \cite{Asil2016}.
 This highlights lack of data comparability as an empirically relevant and concerning issue that there is no way around: When conducting internal rankings, the issue of data structure heterogeneity should always be addressed.
@@
 Though the PCADSC toolbox comprises three diagnostic plots as of now, the PCA decomposition can of course be utilized further to bring about new PCADSC plots that illustrate more aspects of data structure differences.
-For instance, the scree plot (as introduced by \cite{Cattell1966}), which plots the ordered eigenvalues against their component numbers, could easily be recreated as a PCADSC tool by creating scree plots for both data subsets in the same coordinate system, yielding a simple graphical comparison.
+For instance, the scree plot (as introduced by Cattell \cite{Cattell1966}), which plots the ordered eigenvalues against their component numbers, could easily be recreated as a PCADSC tool by creating scree plots for both data subsets in the same coordinate system, yielding a simple graphical comparison.
 The resampling approach used in the CE plot could then be adopted, thus annotating the scree plots with resampled curves.
-This general approach to model-fit evaluation is very helpful in combination with permutation tests, %using cumulative residuals,
-as inspired by Lin et al. \citeyearpar{LinEtAl2002}. Similarly, the angle plot could also be annotated with $p$-values from a permutation test based on appropriate test statistics.
+This general approach to model-fit evaluation is very helpful in combination with permutation tests, as inspired by Lin and colleages \cite{LinEtAl2002}.
+Similarly, the angle plot could also be annotated with $p$-values from a permutation test based on appropriate test statistics.
 
@@
 This limitation is inherited from the PCA method and thus, moving beyond it will entail replacing the PCA framework with a more general one.
-Multiple correspondence analysis (MCA) is often suggested as a categorical extension of PCA \citep{Abdi2010} and including MCA in the PCADSC tools is thus a natural next step.
+Multiple correspondence analysis (MCA) is often suggested as a categorical extension of PCA \cite{Abdi2010} and including MCA in the PCADSC tools is thus a natural next step.
 
-Moreover, though we have illustrated the discriminatory power of PCASDC in one data example, of course further evaluation of its performance is needed.  
+Moreover, though we have illustrated the discriminatory power of PCASDC in one simulated- and on real-life data example, of course further evaluation of its performance is needed.  
 While the best feature of PCADSC is perhaps the intuitive, visual nature of the tools, it is also the biggest weakness.
@@
 Therefore, a systematic evaluation of the performance of the PCADSC methods is not very straight-forward to design: It really depends on whether or not users can detect actual differences in data structures by using the diagnostic plots.
-We suggest that the variability of the methods could be studied further using \textit{Wally plots} \citep{Ekstrøm2014}, which are learning tools developed for teaching students how to interpret residual plots.
+We suggest that the variability of the methods could be studied further using \textit{Wally plots} \cite{Ekstrom2014}, which are learning tools developed for teaching students how to interpret residual plots.
 First in line as a topic for more thorough investigations of performance is the sensitivity towards the sample sizes of the two datasets, $n_1$ and $n_2$.
@@
 
-%\section{Concluding Remarks}
-%\label{sec:conclusion}
+\nolinenumbers
 
-\bibliographystyle{apa}
-\bibliography{bib}
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% or
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here. See http://journals.plos.org/plosone/s/latex for 
+% step-by-step instructions.
+%
 
-\appendix
+\section*{Supporting information}
 
-%\newpage
-%\section{Supplementary tables}
-\begin{table}[h]
-\centering
-\scriptsize
-\rowcolors{1}{}{lightgray}
-\bgroup
-\def\arraystretch{1.5}
-\begin{tabular}{ll}
-\hline
-Scale & Items \\
-\hline
-\rowcolor{lightgray} & How satisfied with life as a whole \\
-\rowcolor{lightgray}\multirow{-2}{*}{Evaluative wellbeing }& How happy are you\\
+% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+\paragraph*{S1 Appendix.}
+\label{appendix.simData}
+{\bf Details about the simulated data.}
 
-\rowcolor{white} & Felt sad, how often in the past week\\
-\rowcolor{white} & Felt depressed, how often in the past week\\
-\rowcolor{white} & Enjoyed life, how often in the past week\\
-\rowcolor{white} & Were happy, how often in the past week\\
-\rowcolor{white} & Felt anxious, how often in the past week\\
-\rowcolor{white}\multirow{-6}{*}{Emotional wellbeing} & Felt calm and peaceful, how often in the past week\\
+\paragraph*{S1 Table.}
+\label{table:items}
+{\bf ESS psychological wellbeing scale definitions.} Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
+Note that before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.
 
-\rowcolor{lightgray} & Free to decide how to live my life\\
-\rowcolor{lightgray} & Little chance to show how capable I am\\
-\rowcolor{lightgray} & Feel accomplishment from what I do\\
-\rowcolor{lightgray} & Interested in what you are doing\\
-\rowcolor{lightgray} & Absorbed in what you are doing\\
-\rowcolor{lightgray} & Enthusiastic about what you are doing\\
-\rowcolor{lightgray} & Feel what I do in life is valuable and worthwhile\\
-\rowcolor{lightgray} & Have a sense of direction\\
-\rowcolor{lightgray} & Always optimistic about my future\\
-\rowcolor{lightgray} & There are lots of things I feel I am good at\\
-\rowcolor{lightgray} & In general feel very positive about myself\\
-\rowcolor{lightgray} & At times feel as if I am a failure\\
-\rowcolor{lightgray} & When things go wrong in my life it takes a long time to get back to normal\\
-\rowcolor{lightgray}\multirow{-14}{*}{Functioning} & Deal with important problems\\
+\begin{thebibliography}{10}
 
-\rowcolor{white} & Felt everything did as effort, how often in the past week\\
-\rowcolor{white} & Sleep was restless, how often in the past week\\
-\rowcolor{white} & Could not get going, how often in the past week\\
-\rowcolor{white}\multirow{-4}{*}{Vitality} & Had a lot of energy, how often in the past week\\
+\bibitem{Brambilla1987}
+Brambilla DJ, McKinlay SM.
+\newblock {A comparison of responses to mailed questionnaires and telephone
+  interviews in a mixed mode health survey.}
+\newblock American journal of epidemiology. 1987;126(5):962--71.
 
-\rowcolor{lightgray} &  Most people can be trusted\\
-\rowcolor{lightgray} &  People try to take advantage\\
-\rowcolor{lightgray} &  Most of the time people are helpful\\
-\rowcolor{lightgray} &  Feel people in local area help one another\\
-\rowcolor{lightgray}\multirow{-5}{*}{Community wellbeing}& Feel close to the people in local area\\
+\bibitem{McHorney1994}
+McHorney CA, Kosinski M, Ware JE.
+\newblock {Comparisons of the costs and quality of norms for the SF-36 health
+  survey collected by mail versus telephone interview: results from a national
+  survey.}
+\newblock Medical care. 1994;32(6):551--67.
 
-\rowcolor{white} & How many with whom you can discuss intimate matters\\
-\rowcolor{white} & Feel appreciated by those you are close to\\
-\rowcolor{white} & Receive help and support\\
-\rowcolor{white}\multirow{-4}{*}{Supportive relationships} & Felt lonely, how often in the past week\\
-\hline
-\end{tabular}
-\egroup
-\caption{Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
-Note that before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
-\label{table:items}
-\end{table}
+\bibitem{Powers2005}
+Powers JR, Mishra G, Young AF.
+\newblock {Differences in mail and telephone responses to self-rated health:
+  use of multiple imputation in correcting for response bias.}
+\newblock Australian and New Zealand journal of public health.
+  2005;29(2):149--54.
+
+\bibitem{Liu2016}
+Liu M.
+\newblock {Comparing data quality between online panel and intercept samples}.
+\newblock Methodological Innovations. 2016;9:2059799116672877.
+\newblock doi:{10.1177/2059799116672877}.
+
+\bibitem{PCADSC}
+Petersen AH, Markussen B. PCADSC: Tools for Principal Component Analysis-Based
+  Data Structure Comparisons; 2017.
+\newblock Available from: \url{https://CRAN.R-project.org/package=PCADSC}.
+
+\bibitem{WHR2016}
+Helliwell J, Layard R, Sachs J.
+\newblock {World Happiness Report 2016 Update}.
+\newblock New York: Sustainable Development Solutions Network. 2016;.
+
+\bibitem{HastieEtAl2009}
+Hastie T, Tibshirani R, Friedman J.
+\newblock 14.5.1.
+\newblock In: The Elements of Statistical Learning. 2nd ed. Springer; 2009.
+
+\bibitem{Veenhoven2012}
+Veenhoven R.
+\newblock {Cross-national differences in happiness: Cultural measurement bias
+  or effect of culture?}
+\newblock International Journal of Wellbeing. 2012;2(4):333--353.
+\newblock doi:{10.5502/ijw.v2.i4.4}.
+
+\bibitem{Lolle2016}
+Lolle HL, Andersen JG.
+\newblock {Measuring Happiness and Overall Life Satisfaction: A Danish Survey
+  Experiment on the Impact of Language and Translation Problems}.
+\newblock Journal of Happiness Studies. 2016;17(4):1337--1350.
+\newblock doi:{10.1007/s10902-015-9646-4}.
+
+\bibitem{ESStopline5}
+Jeffrey K, Abdallah S, Quick A.
+\newblock {Europeans' Personal and Social Wellbeing: Topline Results from Round
+  6 of the European Social Survey}.
+\newblock ESS Topline Results (Series 5). 2015;.
+
+\bibitem{Rodriguez2014}
+Rodr\'iguez-Pose A, von Berlepsch V.
+\newblock {Social Capital and Individual Happiness in Europe}.
+\newblock Journal of Happiness Studies. 2014;15(2):357--386.
+\newblock doi:{10.1007/s10902-013-9426-y}.
+
+\bibitem{Kankaras2014}
+Kankara{\v{s}} M, Moors G.
+\newblock {Analysis of Cross-Cultural Comparability of PISA 2009 Scores}.
+\newblock Journal of Cross-Cultural Psychology. 2014;45(3):381--399.
+\newblock doi:{10.1177/0022022113511297}.
+
+\bibitem{Kreiner2014}
+Kreiner S, Christensen KB.
+\newblock {Analyses of Model Fit and Robustness. A New Look at the PISA Scaling
+  Model Underlying Ranking of Countries According to Reading Literacy}.
+\newblock Psychometrika. 2014;79(2):210--231.
+\newblock doi:{10.1007/s11336-013-9347-z}.
+
+\bibitem{ZwitserEtAl2017}
+Zwitser RJ, Glaser SSF, Maris G.
+\newblock {Monitoring Countries in a Changing World: A New Look at DIF in
+  International Surveys}.
+\newblock Psychometrika. 2017;82(1):210--232.
+
+\bibitem{Asil2016}
+Asil M, Brown GTL.
+\newblock {Comparing OECD PISA Reading in English to Other Languages:
+  Identifying Potential Sources of NonInvariance}.
+\newblock International Journal of Testing. 2016;16(1):71--93.
+\newblock doi:{10.1080/15305058.2015.1064431}.
+
+\bibitem{Cattell1966}
+Cattell RB.
+\newblock {The Scree Test For The Number Of Factors}.
+\newblock Multivariate Behavioral Research. 1966;1(2):245--276.
+
+\bibitem{LinEtAl2002}
+Lin DY, Wei LJ, Ying Z.
+\newblock {Model-Checking Techniques Based on Cumulative Residuals}.
+\newblock Biometrics. 2002;58:1--12.
+
+\bibitem{Abdi2010}
+Abdi H, Williams LJ.
+\newblock Principal component analysis.
+\newblock Wiley Interdisciplinary Reviews: Computational Statistics.
+  2010;2(4):433--459.
+\newblock doi:{10.1002/wics.101}.
+
+\bibitem{Ekstrom2014}
+Ekstr{\o}m CT.
+\newblock {Teaching 'Instant Experience' with Graphical Model Validation
+  Techniques}.
+\newblock Teaching Statistics. 2014;36(1):23--26.
+\newblock doi:{10.1111/test.12027}.
+
+\end{thebibliography}
 

### 2017-06-20 article/PlosOne/coverLetter.tex
Påbegyndt skrivning af coverletter.

\documentclass[titlepage,11pt,twoside]{article}
\usepackage[dvips]{graphicx}

\usepackage{color}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{multirow}
%\usepackage{colortbl}
\usepackage[table]{xcolor}
\definecolor{lightgray}{gray}{0.9}
\usepackage[T1]{fontenc}

\usepackage{caption,setspace}
\captionsetup{font={footnotesize,stretch=1.5}}
\renewcommand{\baselinestretch}{1.2}

\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes, positioning, calc}

\newcommand{\bfU}{\mbox{\boldmath$\mathsf{U}$}}
\newcommand{\bfu}{\mbox{\boldmath$\mathsf{u}$}}
\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
\newcommand{\RR}{\mathbb{R}}
\DeclareMathOperator*{\V}{V}
\DeclareMathOperator*{\argmax}{argmax}
\newcommand{\R}[1]{\texttt{#1}}
\newcommand{\acos}{\text{arccos}}

%\usepackage{helvet}
%\renewcommand{\familydefault}{\sfdefault}
%\renewcommand{\rmdefault}{\sfdefault} % Aria

\usepackage{arev}

%\begin{figure}[h]
%\centerline{\includegraphics{figure03.eps}}
%\caption{Projection of item discrimination vectors onto $V_{\theta_T}$ hyperplance for a six item three-dimensional approximate sample structure.}
%\end{figure}

%\raggedbottom
\flushbottom

%\firstpage{1}
%\setcounter{lastpage}{999}
\setcounter{secnumdepth}{3}

\begin{document}

Dear Editor,

\medskip

Please consider the attached research paper \emph{``Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis''} for publication in PLoS One.

\bigskip

Yours Sincerely,

\bigskip\bigskip

Anne Helby Petersen

Scientific Assistant

Section of Biostatistics

Faculty of Health and Medical Sciences

University of Copenhagen

\begin{itemize}
\item Summarize the study’s contribution to the scientific literature
\item Relate the study to previously published work
\item[Done] Specify the type of article (for example, research article, systematic review, meta-analysis, clinical trial)
\item[Not relevant] Describe any prior interactions with PLOS regarding the submitted manuscript
\item Suggest appropriate Academic Editors to handle your manuscript (see the full list of Academic Editors)
\item[Not relevant] List any opposed reviewers
\end{itemize}
\end{document}

%%%%%%%%%%%%%%%%%%%

% Hvad var meningen med nedenstående diagrammer? 
% De skal nok bare fjernes.

\begin{document}
\center
Model 1: \medskip \\
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=2.5cm,
  thick,main node/.style={rectangle, rounded corners,draw,font=\sffamily\small\normalfont},
  sq node/.style={rectangle,draw,font=\sffamily\small\normalfont}]
  \node[sq node] (1) {$v_1$};
  \node[sq node, right of = 1] (2) {$v_2$};
  \node[sq node, right of = 2] (3) {$v_3$};
  \node[sq node, right of = 3] (4) {$v_4$};
  \node[sq node, right of = 4] (5) {$v_5$};
  \node[sq node, right of = 5] (6) {$v_6$};
  \node[main node, above right=1cm of 3] (7) {$X_1$};
  \node[main node, above right=1cm of 5] (8) {$X_2$};
  \path[every node/.style={font=\sffamily\small}]
    (7) edge node  {} (3)
    (7) edge node {} (4)
    (8) edge node {} (5)
    (8) edge node {} (6);
 \end{tikzpicture}
\\
\bigskip
Model 2: \medskip \\
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=2.5cm,
  thick,main node/.style={rectangle, rounded corners,draw,font=\sffamily\small\normalfont},
  sq node/.style={rectangle,draw,font=\sffamily\small\normalfont}]
  \node[sq node] (1) {$v_1$};
  \node[sq node, right of = 1] (2) {$v_2$};
  \node[sq node, right of = 2] (3) {$v_3$};
  \node[sq node, right of = 3] (4) {$v_4$};
  \node[sq node, right of = 4] (5) {$v_5$};
  \node[sq node, right of = 5] (6) {$v_6$};
  \node[main node, above=1cm of 2] (7) {$Y_1$};
  \node[main node, above=1cm of 5] (8) {$Y_2$};
  \path[every node/.style={font=\sffamily\small}]
    (7) edge node  {} (1)
    (7) edge node {} (2)
    (7) edge node {} (3)
    (8) edge node {} (4)
    (8) edge node {} (5)
    (8) edge node {} (6);
 \end{tikzpicture}

\end{document}

### 2017-06-21 article/PlosOne/coverLetter.tex
Udkast til coverletter afsluttet.

@@
-\documentclass[titlepage,11pt,twoside]{article}
-\usepackage[dvips]{graphicx}
+\documentclass[12pt,a4paper]{article}
 
-\usepackage{color}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage{hyperref}
-\usepackage{natbib}
-\usepackage{multirow}
-%\usepackage{colortbl}
-\usepackage[table]{xcolor}
-\definecolor{lightgray}{gray}{0.9}
-\usepackage[T1]{fontenc}
+%\usepackage[dvips]{graphicx}
 
-\usepackage{caption,setspace}
-\captionsetup{font={footnotesize,stretch=1.5}}
-\renewcommand{\baselinestretch}{1.2}
+%\usepackage{color}
+%\usepackage{mathtools}
+%\usepackage{amssymb}
+%\usepackage{hyperref}
+%\usepackage{natbib}
+%\usepackage{multirow}
+%\usepackage[table]{xcolor}
+%\definecolor{lightgray}{gray}{0.9}
+%\usepackage[T1]{fontenc}
+%
+%\usepackage{caption,setspace}
+%\captionsetup{font={footnotesize,stretch=1.5}}
+\renewcommand{\baselinestretch}{1.1}
 
-\usepackage{tikz}
-\usetikzlibrary{arrows}
-\usetikzlibrary{shapes, positioning, calc}
+%\usepackage{arev}
 
-\newcommand{\bfU}{\mbox{\boldmath$\mathsf{U}$}}
-\newcommand{\bfu}{\mbox{\boldmath$\mathsf{u}$}}
-\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
-\newcommand{\RR}{\mathbb{R}}
-\DeclareMathOperator*{\V}{V}
-\DeclareMathOperator*{\argmax}{argmax}
-\newcommand{\R}[1]{\texttt{#1}}
-\newcommand{\acos}{\text{arccos}}
-
-%\usepackage{helvet}
-%\renewcommand{\familydefault}{\sfdefault}
-%\renewcommand{\rmdefault}{\sfdefault} % Aria
-
-\usepackage{arev}
-
-%\begin{figure}[h]
-%\centerline{\includegraphics{figure03.eps}}
-%\caption{Projection of item discrimination vectors onto $V_{\theta_T}$ hyperplance for a six item three-dimensional approximate sample structure.}
-%\end{figure}
-
-%\raggedbottom
-\flushbottom
-
-%\firstpage{1}
-%\setcounter{lastpage}{999}
-\setcounter{secnumdepth}{3}
+\pagestyle{empty}
 
@@
 Please consider the attached research paper \emph{``Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis''} for publication in PLoS One.
+As Academic Editor to handle the manuscript we suggest Ulrich S.\ Tran, Jakob Pietschnig, or Christopher M.\ Danforth.
+
+\medskip
+
+Data comparability is a recurring topic in statistics, in particular in psychometrics and social sciences.
+An example of this is the \emph{PISA global education rankings}, where  it a priori is not clear whether it makes sense to compare children's reading capabilities across different cultures and languages.
+There exist methods to investigate data comparability within the framework of statistical models, e.g.\ \emph{differential item functioning} within \emph{item response theory}.
+However, without assuming a statistical model practitioners often report simple data summaries.
+In the submitted paper we propose three graphical tool that can be used in an initial exploratory comparison of the data structure of two multivariate datasets prior to the formulation of a statistical model.
+The tools are implemented in an R package publicly available on CRAN.
+The methodology is exemplified on 6 psychological well-being scores (derived from the \emph{European Social Survey}) believed to measuring the concept of \emph{happiness}.
+Using the graphical tools we are able to visualize a difference between the measured conception of happiness in Denmark and Bulgaria, hence verifying the difficulty in transferring the concept of happiness across countries as also found by other authors.
+On the other hand, the graphical tools also find that Denmark and Sweden have comparable data structures in accordance with the understanding of similarity within the Nordic countries.
+
+\medskip
+
+We look forward to your reply.
 
@@
 
-\begin{itemize}
-\item Summarize the study’s contribution to the scientific literature
-\item Relate the study to previously published work
-\item[Done] Specify the type of article (for example, research article, systematic review, meta-analysis, clinical trial)
-\item[Not relevant] Describe any prior interactions with PLOS regarding the submitted manuscript
-\item Suggest appropriate Academic Editors to handle your manuscript (see the full list of Academic Editors)
-\item[Not relevant] List any opposed reviewers
-\end{itemize}
 \end{document}
 
-%%%%%%%%%%%%%%%%%%%
+Summarize the study’s contribution to the scientific literature.
 
-% Hvad var meningen med nedenstående diagrammer? 
-% De skal nok bare fjernes.
+Relate the study to previously published work.
 
-\begin{document}
-\center
-Model 1: \medskip \\
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=2.5cm,
-  thick,main node/.style={rectangle, rounded corners,draw,font=\sffamily\small\normalfont},
-  sq node/.style={rectangle,draw,font=\sffamily\small\normalfont}]
-  \node[sq node] (1) {$v_1$};
-  \node[sq node, right of = 1] (2) {$v_2$};
-  \node[sq node, right of = 2] (3) {$v_3$};
-  \node[sq node, right of = 3] (4) {$v_4$};
-  \node[sq node, right of = 4] (5) {$v_5$};
-  \node[sq node, right of = 5] (6) {$v_6$};
-  \node[main node, above right=1cm of 3] (7) {$X_1$};
-  \node[main node, above right=1cm of 5] (8) {$X_2$};
-  \path[every node/.style={font=\sffamily\small}]
-    (7) edge node  {} (3)
-    (7) edge node {} (4)
-    (8) edge node {} (5)
-    (8) edge node {} (6);
- \end{tikzpicture}
-\\
-\bigskip
-Model 2: \medskip \\
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=2.5cm,
-  thick,main node/.style={rectangle, rounded corners,draw,font=\sffamily\small\normalfont},
-  sq node/.style={rectangle,draw,font=\sffamily\small\normalfont}]
-  \node[sq node] (1) {$v_1$};
-  \node[sq node, right of = 1] (2) {$v_2$};
-  \node[sq node, right of = 2] (3) {$v_3$};
-  \node[sq node, right of = 3] (4) {$v_4$};
-  \node[sq node, right of = 4] (5) {$v_5$};
-  \node[sq node, right of = 5] (6) {$v_6$};
-  \node[main node, above=1cm of 2] (7) {$Y_1$};
-  \node[main node, above=1cm of 5] (8) {$Y_2$};
-  \path[every node/.style={font=\sffamily\small}]
-    (7) edge node  {} (1)
-    (7) edge node {} (2)
-    (7) edge node {} (3)
-    (8) edge node {} (4)
-    (8) edge node {} (5)
-    (8) edge node {} (6);
- \end{tikzpicture}
+Specify the type of article (for example, research article, systematic review, meta-analysis, clinical trial).
 
-\end{document}
+Describe any prior interactions with PLOS regarding the submitted manuscript.
+
+Suggest appropriate Academic Editors to handle your manuscript (see the full list of Academic Editors).
+
+List any opposed reviewers.
+\end{itemize}

### 2017-06-22 article/PlosOne/coverLetter.tex
Slight editing

@@
 There exist methods to investigate data comparability within the framework of statistical models, e.g.\ \emph{differential item functioning} within \emph{item response theory}.
-However, without assuming a statistical model practitioners often report simple data summaries.
-In the submitted paper we propose three graphical tool that can be used in an initial exploratory comparison of the data structure of two multivariate datasets prior to the formulation of a statistical model.
+However, without assuming a statistical model, practitioners often report simple data summaries in the absense of a systematic framework for understanding and describing data structure differences.
+In the submitted paper, we propose three graphical tool that can be used in an initial exploratory comparison of the data structure of two multivariate datasets prior to the formulation of a statistical model.
 The tools are implemented in an R package publicly available on CRAN.
-The methodology is exemplified on 6 psychological well-being scores (derived from the \emph{European Social Survey}) believed to measuring the concept of \emph{happiness}.
-Using the graphical tools we are able to visualize a difference between the measured conception of happiness in Denmark and Bulgaria, hence verifying the difficulty in transferring the concept of happiness across countries as also found by other authors.
+The methodology is exemplified on 6 psychological well-being scores (derived from the \emph{European Social Survey}) believed to be measuring the concept of \emph{happiness}.
+Using the graphical tools we are able to visualize a difference between the measured conception of happiness in Denmark and Bulgaria, hence verifying the difficulty in transferring the concept of happiness across countries, as also found by other authors.
 On the other hand, the graphical tools also find that Denmark and Sweden have comparable data structures in accordance with the understanding of similarity within the Nordic countries.
@@
 
-Scientific Assistant
+Research Assistant
 

### 2017-11-22 article/APM/article.tex
version til Applied Psyhcological Measurement

--- 2017-06-18 article/PlosOne/article_v1.1.tex
@@
-% Template for PLoS
-% Version 3.4 January 2017
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
+\documentclass[a4paper,14pt]{article}
+\usepackage[danish]{babel}
+\usepackage{authblk}
+\usepackage{geometry}
+\usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{amsthm}
+\usepackage{mathrsfs}
+\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
+\usepackage{apalike}
 \usepackage[utf8x]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
 \usepackage{nameref,hyperref}
-
 % line numbers
-\usepackage[right]{lineno}
-
+%\usepackage[right]{lineno}
 % ligatures disabled
@@
 \DisableLigatures[f]{encoding = *, family = * }
-
 % color can be used to apply background shading to table cells only
 \usepackage[table]{xcolor}
-
 % array package and thick rules for tables
 \usepackage{array}
-
 % create "+" rule type for thick vertical lines
@@
 
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
+\usepackage{setspace}
+\doublespacing
 
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
+\newcommand{\RR}{\mathbb{R}}
+\DeclareMathOperator*{\V}{V}
+\DeclareMathOperator*{\argmax}{argmax}
+\newcommand{\R}[1]{\texttt{#1}}
+\newcommand{\acos}{\text{arccos}}
 
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
+\renewcommand{\baselinestretch}{2}
+
+\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
+T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
+\def\volumeyear{2017}
+
+\begin{document}
+
+%\runninghead{PCADSC}
+
+\title{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}
+
+\author[1]{Anne Helby Petersen}
+\author[1]{Bo Markussen}
+\author[3]{Karl Bang Christensen}
+\affil[1]{Department of Public Health, University of Copenhagen, Copenhagen, Denmark}
+\affil[2]{Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}
+
+\maketitle
+
+%\corrauth{Anne Helby Petersen}
+%\email{ahpe@sund.ku.dk}
+
+\newpage
+
+\begin{abstract}
+Researchers often encounter data that is divided into distinct subsets.
+This might be due to multi-center sampling frames, or to variations in instruments, questionnaire item ordering or mode of administration.
+Often, one wishes to combine such data into one and conduct only a single analysis, but this is only meaningful if the subsets are similar in structure.
+Often, procedures for assessing whether such a data collapsing is allowed are ad hoc and require making full model assumptions.
+We present visual diagnostic tools for investigating differences in structure for two datasets containing different observations of the same variables.
+This approach, referred to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC), requires no directional nor hierarchical assumptions on the data and consists of three plots based on Principal Component Analysis (PCA), that address differences in the structures of the covariance matrices of the two datasets to yield an  understanding of where differences occur and why so.
+The functionality of the PCADSC tools is presented by use of data from the European Social Survey on psychological well-being in three European countries, namely Denmark, Sweden and Bulgaria.
+Here, we find that the concepts of psychological well-being are not the same when comparing Denmark and Bulgaria, and thus a comparison of e.g.
+mean scores in different well-being aspects is not meaningful.
+However, when comparing the two Scandinavian countries, Denmark and Sweden, we find very similar data structures and thus very similar concepts of well-being.
+Therefore, inter-country comparisons are warranted for these two countries.
+\end{abstract}
+
+%\keywords{Class file, \LaTeXe, \textit{SAGE Publications}}
 
@@
 % Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Leave date blank
 \date{}
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-\setlength{\headheight}{27.023pt}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\sf PLOS}
 
@@
 
-%MY PREAMBLE
-\usepackage{multirow}
-\newcommand{\RR}{\mathbb{R}}
-\DeclareMathOperator*{\V}{V}
-\DeclareMathOperator*{\argmax}{argmax}
-\newcommand{\R}[1]{\texttt{#1}}
-\newcommand{\acos}{\text{arccos}}
-
-\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
-  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
-  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
-  \renewenvironment{#1}%
-     {\linenomath\csname old#1\endcsname}%
-     {\csname oldend#1\endcsname\endlinenomath}}% 
-\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
-  \patchAmsMathEnvironmentForLineno{#1}%
-  \patchAmsMathEnvironmentForLineno{#1*}}%
-\AtBeginDocument{%
-\patchBothAmsMathEnvironmentsForLineno{equation}%
-\patchBothAmsMathEnvironmentsForLineno{align}%
-\patchBothAmsMathEnvironmentsForLineno{flalign}%
-\patchBothAmsMathEnvironmentsForLineno{alignat}%
-\patchBothAmsMathEnvironmentsForLineno{gather}%
-\patchBothAmsMathEnvironmentsForLineno{multline}%
-}
-
-\renewcommand{\baselinestretch}{2}
-
-%% END MACROS SECTION
-
-\begin{document}
-\vspace*{0.2in}
-
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Anne Helby Petersen\textsuperscript{1*\textcurrency},
-Bo Markussen\textsuperscript{2\textcurrency},
-Karl Bang Christensen\textsuperscript{1\textcurrency},
-\\
-\bigskip
-\textbf{1} Department of Public Health, University of Copenhagen, Copenhagen, Denmark
-\\
-\textbf{2} Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark
-\\
-\bigskip
-
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-% 
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-\textcurrency These authors contributed equally to this work.
-
-% Current address notes
-%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
-% \textcurrency b Insert second current address 
-% \textcurrency c Insert third current address
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* ahpe@sund.ku.dk
-
-\end{flushleft}
-% Please keep the abstract below 300 words CHECK
-\section*{Abstract}
-In empirical psychometric work, one often encounters data that is somehow divided into two distinct subsets.
-This might be due to multi-center sampling frames, or to variations in instruments, orders of questionnaire item, modes of administration etc.
-Often, one wishes to combine such data into one and conduct only a single analysis, but this is only meaningful if the two subsets are similar in structures, that is, if they measure the same underlying concepts.
-However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc and require making full model assumptions.
-We present a suite of three visual, non-parametric diagnostic tools for investigating differences in data structures for two datasets containing different observations of the same variables, refered to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC).
-This approach requires no directional nor hierarchical assumptions on the data.
-Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets, and by targeting different aspects of the PCA information, the three plots give a thorough and deep understanding of where differences occur and why so.
-
-The functionality of the PCADSC tools is presented by use of data from the European Social Survey on psycological well-being in three European countries, namely Denmark, Sweden and Bulgaria.
-Here, we find that the concepts of psychological well-being are not the same when comparing Denmark and Bulgaria, and thus a comparison of e.g.
-mean scores in different well-being aspects is not meaningful.
-However, when comparing the two Scandinavian countries, Denmark and Sweden, we find very similar data structures and thus very similar concepts of well-being.
-Therefore, inter-country comparisons are warranted for these two countries.
-
-\linenumbers
-
-\section*{Introduction}
-\label{sec:introduction}
+\section*{Introduction}\label{sec:introduction}
 
@@
 However, useful tools for initial, exploratory investigations of data comparability that help researchers identify potential problems or challenges early in the study design development are mostly absent.
-What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
-Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables.
-On the other hand, e.g.
-entry-by-entry comparisons of the two empirical correlation matrices quickly become unmanageable as the number of variables increase.
+What is needed is a procedure that compares differences in overall data structures in two subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
+Simple methods like variable-by-variable tests of distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables, but entry-by-entry comparisons of two empirical correlation matrices quickly become unmanageable as the number of variables increase.
 
@@
 This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for everyone to use.
-However, large scale publicly available datasets, such as the PISA data and European Social Survey (ESS) data, are often used by data analysts that are far removed from the data producers and therefore, problem-specific recommendations about e.g.
+However, large scale publicly available datasets, such as the PISA (Programme for International Student Assessment) data and European Social Survey (ESS) data, are often used by data analysts that are far removed from the data producers and therefore, problem-specific recommendations about e.g.
 potential instrument-induced challenges in the datasets may not be available for these data analysts.
@@
 Even in single-population survey studies, a mixture of modes of administration is often used, e.g.
-mail and telephone, and while this can improve response rates, it can also be quite problematic \cite{Brambilla1987, McHorney1994}, and differences in response behavior can lead to biased results.
+mail and telephone, and while this can improve response rates, it can also be quite problematic \cite{Brambilla1987,McHorney1994}, and differences in response behavior can lead to biased results.
 Powers, Mishra and Young report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful \cite{Powers2005}, so the comparability issue is not just a statistical puzzle.
-Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu, who warns against the combination of online panel data (i.e.
-pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions) \cite{Liu2016}.
+Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu (2016), who warns against combining online panel data with intercept samples (a pool of respondents obtained through banners, ads, or promotions).\nocite{Liu2016}
 
-The comparability issue that is present in all the examples presented so far can be summarized as follows: Assume that we have two datasets with the same variables, but different observations, or alternatively (and equivalently) a single dataset with a subset-inducing variable.
-Assume next that we wish to compare data structures without specifying a model, or even a variable of interest.
-The central question is then whether the two datasets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be somehow be dealt with, either in terms of modeling choices or simply by conducting several separate analyses.
+The comparability issue in these examples can be summarized as follows: Assume that we have two datasets with the same variables, but different observations (often represented as a single dataset with a subset-inducing variable) and that we wish to compare them without specifying a model, or even a variable of interest.
+The central question is then whether the two datasets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be dealt with in later statistical modelling.
 
-We propose a suite of three new tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC).
-These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
+We propose three visual tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC).
+These methods uses principal component decomposition of the empirical covariance matrix in the two subsets to create intuitive visualizations of data structure differences.
 The proposed tools are implemented in the statistical software \texttt{R} in our package, \texttt{PCADSC} \cite{PCADSC}.
@@
 
-\section*{PCA-based tools for data structure comparisons}
-\label{sec:pcadscintro}
-As mentioned above, the purpose of PCADSC is to compare overall data structures in two subsets of a dataset.
-But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context.
-One such definition is the structure of the covariance matrix of the dataset.
-If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the joint distribution of all the variables.
-%This gives it a very nice interpretation as a measure of the overall structure. If we do not accept 
-And without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data.
-All in all, the empirical covariance matrix is a reasonable place to start looking for differences in \textit{overall data structures}.
+\section*{PCA-based tools for data structure comparisons}\label{sec:pcadscintro}
 
-A naive approach for data structure comparisons might therefore be to compute the empirical covariance matrices on each of the two data subsets and simply compare these matrices entry-by-entry.
-Though the idea perhaps sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices.
-There is simply too much information to consider at once.
-However, by use of linear algebra, we can decompose and recompose the covariance matrix such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance.
-In particular, we find a new representation of the covariance matrix that makes it possible to gain an overview of the most interesting aspects of the data.
-One such decomposition strategy is known as \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.
+PCADSC compares the covariance matrices of the two subsets of a dataset.
+If all variables in the datasets are jointly normal with known means, these are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing interrelations between variables.
+This makes the empirical covariance matrix a reasonable place to start looking for differences in data structures.
+Computing empirical covariance matrices for the two subsets and comparing these entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets.
 
@@
 We assume in the following that all variables of $X$ have a numerical interpretation, e.g.
-by being continuous or ordinal and categorical.
+by being continuous or ordinal categorical.
 For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost.
@@
 Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
-\begin{equation*}
+
+\begin{equation}
 \min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
 \min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
-\end{equation*}
+\end{equation}
+
 The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \cite{HastieEtAl2009}.
 More specifically, the rank-q-reconstruction error is attained when we choose
+
 \begin{equation*}
@@
 \end{equation*}
-where $\eta_1, ..., \eta_q$ are the $q$ first eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$.
+where $\eta_1, ..., \eta_q$ are the first $q$ eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$.
 In the PCA framework, we refer to the eigenvectors as \textit{loadings}.
@@
 All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets.
-Usually, when performing PCA with other purposes in mind, the main interest lies in the \textit{scores}, i.e.
-the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings.
-But where the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables.
-If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both datasets.
-This can be the case while the two sets of scores could be arbitrarily different, which could e.g.\ happen if the two datasets were taken from two different populations of subjects.
+Usually, in PCA, main interest lies in the \textit{scores}, i.e.
+the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings, but the variation components and the accompanying loadings also describe aspects of the data.
+If two different datasets with the same variables, but different samples of observations, have similar loadings, then the variables can be argued to measure the same underlying quantities in both datasets.
+Similar loadings can occur in situations where the two sets of scores are different, e.g.\ if the two datasets come from two different populations of subjects.
 On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.
@@
 \caption{{\bf CE plots for the simulated datasets.} Dataset A, which includes no differences in data structures, is found in the top panel, and dataset B, where the observations stem from two different underlying models, is presented in the bottom panel.
-Both plots are annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
+Both plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures.
 The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.
@@
 More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors.
-The width of the $j$th colored segments are given by
+The width of the $j$th colored segment is given by
 \begin{align*}
@@
 
-\section*{Comparing psychological well-being: A data example}
-\label{sec:dataexample}
-We will now turn to a concrete data example in order to illustrate the capabilities of the tools presented above.
-We use data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
+\section*{Comparing psychological well-being: A data example}\label{sec:dataexample}
+
+We illustrate the three tools using data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
 The topic of our investigation is motivated by an increasingly popular tendency to publish rankings of countries in fields as different as educational quality (e.g.
@@
 the UN \textit{World Happiness Report} project).
-From a methodological point of view, such international rankings are very problematic, as they rely on the fundamental assumption that the measured concepts are inherently the same across countries.
+From a methodological point of view, such international rankings are problematic, as they rely on the fundamental assumption that the measured concepts are inherently the same across countries.
 The PCADSC tools qualify as a suite of methods for exploring the validity of this assumption empirically.
@@
 In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though Veenhoven presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability \cite{Veenhoven2012}, while Lolle \& Andersen show highly potent translation issues for the term \textit{happiness} \cite{Lolle2016}.
-But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
+
+% But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
 If not, the very idea of translation is meaningless, since different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socio-economic settings.
 This is in fact a question concerning comparability of data structures.
+
 If two countries differ e.g.
@@
 In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g.
-conclude using Table~\ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally a lot happier than the least happy people in Sweden.
+conclude using Table~\ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally happier than the least happy people in Sweden.
 A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.
@@
 
-\section*{Discussion}
-\label{sec.Discussion}
+\section*{Discussion}\label{sec.Discussion}
+
 Three new tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if two datasets can be combined for analysis were proposed and discussed in this paper.
@@
 
-\nolinenumbers
-
 % Either type in your references using
@@
 Note that before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.
+
+\newpage
+
+\bibliographystyle{apalike}
+\bibliography{bib}
+
+\end{document}
 
@@
 \end{thebibliography}
-
-\end{document}

### 2017-11-22 article/APM/coverLetter.tex
version til Applied Psyhcological Measurement

--- 2017-06-22 article/PlosOne/coverLetter.tex
@@
 
-Please consider the attached research paper \emph{``Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis''} for publication in PLoS One.
-As Academic Editor to handle the manuscript we suggest Ulrich S.\ Tran, Jakob Pietschnig, or Christopher M.\ Danforth.
+Please consider the attached research paper \emph{``Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis''} for publication in Applied Psychological Measurement.
 
@@
 \end{document}
-
-Summarize the study’s contribution to the scientific literature.
-
-Relate the study to previously published work.
-
-Specify the type of article (for example, research article, systematic review, meta-analysis, clinical trial).
-
-Describe any prior interactions with PLOS regarding the submitted manuscript.
-
-Suggest appropriate Academic Editors to handle your manuscript (see the full list of Academic Editors).
-
-List any opposed reviewers.
-\end{itemize}

### 2018-01-09 article/APM/article.tex
version submittet til APM 9/1-2018

@@
 \usepackage{mathrsfs}
-\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
+%\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
 \usepackage{apalike}
 \usepackage[utf8x]{inputenc}
-\usepackage{nameref,hyperref}
+%\usepackage{nameref,hyperref}
 % line numbers
@@
 
-\author[1]{Anne Helby Petersen}
-\author[1]{Bo Markussen}
-\author[3]{Karl Bang Christensen}
-\affil[1]{Department of Public Health, University of Copenhagen, Copenhagen, Denmark}
-\affil[2]{Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}
+\author[1]{}
+%\author[1]{Anne Helby Petersen}
+%\author[1]{Bo Markussen}
+%\author[3]{Karl Bang Christensen}
+%\affil[1]{Department of Public Health, University of Copenhagen, Copenhagen, Denmark}
+%\affil[2]{Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}
 
@@
 
+\renewcommand{\abstractname}{Abstract}
+\renewcommand{\figurename}{Figure}
+\renewcommand{\tablename}{Table}
+
 \begin{abstract}
@@
 Often, one wishes to combine such data into one and conduct only a single analysis, but this is only meaningful if the subsets are similar in structure.
-Often, procedures for assessing whether such a data collapsing is allowed are ad hoc and require making full model assumptions.
+Common procedures for assessing whether such a data collapsing is allowed are ad hoc and require making full model assumptions.
 We present visual diagnostic tools for investigating differences in structure for two datasets containing different observations of the same variables.
@@
 \end{abstract}
-
-%\keywords{Class file, \LaTeXe, \textit{SAGE Publications}}
 
@@
 We propose three visual tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC).
-These methods uses principal component decomposition of the empirical covariance matrix in the two subsets to create intuitive visualizations of data structure differences.
-The proposed tools are implemented in the statistical software \texttt{R} in our package, \texttt{PCADSC} \cite{PCADSC}.
+These methods use principal component decomposition of the empirical covariance matrix in the two sub sets to create intuitive visualizations of data structure differences.
+The proposed tools are implemented in the \texttt{R} package \texttt{PCADSC} \cite{PCADSC}.
 
@@
 PCADSC compares the covariance matrices of the two subsets of a dataset.
-If all variables in the datasets are jointly normal with known means, these are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing interrelations between variables.
+If all variables in the datasets are jointly normal with known means, the covariance matrices are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing interrelations between variables.
 This makes the empirical covariance matrix a reasonable place to start looking for differences in data structures.
@@
 While we present the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions.
-These plots are available in Figs~\ref{plot.simCE}, \ref{plot.simAngle} and \ref{plot.simChroma} for the CE-, angle- and chroma plots, respectively, and they are based on two simulated datasets:
+These plots are available in Figures \ref{plot.simCE}, \ref{plot.simAngle} and \ref{plot.simChroma} for the CE-, angle- and chroma plots, respectively, and they are based on two simulated datasets:
 \begin{description}
-\item[Dataset A:] This dataset contains 1000 independent simulations from the same underlying model, namely \textit{model 1} from Fig~\ref{fig.simGraphs}.
+\item[Dataset A:] This dataset contains 1000 independent simulations from the same underlying model, namely \textit{model 1} from Figure~\ref{fig:simGraphs}.
 The observations are randomly divided into two groups.
-\item[Dataset B:] This dataset contains 500 independent simulations from \textit{model 1} from Fig~\ref{fig.simGraphs} and 500 independent simulations from \textit{model 2} in the same figure.
+\item[Dataset B:] This dataset contains 500 independent simulations from \textit{model 1} from Figure~\ref{fig:simGraphs} and 500 independent simulations from \textit{model 2} in the same figure.
 The observations are naturally divided into groups corresponding to which model they were simulated from.
 \end{description}
-Further details about the data simulations are available in the \nameref{appendix.simData}.
+
+In \textit{Model 1} the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable $X_1$.
+Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models.
+Further details about data simulation is also available in the supplementary material.
 In dataset A, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures.
-In dataset B, on the other hand, there are two fundamentally different, underlying datastructures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.
+In dataset B, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.
 
 \begin{figure}[!h]
-\caption{{\bf Graphical representations of the two models used for simulating example data for the PCADSC plots.} Square nodes represent observed variables, while rounded nodes represent latent variables.
-Arrows are used to illustrate causal links, and thus we e.g.
-find in \textit{Model 1} that the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable, namely $X_1$.
-Further details on the models and the simulated data are available in the \nameref{appendix.simData}, including covariance matrices for the two models.}
-\label{fig.simGraphs}
+\caption{[FIGURE 1 HERE]}
+% Graphical representations of models used for simulating example data.
+Square nodes represent observed variables, while rounded nodes represent latent variables.
+Arrows are used to illustrate causal links.
+\label{fig:simGraphs}
 \end{figure}
@@
 Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does.
-Two examples of cumulative eigenvalue plots are available in Fig~\ref{plot.simCE}.
+Two examples of cumulative eigenvalue plots are available in Figure \ref{fig:plot.simCE}.
 
 \begin{figure}[!h]
-\caption{{\bf CE plots for the simulated datasets.} Dataset A, which includes no differences in data structures, is found in the top panel, and dataset B, where the observations stem from two different underlying models, is presented in the bottom panel.
-Both plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures.
+\caption{[FIGURE 2 HERE]}
+\label{plot.simCE}
+% CE plots for the simulated datasets.
+Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel.
+Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures.
 The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.
+\end{figure}
+
 In the top panel, we see a cumulative eigenvalue curve that falls well within the confidence region, thereby suggesting no difference in eigenvalues (as expected).
 This also agrees with the large $p$-values and the fact that only a single underlying model was used for generating this data.
-In the bottom panel, the conclusion is reversed, which is also in accordance with the nature of the simulated data, which contains two distinct groups.}
-\label{plot.simCE}
-\end{figure}
-
+In the bottom panel, the conclusion is reversed, which is also in accordance with the nature of the simulated data, which contains two distinct groups.
 In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
@@
 Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.
-Fig~\ref{plot.simAngle} presents two examples of angle plots based on simulated data.
+Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data.
 
 \begin{figure}[!h]
-\caption{{\bf Angle plots for the simulated datasets.}Dataset A is used in the top panel and dataset B is used in the bottom panel.
-The blue show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.
-For dataset A, we find very short off-diagonal arrows, suggesting that the two PCA decompositions agree on the relative placement of the information in  the data among the 6 PCs.
-For dataset B, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}
+\caption{[FIGURE 3 HERE]}
+% Angle plots for the simulated datasets.
+Dataset A in top panel and dataset B in bottom panel.
+Blue arrows show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.
 \label{plot.simAngle}
 \end{figure}
+
+For dataset A, we find very short off-diagonal arrows, suggesting that the two PCA decompositions agree on the relative placement of the information in  the data among the 6 PCs.
+For dataset B, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.
 
@@
 In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$.
-In Fig~ \ref{plot.simChroma}, two examples of chroma plots are available.
+In Figure  \ref{plot.simChroma}, two examples of chroma plots are available.
 
 \begin{figure}[!h]
-\caption{Chroma plots for the simulated datasets, with dataset A in the top and dataset B in the bottom.
-Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance ($\tilde\sigma_{ij}^c$).
-For dataset A, we see very similar visual patterns across the two panels, corresponding well with the fact that both groups are drawn from the same underlying model.
-For dataset B, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups.}
+\caption{[FIGURE 4 HERE]}
+% Chroma plots for the simulated datasets.
+Dataset A in top panel and dataset B in bottom panel.
+Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.
 \label{plot.simChroma}
@@
 
+For dataset A, we see very similar visual patterns across the two panels, corresponding well with the fact that both groups are drawn from the same underlying model.
+For dataset B, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups.
 The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.
@@
 In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though Veenhoven presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability \cite{Veenhoven2012}, while Lolle \& Andersen show highly potent translation issues for the term \textit{happiness} \cite{Lolle2016}.
-
-% But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
-If not, the very idea of translation is meaningless, since different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socio-economic settings.
-This is in fact a question concerning comparability of data structures.
 
@@
 
-All computations and figures presented in this section were created using our \R{R} package \R{PCADSC} \cite{PCADSC}.
+All computations and figures presented in this section were created using the \R{R} package \R{PCADSC} \cite{PCADSC}.
 
@@
 These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}.
-More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in the \nameref{table:items}.
+More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in the supplementray material.
 We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10.
@@
 Community wellbeing     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
-Supportive relationships & 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
+Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
 \hline
@@
 \subsection*{Comparing Denmark and Bulgaria}
-Fig~\ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales.
+Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales.
 The CE plot show a remarkable degree of lacking comparability: The cumulative differences in the eigenvalues by far exceed what could come about randomly if there really were no difference in the data structures.
@@
 \begin{figure}[!h]
-\caption{{\bf CE- and angle plots comparing Danish and Bulgarian psychological well-being.} The CE plot (top) and the angle plot (bottom) resulting from comparing Bulgarian and Danish data on psychosocial well-being.
+\caption{[FIGURE 5 HERE]}
+% CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being.
 The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
-In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
+In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.
 \label{plotBG.cehair}
@@
 
-The chroma plot in Fig~\ref{plotBG.pancake} allows us to look closer into these components.
+The chroma plot in Figure \ref{plotBG.pancake} allows us to look closer into these components.
 Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish.
@@
 \begin{figure}[!h]
-\caption{{\bf Chroma plot for comparing Danish and Bulgarian psychological well-being.} A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data.
-The component-bars are annotated with their relative variance contributions (denoted $\tilde\sigma$ in the previous section).}
+\caption{[FIGURE 6 HERE]}
+% Chroma plot for comparing Danish and Bulgarian data on psychological well-being.
+A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data.
+The component-bars are annotated with their relative variance contributions.
 \label{plotBG.pancake}
@@
 We now turn to the comparison of Denmark and Sweden in terms of psychological well-being.
-Fig~\ref{plotSE.cehair} shows the CE- and angle plots for these two countries.
+Figure \ref{plotSE.cehair} shows the CE- and angle plots for these two countries.
 In the CE plot, we now find the cumulative eigenvalue curve to be just within the acceptance region of the null-hypothesis.
@@
 \begin{figure}[!h]
-\caption{{\bf CE- and angle plots comparing Danish and Swedish psychological well-being.} A CE (top) and a angle (bottom) plot for comparing the Danish and the Swedish psychological well-being data.
-The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
+\caption{[FIGURE 7 HERE]}
+% A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being.
+The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.
 \label{plotSE.cehair}
@@
 
-The angle plot in Fig~\ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent.
+The angle plot in Figure \ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent.
 This implies that if one already has e.g.
@@
 
-Looking at the chroma plot in Fig~\ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80\% of the variance in both datasets), and slight, but increasing, differences in the remaining three components.
+Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80\% of the variance in both datasets), and slight, but increasing, differences in the remaining three components.
 We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25\% of the variance in both datasets.
@@
 \begin{figure}[!h]
-\caption{{\bf Chroma plot comparing Danish and Bulgarian psychological well-being.} A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples.
-Note that for each component, the bar is annotated with its cumulative variance contribution (denoted $\tilde\sigma^c$ in the previous section), that is, how much variance can be explained by having information of this and the preceding components.}
+\caption{[FIGURE 8 HERE]}
+% Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples.
+For each component, the bar is annotated with its cumulative variance contribution, that is, how much variance can be explained by having information of this and the preceding components.
 \label{plotSE.pancake}
@@
 For instance, the PISA tests have repeatedly been criticized for not being meaningful objects of international comparisons, especially due to problems with differential item functioning \cite{Kankaras2014,Kreiner2014,ZwitserEtAl2017} and translation problems \cite{Asil2016}.
-This highlights lack of data comparability as an empirically relevant and concerning issue that there is no way around: When conducting internal rankings, the issue of data structure heterogeneity should always be addressed.
+This highlights lack of data comparability as an empirically relevant and concerning issue that there is no way around: When conducting rankings, the issue of data structure heterogeneity should always be addressed.
 
@@
 The resampling approach used in the CE plot could then be adopted, thus annotating the scree plots with resampled curves.
-This general approach to model-fit evaluation is very helpful in combination with permutation tests, as inspired by Lin and colleages \cite{LinEtAl2002}.
+This general approach to model-fit evaluation is very helpful in combination with permutation tests, as inspired by Lin and colleagues \cite{LinEtAl2002}.
 Similarly, the angle plot could also be annotated with $p$-values from a permutation test based on appropriate test statistics.
@@
 
-\section*{Supporting information}
+\section*{Supplementary material}
 
@@
 \bibliography{bib}
-
 \end{document}
-
-\begin{thebibliography}{10}
-
-\bibitem{Brambilla1987}
-Brambilla DJ, McKinlay SM.
-\newblock {A comparison of responses to mailed questionnaires and telephone
-  interviews in a mixed mode health survey.}
-\newblock American journal of epidemiology. 1987;126(5):962--71.
-
-\bibitem{McHorney1994}
-McHorney CA, Kosinski M, Ware JE.
-\newblock {Comparisons of the costs and quality of norms for the SF-36 health
-  survey collected by mail versus telephone interview: results from a national
-  survey.}
-\newblock Medical care. 1994;32(6):551--67.
-
-\bibitem{Powers2005}
-Powers JR, Mishra G, Young AF.
-\newblock {Differences in mail and telephone responses to self-rated health:
-  use of multiple imputation in correcting for response bias.}
-\newblock Australian and New Zealand journal of public health.
-  2005;29(2):149--54.
-
-\bibitem{Liu2016}
-Liu M.
-\newblock {Comparing data quality between online panel and intercept samples}.
-\newblock Methodological Innovations. 2016;9:2059799116672877.
-\newblock doi:{10.1177/2059799116672877}.
-
-\bibitem{PCADSC}
-Petersen AH, Markussen B. PCADSC: Tools for Principal Component Analysis-Based
-  Data Structure Comparisons; 2017.
-\newblock Available from: \url{https://CRAN.R-project.org/package=PCADSC}.
-
-\bibitem{WHR2016}
-Helliwell J, Layard R, Sachs J.
-\newblock {World Happiness Report 2016 Update}.
-\newblock New York: Sustainable Development Solutions Network. 2016;.
-
-\bibitem{HastieEtAl2009}
-Hastie T, Tibshirani R, Friedman J.
-\newblock 14.5.1.
-\newblock In: The Elements of Statistical Learning. 2nd ed. Springer; 2009.
-
-\bibitem{Veenhoven2012}
-Veenhoven R.
-\newblock {Cross-national differences in happiness: Cultural measurement bias
-  or effect of culture?}
-\newblock International Journal of Wellbeing. 2012;2(4):333--353.
-\newblock doi:{10.5502/ijw.v2.i4.4}.
-
-\bibitem{Lolle2016}
-Lolle HL, Andersen JG.
-\newblock {Measuring Happiness and Overall Life Satisfaction: A Danish Survey
-  Experiment on the Impact of Language and Translation Problems}.
-\newblock Journal of Happiness Studies. 2016;17(4):1337--1350.
-\newblock doi:{10.1007/s10902-015-9646-4}.
-
-\bibitem{ESStopline5}
-Jeffrey K, Abdallah S, Quick A.
-\newblock {Europeans' Personal and Social Wellbeing: Topline Results from Round
-  6 of the European Social Survey}.
-\newblock ESS Topline Results (Series 5). 2015;.
-
-\bibitem{Rodriguez2014}
-Rodr\'iguez-Pose A, von Berlepsch V.
-\newblock {Social Capital and Individual Happiness in Europe}.
-\newblock Journal of Happiness Studies. 2014;15(2):357--386.
-\newblock doi:{10.1007/s10902-013-9426-y}.
-
-\bibitem{Kankaras2014}
-Kankara{\v{s}} M, Moors G.
-\newblock {Analysis of Cross-Cultural Comparability of PISA 2009 Scores}.
-\newblock Journal of Cross-Cultural Psychology. 2014;45(3):381--399.
-\newblock doi:{10.1177/0022022113511297}.
-
-\bibitem{Kreiner2014}
-Kreiner S, Christensen KB.
-\newblock {Analyses of Model Fit and Robustness. A New Look at the PISA Scaling
-  Model Underlying Ranking of Countries According to Reading Literacy}.
-\newblock Psychometrika. 2014;79(2):210--231.
-\newblock doi:{10.1007/s11336-013-9347-z}.
-
-\bibitem{ZwitserEtAl2017}
-Zwitser RJ, Glaser SSF, Maris G.
-\newblock {Monitoring Countries in a Changing World: A New Look at DIF in
-  International Surveys}.
-\newblock Psychometrika. 2017;82(1):210--232.
-
-\bibitem{Asil2016}
-Asil M, Brown GTL.
-\newblock {Comparing OECD PISA Reading in English to Other Languages:
-  Identifying Potential Sources of NonInvariance}.
-\newblock International Journal of Testing. 2016;16(1):71--93.
-\newblock doi:{10.1080/15305058.2015.1064431}.
-
-\bibitem{Cattell1966}
-Cattell RB.
-\newblock {The Scree Test For The Number Of Factors}.
-\newblock Multivariate Behavioral Research. 1966;1(2):245--276.
-
-\bibitem{LinEtAl2002}
-Lin DY, Wei LJ, Ying Z.
-\newblock {Model-Checking Techniques Based on Cumulative Residuals}.
-\newblock Biometrics. 2002;58:1--12.
-
-\bibitem{Abdi2010}
-Abdi H, Williams LJ.
-\newblock Principal component analysis.
-\newblock Wiley Interdisciplinary Reviews: Computational Statistics.
-  2010;2(4):433--459.
-\newblock doi:{10.1002/wics.101}.
-
-\bibitem{Ekstrom2014}
-Ekstr{\o}m CT.
-\newblock {Teaching 'Instant Experience' with Graphical Model Validation
-  Techniques}.
-\newblock Teaching Statistics. 2014;36(1):23--26.
-\newblock doi:{10.1111/test.12027}.
-
-\end{thebibliography}

### 2018-03-08 article/Old versions/APM/S2_table.tex
Redid folder structure + added first print/summary functions

--- 2017-04-25 article/article_v0.9.5.tex
@@
-\documentclass[titlepage,11pt,twoside]{article}
+\documentclass{article}
 \usepackage[dvips]{graphicx}
+\usepackage{fullpage}
 
-\usepackage[myheadings]{fullpage}
-\usepackage{pmetrika}
-%\usepackage{pmbib}
-
+\usepackage{float}
+\restylefloat{table}
 \usepackage{color}
@@
 \usepackage{hyperref}
-\usepackage{natbib}
 \usepackage{multirow}
-%\usepackage{colortbl}
 \usepackage[table]{xcolor}
-\definecolor{lightgray}{gray}{0.9}
+\definecolor{lightgray}{HTML}{e6e6e6}
 \usepackage[T1]{fontenc}
@@
 \captionsetup{font={footnotesize,stretch=1.5}}
-\renewcommand{\baselinestretch}{2}
+%\renewcommand{\baselinestretch}{2}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
 
@@
 
-%\begin{figure}[h]
-%\centerline{\includegraphics{figure03.eps}}
-%\caption{Projection of item discrimination vectors onto $V_{\theta_T}$ hyperplance for a six item three-dimensional approximate sample structure.}
-%\end{figure}
-
-%\raggedbottom
-\flushbottom
-
-%\firstpage{1}
-%\setcounter{lastpage}{999}
-\setcounter{secnumdepth}{3}
-
 \begin{document}
 
-\begin{titlepage}
-\linespacing{1}
-
-\title{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}
-
-\author{Anne H. Petersen}
-\affil{Department of Public Health, University of Copenhagen}
-
-\author{Bo Markussen}
-\affil{Department of Mathematical Sciences, University of Copenhagen}
-
-\author{Karl Bang Christensen}
-\affil{Department of Public Health, University of Copenhagen}
-
-\vspace{\fill}
-
-\end{titlepage}\vspace*{24pt}
-
-%\RepeatTitle{Exploratory data structure comparisons: Three new tools based on Principal Component Analysis}
-
-\begin{center}\vskip3pt
-
-\vspace{32pt}
-
-Abstract\vskip3pt
-
-\end{center}
-
-\begin{abstract}
-\linespacing{1}
-
-In empirical psychometric work, one often encounters data that is somehow divided into two distinct subsets.
-This might be due to multi-center sampling frames, or to variations in instruments, orders of questionnaire item, modes of administration etc.
-Often, one wishes to combine such data into one an conduct only a single analysis, but this is only meaningful if the two subsets are similar in structures, that is, if they measure the same underlying concepts.
-However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc or require making full model assumptions.
-We present a suite of three visual, non-parametric diagnostic tools for investigating differences in data structures for two datasets containing different observations of the same variables, refered to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC).
-This approach requires no directional nor hierarchical assumptions on the data.
-Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets, and by targeting different aspects of the PCA information, the three plots give a thorough and deep understanding of where differences occur and why it might be the case.
-
-The functionality of the PCADSC tools is presented by use of data from the European Social Survey on psycological well-being in three European countries, namely Denmark, Sweden and Bulgaria.
-Here, we find that the concepts of psychological well-being are not the same when comparing Denmark and Bulgaria, and thus a comparison of e.g.
-mean scores in different well-being aspects is not meaningful.
-However, when comparing the two Scandinavian countries, Denmark and Sweden, we find very similar data structures and thus very similar concepts of well-being.
-Therefore, inter-country comparisons are warranted for these two countries.
-
-\begin{keywords}
-multi-center studies, data heterogeneity, data structure comparisons, principal component analysis (PCA), European Social Survey.
-\end{keywords}
-\end{abstract}
-
-\vspace{\fill}\newpage
-
-\section{Introduction}
-\label{sec:introduction}
-
-Data comparability is a classical topic in psychometrics and, more generally, in statistics.
-Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability (and thus collapsibility) needs to be assessed empirically.
-This happens e.g.
-when data are collected across several centers (or even countries) or when different versions of an instrument, the mode of administration or the order of survey questionnaire items are applied.
-All of these practices are essential parts of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions.
-However, if we do not address the question of collapsibility among the data subsets arriving from e.g.
-different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.
-
-Sophisticated methods for addressing this question are available when we are willing to assume a statistical model.
-But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment %very
-ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a %valid 
-general data structure comparison method, but rather a fitted-model comparison method.
-It addresses the interplay between the model and the data, not the data alone.
-However, useful tools for initial, exploratory investigations of data comparability data that help researchers identify potential problems or challenges early in the study design development are mostly absent.
-What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
-Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables.
-On the other hand, e.g.
-entry-by-entry comparisons of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase.
-
-While the issue is not in any way new, its relative importance is increasing due to new developments in data availability and data collection.
-Classical statistical methods are generally aimed at analyzing data from designed experiments and historically, statistical analyses have been conducted by researchers who knew the design and the source of the dataset well.
-However, the origin stories of datasets have changed over time and today, a lot of data are accumulated without a specific purpose in mind, as data collection and sharing has become easier and more affordable through the development of new technologies.
-This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for %all
-everyone to use.
-However, large scale publicly available datasets, such as the PISA data and European Social Survey (ESS) data, are often used by data analysts that are far removed from the data producers and therefore, problem-specific recommendations about e.g.
-potential instrument-induced challenges in the datasets %are not
-may not be available for these data analysts.
-For this reason, a thorough investigation of data structure comparability becomes a crucial step in any meaningful data analysis when using such open-source data.
-
-The data structure comparison problem is however not limited to the context of new data sharing platforms.
-Even in single-population survey studies, a mixture of modes of administration is often used, e.g.
-mail and telephone, and while this can improve response rates, it can also be quite problematic \citep{Brambilla1987, McHorney1994}, and differences in response behavior can lead to biased results.
-Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful, so the comparability issue is not just a statistical puzzle.
-Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu \citeyearpar{Liu2016}, who warns against the combination of online panel data (i.e.
-pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions).
-
-The comparability issue that is present in all the examples presented so far can be summarized as follows: Assume that we have two datasets with the same variables, but different observations, or alternatively (and equivalently) a single dataset with a subset-inducing variable.
-Assume next that we wish to compare data structures without specifying a model, or even a variable of interest.
-The central question is then whether the two datasets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be somehow be dealt with, either in terms of modeling choices or simply by conducting several separate analyses.
-
-We propose a suite of three new tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC).
-These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
-%This yields a solution that is largely independent of the sizes of the two subsets of data. Anne: Det ved vi da ikke om det gør? Er det ikke netop det, vi siger, at man burde teste som næste step?
-The proposed tools are implemented in the statistical software \texttt{R} in our package, \texttt{PCADSC} \citep{PCADSC}.
-
-This manuscript describes the procedures, including a brief introduction to principal component analysis (PCA), and presents a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey.
-More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is completely ignored e.g.
-when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project \citep{WHR2016}.
-
-\section{PCA-based tools for data structure comparisons}
-\label{sec:pcadscintro}
-As mentioned above, the purpose of PCADSC is to compare overall data structures in two subsets of a dataset.
-But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context.
-One such definition is the structure of the covariance matrix of the dataset.
-If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables.
-%This gives it a very nice interpretation as a measure of the overall structure. If we do not accept 
-And without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data.
-All in all, the empirical covariance matrix is a reasonable place to start looking for differences in \textit{overall data structures}.
-
-A naive approach for data structure comparisons might therefore be to compute the empirical covariance matrices on each of the two data subsets and simply compare these matrices entry-by-entry.
-Though the idea perhaps sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices.
-There is simply too much information to consider at once.
-However, by use of linear algebra, we can decompose and recompose the covariance matrix such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance.
-In particular, we find a new representation of the covariance matrix that makes it possible to gain an overview of the most interesting aspects of the data.
-One such decomposition strategy is known as \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.
-
-At this point a general remark about standardization is in place.
-PCA deconstructs the covariance matrix in components according to the most explained variance.
-This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will always be deemed highly influential, no matter the structure of the data.
-Therefore, the variables are typically standardized prior to performing PCA.
-Note that the covariance matrix of the standardized variables is the same as the correlation matrix of the original variables, so post-standardization PCA simply corresponds to performing data structure comparisons of the correlation matrices rather than the covariance matrices.
-The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation.
-In this paper we will adopt this standardization policy, and in particular, we will allow for differences between scalings of the original, raw variables in the two data subsets.
-
-%\hl{Bo: I næste afsnit bruges stadigvæk ``covariance''. Det er vel ok?} Anne: Synes det fungerer fint.
-
-\subsection{A brief introduction to principal component analysis}
-Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = (\bar{x}_1, ..., \bar{x}_d)^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$.
-We assume in the following that all variables of $X$ have a numerical interpretation, e.g.
-by being continuous or ordinal and categorical.
-For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost.
-More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the original $d$-dimensional one.
-Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
-\begin{equation*}
-\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
-\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
-\end{equation*}
-The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \citep{HastieEtAl2009}.
-More specifically, the rank-q-reconstruction error is attained when we choose
-\begin{equation*}
-\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\},
-\end{equation*}
-where $\eta_1, ..., \eta_q$ are the $q$ first eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$.
-In the PCA framework, we refer to the eigenvectors as \textit{loadings}.
-The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
-\begin{equation*}
-\text{trace}(S) = \sum_{j=1}^d \hat{V}(x_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
-\end{equation*}
-where $x_j \in \RR^n$ 
-%\hl{(Bo: $X_1$ og $X_2$ bruges i en anden betydning i afsnit 2.2. Omformuler?)} Anne: Hov, ja, hermed x_j så det passer med notationen ovenfor.
-denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
-This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
-And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are %different,
-distinct, the representation hereby obtained is a valid object of inference.
-If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
-
-The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
-\begin{align*}
-\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
-\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2,
-\end{align*}
-where the initial subspace is defined as $\hat{K}_0 = \{0\}$.
-It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
-This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$.
-This implies that the loadings and the variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.
-
-\subsection{Using PCA for data structure comparisons}
-All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets.
-Usually, when performing PCA with other purposes in mind, the main interest lies in the \textit{scores}, i.e.
-the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings.
-But where the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables.
-If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both datasets.
-This can be the case while the two sets of scores could be arbitrarily different, which could e.g.\ happen if the two datasets were taken from two different populations of subjects.
-On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.
-
-The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables.
-We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the apparently same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$.
-%
-Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For these %two
-three datasets, we complete the following steps:
-\begin{enumerate}
-\item Within each dataset, standardize each of the variables to have mean zero and unit standard deviation.
-Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ and $\tilde{x}_j \in \RR^d$ for $j=1,\dotsc,n$ be the standardized datasets.
-\item Form the principal component analyses
-\begin{align*}
-S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} &
-S &= \frac{1}{n-1} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top,
-\end{align*}
-thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda_k$ for $k=1,\dotsc,d$.
-\end{enumerate}
-The hereby obtained PCA decompositions of the correlation matrices can then be compared.
-We present three diagnostics plots that are designed to shine a light on different types and levels of data structure differences.
-These three plots are:
-\begin{description}
-\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA.
-The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
-\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e.
-loading and variance component pairs) that are most similar and most different across the two datasets
-\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
-\end{description}
-For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.
-
-Note that the standardization implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
-\begin{equation*}
-\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d.
-\end{equation*}
-This identity will simplify some expressions below.
-
-\subsubsection{The cumulative eigenvalue plot}
-The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix.
-%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
-And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
-In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
-\begin{align*}
-(0,0), &&
-(\lambda_1,\lambda_{11}-\lambda_{12}), &&
-(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
-\ldots, &&
-\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
-\end{align*}
-This may be seen as a cumulative Bland-Altman plot \citep{AltmanBland83,LinEtAl2002} for the variation components.
-Note that due to the standardization, the last point will always be equal to $(d,0)$.
-Thus, this curve will begin and end at the x-axis.
-And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are.
-Moreover, a positive cumulative differences implies that dataset 1 holds more information in the first components than dataset 2 does.
-
-In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
-\begin{align*}
-\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
-\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)^2.
-\end{align*}
-We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_1$ and $n_2$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics.
-This should be done a large (e.g.
-10000) number of times.
-Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to %even larger
-test statistics at least as large as %than 
-the one we found for the original datasets.
-
-The permutation test results are also used to visualize the uncertainty of the CE curve in the plots.
-In the CE plots shown in the following section, we plot the observed curve together with 20 of the resampled curves, as well as a shaded region visualizing pointwise 95 \% coverage intervals.
-If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicates differences between the two datasets.
-
-%Whether the excursions in the observed curve are large or within the range of sample variation can be quantified by a permutation test.
-The idea is that we randomly reallocate the $n_1+n_2$ observations in the combined dataset to two datasets with $n_1$ and $n_2$ observations, respectively, and then repeat the procedure described above.
-The random reallocation by construction ensures that the two resampled datasets are alike except their sample size and sampling variation.
-In the \emph{CumEigenPlot} we make 1000 independent random reallocations, and plot  P-values for the null hypothesis that the variation components are the same are easily provided for any appropriate test statistic by the resampling procedure as well.
-
-\subsubsection{The angle plot}
-This plot simultaneously compares the variation components and the loadings.
-Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
-Then the empirical correlation matrix for the first dataset, $S_1$, has the following orthogonal decomposition in the coordinate system of the second dataset
-%\begin{equation*}
-%S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
-%= \lambda_{\max} \sum_{k=1}^d
-%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
-%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
-%\end{equation*}
-\begin{equation*}
-S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
-= \lambda_{\max} \sum_{k=1}^d
-\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
-\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)^\top,
-\end{equation*}
-and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset %\hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?
-Bo: Jeg har ikke noget bud på en sådan udvidelse lige nu og her --- skal vi ikke bare sige at det er ok i sin nuværende form?}.
-We propose to visualize these two decompositions in a $d \times d$ grid display.
-In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell.
-The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal.
-To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows.
-The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
-\begin{align*}
-\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
-\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
-\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
-\end{align*}
-Note that for two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the sample correlation.
-Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance component contributions.
-The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors.
-This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal.
-Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.
-
-In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the first dataset in the coordinate system of the second dataset.
-Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the second dataset in the coordinate system of the first dataset.
-%\hl{Can we expand on this in a non-technical way?
-- Karl: it seems that we cannot}
-If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
-Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
-And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.
-
-\subsubsection{The chroma plot}
-This plot compares the loadings of the two datasets.
-The chroma plot consists of two panels, one for each dataset, made up of colored bars.
-These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions.
-More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors.
-The width of the $j$th colored segments are given by
-\begin{align*}
-\omega_{1ij} &= \frac{\lvert\eta_{1ij}\rvert}{\sum_{k=1}^d \lvert\eta_{1ik}\rvert}, &
-\omega_{2ij} &= \frac{\lvert\eta_{2ij}\rvert}{\sum_{k=1}^d \lvert\eta_{2ik}\rvert},
-\end{align*}
-where $\eta_{1ij}$ and $\eta_{2ij}$ denotes the $j$th entry of $\eta_{1i}$ and $\eta_{2i}$, respectively.
-Due to the indeterminacy of the sign, all the signs are removed from the coefficients in the loadings.
-The bars are ordered according to the variation components and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance component contributions,
-\begin{align*}
-\tilde\sigma^c_{1i} &= \frac{\sum_{j=1}^i \lambda_{1j}}{\sum_{k=1}^d \lambda_{1k}} = \frac{\sum_{j=1}^i \lambda_{1j}}{d}, &
-\tilde\sigma^c_{2i} &= \frac{\sum_{j=1}^i \lambda_{2j}}{\sum_{k=1}^d \lambda_{2k}} = \frac{\sum_{j=1}^i \lambda_{2j}}{d}.
-\end{align*}
-Especially when $d$ is large, we recommend plotting only a select set of interesting principal components, e.g.
-as identified by use of the angle plot.
-In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$.
-
-The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.
-For each component, the loadings describe how influential the different variables are on that component.
-Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such as \textit{"variable $x$ is generally more influential in subset 1 than it is in subset 2"}, thereby helping us to understand where and why potential data structure differences are found.
-
-\section{European differences in psychological well-being: A data example}
-\label{sec:dataexample}
-We will now turn to a concrete data example in order to illustrate the capabilities of the tools presented above.
-We use data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
-The topic of our investigation is motivated by an increasingly popular tendency to publish rankings of countries in fields as different as educational quality (e.g.
-the PISA tests) and citizen happiness (e.g.
-the UN \textit{World Happiness Report} project).
-From a methodological point of view, such international rankings are very problematic, as they rely on the fundamental assumption that the measured concepts are inherently the same across countries.
-The PCADSC tools qualify as a suite of methods for exploring the validity of this assumption empirically.
-
-In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though \cite{Veenhoven2012} presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability, while \cite{Lolle2016} shows highly potent translation issues for the term \textit{happiness}.
-But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
-If not, the very idea of translation is meaningless, since different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socio-economic settings.
-This is in fact a question concerning comparability of data structures.
-If two countries differ e.g.
-in terms of how social networks are typically build and structured, with one emphasizing family relations and the other mostly focusing on other social relations, having a weak family connection does not have the same implications in the first country as it does in the second one.
-More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being.
-The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures.
-And therefore, comparing the two countries in these measures is not a meaningful endeavor.
-
-In this section, we use the PCADSC tools to unveil international differences in one aspect of happiness, namely psychological well-being.
-Our starting point is Denmark, a small, northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \citep{WHR2016}, and we wish to investigate if this title is really meaningful at all.
-In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria.
-Though both countries are European and thus not geographically nor culturally as far apart as some other countries might be, these two countries have previously been highlighted to be very different in terms of what defines happiness \citep{ESStopline5}.
-Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \citep{Rodriguez2014}, with a much less pronounced relationship between the two in Northern European countries, compared to other European countries.
-In particular, interpersonal relations should play a less important role in Denmark, compared to Bulgaria.
-Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.
-
-We also compare the Danish data with Swedish data in order to illustrate that the PCADSC tools actually do have some discriminatory power.
-Denmark and Sweden are both Northern European countries and are often deemed very similar in terms of culture and history.
-Therefore, we expect fundamental concepts such as psychological well-being to be similar across these two countries.
-
-All computations and figures presented in this section were created using our \R{R} package \R{PCADSC} \citep{PCADSC}.
-
-\subsection{Data}
-
-\begin{table}[t]
-\centering
-\begin{tabular}{lccccccccc}
-  \hline
-  & \multicolumn{3}{c}{Denmark} & \multicolumn{3}{c}{Bulgaria} & \multicolumn{3}{c}{Sweden} \\
- & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \\
-  \hline
- Evaluative wellbeing & 8.00 & 8.75 & 9.50 & 3.50 & 5.00 & 7.00 & 7.00 & 8.00 & 9.00 \\
- Emotional wellbeing & 7.22 & 8.33 & 8.89 & 5.00 & 6.67 & 7.78 & 6.67 & 7.78 & 8.89 \\
-Functioning & 6.93 & 7.57 & 8.21 & 5.50 & 6.68 & 7.68 & 6.39 & 7.04 & 7.68 \\
-Vitality & 6.67 & 7.50 & 8.33 & 5.83 & 7.50 & 8.33 & 6.67 & 7.50 & 9.17 \\
- Community wellbeing & 5.83 & 6.77 & 7.57 & 3.70 & 4.67 & 5.70 & 5.66 & 6.57 & 7.37 \\
-Supportive relationships & 7.42 & 8.25 & 8.92 & 6.17 & 7.25 & 8.08 & 7.42 & 8.25 & 8.75 \\
-   \hline
-\end{tabular}
-\caption{The 1st quartile, the median and the third quartile of the distrubutions of each of the six dimensions of psychological well-being, stratified by country.
-Note that the scales are constructed such that they all run from 0-10.}
-\label{tableDistr}
-\end{table}
-
-The ESS 2012 data contains a total of 626 variables collected from 54673 citizens of 29 countries.
-Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
-These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}.
-More details on these scales can be found in \citep{ESStopline5} and the relationship between questionnaire items and scales is summarized in Table \ref{table:items} in the appendices.
-We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10.
-For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below.
-This gives us $n_{DK} = 1498$ observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations.
-Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
-
-\subsection{Comparing Denmark and Bulgaria}
-\begin{figure}
-\centerline{\includegraphics[scale = 0.65]{essDKBGce.pdf}}
-\centerline{\includegraphics[scale = 0.65]{essDKBGhair.pdf}}
-\caption{The CE plot (top) and the angle plot (bottom) resulting from comparing Bulgarian and Danish data on psychosocial well-being.
-The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
-In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
-\label{plotBG.cehair}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[scale = 0.7]{essDKBGpancake234.pdf}}
-\caption{A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data.
-The component-bars are annotated with their relative variance contributions (denoted $\tilde\sigma_i$ in the above).}
-\label{plotBG.pancake}
-\end{figure}
-
-%\begin{figure}
-%\centerline{\includegraphics[scale = 0.7]{essDKBGWallyPCADSC234.pdf}}
-%\caption{\hl{something.}}
-%\label{plotBG.wally}
-%\end{figure}
-
-Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales.
-The CE plot show a remarkable degree of lacking comparability: The cumulative differences in the eigenvalues by far exceed what could come about randomly if there really were no difference in the data structures.
-This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero.
-
-Moving on to the angle plot, we find that the differences are primarily to be found in the second, third and fourth principal components (PCs).
-The blue arrows visualize the decomposition of the principal components for the Bulgarian dataset in the coordinate system of the Danish dataset.
-We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3.
-The red arrows visualize the decomposition of the principal components for the Danish dataset in the coordinate system of the Bulgarian dataset.
-Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3.
-Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.
-
-The chroma blot in Figure \ref{plotBG.pancake} allows us to look closer into these components.
-Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish.
-In the Danish data, on the other hand, we find that \textit{Vitality} and \textit{Emotional well-being} seem to play bigger roles, as they appear with larger loadings in more high-ranking components in this sample, relative to the Bulgarian.
-
-All in all, we find that psychological well-being does \textit{not} seem to be the same concept in Bulgaria and Denmark.
-The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalues) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle- and chroma plots).
-In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, corresponding well with the previous findings mentioned above.
-Thus, the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics (\cite{ESStopline5}).
-Moreover, the two countries cannot be ranked in terms of which country is "the most happy", at least not by referring to psychological well-being dimensions such as those encountered here.
-
-% While the first principal component, which is responsible for explaining 50-60 \% of the variance in the data, is very similar for the two countries, we see quite large differences in the remaining components.
-In the second component, we see that the two countries disagree in the relative importance of the scales \textit{Community wellbeing} and \textit{Vitality}.
-In the third and fourth components, general disagreement  is found.
-All in all, components 2-4, representing almost half of the variability in the data, are not very similar across the two countries.
-Moreover, the two subsets of the data also differ with respect to how much variance is explained by each component, and the difference is particularly big for the first component.
-This component has approximately 15 \%  more explanatory power in the Bulgarian subsample than it does in the Danish.
-
-%{\color{red}
-%Figure \ref{plotESSHairplot} shows the hairplot.
-The blue arrows visualize the decomposition of the principal components for the %first dataset in the coordinate system of the second dataset.
-We see that PC2 also loads on PC3, that PC3 also loads on PC4, and %that PC4 also loads on PC2 and PC3.
-The red arrows visualize the decomposition of the principal components for the second dataset %in the coordinate system of the first dataset.
-We see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that %PC4 also loads on PC3.
-For PC1, PC5 and PC6 the main difference is in the size of the variation component.
-%}
-
-%But did we really illustrate a data structure difference due to country differences or did we just illustrate the variability of the results of the PCADSC method?
-In order to investigate this further, we look at a so-called \textit{Wally plot} (\hl{ref: Claus Ekstrøm}).
-In this plot, we compare the results of PCADSC conducted with grouping by country with several random, but similar grouping variables.
-Specifically, we produce 7 PCADSC plots where the country variable was replaced by a randomly generated variable that divides the observations into two groups of the same sizes as the country samples.
-The results are illustrated in Figure \ref{plotESSPCADSCWally}.
-Here, we see that the differences in the second component from the original PCADSC results are not matched in any of the randomly grouped PCADSC runs.
-In fact, the 7 runs are remarkably similar, thereby illustrating that PCADSC seems to be very robust with respect to random groupings: The signal in the data is not blurred by the random subdivisions.
-When it comes to the differences in the third component for the two groups, we find much larger variability in the 7 random runs.
-\hl{more comments here...
-Wait until we are sure exactly what we think about the results and what other PCA-based methods, we will do before/after.
-Particularly, how do we deal with eigen value differences?}
-
-\subsection{Comparing Denmark and Sweden}
-\begin{figure}
-\centerline{\includegraphics[scale = 0.65]{essDKSEce.pdf}}
-\centerline{\includegraphics[scale = 0.65]{essDKSEhair.pdf}}
-\caption{A CE (top) and a angle (bottom) plot for comparing the Danish and the Swedish psychological well-being data.
-The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
-\label{plotSE.cehair}
-\end{figure}
-
-\begin{figure}
-\center
-\centerline{\includegraphics[scale = 0.7]{essDKSEpancake.pdf}}
-\caption{A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples.
-Note that for each component, the bar is annotated with its cumulative variance contribution (denoted $\tilde\sigma^c_i$ in the above), that is, how much variance can be explained by having information of this and the preceding components.}
-\label{plotSE.pancake}
-\end{figure}
-
-%\begin{figure}
-%\center
-%\includegraphics[scale = 0.7]{essDKSEWallyPCADSC.pdf}
-%\caption{\hl{something.}}
-%\label{plotSE.wally}
-%\end{figure}
-
-We now turn to the comparison of Denmark and Sweden in terms of psychological well-being.
-Figure \ref{plotSE.cehair} shows the CE- and angle plots for these two countries.
-In the CE plot, we now find the cumulative eigenvalue curve to be just within the acceptance region of the null-hypothesis.
-This is also reflected by the two tests, which now produce $p$-values of $p_\text{KS} = 0.17$ and $p_\text{CvM} = 0.11$, respectively.
-Thus, it is not unreasonable to assume equal variance components in the two datasets.
-
-% but not with overwhelming evidence.
-% Bo: absence of proof \neq proof of absence.
-
-The angle plot in Figure \ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent.
-This implies that if one already has e.g.
-the information held in the first PC from the Danish data, this information is in itself mostly sufficient to describe the first PC of the Swedish data.
-
-Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80 \% of the variance in both datasets), and slight, but increasing, differences in the remaining three components.
-We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25 \% of the variance in both datasets.
-In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g.
-conclude using Table \ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally a lot happier than the least happy people in Sweden.
-A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.
-
-\section{Discussion}
-\label{sec.Discussion}
-Three new tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if two datasets can be combined for analysis were proposed and discussed in this paper.
-They all employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
-% yielding a solution that is largely independent of the sizes  the two datasets.
-Anne: We don't know if that is true.
-
-In an analysis of data from the 2012 version of the European Social Survey (ESS) project, we illustrated that the PCADSC tools help inform analyses of inter-country differences in psychological well-being and happiness.
-It should be noted that even though concerns about pooling data from different countries are not yet very prevalent when it comes to happiness rankings, the subject has been very well scrutinized in the field of international educational rankings.
-For instance, the PISA tests have repeatedly been criticized for not being meaningful objects of international comparisons, especially due to problems with differential item functions \citep{Kankaras2014,Kreiner2014,ZwitserEtAl2017} and translation problems \citep{Asil2016}.
-This highlights lack of data comparability as a empirically relevant and concerning issue that there is no way around: When conducting internal rankings, the issue of data structure heterogeneity should always be addressed.
-
-Though the PCADSC toolbox comprises three diagnostic plots as of now, the PCA decomposition can of course be utilized further to bring about new PCADSC plots that illustrate more aspects of data structure differences.
-For instance, the scree plot (as introduced by \cite{Cattell1966}), which plots the ordered eigenvalues against their component numbers, could easily be recreated as a PCADSC tool by creating scree plots for both data subsets in the same coordinate system, yielding a simple graphical comparison.
-The resampling approach used in the CE plot could then be adopted, thus annotating the scree plots with resampled curves.
-This general approach to model-fit evaluation is very helpful in combination with permutation tests, %using cumulative residuals,
-as inspired by Lin et al. \citeyearpar{LinEtAl2002}. Similarly, the angle plot could also be annotated with p-values from a permutation test based on appropriate test statistics.
-
-The proposed methods also have limitations.
-First of all, as PCA is performed after standardization of the variables, the PCADSC methods cannot detect differences in mean values nor in marginal variances - this information is thrown away as the very first step.
-However, such differences are of a fundamentally different nature than those we have discussed so far.
-Differences in mean values are often the main interest of the analysis and should of course not be regarded as a data comparability issue.
-Differences in marginal variances, on the other hand, can pose some modeling challenges, but none that are not generally solvable by use of random effects models.
-We therefore do not consider this property of PCADSC to be a major drawback.
-
-A more prominent limitation is the fact that the PCASDC methods are only valid for variables that have a numerical interpretation.
-In particular, this means that the methods can not be used for data with nominal categorical variables.
-This limitation is inherited from the PCA method and thus, moving beyond it will entail replacing the PCA framework with a more general one.
-Multiple correspondence analysis (MCA) is often suggested as a categorical extension of PCA \citep{Abdi2010} and including MCA in the PCADSC tools is thus a natural next step.
-
-Moreover, though we have illustrated the discriminatory power of PCASDC in one data example, of course further evaluation of its performance is needed.  
-While the best feature of PCADSC is perhaps the intuitive, visual nature of the tools, it is also the biggest weakness.
-For statistical methods that result in a number (or a set of numbers), the standard way of evaluating performance is to report on simulation studies, but in the case of statistical methods that report graphics this is not feasible.
-Therefore, a systematic evaluation of the performance of the PCADSC methods is not very straight-forward to design: It really depends on whether or not users can detect actual differences in data structures by using the diagnostic plots.
-We suggest that the variability of the methods could be studied further using \textit{Wally plots} \citep{Ekstrøm2014}, which are learning tools developed for teaching students how to interpret residual plots.
-First in line as a topic for more thorough investigations of performance is the sensitivity towards the sample sizes of the two datasets, $n_1$ and $n_2$.
-
-All in all, PCADSC represents a first step towards addressing a known issue that has otherwise been met with unsatisfactory ad-hoc methods whose assumptions rest upon the very hypothesis they are testing.
-The PCADSC tools perform quite well in the data example provided here, yielding conclusions that are in accordance with well-established theories and they are simple to use and test further using the PCADSC \texttt{R} package.
-
-%\section{Concluding Remarks}
-%\label{sec:conclusion}
-
-\bibliographystyle{apa}
-\bibliography{bib}
-
-\appendix
-
-%\newpage
-%\section{Supplementary tables}
-\begin{table}[h]
+\section*{S2 Table: ESS psychological wellbeing scale definitions}
+\begin{table}[H]
 \centering
@@
 
+\begin{thebibliography}{10}
+
+\bibitem{ESStopline5}
+Jeffrey K, Abdallah S, Quick A.
+\newblock {Europeans' Personal and Social Wellbeing: Topline Results from Round
+  6 of the European Social Survey}.
+\newblock ESS Topline Results (Series 5). 2015;.
+
+\end{thebibliography}
+
 \end{document}

### 2018-03-08 article/Old versions/article_v0.01.tex
Redid folder structure + added first print/summary functions

--- 2017-03-15 article/article_v0.01.tex
@@
 
-\section{Data example stuff}
+\section{European differences in psychological well-being: A data example}
 \label{sec:dataexample}
-We will now turn to a concrete data example in order to illustrate the possibilities of each of the methods presented above \hl{more than just PCADSC?}.
-We use data from the 2012 version of the European Social Survey (ESS) project, a very large dataset that is freely available online at \url{www.europeansocialsurvey.org}.
+We will now turn to a concrete data example in order to illustrate the capabilities of the methods presented above.
+We use data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
+This investigation is motivated by an increasingly popular new tendency to publish miscellaneous rankings of countries in fields as different as educational quality (\hl{ref: PISA stuff}), \hl{XXX} and citizen happiness (\hl{ref OECD?}).
+From a methodological point of view, such international rankings are very concerning, as they rely on the fundamental assumption  that the measured concepts are inherently the same across countries.
+The PCADSC tools qualify as a suite of methods for exploring the validity of this assumption empirically.
+
+In the rankings of happiness, Denmark has repeatedly been awarded with the title of "happiest country" (\hl{ref?}.
+However, little critical discussion has been aimed at whether or not this title is in any way meaningful.
+Does there exist such a thing as a universal, internationally valid concept of happiness?
+Or do different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socioeconomic settings?
+This is in fact a question concerning comparability of data structures.
+If two countries differ e.g.
+in terms of how social networks are typically build and structured, with one emphasizing family relations and the other mostly focusing on other social relations, having a weak family connection does not have the same implications in the first country as it does in the second one.
+More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being.
+The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures.
+And therefore, comparing the two countries in these measures is not a meaningful endeavor.
+
+In this section we will use the PCADSC tools to unveil such differences between Denmark and Bulgaria.
+These two countries have previously been highlighted to be very different in terms of what defines happiness (\hl{REF: ESS6 Topline Results Series 5}).
+This might be the result of large, cultural and socioeconomic differences between the two countries.
+Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.
+
+We also compare the Danish data with Swedish data in order to illustrate that the PCADSC tools actually do have some discriminatory power.
+Denmark and Sweden are both Scandinavian countries and are often deemed very similar in terms of culture and history.
+Therefore, we expect fundamental concepts such as psychological well-being to similar across these two countries.
+
+%As Denmark has repeatedly been awarded with the title of "the happiest country", we have included this country
+
+%Specifically, we take on a critical perspective on
+
+%, a very large dataset that is freely available online at \url{www.europeansocialsurvey.org}.
 As the name suggests, the data comes from a survey that was conducted primarily in Europe aiming to collect information about the social conditions of the citizens \hl{reference?}.

### 2018-03-08 article/Old versions/article_v1.tex
Redid folder structure + added first print/summary functions

--- 2017-05-17 article/article_v0.9.5.tex
@@
 \renewcommand{\baselinestretch}{2}
+
+\usepackage{tikz}
+\usetikzlibrary{arrows}
+\usetikzlibrary{shapes, positioning, calc}
 
@@
 
-%\RepeatTitle{Exploratory data structure comparisons: Three new tools based on Principal Component Analysis}
+%\Large{\textsc{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}}
 
@@
 \subsection{Using PCA for data structure comparisons}
+\begin{figure}
+\center 
+Model 1: \medskip \\
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=2.5cm,
+  thick,main node/.style={rectangle, rounded corners,draw,font=\sffamily\small\normalfont},
+  sq node/.style={rectangle,draw,font=\sffamily\small\normalfont}]
+  \node[sq node] (1) {$v_1$};
+  \node[sq node, right of = 1] (2) {$v_2$};
+  \node[sq node, right of = 2] (3) {$v_3$};
+  \node[sq node, right of = 3] (4) {$v_4$};
+  \node[sq node, right of = 4] (5) {$v_5$};
+  \node[sq node, right of = 5] (6) {$v_6$};
+  \node[main node, above right=1cm of 3] (7) {$X_1$};
+  \node[main node, above right=1cm of 5] (8) {$X_2$};
+  \path[every node/.style={font=\sffamily\small}]
+    (7) edge node  {} (3)
+    (7) edge node {} (4)
+    (8) edge node {} (5)
+    (8) edge node {} (6);
+ \end{tikzpicture}
+\\
+\bigskip
+Model 2: \medskip \\
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=2.5cm,
+  thick,main node/.style={rectangle, rounded corners,draw,font=\sffamily\small\normalfont},
+  sq node/.style={rectangle,draw,font=\sffamily\small\normalfont}]
+  \node[sq node] (1) {$v_1$};
+  \node[sq node, right of = 1] (2) {$v_2$};
+  \node[sq node, right of = 2] (3) {$v_3$};
+  \node[sq node, right of = 3] (4) {$v_4$};
+  \node[sq node, right of = 4] (5) {$v_5$};
+  \node[sq node, right of = 5] (6) {$v_6$};
+  \node[main node, above=1cm of 2] (7) {$Y_1$};
+  \node[main node, above=1cm of 5] (8) {$Y_2$};
+  \path[every node/.style={font=\sffamily\small}]
+    (7) edge node  {} (1)
+    (7) edge node {} (2)
+    (7) edge node {} (3)
+    (8) edge node {} (4)
+    (8) edge node {} (5)
+    (8) edge node {} (6);
+ \end{tikzpicture}
+\caption{Graphical representations of the two models that are used for simulating example data for the PCADSC plots.
+Square nodes represent observed variables, while rounded nodes represent latent variables.
+Arrows are used to illustrate causal links, and thus we e.g.
+find in \textit{Model 1} that the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable, namely $X_1$.
+Further details on the models and the simulated data are available in Appendix \ref{appendix.simData}, including covariance matrices for the two models.}
+\label{fig.simGraphs}
+\end{figure}
+
 All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets.
@@
 
+While we present the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions.
+These plots are available in Figures \ref{plot.simCE}, \ref{plot.simAngle} and \ref{plot.simChroma} for the CE-, angle- and chroma plots, respectively, and they are based on two simulated datasets:
+\begin{description}
+\item[Dataset A:] This dataset contains 1000 independent simulations from the same underlying model, namely \textit{model 1} from Figure \ref{fig.simGraphs}.
+The observations are randomly divided into two groups.
+\item[Dataset B:] This dataset contains 500 independent simulations from \textit{model 1} from Figure \ref{fig.simGraphs} and 500 independent simulations from \textit{model 2} in the same figure.
+The observations are naturally divided into groups corresponding to which model they were simulated from.
+\end{description} 
+Further details about the data simulations are available in Appendix \ref{appendix.simData}.
+In dataset A, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures.
+In dataset B, on the other hand, there are two fundamentally different, underlying datastructures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.
+
 \subsubsection{The cumulative eigenvalue plot}
+\begin{figure}
+\centerline{\includegraphics[scale = 0.65]{simCE2.pdf}}
+\centerline{\includegraphics[scale = 0.65]{simCE1.pdf}}
+\caption{CE plots for the two simulated datasets.
+Dataset A, which includes no differences in data structures, is found in the top panel, and dataset B, where the observations stem from two different underlying models, is presented in the bottom panel.
+Both plots are annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
+The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.
+In the top panel, we see a cumulative eigenvalue curve that falls well within the confidence region, thereby suggesting no difference in eigenvalues (as expected).
+This also agrees with the large $p$-values and the fact that only a single underlying model was used for generating this data.
+In the bottom panel, the conclusion is reversed, which is also in accordance with the nature of the simulated data, which contains two distinct groups.}
+\label{plot.simCE}
+\end{figure}
+
 The cumulative eigenvalue (CE) plot compares the variation components, i.e.
@@
 Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does.
+Two examples of cumulative eigenvalue plots are available in Figure \ref{plot.simCE}.
 
@@
 \subsubsection{The angle plot}
+\begin{figure}
+\centerline{\includegraphics[scale = 0.65]{simAngle2.pdf}}
+\centerline{\includegraphics[scale = 0.65]{simAngle1.pdf}}
+\caption{Angle plots for the two simulated datasets, with dataset A in the top panel and dataset B in the bottom panel.
+The blue show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.
+For dataset A, we find very short off-diagonal arrows, suggesting that the two PCA decompositions agree on the relative placement of the information in  the data among the 6 PCs.
+For dataset B, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}
+\label{plot.simAngle}
+\end{figure}
+
 This plot simultaneously compares the variation components and the loadings.
@@
 Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.
+Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data.
 
@@
 \subsubsection{The chroma plot}
+\begin{figure}
+\centerline{\includegraphics[scale = 0.65]{simChroma2.pdf}}
+\centerline{\includegraphics[scale = 0.65]{simChroma1.pdf}}
+\caption{Chroma plots for the two simulated datasets, with dataset A in the top and dataset B in the bottom.
+Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance ($\tilde\sigma_{ij}^c$).
+For dataset A, we see very similar visual patterns across the two panels, corresponding well with the fact that both groups are drawn from the same underlying model.
+For dataset B, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups.}
+\label{plot.simChroma}
+\end{figure}
+
 This plot compares the loadings of the two datasets.
@@
 In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$.
+In Figure \ref{plot.simChroma}, two examples of chroma plots are available.
 
@@
 
-Moreover, though we have illustrated the discriminatory power of PCASDC in one data example, of course further evaluation of its performance is needed.  
+Moreover, though we have illustrated the discriminatory power of PCASDC in one simulated- and on real-life data example, of course further evaluation of its performance is needed.  
 While the best feature of PCADSC is perhaps the intuitive, visual nature of the tools, it is also the biggest weakness.
@@
 \appendix
+
+\section{Details about the simulated data}
+\label{appendix.simData}
+In Section \ref{sec:pcadscintro}, we used simulated data for presenting the PCADSC plots.
+Dataset A consisted of 1000 independent simulated realizations from the $N(0, \Sigma_1)$-distribution, while dataset B consisted of 500 independent realizations from $N(0, \Sigma_1)$ and 500 independent realizations from $N(0, \Sigma_2)$.
+Dataset A was furthermore appended with a grouping variable, randomly dividing the observations into two groups.
+Dataset B also included a grouping variable, and this variable contained information about which of the two normal distributions each observation was simulated from.
+The covariance matrices, $\Sigma_1$ and $\Sigma_2$, were defined by
+
+$$\Sigma_1 = \begin{pmatrix}
+    1.0  &  0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
+ 0.0  &  1.0 & 0.0 & 0.0&  0.0 & 0.0 \\
+ 0.0   & 0.0 &  1.0 & 0.7 & 0.0 & 0.0 \\
+ 0.0 &  0.0 & 0.7 & 1.0 & 0.0 & 0.0 \\
+0.0 &    0.0 &  0.0 & 0.0 & 1.0 & 0.4 \\
+ 0.0 & 0.0 &  0.0 & 0.0 & 0.4 & 1.0
+\end{pmatrix}
+\; \text{ and } \;
+\Sigma_2 = \begin{pmatrix}
+ 1.0 & 0.2 & 0.1 & 0.0 & 0.0 & 0.0 \\
+ 0.2 & 1.0 & 0.1 & 0.0 & 0.0 & 0.0 \\
+ 0.1 & 0.1 & 1.0 & 0.0 & 0.0 & 0.0 \\
+ 0.0 & 0.0 & 0.0 & 1.0 & 0.3 & 0.1 \\
+ 0.0 & 0.0 & 0.0 & 0.3 & 1.0 & 0.2 \\
+ 0.0 & 0.0 & 0.0 & 0.1 & 0.2 & 1.0
+\end{pmatrix}
+$$,
+
+respectively.
 

### 2018-03-08 article/Old versions/slides.tex
Redid folder structure + added first print/summary functions

\documentclass[a4]{beamer}
%\usepackage[pdftex]{graphicx}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
%\usepackage{mathtools}
%\usepackage{amssymb}
\usepackage{underscore}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage{multirow}
%\usepackage{wrapfig}
%\usepackage[skip=1.5pt,font=scriptsize]{caption}
%\usepackage{enumitem}
\usetheme{Frankfurt}
\usecolortheme{seahorse}
\setbeamertemplate{frametitle}[default][colsep=-4bp,rounded=false,shadow=false]
\setbeamertemplate{title page}[default][colsep=-4bp,rounded=false,shadow=false]
\addtobeamertemplate{navigation symbols}{}{ \hspace{1em}    \usebeamerfont{footline}%
    \insertframenumber / \inserttotalframenumber }
\setbeamertemplate{itemize items}[circle]

\newcommand{\bofr}{\mathcal{B}(\mathbb{R})}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\bs}{\backslash}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\BB}{\mathbb{B}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\FF}{\mathbb{F}}
\newcommand{\la}{\lambda}
\newcommand{\coola}{\mathcal{A}}
\newcommand{\coolm}{\mathcal{M}}
\newcommand{\cooll}{\mathcal{L}}
\newcommand{\ipl}{\langle}
\newcommand{\ipr}{\rangle}
\newcommand{\limn}{\lim_{n \to \infty}}
\newcommand{\eq}[1]{\begin{align*} #1 \end{align*}}
\newcommand{\lr}{\Leftrightarrow}
\newcommand{\set}[1]{\left\lbrace #1 \right\rbrace}
\renewcommand{\d}{\,\text{d}}
\newcommand{\pt}[1]{\left(#1\right)}
\newcommand{\fork}[1]{\left\{\begin{array}{lr}#1\end{array}\right.} % gaffel-funktion
\newcommand{\tuad}[1]{\qquad\ \text{#1}\ }
\newcommand{\ND}{\mathcal{N}}
\newcommand{\Perp}{\perp \! \! \! \perp}
\newcommand{\ddu}{\frac{\partial}{\partial u}}
\newcommand{\ddv}{\frac{\partial}{\partial v}}
\newcommand{\ddx}{\frac{\partial}{\partial x}}
\newcommand{\dds}{\frac{\partial}{\partial s}}
\newcommand{\ddt}{\frac{\partial}{\partial t}}
\newcommand{\dd}{\partial}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\var}[1]{\textsc{\lowercase{#1}}}

%\newenvironment{myindentpar}[1]%
%	{\begin{list}{}%
%             {\setlength{\leftmargin}{#1}}%
%             \item[]%
%     }
%    {\end{list}}

\DeclareMathOperator*{\plimo}{plim}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\E}{E}
\DeclareMathOperator*{\V}{V}
\DeclareMathOperator*{\T}{T}
\newcommand{\plim}{\plimo_{n \to \infty}}
\usepackage{wallpaper}
%\definecolor{blue(pigment)}{rgb}{0.2, 0.2, 0.6}
%\definecolor{darkcerulean}{rgb}{0.03, 0.27, 0.49}
%\setbeamercolor{frametitle}{bg=blue(pigment)}

%\begin{center}
%\includegraphics[scale=0.3]{residualplot_logengel.jpeg} 
%\end{center}

\title{Exploratory data structure comparisons}
\subtitle{A few tools based on PCA}
\author{Anne H. Petersen, Bo Markussen \& Karl Bang Christensen}
\date{March 15, 2017}

\begin{document}
\maketitle

\section{Set-up}

\subsection{Problem}
\begin{frame}
\frametitle{Problem}
\begin{itemize}
\item Let $X_k$ be a $n_k \times d$ matrix of data for $k \in \{1, 2\}$ with (the same) ordinal or numeric variables for both $k$. 
\item We ask: Are the structures of $X_1$ and $X_2$ similar? Can we e.g. combine the two into one dataset $X = \begin{pmatrix} X_1 \\ X_2 \end{pmatrix}$ with $n = n_1 + n_2$ observations of the $d$ variables without problems?
\item Example: $X_1$ stems from a survey administered by telephone, $X_2$ stems from the same survey administered as a web questionnaire. Can we analyze the "full" dataset jointly?
\item Example: $X_1$ represents data from a study conducted in Denmark, $X_2$ represents the same measurements from a study in Bulgaria. Can we combine the two?
\end{itemize}
\end{frame}

\subsection{Requirements}
\begin{frame}
\frametitle{Requirements for a candidate method}
\begin{itemize}
\item New origin stories of datasets add a new challenge to this question: Increasingly often, data is \textit{not} collected with a specific endpoint in mind (e.g. PISA, the European Social Survey (ESS), traffic data from websites, ...)
	\begin{itemize}
		\item[$\Rightarrow$] We wish to answer the question \textit{before} specifying what we are going to use the data for
	\end{itemize}
\item In particular, we do not want to specify a model (yet) 
	\begin{itemize}
		\item[$\Rightarrow$] Cannot use e.g. IRT methods, regression methods  
	\end{itemize}
\item We do not want just a test, but instead tools for intuitive, exploratory investigations that might lead to insights as to \textit{why} the differences occur 
\end{itemize}
\end{frame}

\subsection{Strategy}
\begin{frame}
\frametitle{Solution strategy}
\begin{itemize}
\item Focus on the covariance structures:
\begin{itemize}
\item Let $S_1 = \tilde{\V}(X_1)$ and $S_2 = \tilde{\V}(X_2)$ denote the empirical covariance matrices of the two datasets
\item $S_k$ describes the interplay between variables. And for Gaussian (zero-mean) variables: A sufficient statistic for the joint distribution.
\item In principle, we could compare $S_1$ and $S_2$ entry by entry, but this strategy does not scale well
\end{itemize}
\item Deconstruct each covariance matrix such that we can deal with the most informative components first and ignore noise
\begin{itemize}
\item[$\rightarrow$] Use Principal Component Analysis (PCA)
\end{itemize}
\end{itemize}
\end{frame}

\section{PCA}

\begin{frame}
\frametitle{Principal component analysis: A greedy interpretation}
\begin{itemize}
\item Let $U \subset \RR^d$ be the set of unit vectors of dimension $d$. For $j \in 1, ..., d$, define: 
\begin{itemize}
\item[] The $j$th \textit{loading vector}: \onslide<2->{\blue{(eigenvectors)}}
$$\eta_j := \text{argmax}_{u \in U: u \perp \hat{K}_{j-1}} \tilde{\V}(u^\top X^\top)$$
\item[] The $j$th \textit{variance component}: \onslide<2->{\blue{(eigenvalues)}}
$$\lambda_j := \tilde{\V}(\eta_j^\top X^\top)$$
\end{itemize}
where $K_0 = \emptyset$ and $K_j = \text{span}\{\eta_1, ..., \eta_j\}$. 
\item Now, $\eta_1$ is the linear transformation of dimension $d \times 1$ of the data that explains the largest possible amount of the variance, and this amount is $\frac{\lambda_1}{tr{S}}$.
\item It holds that $S = \sum_{j=1}^d \lambda_j \eta_j \eta_j^\top$
\end{itemize}
\end{frame}

\section{PCADSC tools}

\subsection{PCADSC}
\begin{frame}
\frametitle{Principle Component Analysis-based Data Structure Comparisons (PCADSC)}
\begin{itemize}
\item PCADSC: Conduct PCA decomposition on each of the (standardized) datasets $X_1$ and $X_2$ and compare the results
\item We provide three tools for visualizing the results of PCA in order to compare data structures:
	\begin{itemize}
		\item The cumulated eigenvalue (CE) plot \onslide<2->{\blue{- compare eigenvalues}}
		\item The hair plot \onslide<3->{\blue{- explain $S_1$ from $S_2$ (and $S_2$ from $S_1$)}}
		\item The pancake plot \onslide<4->{\blue{- compare loadings}}
	\end{itemize}
\item Available in \texttt{R}-package at \url{www.github.com/AnnePetersen1/PCADSC}
\onslide<5->{\item First step of all procedures:
	\begin{itemize}
		\item[1.]  For $k \in \{1, 2\}$, standardize $X_k$ and perform PCA, thereby obtaining $\eta^k_1, ..., \eta^k_d$ and $\lambda^k_1, ..., \lambda^k_d$
	\end{itemize}}  
\end{itemize}
\end{frame}

\subsection{The CE plot}
\begin{frame}
\frametitle{The CE plot}
\begin{itemize}
\item Let $\lambda_1, ..., \lambda_d$ be the eigenvalues of the covariance matrix of the combined dataset, $X$. 
\item Draw a piecewise linear curve through the points
\begin{align*}
&(0,0), 
(\lambda_1,\lambda^1_1-\lambda^2_1), 
(\lambda_1 + \lambda_2,\lambda^1_1+\lambda^1_2-\lambda^2_1-\lambda^2_2), \\ 
&\ldots, \bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda^1_j - \lambda^2_j \bigg).
\end{align*}
\item Get an idea of the variability of the results using repeated random splits:
\begin{itemize}
\item E.g. 10000 times, divide $X$ randomly into two subsets of size $n_1$ and $n_2$, respectively and perform the steps from above
\item Draw a shaded region representing a "bootstrapped" pointwise confidence interval obtained in this way
\item Draw a subset of the random curves (e.g. 20) to illustrate how they vary
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The CE plot}
\begin{figure}
\centering
    \includegraphics[scale = 0.3]{essDKSEce.pdf}
\end{figure}
\small
Annotations: $p$-values from two test statistics:
\begin{itemize}
\item Kolmogorov-Smirnov: $\max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda^1_j -  \lambda^2_j \bigg\rvert$
\item Cramér-von Mises: $\sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda^1_j - \lambda^2_j  \bigg)^2$
\end{itemize}
\end{frame}

\subsection{The hair plot}
\begin{frame}
\frametitle{The hair plot}
   \begin{itemize}
   \item Let $\lambda_\text{max} = \max\{\lambda^1_1, \lambda^2_1\}$. 
   \item Define $\mu_{jk} = \sqrt{\frac{\lambda^1_k}{\lambda_\text{max}}} \left|(\eta^1_k)^\top \eta^2_j \right|$, $\nu_{jk} = \sqrt{\frac{\lambda^2_j}{\lambda_\text{max}}} \left|(\eta^2_j)^\top \eta^1_k \right|$,
   $\theta_{jk} = \arccos \left( \left|(\eta^1_k)^\top \eta^2_j \right| \right)$
   \item In the $jk$th position in a $d \times d$ grid, draw two arrows (hairs): 
   \begin{itemize}
   	\item A blue arrow with length $\mu_{jk}$ drawn at an angle of $\theta_{jk}$ counter-clockwise from the diagonal
   	\item A red arrow with length $\mu_{jk}$ drawn at an angle of $\theta_{jk}$ clockwise from the diagonal
   \end{itemize} 
  \item Note: If $\eta^1_k \perp \eta^2_j$, then $\mu_{jk} = \nu_{jk} =  0$. If $\eta^1_k = \eta^2_j$, then $\theta_{jk} = \arccos(1) = 0$ (due to unit length). 
  \item Thus: Identical structures imply zero-length arrows for $j \neq k$ and identical red and blue arrows at the diagonal for $j = k$.  
   \end{itemize}
\end{frame}

\begin{frame}
\frametitle{The hair plot}
\begin{figure}
\centering
    \includegraphics[scale = 0.4]{essDKBGhair.pdf}
\end{figure}
\end{frame}

\subsection{The pancake plot}
\begin{frame}
\frametitle{The pancake plot}
For $k \in \{1, 2\}$, do:
   \begin{itemize}
   \item[2.] For $i \in 1, ..., d$, normalize $\eta^k_i$ and add a bar to the plot consisting of differently colored "pancakes" whose widths correspond to their standardized loading weights. 
   \item[3.] Annotate each bar with the amount of (cumulated) explained variance when using the information from this and the previous components, i.e. 
   $$\frac{\sum_{j = 1}^i \eta^k_j}{\sum_{j=1}^d \eta^k_j}$$
   % \grey{\quad \left(\text{or } \frac{ \eta^k_j}{\sum_{j=1}^d \eta^k_j} \right) } 
   \end{itemize}   
\end{frame}

\begin{frame}
\frametitle{The pancake plot}
    \includegraphics[scale = 0.5]{essDKBGpancake.pdf}
\end{frame}

\section{Data example}
\subsection{setup}
\begin{frame}
\frametitle{Data example - set-up}
\begin{itemize}
\item Data on psychological well-being from the 2012 version of the European Social Survey (ESS)
\item We use 35 items for producing 6 distinct psychological well-being scales
\item ESS report: Bulgaria (BG) and Denmark (DK) are particularly different in the interplay between different aspects of psychological well-being (evaluated at aggregated country-level)
\item Postulate: Denmark and Sweden (SE) are quite similar in what defines personal happiness and psychological well-being
\item Strategy: Run all three methods on the DK vs. BG and DK vs. SE comparisons and expect to find \textit{different} data structures for DK vs. BG and \textit{similar} structures for DK vs. SE
\item Only use complete cases. This results in $n_\text{DK} = 1498$, $n_\text{BG} = 1798$ and $n_\text{SE} = 1736$ observations, respectively. 
\end{itemize}
\end{frame}

\subsection{DK vs. BG}
\begin{frame}
\frametitle{DK vs. BG: CE plot}
\includegraphics[scale = 0.5]{essDKBGce.pdf}
\end{frame}

\begin{frame}
\frametitle{DK vs. BG: hair plot}
\includegraphics[scale = 0.5]{essDKBGhair.pdf}
\end{frame}

\begin{frame}
\frametitle{DK vs. BG: Pancake plot}
\includegraphics[scale = 0.5]{essDKBGpancake234.pdf}
\end{frame}

\begin{frame}
\frametitle{DK vs. BG: Pancake Wally plot}
\includegraphics[scale = 0.4]{essDKBGWallyPCADSC234.pdf}
\end{frame}

\subsection{DK vs. SE}
\begin{frame}
\frametitle{DK vs. SE: CE plot}
\includegraphics[scale = 0.5]{essDKSEce.pdf}
\end{frame}

\begin{frame}
\frametitle{DK vs. SE: hair plot}
\includegraphics[scale = 0.5]{essDKSEhair.pdf}
\end{frame}

\begin{frame}
\frametitle{DK vs. SE: Pancake plot}
\includegraphics[scale = 0.5]{essDKSEpancake.pdf}
\end{frame}

\begin{frame}
\frametitle{DK vs. SE: Pancake Wally plot}
\includegraphics[scale = 0.4]{essDKSEWallyPCADSC.pdf}
\end{frame}

\section{Next steps}
\begin{frame}
\frametitle{Ideas for next steps}
\begin{itemize}
\item Investigate sensitivity towards the sample sizes $n_1$ and $n_2$
\item Limitations: What sorts of problems can never be found using PCADSC? 
\begin{itemize}
\item Differences in scaling, as we standardize all variables
\item More?
\end{itemize}
\item Interpretation for binary variables?
\item Any meaningful way to allow for nominal, categorical variables?
\item ... ?
\end{itemize}

\end{frame}

\end{document}

### 2018-03-08 ESS/ess_notes.R

@@ -214,6 +214,10 @@ ggsave(plot = fig6_chroma, file = "./article/PlosOne/Fig6.eps",
        device = "eps", dpi = 400, width = 7, height = 4)
 
 
+ggsave(plot = fig5_CE, file = "P:/PCADSC/CSP/CEBGDK.pdf", width = 7, height = 5)
+ggsave(plot = fig5_angle, file = "P:/PCADSC/CSP/angleBGDK.pdf", width = 7, height = 5)
+ggsave(plot = fig6_chroma, file = "P:/PCADSC/CSP/chromaBGDK.pdf", width = 7, height = 5)
+
 #DK/SE
 set.seed(1234)
 a1 <- PCADSC(data_SEDK, "cntry",  c("EvaluativeWellbeing",
@@ -235,3 +239,7 @@ ggsave(plot = grid.arrange(fig7_CE, fig7_angle, nrow = 2), file = "./article/Plo
 
 ggsave(plot = fig8_chroma, file = "./article/PlosOne/Fig8.eps",
        device = "eps", dpi = 400, width = 7, height = 4)
+
+ggsave(plot = fig7_CE, file = "P:/PCADSC/CSP/CESEDK.pdf", width = 7, height = 5)
+ggsave(plot = fig7_angle, file = "P:/PCADSC/CSP/angleSEDK.pdf", width = 7, height = 5)
+ggsave(plot = fig8_chroma, file = "P:/PCADSC/CSP/chromaSEDK.pdf", width = 7, height = 5)

### 2018-03-08 ESS/sim_notes.R

@@ -63,11 +63,26 @@ ggsave(plot = grid.arrange(CE2, CE1, nrow = 2), file = "./article/PlosOne/Fig2.e
        device = "eps", dpi = 400, width = 7, height = 8)
 
 
+
 ggsave(plot = grid.arrange(angle2, angle1, nrow = 2), file = "./article/PlosOne/Fig3.eps",
        device = "eps", dpi = 400, width = 7, height = 8)
 
 ggsave(plot = grid.arrange(chroma2, chroma1, nrow = 2), file = "./article/PlosOne/Fig4.eps",
        device = "eps", dpi = 400, width = 7, height = 8)
+
+
+CE2b <- CE2 + ggtitle("Simulated data A")
+ggsave(plot = CE2b, file = "P:/PCADSC/CSP/CEsim.pdf",
+       width = 8, height = 6)
+
+
+angle1b <- angle1 + ggtitle("Simulated data B")
+ggsave(plot = angle1b, file = "P:/PCADSC/CSP/anglesim.pdf",
+       width = 8, height = 6)
+
+chroma1b <- chroma1 + ggtitle("Simulated data B")
+ggsave(plot = chroma1b, file = "P:/PCADSC/CSP/chromasim.pdf",
+       width = 8, height = 6)
 
 
 #ggsave(plot = angle1, file = "./article/simAngle1.pdf",

### 2018-03-08 R/print.R

@@ -1,0 +1,17 @@
+#'@import pander pander
+#'@export
+print.CEInfo <- function(x) {
+  d <- x$d
+  outtab <- matrix(round(c(x$xVals, x$y.obs), 2), 2, d + 1, byrow = TRUE,
+                   dimnames = c(list(c("Joint cumulative eigenvalues",
+                                       "Cumulative difference in eigenvalues"),
+                                     paste("PCs: ", 0:d, sep = ""))))
+  cat(pander(outtab, justify = paste(c("l", rep("r", d + 1)), collapse = ""),
+      keep.trailing.zeros = TRUE))
+}
+
+
+print.angleInfo <- function(x) {
+
+
+}

### 2018-03-08 supplementary_code/benchmarking.R

@@ -1,0 +1,155 @@
+#Angle plot
+microbenchmark({
+  data(iris)
+  iris$group <- "setosa"
+  iris$group[iris$Species != "setosa"] <- "non-setosa"
+
+  irisPCADSC <- PCADSC(iris, "group", setdiff(names(iris), c("group", "Species")))
+  irisPCADSC2 <- PCADSC(iris, "group", setdiff(names(iris), c("group", "Species")), doAngle = FALSE)
+  irisPCADSC2 <- doAngle(irisPCADSC2)
+  anglePlot(irisPCADSC)
+  anglePlot(irisPCADSC2)
+  }, times = 10)
+#min       lq     mean   median       uq      max neval
+#4.061592 4.085913 4.364346 4.210754 4.329815 5.858282    10
+
+
+#CE plot
+microbenchmark({
+  data(iris)
+  iris$group <- "setosa"
+  iris$group[iris$Species != "setosa"] <- "non-setosa"
+
+  irisPCADSC <- PCADSC(iris, "group", setdiff(names(iris), c("group", "Species")))
+  irisPCADSC2 <- PCADSC(iris, "group", setdiff(names(iris), c("group", "Species")),
+                        doCE = FALSE)
+  irisPCADSC2 <- doCE(irisPCADSC2)
+
+  CEPlot(irisPCADSC)
+  CEPlot(irisPCADSC2)
+}, times = 10)
+#min      lq     mean   median       uq      max neval
+#3.612281 3.70354 3.852662 3.841499 3.892452 4.158875    10
+
+#Chroma plot
+microbenchmark({
+  data(iris)
+  iris$group <- "setosa"
+  iris$group[iris$Species != "setosa"] <- "non-setosa"
+
+  irisPCADSC <- PCADSC(iris, "group", setdiff(names(iris), c("group", "Species")))
+  irisPCADSC2 <- PCADSC(iris, "group", setdiff(names(iris), c("group", "Species")),
+                        doChroma = FALSE)
+  irisPCADSC2 <- doChroma(irisPCADSC2)
+
+  chromaPlot(irisPCADSC)
+  chromaPlot(irisPCADSC)
+
+  chromaPlot(irisPCADSC, splitLabels = list("non-setosa" = "Not Setosa",
+                                            "setosa" = "Setosa"))
+
+  chromaPlot(irisPCADSC, useComps = c(1,4), varAnnotation = "no")
+  chromaPlot(irisPCADSC, cvCO = 0.8)
[…]

### 2018-03-13 article/Old versions/MPS/coverLetter.tex
Added MPS submission

--- 2017-06-22 article/PlosOne/coverLetter.tex
@@
 \renewcommand{\baselinestretch}{1.1}
-
+\setlength{\parindent}{0pt}
 %\usepackage{arev}
@@
 
-\medskip
+\bigskip
 
-Please consider the attached research paper \emph{``Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis''} for publication in PLoS One.
-As Academic Editor to handle the manuscript we suggest Ulrich S.\ Tran, Jakob Pietschnig, or Christopher M.\ Danforth.
+Please consider the attached research paper \emph{``Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis''} for publication in Mathematical Population Studies.
 
-\medskip
+\bigskip
 
-Data comparability is a recurring topic in statistics, in particular in psychometrics and social sciences.
-An example of this is the \emph{PISA global education rankings}, where  it a priori is not clear whether it makes sense to compare children's reading capabilities across different cultures and languages.
+Data comparability is a recurring topic in statistics, e.g.
+when multi-center studies are conducted to compare populations across countries.
 There exist methods to investigate data comparability within the framework of statistical models, e.g.\ \emph{differential item functioning} within \emph{item response theory}.
 However, without assuming a statistical model, practitioners often report simple data summaries in the absense of a systematic framework for understanding and describing data structure differences.
-In the submitted paper, we propose three graphical tool that can be used in an initial exploratory comparison of the data structure of two multivariate datasets prior to the formulation of a statistical model.
+In the submitted paper, we propose three graphical tools that can be used in an exploratory comparison of the data structure of two multivariate datasets prior to the formulation of a statistical model.
 The tools are implemented in an R package publicly available on CRAN.
@@
 Using the graphical tools we are able to visualize a difference between the measured conception of happiness in Denmark and Bulgaria, hence verifying the difficulty in transferring the concept of happiness across countries, as also found by other authors.
-On the other hand, the graphical tools also find that Denmark and Sweden have comparable data structures in accordance with the understanding of similarity within the Nordic countries.
+On the other hand, the graphical tools also find that Denmark and Sweden have comparable data structures in accordance with the understanding of similarity within the Scandinavian countries.
 
-\medskip
+\bigskip
 
@@
 
+\bigskip\bigskip
+
+Yours sincerely,
+
 \bigskip
-
-Yours Sincerely,
-
-\bigskip\bigskip
 

### 2018-03-13 article/Old versions/MPS/titlepage.tex
Added MPS submission

--- 2017-04-25 article/article_v0.9.5.tex
@@
-\documentclass[titlepage,11pt,twoside]{article}
+\documentclass[titlepage,12pt,twoside]{article}
 \usepackage[dvips]{graphicx}
@@
 \usepackage[myheadings]{fullpage}
-\usepackage{pmetrika}
-%\usepackage{pmbib}
 
+\usepackage{authblk}
 \usepackage{color}
@@
 \usepackage{hyperref}
-\usepackage{natbib}
-\usepackage{multirow}
 %\usepackage{colortbl}
-\usepackage[table]{xcolor}
-\definecolor{lightgray}{gray}{0.9}
 \usepackage[T1]{fontenc}
@@
 \usepackage{caption,setspace}
-\captionsetup{font={footnotesize,stretch=1.5}}
-\renewcommand{\baselinestretch}{2}
 
@@
 
+\title{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}
+%\shorttitle{PCA-based data comparisons}
+
+\author[1]{Anne Helby Petersen}
+\affil[1]{Department of Public Health, University of Copenhagen}
+\affil{\href{mailto:ahpe@sund.ku.dk}{\nolinkurl{ahpe@sund.ku.dk}} (Corresponding author)}
+
+\author[2]{Bo Markussen}
+\affil[2]{Department of Mathematical Sciences, University of Copenhagen}
+
+\author[2]{Karl Bang Christensen}
+\vspace{\fill}
+
+\setlength{\parindent}{0pt}
+
 \begin{document}
 
-\begin{titlepage}
-\linespacing{1}
+\thispagestyle{empty}
 
-\title{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}
+\centering
 
-\author{Anne H. Petersen}
-\affil{Department of Public Health, University of Copenhagen}
+{\huge{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}}
 
-\author{Bo Markussen}
-\affil{Department of Mathematical Sciences, University of Copenhagen}
+\bigskip 
+Short title: PCA-based data comparisons
+\bigskip\bigskip
 
-\author{Karl Bang Christensen}
-\affil{Department of Public Health, University of Copenhagen}
+Anne Helby Petersen \\
+\href{mailto:ahpe@sund.ku.dk}{\nolinkurl{ahpe@sund.ku.dk}} (Corresponding author)\\
+Department of Public Health, University of Copenhagen
 
-\vspace{\fill}
+\bigskip
 
-\end{titlepage}\vspace*{24pt}
+Bo Markussen \\
+\href{mailto:bomar@math.ku.dk}{\nolinkurl{bomar@math.ku.dk}} \\
+Department of Mathematical Sciences, University of Copenhagen
 
-%\RepeatTitle{Exploratory data structure comparisons: Three new tools based on Principal Component Analysis}
+\bigskip
 
-\begin{center}\vskip3pt
-
-\vspace{32pt}
-
-Abstract\vskip3pt
-
-\end{center}
-
-\begin{abstract}
-\linespacing{1}
-
-In empirical psychometric work, one often encounters data that is somehow divided into two distinct subsets.
-This might be due to multi-center sampling frames, or to variations in instruments, orders of questionnaire item, modes of administration etc.
-Often, one wishes to combine such data into one an conduct only a single analysis, but this is only meaningful if the two subsets are similar in structures, that is, if they measure the same underlying concepts.
-However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc or require making full model assumptions.
-We present a suite of three visual, non-parametric diagnostic tools for investigating differences in data structures for two datasets containing different observations of the same variables, refered to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC).
-This approach requires no directional nor hierarchical assumptions on the data.
-Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets, and by targeting different aspects of the PCA information, the three plots give a thorough and deep understanding of where differences occur and why it might be the case.
-
-The functionality of the PCADSC tools is presented by use of data from the European Social Survey on psycological well-being in three European countries, namely Denmark, Sweden and Bulgaria.
-Here, we find that the concepts of psychological well-being are not the same when comparing Denmark and Bulgaria, and thus a comparison of e.g.
-mean scores in different well-being aspects is not meaningful.
-However, when comparing the two Scandinavian countries, Denmark and Sweden, we find very similar data structures and thus very similar concepts of well-being.
-Therefore, inter-country comparisons are warranted for these two countries.
-
-\begin{keywords}
-multi-center studies, data heterogeneity, data structure comparisons, principal component analysis (PCA), European Social Survey.
-\end{keywords}
-\end{abstract}
-
-\vspace{\fill}\newpage
-
-\section{Introduction}
-\label{sec:introduction}
-
-Data comparability is a classical topic in statistics and a recurring one in psychometrics.
-Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability (and thus collapsibility) needs to be assessed empirically.
-This happens e.g.
-when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration or the order of survey questionnaire items are applied.
-All of these practices are essential parts of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions.
-However, if we do not address the question of collapsibility among the data subsets arriving from e.g.
-different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.
-
-Sophisticated methods for addressing this question are available when we are willing to assume a statistical model.
-But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment %very
-ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a %valid 
-general data structure comparison method, but rather a fitted-model comparison method.
-It addresses the interplay between the model and the data, not the data alone.
-However, useful tools for initial, exploratory investigations of data comparability data that help researchers identify potential problems or challenges early in the study design development are mostly absent.
-What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
-Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables.
-On the other hand, e.g.
-entry-by-entry comparisons of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase.
-
-While the issue is not in any way new, its relative importance is increasing due to new developments in data availability and data collection.
-Classical statistical methods are generally aimed at analyzing data from designed experiments and historically, statistical analyses have been conducted by researchers who knew the design and the source of the dataset well.
-However, the origin stories of datasets have changed over time and today, a lot of data are accumulated without a specific purpose in mind, as data collection and sharing has become easier and more affordable through the development of new technologies.
-This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for %all
-everyone to use.
-However, large scale publicly available datasets, such as the PISA data and European Social Survey (ESS) data, are often used by data analysts that are far removed from the data producers and therefore, problem-specific recommendations about e.g.
-potential instrument-induced challenges in the datasets %are not
-may not be available for these data analysts.
-For this reason, a thorough investigation of data structure comparability becomes a crucial step in any meaningful data analysis when using such open-source data.
-
-The data structure comparison problem is however not limited to the context of new data sharing platforms.
-Even in single-population survey studies, a mixture of modes of administration is often used, e.g.
-mail and telephone, and while this can improve response rates, it can also be quite problematic \citep{Brambilla1987, McHorney1994}, and differences in response behavior can lead to biased results.
-Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful, so the comparability issue is not just a statistical puzzle.
-Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu \citeyearpar{Liu2016}, who warns against the combination of online panel data (i.e.
-pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions).
-
-The comparability issue that is present in all the examples presented so far can be summarized as follows: Assume that we have two datasets with the same variables, but different observations, or alternatively (and equivalently) a single dataset with a subset-inducing variable.
-Assume next that we wish to compare data structures without specifying a model, or even a variable of interest.
-The central question is then whether the two datasets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be somehow be dealt with, either in terms of modeling choices or simply by conducting several separate analyses.
-
-We propose a suite of three new tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC).
-These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
-%This yields a solution that is largely independent of the sizes of the two subsets of data. Anne: Det ved vi da ikke om det gør? Er det ikke netop det, vi siger, at man burde teste som næste step?
-The proposed tools are implemented in the statistical software \texttt{R} in our package, \texttt{PCADSC} \citep{PCADSC}.
-
-This manuscript describes the procedures, including a brief introduction to principal component analysis (PCA), and presents a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey.
-More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is completely ignored e.g.
-when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project \citep{WHR2016}.
-
-\section{PCA-based tools for data structure comparisons}
-\label{sec:pcadscintro}
-As mentioned above, the purpose of PCADSC is to compare overall data structures in two subsets of a dataset.
-But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context.
-One such definition is the structure of the covariance matrix of the dataset.
-If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables.
-%This gives it a very nice interpretation as a measure of the overall structure. If we do not accept 
-And without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data.
-All in all, the empirical covariance matrix is a reasonable place to start looking for differences in \textit{overall data structures}.
-
-A naive approach for data structure comparisons might therefore be to compute the empirical covariance matrices on each of the two data subsets and simply compare these matrices entry-by-entry.
-Though the idea perhaps sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices.
-There is simply too much information to consider at once.
-However, by use of linear algebra, we can decompose and recompose the covariance matrix such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance.
-In particular, we find a new representation of the covariance matrix that makes it possible to gain an overview of the most interesting aspects of the data.
-One such decomposition strategy is known as \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.
-
-At this point a general remark about standardization is in place.
-PCA deconstructs the covariance matrix in components according to the most explained variance.
-This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will always be deemed highly influential, no matter the structure of the data.
-Therefore, the variables are typically standardized prior to performing PCA.
-Note that the covariance matrix of the standardized variables is the same as the correlation matrix of the original variables, so post-standardization PCA simply corresponds to performing data structure comparisons of the correlation matrices rather than the covariance matrices.
-The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation.
-In this paper we will adopt this standardization policy, and in particular, we will allow for differences between scalings of the original, raw variables in the two data subsets.
-
-%\hl{Bo: I næste afsnit bruges stadigvæk ``covariance''. Det er vel ok?} Anne: Synes det fungerer fint.
-
-\subsection{A brief introduction to principal component analysis}
-Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = (\bar{x}_1, ..., \bar{x}_d)^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$.
-We assume in the following that all variables of $X$ have a numerical interpretation, e.g.
-by being continuous or ordinal and categorical.
-For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost.
-More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the original $d$-dimensional one.
-Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
-\begin{equation*}
-\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
-\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
-\end{equation*}
-The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \citep{HastieEtAl2009}.
-More specifically, the rank-q-reconstruction error is attained when we choose
-\begin{equation*}
-\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\},
-\end{equation*}
-where $\eta_1, ..., \eta_q$ are the $q$ first eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$.
-In the PCA framework, we refer to the eigenvectors as \textit{loadings}.
-The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
-\begin{equation*}
-\text{trace}(S) = \sum_{j=1}^d \hat{V}(x_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
-\end{equation*}
-where $x_j \in \RR^n$ 
-%\hl{(Bo: $X_1$ og $X_2$ bruges i en anden betydning i afsnit 2.2. Omformuler?)} Anne: Hov, ja, hermed x_j så det passer med notationen ovenfor.
-denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
-This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
-And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are %different,
-distinct, the representation hereby obtained is a valid object of inference.
-If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
-
-The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
-\begin{align*}
-\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
-\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2,
-\end{align*}
-where the initial subspace is defined as $\hat{K}_0 = \{0\}$.
-It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
-This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$.
-This implies that the loadings and the variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.
-
-\subsection{Using PCA for data structure comparisons}
-All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets.
-Usually, when performing PCA with other purposes in mind, the main interest lies in the \textit{scores}, i.e.
-the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings.
-But where the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables.
-If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both datasets.
-This can be the case while the two sets of scores could be arbitrarily different, which could e.g.\ happen if the two datasets were taken from two different populations of subjects.
-On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.
-
-The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables.
-We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the apparently same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$.
-%
-Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For these %two
-three datasets, we complete the following steps:
-\begin{enumerate}
-\item Within each dataset, standardize each of the variables to have mean zero and unit standard deviation.
-Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ and $\tilde{x}_j \in \RR^d$ for $j=1,\dotsc,n$ be the standardized datasets.
-\item Form the principal component analyses
-\begin{align*}
-S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} &
-S &= \frac{1}{n-1} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top,
-\end{align*}
-thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda_k$ for $k=1,\dotsc,d$.
-\end{enumerate}
-The hereby obtained PCA decompositions of the correlation matrices can then be compared.
-We present three diagnostics plots that are designed to shine a light on different types and levels of data structure differences.
-These three plots are:
-\begin{description}
-\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA.
-The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
-\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e.
-loading and variance component pairs) that are most similar and most different across the two datasets
-\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
-\end{description}
-For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.
-
-Note that the standardization implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
-\begin{equation*}
-\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d.
-\end{equation*}
-This identity will simplify some expressions below.
-
-\subsubsection{The cumulative eigenvalue plot}
-The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix.
-%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
-And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
-In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
-\begin{align*}
-(0,0), &&
-(\lambda_1,\lambda_{11}-\lambda_{12}), &&
-(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
-\ldots, &&
-\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
-\end{align*}
-This may be seen as a cumulative Bland-Altman plot \citep{AltmanBland83,LinEtAl2002} for the variation components.
-Note that due to the standardization, the last point will always be equal to $(d,0)$.
-Thus, this curve will begin and end at the x-axis.
-And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are.
-Moreover, a positive cumulative differences implies that dataset 1 holds more information in the first components than dataset 2 does.
-
-In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
-\begin{align*}
-\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
-\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)^2.
-\end{align*}
-We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_1$ and $n_2$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics.
-This should be done a large (e.g.
-10000) number of times.
-Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to %even larger
-test statistics at least as large as %than 
-the one we found for the original datasets.
-
-The permutation test results are also used to visualize the uncertainty of the CE curve in the plots.
-In the CE plots shown in the following section, we plot the observed curve together with 20 of the resampled curves, as well as a shaded region visualizing pointwise 95 \% coverage intervals.
-If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicates differences between the two datasets.
-
-%Whether the excursions in the observed curve are large or within the range of sample variation can be quantified by a permutation test.
-The idea is that we randomly reallocate the $n_1+n_2$ observations in the combined dataset to two datasets with $n_1$ and $n_2$ observations, respectively, and then repeat the procedure described above.
-The random reallocation by construction ensures that the two resampled datasets are alike except their sample size and sampling variation.
-In the \emph{CumEigenPlot} we make 1000 independent random reallocations, and plot  P-values for the null hypothesis that the variation components are the same are easily provided for any appropriate test statistic by the resampling procedure as well.
-
-\subsubsection{The angle plot}
-This plot simultaneously compares the variation components and the loadings.
-Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
-Then the empirical correlation matrix for the first dataset, $S_1$, has the following orthogonal decomposition in the coordinate system of the second dataset
-%\begin{equation*}
-%S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
-%= \lambda_{\max} \sum_{k=1}^d
-%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
-%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
-%\end{equation*}
-\begin{equation*}
-S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
-= \lambda_{\max} \sum_{k=1}^d
-\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
-\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)^\top,
-\end{equation*}
-and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset %\hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?
-Bo: Jeg har ikke noget bud på en sådan udvidelse lige nu og her --- skal vi ikke bare sige at det er ok i sin nuværende form?}.
-We propose to visualize these two decompositions in a $d \times d$ grid display.
-In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell.
-The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal.
-To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows.
-The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
-\begin{align*}
-\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
-\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
-\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
-\end{align*}
-Note that for two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the sample correlation.
-Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance component contributions.
-The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors.
-This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal.
-Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.
-
-In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the first dataset in the coordinate system of the second dataset.
-Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the second dataset in the coordinate system of the first dataset.
-%\hl{Can we expand on this in a non-technical way?
-- Karl: it seems that we cannot}
-If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
-Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
-And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.
-
-\subsubsection{The chroma plot}
-This plot compares the loadings of the two datasets.
-The chroma plot consists of two panels, one for each dataset, made up of colored bars.
-These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions.
-More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors.
-The width of the $j$th colored segments are given by
-\begin{align*}
-\omega_{1ij} &= \frac{\lvert\eta_{1ij}\rvert}{\sum_{k=1}^d \lvert\eta_{1ik}\rvert}, &
-\omega_{2ij} &= \frac{\lvert\eta_{2ij}\rvert}{\sum_{k=1}^d \lvert\eta_{2ik}\rvert},
-\end{align*}
-where $\eta_{1ij}$ and $\eta_{2ij}$ denotes the $j$th entry of $\eta_{1i}$ and $\eta_{2i}$, respectively.
-Due to the indeterminacy of the sign, all the signs are removed from the coefficients in the loadings.
-The bars are ordered according to the variation components and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance component contributions,
-\begin{align*}
-\tilde\sigma^c_{1i} &= \frac{\sum_{j=1}^i \lambda_{1j}}{\sum_{k=1}^d \lambda_{1k}} = \frac{\sum_{j=1}^i \lambda_{1j}}{d}, &
-\tilde\sigma^c_{2i} &= \frac{\sum_{j=1}^i \lambda_{2j}}{\sum_{k=1}^d \lambda_{2k}} = \frac{\sum_{j=1}^i \lambda_{2j}}{d}.
-\end{align*}
-Especially when $d$ is large, we recommend plotting only a select set of interesting principal components, e.g.
-as identified by use of the angle plot.
-In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$.
-
-The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.
-For each component, the loadings describe how influential the different variables are on that component.
-Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such as \textit{"variable $x$ is generally more influential in subset 1 than it is in subset 2"}, thereby helping us to understand where and why potential data structure differences are found.
-
-\section{Comparing psychological well-being: A data example}
-\label{sec:dataexample}
-We will now turn to a concrete data example in order to illustrate the capabilities of the tools presented above.
-We use data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
-The topic of our investigation is motivated by an increasingly popular tendency to publish rankings of countries in fields as different as educational quality (e.g.
-the PISA project) and citizen happiness (e.g.
-the UN \textit{World Happiness Report} project).
-From a methodological point of view, such international rankings are very problematic, as they rely on the fundamental assumption that the measured concepts are inherently the same across countries.
-The PCADSC tools qualify as a suite of methods for exploring the validity of this assumption empirically.
-
-In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though \cite{Veenhoven2012} presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability, while \cite{Lolle2016} shows highly potent translation issues for the term \textit{happiness}.
-But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
-If not, the very idea of translation is meaningless, since different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socio-economic settings.
-This is in fact a question concerning comparability of data structures.
-If two countries differ e.g.
-in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one.
-More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being.
-The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures.
-And therefore, comparing the two countries in these measures is not a meaningful endeavor.
-
-In this section, we use the PCADSC tools to unveil international differences in one aspect of happiness, namely psychological well-being.
-Our starting point is Denmark, a small, northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \citep{WHR2016}, and we wish to investigate if this title is really meaningful at all.
-In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria.
-Though both countries are European and thus not geographically nor culturally as far apart as some other countries might be, these two countries have previously been highlighted to be very different in terms of what defines happiness \citep{ESStopline5}.
-Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \citep{Rodriguez2014}, with a much less pronounced relationship between the two in Northern European countries, compared to other European countries.
-In particular, interpersonal relations should play a less important role in Denmark, compared to Bulgaria.
-Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.
-
-We also compare the Danish data with Swedish data in order to illustrate that the PCADSC tools actually do have some discriminatory power.
-Denmark and Sweden are both Northern European countries and are often deemed very similar in terms of culture and history.
-Therefore, we expect fundamental concepts such as psychological well-being to be similar across these two countries.
-
-All computations and figures presented in this section were created using our \R{R} package \R{PCADSC} \citep{PCADSC}.
-
-\subsection{Data}
-
-\begin{table}[t]
-\centering
-%\begin{tabular}{lccccccccc}
-%  \hline
-%  & \multicolumn{3}{c}{Denmark} & \multicolumn{3}{c}{Bulgaria} & \multicolumn{3}{c}{Sweden} \\
-% & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \\
-%  \hline
-% Evaluative wellbeing & 8.00 & 8.75 & 9.50 & 3.50 & 5.00 & 7.00 & 7.00 & 8.00 & 9.00 \\
-% Emotional wellbeing & 7.22 & 8.33 & 8.89 & 5.00 & 6.67 & 7.78 & 6.67 & 7.78 & 8.89 \\
-%Functioning & 6.93 & 7.57 & 8.21 & 5.50 & 6.68 & 7.68 & 6.39 & 7.04 & 7.68 \\
-%Vitality & 6.67 & 7.50 & 8.33 & 5.83 & 7.50 & 8.33 & 6.67 & 7.50 & 9.17 \\
-% Community wellbeing & 5.83 & 6.77 & 7.57 & 3.70 & 4.67 & 5.70 & 5.66 & 6.57 & 7.37 \\
-%Supportive relationships & 7.42 & 8.25 & 8.92 & 6.17 & 7.25 & 8.08 & 7.42 & 8.25 & 8.75 \\
-%   \hline
-%\end{tabular}
-\begin{tabular}{lccc}
-  \hline
-  & Denmark & Bulgaria & Sweden \\
-  \hline
- Evaluative wellbeing & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
- Emotional wellbeing & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
-Functioning & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
-Vitality & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
- Community wellbeing & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
-Supportive relationships & 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
-   \hline
-\end{tabular}
-
-\caption{Median values of each of the six dimensions of psychological well-being, stratified by country.
-25 and 75 percentiles are listed in parentheses.
-Note that the scales are constructed such that they all run from 0-10.}
-\label{tableDistr}
-\end{table}
-
-The ESS 2012 data contains a total of 626 variables collected from 54673 citizens of 29 countries.
-Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
-These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}.
-More details on these scales can be found in \citep{ESStopline5} and the relationship between questionnaire items and scales is summarized in Table \ref{table:items} in the appendices.
-We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10.
-For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below.
-This gives us $n_{DK} = 1498$ observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations.
-Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
-
-\subsection{Comparing Denmark and Bulgaria}
-\begin{figure}
-\centerline{\includegraphics[scale = 0.65]{essDKBGce.pdf}}
-\centerline{\includegraphics[scale = 0.65]{essDKBGhair.pdf}}
-\caption{The CE plot (top) and the angle plot (bottom) resulting from comparing Bulgarian and Danish data on psychosocial well-being.
-The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
-In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
-\label{plotBG.cehair}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[scale = 0.7]{essDKBGpancake234.pdf}}
-\caption{A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data.
-The component-bars are annotated with their relative variance contributions (denoted $\tilde\sigma_i$ in the above).}
-\label{plotBG.pancake}
-\end{figure}
-
-%\begin{figure}
-%\centerline{\includegraphics[scale = 0.7]{essDKBGWallyPCADSC234.pdf}}
-%\caption{\hl{something.}}
-%\label{plotBG.wally}
-%\end{figure}
-
-Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales.
-The CE plot show a remarkable degree of lacking comparability: The cumulative differences in the eigenvalues by far exceed what could come about randomly if there really were no difference in the data structures.
-This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero.
-
-Moving on to the angle plot, we find that the differences are primarily to be found in the second, third and fourth principal components (PCs).
-The blue arrows visualize the decomposition of the principal components for the Bulgarian dataset in the coordinate system of the Danish dataset.
-We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3.
-The red arrows visualize the decomposition of the principal components for the Danish dataset in the coordinate system of the Bulgarian dataset.
-Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3.
-Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.
-
-The chroma blot in Figure \ref{plotBG.pancake} allows us to look closer into these components.
-Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish.
-In the Danish data, on the other hand, we find that \textit{Vitality} and \textit{Emotional well-being} seem to play bigger roles, as they appear with larger loadings in more high-ranking components in this sample, relative to the Bulgarian.
-
-All in all, we find that psychological well-being does \textit{not} seem to be the same concept in Bulgaria and Denmark.
-The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalues) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle- and chroma plots).
-In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, corresponding well with the previous findings mentioned above.
-Thus, the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics (\cite{ESStopline5}).
-Moreover, the two countries cannot be ranked in terms of which country is "the most happy", at least not by referring to psychological well-being dimensions such as those encountered here.
-
-% While the first principal component, which is responsible for explaining 50-60 \% of the variance in the data, is very similar for the two countries, we see quite large differences in the remaining components.
-In the second component, we see that the two countries disagree in the relative importance of the scales \textit{Community wellbeing} and \textit{Vitality}.
-In the third and fourth components, general disagreement  is found.
-All in all, components 2-4, representing almost half of the variability in the data, are not very similar across the two countries.
-Moreover, the two subsets of the data also differ with respect to how much variance is explained by each component, and the difference is particularly big for the first component.
-This component has approximately 15 \%  more explanatory power in the Bulgarian subsample than it does in the Danish.
-
-%{\color{red}
-%Figure \ref{plotESSHairplot} shows the hairplot.
-The blue arrows visualize the decomposition of the principal components for the %first dataset in the coordinate system of the second dataset.
-We see that PC2 also loads on PC3, that PC3 also loads on PC4, and %that PC4 also loads on PC2 and PC3.
-The red arrows visualize the decomposition of the principal components for the second dataset %in the coordinate system of the first dataset.
-We see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that %PC4 also loads on PC3.
-For PC1, PC5 and PC6 the main difference is in the size of the variation component.
-%}
-
-%But did we really illustrate a data structure difference due to country differences or did we just illustrate the variability of the results of the PCADSC method?
-In order to investigate this further, we look at a so-called \textit{Wally plot} (\hl{ref: Claus Ekstrøm}).
-In this plot, we compare the results of PCADSC conducted with grouping by country with several random, but similar grouping variables.
-Specifically, we produce 7 PCADSC plots where the country variable was replaced by a randomly generated variable that divides the observations into two groups of the same sizes as the country samples.
-The results are illustrated in Figure \ref{plotESSPCADSCWally}.
-Here, we see that the differences in the second component from the original PCADSC results are not matched in any of the randomly grouped PCADSC runs.
-In fact, the 7 runs are remarkably similar, thereby illustrating that PCADSC seems to be very robust with respect to random groupings: The signal in the data is not blurred by the random subdivisions.
-When it comes to the differences in the third component for the two groups, we find much larger variability in the 7 random runs.
-\hl{more comments here...
-Wait until we are sure exactly what we think about the results and what other PCA-based methods, we will do before/after.
-Particularly, how do we deal with eigen value differences?}
-
-\subsection{Comparing Denmark and Sweden}
-\begin{figure}
-\centerline{\includegraphics[scale = 0.65]{essDKSEce.pdf}}
-\centerline{\includegraphics[scale = 0.65]{essDKSEhair.pdf}}
-\caption{A CE (top) and a angle (bottom) plot for comparing the Danish and the Swedish psychological well-being data.
-The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
-\label{plotSE.cehair}
-\end{figure}
-
-\begin{figure}
-\center
-\centerline{\includegraphics[scale = 0.7]{essDKSEpancake.pdf}}
-\caption{A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples.
-Note that for each component, the bar is annotated with its cumulative variance contribution (denoted $\tilde\sigma^c_i$ in the above), that is, how much variance can be explained by having information of this and the preceding components.}
-\label{plotSE.pancake}
-\end{figure}
-
-%\begin{figure}
-%\center
-%\includegraphics[scale = 0.7]{essDKSEWallyPCADSC.pdf}
-%\caption{\hl{something.}}
-%\label{plotSE.wally}
-%\end{figure}
-
-We now turn to the comparison of Denmark and Sweden in terms of psychological well-being.
-Figure \ref{plotSE.cehair} shows the CE- and angle plots for these two countries.
-In the CE plot, we now find the cumulative eigenvalue curve to be just within the acceptance region of the null-hypothesis.
-This is also reflected by the two tests, which now produce $p$-values of $p_\text{KS} = 0.17$ and $p_\text{CvM} = 0.11$, respectively.
-Thus, it is not unreasonable to assume equal variance components in the two datasets.
-
-% but not with overwhelming evidence.
-% Bo: absence of proof \neq proof of absence.
-
-The angle plot in Figure \ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent.
-This implies that if one already has e.g.
-the information held in the first PC from the Danish data, this information is in itself mostly sufficient to describe the first PC of the Swedish data.
-
-Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80 \% of the variance in both datasets), and slight, but increasing, differences in the remaining three components.
-We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25 \% of the variance in both datasets.
-In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g.
-conclude using Table \ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally a lot happier than the least happy people in Sweden.
-A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.
-
-\section{Discussion}
-\label{sec.Discussion}
-Three new tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if two datasets can be combined for analysis were proposed and discussed in this paper.
-They all employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
-% yielding a solution that is largely independent of the sizes  the two datasets.
-Anne: We don't know if that is true.
-
-In an analysis of data from the 2012 version of the European Social Survey (ESS) project, we illustrated that the PCADSC tools help inform analyses of inter-country differences in psychological well-being and happiness.
-It should be noted that even though concerns about pooling data from different countries are not yet very prevalent when it comes to happiness rankings, the subject has been very well scrutinized in the field of international educational rankings.
-For instance, the PISA tests have repeatedly been criticized for not being meaningful objects of international comparisons, especially due to problems with differential item functions \citep{Kankaras2014,Kreiner2014,ZwitserEtAl2017} and translation problems \citep{Asil2016}.
-This highlights lack of data comparability as a empirically relevant and concerning issue that there is no way around: When conducting internal rankings, the issue of data structure heterogeneity should always be addressed.
-
-Though the PCADSC toolbox comprises three diagnostic plots as of now, the PCA decomposition can of course be utilized further to bring about new PCADSC plots that illustrate more aspects of data structure differences.
-For instance, the scree plot (as introduced by \cite{Cattell1966}), which plots the ordered eigenvalues against their component numbers, could easily be recreated as a PCADSC tool by creating scree plots for both data subsets in the same coordinate system, yielding a simple graphical comparison.
-The resampling approach used in the CE plot could then be adopted, thus annotating the scree plots with resampled curves.
-This general approach to model-fit evaluation is very helpful in combination with permutation tests, %using cumulative residuals,
-as inspired by Lin et al. \citeyearpar{LinEtAl2002}. Similarly, the angle plot could also be annotated with p-values from a permutation test based on appropriate test statistics.
-
-The proposed methods also have limitations.
-First of all, as PCA is performed after standardization of the variables, the PCADSC methods cannot detect differences in mean values nor in marginal variances - this information is thrown away as the very first step.
-However, such differences are of a fundamentally different nature than those we have discussed so far.
-Differences in mean values are often the main interest of the analysis and should of course not be regarded as a data comparability issue.
-Differences in marginal variances, on the other hand, can pose some modeling challenges, but none that are not generally solvable by use of random effects models.
-We therefore do not consider this property of PCADSC to be a major drawback.
-
-A more prominent limitation is the fact that the PCASDC methods are only valid for variables that have a numerical interpretation.
-In particular, this means that the methods can not be used for data with nominal categorical variables.
-This limitation is inherited from the PCA method and thus, moving beyond it will entail replacing the PCA framework with a more general one.
-Multiple correspondence analysis (MCA) is often suggested as a categorical extension of PCA \citep{Abdi2010} and including MCA in the PCADSC tools is thus a natural next step.
-
-Moreover, though we have illustrated the discriminatory power of PCASDC in one data example, of course further evaluation of its performance is needed.  
-While the best feature of PCADSC is perhaps the intuitive, visual nature of the tools, it is also the biggest weakness.
-For statistical methods that result in a number (or a set of numbers), the standard way of evaluating performance is to report on simulation studies, but in the case of statistical methods that report graphics this is not feasible.
-Therefore, a systematic evaluation of the performance of the PCADSC methods is not very straight-forward to design: It really depends on whether or not users can detect actual differences in data structures by using the diagnostic plots.
-We suggest that the variability of the methods could be studied further using \textit{Wally plots} \citep{Ekstrøm2014}, which are learning tools developed for teaching students how to interpret residual plots.
-First in line as a topic for more thorough investigations of performance is the sensitivity towards the sample sizes of the two datasets, $n_1$ and $n_2$.
-
-All in all, PCADSC represents a first step towards addressing a known issue that has otherwise been met with unsatisfactory ad-hoc methods whose assumptions rest upon the very hypothesis they are testing.
-The PCADSC tools perform quite well in the data example provided here, yielding conclusions that are in accordance with well-established theories and they are simple to use and test further using the PCADSC \texttt{R} package.
-
-%\section{Concluding Remarks}
-%\label{sec:conclusion}
-
-\bibliographystyle{apa}
-\bibliography{bib}
-
-\appendix
-
-%\newpage
-%\section{Supplementary tables}
-\begin{table}[h]
-\centering
-\scriptsize
-\rowcolors{1}{}{lightgray}
-\bgroup
-\def\arraystretch{1.5}
-\begin{tabular}{ll}
-\hline
-Scale & Items \\
-\hline
-\rowcolor{lightgray} & How satisfied with life as a whole \\
-\rowcolor{lightgray}\multirow{-2}{*}{Evaluative wellbeing }& How happy are you\\
-
-\rowcolor{white} & Felt sad, how often in the past week\\
-\rowcolor{white} & Felt depressed, how often in the past week\\
-\rowcolor{white} & Enjoyed life, how often in the past week\\
-\rowcolor{white} & Were happy, how often in the past week\\
-\rowcolor{white} & Felt anxious, how often in the past week\\
-\rowcolor{white}\multirow{-6}{*}{Emotional wellbeing} & Felt calm and peaceful, how often in the past week\\
-
-\rowcolor{lightgray} & Free to decide how to live my life\\
-\rowcolor{lightgray} & Little chance to show how capable I am\\
-\rowcolor{lightgray} & Feel accomplishment from what I do\\
-\rowcolor{lightgray} & Interested in what you are doing\\
-\rowcolor{lightgray} & Absorbed in what you are doing\\
-\rowcolor{lightgray} & Enthusiastic about what you are doing\\
-\rowcolor{lightgray} & Feel what I do in life is valuable and worthwhile\\
-\rowcolor{lightgray} & Have a sense of direction\\
-\rowcolor{lightgray} & Always optimistic about my future\\
-\rowcolor{lightgray} & There are lots of things I feel I am good at\\
-\rowcolor{lightgray} & In general feel very positive about myself\\
-\rowcolor{lightgray} & At times feel as if I am a failure\\
-\rowcolor{lightgray} & When things go wrong in my life it takes a long time to get back to normal\\
-\rowcolor{lightgray}\multirow{-14}{*}{Functioning} & Deal with important problems\\
-
-\rowcolor{white} & Felt everything did as effort, how often in the past week\\
-\rowcolor{white} & Sleep was restless, how often in the past week\\
-\rowcolor{white} & Could not get going, how often in the past week\\
-\rowcolor{white}\multirow{-4}{*}{Vitality} & Had a lot of energy, how often in the past week\\
-
-\rowcolor{lightgray} &  Most people can be trusted\\
-\rowcolor{lightgray} &  People try to take advantage\\
-\rowcolor{lightgray} &  Most of the time people are helpful\\
-\rowcolor{lightgray} &  Feel people in local area help one another\\
-\rowcolor{lightgray}\multirow{-5}{*}{Community wellbeing}& Feel close to the people in local area\\
-
-\rowcolor{white} & How many with whom you can discuss intimate matters\\
-\rowcolor{white} & Feel appreciated by those you are close to\\
-\rowcolor{white} & Receive help and support\\
-\rowcolor{white}\multirow{-4}{*}{Supportive relationships} & Felt lonely, how often in the past week\\
-\hline
-\end{tabular}
-\egroup
-\caption{Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
-Note that before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
-\label{table:items}
-\end{table}
+Karl Bang Christensen \\
+\href{mailto:kach@sund.ku.dk}{\nolinkurl{kach@sund.ku.dk}} \\
+Department of Public Health, University of Copenhagen
 

### 2018-03-13 article/Old versions/Psychometrica/article_titlePageOnly.tex
Added MPS submission

--- 2017-04-25 article/article_v0.9.5.tex
@@
 \affil{Department of Public Health, University of Copenhagen}
+\affil{\href{mailto:ahpe@sund.ku.dk}{\nolinkurl{ahpe@sund.ku.dk}} (Corresponding author)}
 
@@
 \affil{Department of Mathematical Sciences, University of Copenhagen}
+\affil{\href{mailto:bomar@math.ku.dk}{\nolinkurl{bomar@math.ku.dk}}}
 
@@
 \affil{Department of Public Health, University of Copenhagen}
+\affil{\href{mailto:kach@sund.ku.dk}{\nolinkurl{kach@sund.ku.dk}}}
 
@@
 
-\end{titlepage}\vspace*{24pt}
-
-%\RepeatTitle{Exploratory data structure comparisons: Three new tools based on Principal Component Analysis}
-
-\begin{center}\vskip3pt
-
-\vspace{32pt}
-
-Abstract\vskip3pt
-
-\end{center}
-
-\begin{abstract}
-\linespacing{1}
-
-In empirical psychometric work, one often encounters data that is somehow divided into two distinct subsets.
-This might be due to multi-center sampling frames, or to variations in instruments, orders of questionnaire item, modes of administration etc.
-Often, one wishes to combine such data into one an conduct only a single analysis, but this is only meaningful if the two subsets are similar in structures, that is, if they measure the same underlying concepts.
-However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc or require making full model assumptions.
-We present a suite of three visual, non-parametric diagnostic tools for investigating differences in data structures for two datasets containing different observations of the same variables, refered to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC).
-This approach requires no directional nor hierarchical assumptions on the data.
-Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets, and by targeting different aspects of the PCA information, the three plots give a thorough and deep understanding of where differences occur and why it might be the case.
-
-The functionality of the PCADSC tools is presented by use of data from the European Social Survey on psycological well-being in three European countries, namely Denmark, Sweden and Bulgaria.
-Here, we find that the concepts of psychological well-being are not the same when comparing Denmark and Bulgaria, and thus a comparison of e.g.
-mean scores in different well-being aspects is not meaningful.
-However, when comparing the two Scandinavian countries, Denmark and Sweden, we find very similar data structures and thus very similar concepts of well-being.
-Therefore, inter-country comparisons are warranted for these two countries.
-
-\begin{keywords}
-multi-center studies, data heterogeneity, data structure comparisons, principal component analysis (PCA), European Social Survey.
-\end{keywords}
-\end{abstract}
-
-\vspace{\fill}\newpage
-
-\section{Introduction}
-\label{sec:introduction}
-
-Data comparability is a classical topic in statistics and a recurring one in psychometrics.
-Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability (and thus collapsibility) needs to be assessed empirically.
-This happens e.g.
-when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration or the order of survey questionnaire items are applied.
-All of these practices are essential parts of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions.
-However, if we do not address the question of collapsibility among the data subsets arriving from e.g.
-different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.
-
-Sophisticated methods for addressing this question are available when we are willing to assume a statistical model.
-But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment %very
-ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a %valid 
-general data structure comparison method, but rather a fitted-model comparison method.
-It addresses the interplay between the model and the data, not the data alone.
-However, useful tools for initial, exploratory investigations of data comparability data that help researchers identify potential problems or challenges early in the study design development are mostly absent.
-What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
-Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables.
-On the other hand, e.g.
-entry-by-entry comparisons of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase.
-
-While the issue is not in any way new, its relative importance is increasing due to new developments in data availability and data collection.
-Classical statistical methods are generally aimed at analyzing data from designed experiments and historically, statistical analyses have been conducted by researchers who knew the design and the source of the dataset well.
-However, the origin stories of datasets have changed over time and today, a lot of data are accumulated without a specific purpose in mind, as data collection and sharing has become easier and more affordable through the development of new technologies.
-This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for %all
-everyone to use.
-However, large scale publicly available datasets, such as the PISA data and European Social Survey (ESS) data, are often used by data analysts that are far removed from the data producers and therefore, problem-specific recommendations about e.g.
-potential instrument-induced challenges in the datasets %are not
-may not be available for these data analysts.
-For this reason, a thorough investigation of data structure comparability becomes a crucial step in any meaningful data analysis when using such open-source data.
-
-The data structure comparison problem is however not limited to the context of new data sharing platforms.
-Even in single-population survey studies, a mixture of modes of administration is often used, e.g.
-mail and telephone, and while this can improve response rates, it can also be quite problematic \citep{Brambilla1987, McHorney1994}, and differences in response behavior can lead to biased results.
-Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful, so the comparability issue is not just a statistical puzzle.
-Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu \citeyearpar{Liu2016}, who warns against the combination of online panel data (i.e.
-pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions).
-
-The comparability issue that is present in all the examples presented so far can be summarized as follows: Assume that we have two datasets with the same variables, but different observations, or alternatively (and equivalently) a single dataset with a subset-inducing variable.
-Assume next that we wish to compare data structures without specifying a model, or even a variable of interest.
-The central question is then whether the two datasets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be somehow be dealt with, either in terms of modeling choices or simply by conducting several separate analyses.
-
-We propose a suite of three new tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC).
-These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
-%This yields a solution that is largely independent of the sizes of the two subsets of data. Anne: Det ved vi da ikke om det gør? Er det ikke netop det, vi siger, at man burde teste som næste step?
-The proposed tools are implemented in the statistical software \texttt{R} in our package, \texttt{PCADSC} \citep{PCADSC}.
-
-This manuscript describes the procedures, including a brief introduction to principal component analysis (PCA), and presents a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey.
-More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is completely ignored e.g.
-when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project \citep{WHR2016}.
-
-\section{PCA-based tools for data structure comparisons}
-\label{sec:pcadscintro}
-As mentioned above, the purpose of PCADSC is to compare overall data structures in two subsets of a dataset.
-But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context.
-One such definition is the structure of the covariance matrix of the dataset.
-If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables.
-%This gives it a very nice interpretation as a measure of the overall structure. If we do not accept 
-And without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data.
-All in all, the empirical covariance matrix is a reasonable place to start looking for differences in \textit{overall data structures}.
-
-A naive approach for data structure comparisons might therefore be to compute the empirical covariance matrices on each of the two data subsets and simply compare these matrices entry-by-entry.
-Though the idea perhaps sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices.
-There is simply too much information to consider at once.
-However, by use of linear algebra, we can decompose and recompose the covariance matrix such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance.
-In particular, we find a new representation of the covariance matrix that makes it possible to gain an overview of the most interesting aspects of the data.
-One such decomposition strategy is known as \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.
-
-At this point a general remark about standardization is in place.
-PCA deconstructs the covariance matrix in components according to the most explained variance.
-This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will always be deemed highly influential, no matter the structure of the data.
-Therefore, the variables are typically standardized prior to performing PCA.
-Note that the covariance matrix of the standardized variables is the same as the correlation matrix of the original variables, so post-standardization PCA simply corresponds to performing data structure comparisons of the correlation matrices rather than the covariance matrices.
-The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation.
-In this paper we will adopt this standardization policy, and in particular, we will allow for differences between scalings of the original, raw variables in the two data subsets.
-
-%\hl{Bo: I næste afsnit bruges stadigvæk ``covariance''. Det er vel ok?} Anne: Synes det fungerer fint.
-
-\subsection{A brief introduction to principal component analysis}
-Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = (\bar{x}_1, ..., \bar{x}_d)^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$.
-We assume in the following that all variables of $X$ have a numerical interpretation, e.g.
-by being continuous or ordinal and categorical.
-For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost.
-More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the original $d$-dimensional one.
-Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
-\begin{equation*}
-\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
-\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
-\end{equation*}
-The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \citep{HastieEtAl2009}.
-More specifically, the rank-q-reconstruction error is attained when we choose
-\begin{equation*}
-\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\},
-\end{equation*}
-where $\eta_1, ..., \eta_q$ are the $q$ first eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$.
-In the PCA framework, we refer to the eigenvectors as \textit{loadings}.
-The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
-\begin{equation*}
-\text{trace}(S) = \sum_{j=1}^d \hat{V}(x_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
-\end{equation*}
-where $x_j \in \RR^n$ 
-%\hl{(Bo: $X_1$ og $X_2$ bruges i en anden betydning i afsnit 2.2. Omformuler?)} Anne: Hov, ja, hermed x_j så det passer med notationen ovenfor.
-denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
-This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
-And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are %different,
-distinct, the representation hereby obtained is a valid object of inference.
-If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
-
-The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
-\begin{align*}
-\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
-\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2,
-\end{align*}
-where the initial subspace is defined as $\hat{K}_0 = \{0\}$.
-It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
-This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$.
-This implies that the loadings and the variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.
-
-\subsection{Using PCA for data structure comparisons}
-All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets.
-Usually, when performing PCA with other purposes in mind, the main interest lies in the \textit{scores}, i.e.
-the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings.
-But where the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables.
-If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both datasets.
-This can be the case while the two sets of scores could be arbitrarily different, which could e.g.\ happen if the two datasets were taken from two different populations of subjects.
-On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.
-
-The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables.
-We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the apparently same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$.
-%
-Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For these %two
-three datasets, we complete the following steps:
-\begin{enumerate}
-\item Within each dataset, standardize each of the variables to have mean zero and unit standard deviation.
-Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ and $\tilde{x}_j \in \RR^d$ for $j=1,\dotsc,n$ be the standardized datasets.
-\item Form the principal component analyses
-\begin{align*}
-S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} &
-S &= \frac{1}{n-1} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top,
-\end{align*}
-thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda_k$ for $k=1,\dotsc,d$.
-\end{enumerate}
-The hereby obtained PCA decompositions of the correlation matrices can then be compared.
-We present three diagnostics plots that are designed to shine a light on different types and levels of data structure differences.
-These three plots are:
-\begin{description}
-\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA.
-The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
-\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e.
-loading and variance component pairs) that are most similar and most different across the two datasets
-\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
-\end{description}
-For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.
-
-Note that the standardization implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
-\begin{equation*}
-\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d.
-\end{equation*}
-This identity will simplify some expressions below.
-
-\subsubsection{The cumulative eigenvalue plot}
-The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix.
-%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset.
-And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
-In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
-\begin{align*}
-(0,0), &&
-(\lambda_1,\lambda_{11}-\lambda_{12}), &&
-(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
-\ldots, &&
-\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
-\end{align*}
-This may be seen as a cumulative Bland-Altman plot \citep{AltmanBland83,LinEtAl2002} for the variation components.
-Note that due to the standardization, the last point will always be equal to $(d,0)$.
-Thus, this curve will begin and end at the x-axis.
-And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are.
-Moreover, a positive cumulative differences implies that dataset 1 holds more information in the first components than dataset 2 does.
-
-In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
-\begin{align*}
-\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
-\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)^2.
-\end{align*}
-We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_1$ and $n_2$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics.
-This should be done a large (e.g.
-10000) number of times.
-Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to %even larger
-test statistics at least as large as %than 
-the one we found for the original datasets.
-
-The permutation test results are also used to visualize the uncertainty of the CE curve in the plots.
-In the CE plots shown in the following section, we plot the observed curve together with 20 of the resampled curves, as well as a shaded region visualizing pointwise 95 \% coverage intervals.
-If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicates differences between the two datasets.
-
-%Whether the excursions in the observed curve are large or within the range of sample variation can be quantified by a permutation test.
-The idea is that we randomly reallocate the $n_1+n_2$ observations in the combined dataset to two datasets with $n_1$ and $n_2$ observations, respectively, and then repeat the procedure described above.
-The random reallocation by construction ensures that the two resampled datasets are alike except their sample size and sampling variation.
-In the \emph{CumEigenPlot} we make 1000 independent random reallocations, and plot  P-values for the null hypothesis that the variation components are the same are easily provided for any appropriate test statistic by the resampling procedure as well.
-
-\subsubsection{The angle plot}
-This plot simultaneously compares the variation components and the loadings.
-Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
-Then the empirical correlation matrix for the first dataset, $S_1$, has the following orthogonal decomposition in the coordinate system of the second dataset
-%\begin{equation*}
-%S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
-%= \lambda_{\max} \sum_{k=1}^d
-%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
-%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
-%\end{equation*}
-\begin{equation*}
-S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
-= \lambda_{\max} \sum_{k=1}^d
-\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
-\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)^\top,
-\end{equation*}
-and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset %\hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?
-Bo: Jeg har ikke noget bud på en sådan udvidelse lige nu og her --- skal vi ikke bare sige at det er ok i sin nuværende form?}.
-We propose to visualize these two decompositions in a $d \times d$ grid display.
-In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell.
-The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal.
-To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows.
-The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
-\begin{align*}
-\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
-\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
-\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
-\end{align*}
-Note that for two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the sample correlation.
-Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance component contributions.
-The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors.
-This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal.
-Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.
-
-In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the first dataset in the coordinate system of the second dataset.
-Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the second dataset in the coordinate system of the first dataset.
-%\hl{Can we expand on this in a non-technical way?
-- Karl: it seems that we cannot}
-If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
-Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
-And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.
-
-\subsubsection{The chroma plot}
-This plot compares the loadings of the two datasets.
-The chroma plot consists of two panels, one for each dataset, made up of colored bars.
-These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions.
-More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors.
-The width of the $j$th colored segments are given by
-\begin{align*}
-\omega_{1ij} &= \frac{\lvert\eta_{1ij}\rvert}{\sum_{k=1}^d \lvert\eta_{1ik}\rvert}, &
-\omega_{2ij} &= \frac{\lvert\eta_{2ij}\rvert}{\sum_{k=1}^d \lvert\eta_{2ik}\rvert},
-\end{align*}
-where $\eta_{1ij}$ and $\eta_{2ij}$ denotes the $j$th entry of $\eta_{1i}$ and $\eta_{2i}$, respectively.
-Due to the indeterminacy of the sign, all the signs are removed from the coefficients in the loadings.
-The bars are ordered according to the variation components and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance component contributions,
-\begin{align*}
-\tilde\sigma^c_{1i} &= \frac{\sum_{j=1}^i \lambda_{1j}}{\sum_{k=1}^d \lambda_{1k}} = \frac{\sum_{j=1}^i \lambda_{1j}}{d}, &
-\tilde\sigma^c_{2i} &= \frac{\sum_{j=1}^i \lambda_{2j}}{\sum_{k=1}^d \lambda_{2k}} = \frac{\sum_{j=1}^i \lambda_{2j}}{d}.
-\end{align*}
-Especially when $d$ is large, we recommend plotting only a select set of interesting principal components, e.g.
-as identified by use of the angle plot.
-In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$.
-
-The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.
-For each component, the loadings describe how influential the different variables are on that component.
-Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such as \textit{"variable $x$ is generally more influential in subset 1 than it is in subset 2"}, thereby helping us to understand where and why potential data structure differences are found.
-
-\section{Comparing psychological well-being: A data example}
-\label{sec:dataexample}
-We will now turn to a concrete data example in order to illustrate the capabilities of the tools presented above.
-We use data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
-The topic of our investigation is motivated by an increasingly popular tendency to publish rankings of countries in fields as different as educational quality (e.g.
-the PISA project) and citizen happiness (e.g.
-the UN \textit{World Happiness Report} project).
-From a methodological point of view, such international rankings are very problematic, as they rely on the fundamental assumption that the measured concepts are inherently the same across countries.
-The PCADSC tools qualify as a suite of methods for exploring the validity of this assumption empirically.
-
-In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though \cite{Veenhoven2012} presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability, while \cite{Lolle2016} shows highly potent translation issues for the term \textit{happiness}.
-But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness.
-If not, the very idea of translation is meaningless, since different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socio-economic settings.
-This is in fact a question concerning comparability of data structures.
-If two countries differ e.g.
-in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one.
-More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being.
-The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures.
-And therefore, comparing the two countries in these measures is not a meaningful endeavor.
-
-In this section, we use the PCADSC tools to unveil international differences in one aspect of happiness, namely psychological well-being.
-Our starting point is Denmark, a small, northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \citep{WHR2016}, and we wish to investigate if this title is really meaningful at all.
-In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria.
-Though both countries are European and thus not geographically nor culturally as far apart as some other countries might be, these two countries have previously been highlighted to be very different in terms of what defines happiness \citep{ESStopline5}.
-Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \citep{Rodriguez2014}, with a much less pronounced relationship between the two in Northern European countries, compared to other European countries.
-In particular, interpersonal relations should play a less important role in Denmark, compared to Bulgaria.
-Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.
-
-We also compare the Danish data with Swedish data in order to illustrate that the PCADSC tools actually do have some discriminatory power.
-Denmark and Sweden are both Northern European countries and are often deemed very similar in terms of culture and history.
-Therefore, we expect fundamental concepts such as psychological well-being to be similar across these two countries.
-
-All computations and figures presented in this section were created using our \R{R} package \R{PCADSC} \citep{PCADSC}.
-
-\subsection{Data}
-
-\begin{table}[t]
-\centering
-%\begin{tabular}{lccccccccc}
-%  \hline
-%  & \multicolumn{3}{c}{Denmark} & \multicolumn{3}{c}{Bulgaria} & \multicolumn{3}{c}{Sweden} \\
-% & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \quad & $Q_1$ & $M$ & $Q_3$ \\
-%  \hline
-% Evaluative wellbeing & 8.00 & 8.75 & 9.50 & 3.50 & 5.00 & 7.00 & 7.00 & 8.00 & 9.00 \\
-% Emotional wellbeing & 7.22 & 8.33 & 8.89 & 5.00 & 6.67 & 7.78 & 6.67 & 7.78 & 8.89 \\
-%Functioning & 6.93 & 7.57 & 8.21 & 5.50 & 6.68 & 7.68 & 6.39 & 7.04 & 7.68 \\
-%Vitality & 6.67 & 7.50 & 8.33 & 5.83 & 7.50 & 8.33 & 6.67 & 7.50 & 9.17 \\
-% Community wellbeing & 5.83 & 6.77 & 7.57 & 3.70 & 4.67 & 5.70 & 5.66 & 6.57 & 7.37 \\
-%Supportive relationships & 7.42 & 8.25 & 8.92 & 6.17 & 7.25 & 8.08 & 7.42 & 8.25 & 8.75 \\
-%   \hline
-%\end{tabular}
-\begin{tabular}{lccc}
-  \hline
-  & Denmark & Bulgaria & Sweden \\
-  \hline
- Evaluative wellbeing & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
- Emotional wellbeing & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
-Functioning & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
-Vitality & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
- Community wellbeing & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
-Supportive relationships & 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
-   \hline
-\end{tabular}
-
-\caption{Median values of each of the six dimensions of psychological well-being, stratified by country.
-25 and 75 percentiles are listed in parentheses.
-Note that the scales are constructed such that they all run from 0-10.}
-\label{tableDistr}
-\end{table}
-
-The ESS 2012 data contains a total of 626 variables collected from 54673 citizens of 29 countries.
-Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
-These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}.
-More details on these scales can be found in \citep{ESStopline5} and the relationship between questionnaire items and scales is summarized in Table \ref{table:items} in the appendices.
-We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10.
-For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below.
-This gives us $n_{DK} = 1498$ observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations.
-Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
-
-\subsection{Comparing Denmark and Bulgaria}
-\begin{figure}
-\centerline{\includegraphics[scale = 0.65]{essDKBGce.pdf}}
-\centerline{\includegraphics[scale = 0.65]{essDKBGhair.pdf}}
-\caption{The CE plot (top) and the angle plot (bottom) resulting from comparing Bulgarian and Danish data on psychosocial well-being.
-The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
-In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
-\label{plotBG.cehair}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[scale = 0.7]{essDKBGpancake234.pdf}}
-\caption{A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data.
-The component-bars are annotated with their relative variance contributions (denoted $\tilde\sigma_i$ in the above).}
-\label{plotBG.pancake}
-\end{figure}
-
-%\begin{figure}
-%\centerline{\includegraphics[scale = 0.7]{essDKBGWallyPCADSC234.pdf}}
-%\caption{\hl{something.}}
-%\label{plotBG.wally}
-%\end{figure}
-
-Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales.
-The CE plot show a remarkable degree of lacking comparability: The cumulative differences in the eigenvalues by far exceed what could come about randomly if there really were no difference in the data structures.
-This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero.
-
-Moving on to the angle plot, we find that the differences are primarily to be found in the second, third and fourth principal components (PCs).
-The blue arrows visualize the decomposition of the principal components for the Bulgarian dataset in the coordinate system of the Danish dataset.
-We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3.
-The red arrows visualize the decomposition of the principal components for the Danish dataset in the coordinate system of the Bulgarian dataset.
-Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3.
-Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.
-
-The chroma blot in Figure \ref{plotBG.pancake} allows us to look closer into these components.
-Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish.
-In the Danish data, on the other hand, we find that \textit{Vitality} and \textit{Emotional well-being} seem to play bigger roles, as they appear with larger loadings in more high-ranking components in this sample, relative to the Bulgarian.
-
-All in all, we find that psychological well-being does \textit{not} seem to be the same concept in Bulgaria and Denmark.
-The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalues) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle- and chroma plots).
-In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, corresponding well with the previous findings mentioned above.
-Thus, the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics (\cite{ESStopline5}).
-Moreover, the two countries cannot be ranked in terms of which country is "the most happy", at least not by referring to psychological well-being dimensions such as those encountered here.
-
-% While the first principal component, which is responsible for explaining 50-60 \% of the variance in the data, is very similar for the two countries, we see quite large differences in the remaining components.
-In the second component, we see that the two countries disagree in the relative importance of the scales \textit{Community wellbeing} and \textit{Vitality}.
-In the third and fourth components, general disagreement  is found.
-All in all, components 2-4, representing almost half of the variability in the data, are not very similar across the two countries.
-Moreover, the two subsets of the data also differ with respect to how much variance is explained by each component, and the difference is particularly big for the first component.
-This component has approximately 15 \%  more explanatory power in the Bulgarian subsample than it does in the Danish.
-
-%{\color{red}
-%Figure \ref{plotESSHairplot} shows the hairplot.
-The blue arrows visualize the decomposition of the principal components for the %first dataset in the coordinate system of the second dataset.
-We see that PC2 also loads on PC3, that PC3 also loads on PC4, and %that PC4 also loads on PC2 and PC3.
-The red arrows visualize the decomposition of the principal components for the second dataset %in the coordinate system of the first dataset.
-We see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that %PC4 also loads on PC3.
-For PC1, PC5 and PC6 the main difference is in the size of the variation component.
-%}
-
-%But did we really illustrate a data structure difference due to country differences or did we just illustrate the variability of the results of the PCADSC method?
-In order to investigate this further, we look at a so-called \textit{Wally plot} (\hl{ref: Claus Ekstrøm}).
-In this plot, we compare the results of PCADSC conducted with grouping by country with several random, but similar grouping variables.
-Specifically, we produce 7 PCADSC plots where the country variable was replaced by a randomly generated variable that divides the observations into two groups of the same sizes as the country samples.
-The results are illustrated in Figure \ref{plotESSPCADSCWally}.
-Here, we see that the differences in the second component from the original PCADSC results are not matched in any of the randomly grouped PCADSC runs.
-In fact, the 7 runs are remarkably similar, thereby illustrating that PCADSC seems to be very robust with respect to random groupings: The signal in the data is not blurred by the random subdivisions.
-When it comes to the differences in the third component for the two groups, we find much larger variability in the 7 random runs.
-\hl{more comments here...
-Wait until we are sure exactly what we think about the results and what other PCA-based methods, we will do before/after.
-Particularly, how do we deal with eigen value differences?}
-
-\subsection{Comparing Denmark and Sweden}
-\begin{figure}
-\centerline{\includegraphics[scale = 0.65]{essDKSEce.pdf}}
-\centerline{\includegraphics[scale = 0.65]{essDKSEhair.pdf}}
-\caption{A CE (top) and a angle (bottom) plot for comparing the Danish and the Swedish psychological well-being data.
-The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
-\label{plotSE.cehair}
-\end{figure}
-
-\begin{figure}
-\center
-\centerline{\includegraphics[scale = 0.7]{essDKSEpancake.pdf}}
-\caption{A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples.
-Note that for each component, the bar is annotated with its cumulative variance contribution (denoted $\tilde\sigma^c_i$ in the above), that is, how much variance can be explained by having information of this and the preceding components.}
-\label{plotSE.pancake}
-\end{figure}
-
-%\begin{figure}
-%\center
-%\includegraphics[scale = 0.7]{essDKSEWallyPCADSC.pdf}
-%\caption{\hl{something.}}
-%\label{plotSE.wally}
-%\end{figure}
-
-We now turn to the comparison of Denmark and Sweden in terms of psychological well-being.
-Figure \ref{plotSE.cehair} shows the CE- and angle plots for these two countries.
-In the CE plot, we now find the cumulative eigenvalue curve to be just within the acceptance region of the null-hypothesis.
-This is also reflected by the two tests, which now produce $p$-values of $p_\text{KS} = 0.17$ and $p_\text{CvM} = 0.11$, respectively.
-Thus, it is not unreasonable to assume equal variance components in the two datasets.
-
-% but not with overwhelming evidence.
-% Bo: absence of proof \neq proof of absence.
-
-The angle plot in Figure \ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent.
-This implies that if one already has e.g.
-the information held in the first PC from the Danish data, this information is in itself mostly sufficient to describe the first PC of the Swedish data.
-
-Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80 \% of the variance in both datasets), and slight, but increasing, differences in the remaining three components.
-We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25 \% of the variance in both datasets.
-In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g.
-conclude using Table \ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally a lot happier than the least happy people in Sweden.
-A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.
-
-\section{Discussion}
-\label{sec.Discussion}
-Three new tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if two datasets can be combined for analysis were proposed and discussed in this paper.
-They all employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
-% yielding a solution that is largely independent of the sizes  the two datasets.
-Anne: We don't know if that is true.
-
-In an analysis of data from the 2012 version of the European Social Survey (ESS) project, we illustrated that the PCADSC tools help inform analyses of inter-country differences in psychological well-being and happiness.
-It should be noted that even though concerns about pooling data from different countries are not yet very prevalent when it comes to happiness rankings, the subject has been very well scrutinized in the field of international educational rankings.
-For instance, the PISA tests have repeatedly been criticized for not being meaningful objects of international comparisons, especially due to problems with differential item functions \citep{Kankaras2014,Kreiner2014,ZwitserEtAl2017} and translation problems \citep{Asil2016}.
-This highlights lack of data comparability as a empirically relevant and concerning issue that there is no way around: When conducting internal rankings, the issue of data structure heterogeneity should always be addressed.
-
-Though the PCADSC toolbox comprises three diagnostic plots as of now, the PCA decomposition can of course be utilized further to bring about new PCADSC plots that illustrate more aspects of data structure differences.
-For instance, the scree plot (as introduced by \cite{Cattell1966}), which plots the ordered eigenvalues against their component numbers, could easily be recreated as a PCADSC tool by creating scree plots for both data subsets in the same coordinate system, yielding a simple graphical comparison.
-The resampling approach used in the CE plot could then be adopted, thus annotating the scree plots with resampled curves.
-This general approach to model-fit evaluation is very helpful in combination with permutation tests, %using cumulative residuals,
-as inspired by Lin et al. \citeyearpar{LinEtAl2002}. Similarly, the angle plot could also be annotated with p-values from a permutation test based on appropriate test statistics.
-
-The proposed methods also have limitations.
-First of all, as PCA is performed after standardization of the variables, the PCADSC methods cannot detect differences in mean values nor in marginal variances - this information is thrown away as the very first step.
-However, such differences are of a fundamentally different nature than those we have discussed so far.
-Differences in mean values are often the main interest of the analysis and should of course not be regarded as a data comparability issue.
-Differences in marginal variances, on the other hand, can pose some modeling challenges, but none that are not generally solvable by use of random effects models.
-We therefore do not consider this property of PCADSC to be a major drawback.
-
-A more prominent limitation is the fact that the PCASDC methods are only valid for variables that have a numerical interpretation.
-In particular, this means that the methods can not be used for data with nominal categorical variables.
-This limitation is inherited from the PCA method and thus, moving beyond it will entail replacing the PCA framework with a more general one.
-Multiple correspondence analysis (MCA) is often suggested as a categorical extension of PCA \citep{Abdi2010} and including MCA in the PCADSC tools is thus a natural next step.
-
-Moreover, though we have illustrated the discriminatory power of PCASDC in one data example, of course further evaluation of its performance is needed.  
-While the best feature of PCADSC is perhaps the intuitive, visual nature of the tools, it is also the biggest weakness.
-For statistical methods that result in a number (or a set of numbers), the standard way of evaluating performance is to report on simulation studies, but in the case of statistical methods that report graphics this is not feasible.
-Therefore, a systematic evaluation of the performance of the PCADSC methods is not very straight-forward to design: It really depends on whether or not users can detect actual differences in data structures by using the diagnostic plots.
-We suggest that the variability of the methods could be studied further using \textit{Wally plots} \citep{Ekstrøm2014}, which are learning tools developed for teaching students how to interpret residual plots.
-First in line as a topic for more thorough investigations of performance is the sensitivity towards the sample sizes of the two datasets, $n_1$ and $n_2$.
-
-All in all, PCADSC represents a first step towards addressing a known issue that has otherwise been met with unsatisfactory ad-hoc methods whose assumptions rest upon the very hypothesis they are testing.
-The PCADSC tools perform quite well in the data example provided here, yielding conclusions that are in accordance with well-established theories and they are simple to use and test further using the PCADSC \texttt{R} package.
-
-%\section{Concluding Remarks}
-%\label{sec:conclusion}
-
-\bibliographystyle{apa}
-\bibliography{bib}
-
-\appendix
-
-%\newpage
-%\section{Supplementary tables}
-\begin{table}[h]
-\centering
-\scriptsize
-\rowcolors{1}{}{lightgray}
-\bgroup
-\def\arraystretch{1.5}
-\begin{tabular}{ll}
-\hline
-Scale & Items \\
-\hline
-\rowcolor{lightgray} & How satisfied with life as a whole \\
-\rowcolor{lightgray}\multirow{-2}{*}{Evaluative wellbeing }& How happy are you\\
-
-\rowcolor{white} & Felt sad, how often in the past week\\
-\rowcolor{white} & Felt depressed, how often in the past week\\
-\rowcolor{white} & Enjoyed life, how often in the past week\\
-\rowcolor{white} & Were happy, how often in the past week\\
-\rowcolor{white} & Felt anxious, how often in the past week\\
-\rowcolor{white}\multirow{-6}{*}{Emotional wellbeing} & Felt calm and peaceful, how often in the past week\\
-
-\rowcolor{lightgray} & Free to decide how to live my life\\
-\rowcolor{lightgray} & Little chance to show how capable I am\\
-\rowcolor{lightgray} & Feel accomplishment from what I do\\
-\rowcolor{lightgray} & Interested in what you are doing\\
-\rowcolor{lightgray} & Absorbed in what you are doing\\
-\rowcolor{lightgray} & Enthusiastic about what you are doing\\
-\rowcolor{lightgray} & Feel what I do in life is valuable and worthwhile\\
-\rowcolor{lightgray} & Have a sense of direction\\
-\rowcolor{lightgray} & Always optimistic about my future\\
-\rowcolor{lightgray} & There are lots of things I feel I am good at\\
-\rowcolor{lightgray} & In general feel very positive about myself\\
-\rowcolor{lightgray} & At times feel as if I am a failure\\
-\rowcolor{lightgray} & When things go wrong in my life it takes a long time to get back to normal\\
-\rowcolor{lightgray}\multirow{-14}{*}{Functioning} & Deal with important problems\\
-
-\rowcolor{white} & Felt everything did as effort, how often in the past week\\
-\rowcolor{white} & Sleep was restless, how often in the past week\\
-\rowcolor{white} & Could not get going, how often in the past week\\
-\rowcolor{white}\multirow{-4}{*}{Vitality} & Had a lot of energy, how often in the past week\\
-
-\rowcolor{lightgray} &  Most people can be trusted\\
-\rowcolor{lightgray} &  People try to take advantage\\
-\rowcolor{lightgray} &  Most of the time people are helpful\\
-\rowcolor{lightgray} &  Feel people in local area help one another\\
-\rowcolor{lightgray}\multirow{-5}{*}{Community wellbeing}& Feel close to the people in local area\\
-
-\rowcolor{white} & How many with whom you can discuss intimate matters\\
-\rowcolor{white} & Feel appreciated by those you are close to\\
-\rowcolor{white} & Receive help and support\\
-\rowcolor{white}\multirow{-4}{*}{Supportive relationships} & Felt lonely, how often in the past week\\
-\hline
-\end{tabular}
-\egroup
-\caption{Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
-Note that before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
-\label{table:items}
-\end{table}
+\end{titlepage}
+%\Large{\textsc{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}}
 

### 2018-05-17 article/Old versions/MPS/PCADSC_article.tex
updated version
updates as outlined in MPS_reviewercomments_response 17MAJ2018

--- 2018-01-09 article/APM/article.tex
@@
-\documentclass[a4paper,14pt]{article}
-\usepackage[danish]{babel}
+\documentclass[a4paper,12pt]{article}
+
+\usepackage{geometry}
+\geometry{margin=1in}
+
+\usepackage[english]{babel}
 \usepackage{authblk}
@@
 \usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
+\usepackage{multirow}
+\usepackage[table]{xcolor}
+\definecolor{lightgray}{HTML}{e6e6e6}
+
+%\DisableLigatures[f]{encoding = *, family = * }
 % color can be used to apply background shading to table cells only
@@
 
-\def\volumeyear{2017}
+\def\volumeyear{2018}
 
@@
 %\affil[2]{Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}
-
+\date{}
 \maketitle
@@
 %\email{ahpe@sund.ku.dk}
-
-\newpage
 
@@
 \begin{abstract}
-Researchers often encounter data that is divided into distinct subsets.
-This might be due to multi-center sampling frames, or to variations in instruments, questionnaire item ordering or mode of administration.
-Often, one wishes to combine such data into one and conduct only a single analysis, but this is only meaningful if the subsets are similar in structure.
-Common procedures for assessing whether such a data collapsing is allowed are ad hoc and require making full model assumptions.
-We present visual diagnostic tools for investigating differences in structure for two datasets containing different observations of the same variables.
-This approach, referred to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC), requires no directional nor hierarchical assumptions on the data and consists of three plots based on Principal Component Analysis (PCA), that address differences in the structures of the covariance matrices of the two datasets to yield an  understanding of where differences occur and why so.
-The functionality of the PCADSC tools is presented by use of data from the European Social Survey on psychological well-being in three European countries, namely Denmark, Sweden and Bulgaria.
-Here, we find that the concepts of psychological well-being are not the same when comparing Denmark and Bulgaria, and thus a comparison of e.g.
-mean scores in different well-being aspects is not meaningful.
-However, when comparing the two Scandinavian countries, Denmark and Sweden, we find very similar data structures and thus very similar concepts of well-being.
-Therefore, inter-country comparisons are warranted for these two countries.
+Datasets are sometimes divided into distinct subsets, e.g.
+due to multi-center sampling, or to variations in instruments, questionnaire item ordering or mode of administration, and the data analyst then needs to assess whether a joint analysis is meaningful.
+The Principal Component Analysis-based Data Structure Comparisons (PCADSC) tools are three new non-parametric, visual diagnostic tools for investigating differences in structure for two subsets of a dataset through covariance matrix comparisons by use of principal component analysis.
+The PCADCS tools are demonstrated in a data example using European Social Survey data on psychological well-being in three countries, Denmark, Sweden and Bulgaria.
+The data structures are found to be different in Denmark and Bulgaria, and thus a comparison of e.g.
+mean psychological well-being scores is not meaningful.
+However, when comparing Denmark and Sweden, very similar data structures, and thus comparable concepts of well-being, are found.
+Therefore, inter-country comparisons are warranted for these countries.
 \end{abstract}
 
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\date{}
-
 %% Include all macros below
-
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
 
@@
 Data comparability is a classical topic in statistics and a recurring one in psychometrics.
-Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability (and thus collapsibility) needs to be assessed empirically.
+Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability needs to be assessed empirically.
 This happens e.g.
@@
 All of these practices are essential parts of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions.
-However, if we do not address the question of collapsibility among the data subsets arriving from e.g.
+However, if we do not address the question of comparability among the data subsets arriving from e.g.
 different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.
@@
 But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data.
-The use of (parametric) models does thus not constitute a %valid
-general data structure comparison method, but rather a fitted-model comparison method.
+The use of (parametric) models does thus not constitute a general data structure comparison method, but rather a fitted-model comparison method.
 It addresses the interplay between the model and the data, not the data alone.
 However, useful tools for initial, exploratory investigations of data comparability that help researchers identify potential problems or challenges early in the study design development are mostly absent.
-What is needed is a procedure that compares differences in overall data structures in two subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
-Simple methods like variable-by-variable tests of distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables, but entry-by-entry comparisons of two empirical correlation matrices quickly become unmanageable as the number of variables increase.
+What is needed are procedures that compares differences in data structures in two subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
+This paper suggests procedures for linear data structures Simple methods like variable-by-variable tests of distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables, but entry-by-entry comparisons of two empirical correlation matrices quickly become unmanageable as the number of variables increase.
 
@@
 We propose three visual tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC).
-These methods use principal component decomposition of the empirical covariance matrix in the two sub sets to create intuitive visualizations of data structure differences.
+These methods use principal component decomposition of the empirical covariance matrix in the two subsets to create intuitive visualizations of data structure differences.
 The proposed tools are implemented in the \texttt{R} package \texttt{PCADSC} \cite{PCADSC}.
@@
 
-\subsection*{A brief introduction to principal component analysis}
-\label{sec:introPCA}
-Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = {(\bar{x}_1, ..., \bar{x}_d)}^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) {(x_i-\bar{x})}^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$.
-We assume in the following that all variables of $X$ have a numerical interpretation, e.g.
-by being continuous or ordinal categorical.
-For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost.
-More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the original $d$-dimensional one.
-Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
-
-\begin{equation}
-\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
-\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
-\end{equation}
-
-The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \cite{HastieEtAl2009}.
-More specifically, the rank-q-reconstruction error is attained when we choose
-
-\begin{equation*}
-\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\},
-\end{equation*}
-where $\eta_1, ..., \eta_q$ are the first $q$ eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$.
-In the PCA framework, we refer to the eigenvectors as \textit{loadings}.
-The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
-\begin{equation*}
-\text{trace}(S) = \sum_{j=1}^d \hat{V}(x_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
-\end{equation*}
-where $x_j \in \RR^n$ denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
-This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
-And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are %different,
-distinct, the representation hereby obtained is a valid object of inference.
-If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
-
-The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
-\begin{align*}
-\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
-\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2,
-\end{align*}
-where the initial subspace is defined as $\hat{K}_0 = \{0\}$.
-It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
-This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$.
-This implies that the loadings and the variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.
-
 \subsection*{Using PCA for data structure comparisons}
 
-All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets.
-Usually, in PCA, main interest lies in the \textit{scores}, i.e.
-the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings, but the variation components and the accompanying loadings also describe aspects of the data.
+PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets.
+We consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ and let $\bar{x} = {(\bar{x}_1, ..., \bar{x}_d)}^\top = \frac{1}{n} \sum_{i=1}^n x_n$ and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) {(x_i-\bar{x})}^\top \in \RR^{d \times d}$ denote their averages and the empirical covariance matrix of the full data matrix, respectively.
+A brief introdution to CFA is included in the supplementary material.
+
+Usually, in PCA, the main interest lies in the \textit{scores}, i.e.
+the coordinates $\eta_j^\top (x_i - \bar{x})$ of the projections observations onto the loadings, but the variation components and the accompanying loadings also describe aspects of the data.
 If two different datasets with the same variables, but different samples of observations, have similar loadings, then the variables can be argued to measure the same underlying quantities in both datasets.
@@
 The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables.
-We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the apparently same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$.
+We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$.
 %
@@
 \item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e.
-loading and variance component pairs) that are most similar and most different across the two datasets
+loading and variance component pairs) are most similar and most different across the two datasets
 \item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
@@
 
-Note that the standardization implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
+Note that the standardization implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equal 1, and thus also that
 \begin{equation*}
@@
 In \textit{Model 1} the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable $X_1$.
-Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models.
+Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models (S1).
 Further details about data simulation is also available in the supplementary material.
@@
 
-\begin{figure}[!h]
-\caption{[FIGURE 1 HERE]}
-% Graphical representations of models used for simulating example data.
+\begin{figure}[h]
+\caption{[FIGURE 1 ABOUT HERE].
+Graphical representations of models used for simulating example data.
 Square nodes represent observed variables, while rounded nodes represent latent variables.
-Arrows are used to illustrate causal links.
+Arrows are used to illustrate causal links.}
 \label{fig:simGraphs}
@@
 Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does.
-Two examples of cumulative eigenvalue plots are available in Figure \ref{fig:plot.simCE}.
+Two examples of cumulative eigenvalue plots are available in Figure \ref{plot.simCE}.
 
-\begin{figure}[!h]
-\caption{[FIGURE 2 HERE]}
-\label{plot.simCE}
-% CE plots for the simulated datasets.
+\begin{figure}[h]
+\caption{[FIGURE 2 ABOUT HERE] CE plots for the simulated datasets.
 Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel.
 Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures.
-The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.
+The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.}
+\label{plot.simCE} 
 \end{figure}
@@
 
-\begin{figure}[!h]
-\caption{[FIGURE 3 HERE]}
-% Angle plots for the simulated datasets.
+\begin{figure}[h]
+\caption{[FIGURE 3 ABOUT HERE] Angle plots for the simulated datasets.
 Dataset A in top panel and dataset B in bottom panel.
-Blue arrows show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.
+Blue arrows show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.}
 \label{plot.simAngle}
@@
 
-\begin{figure}[!h]
-\caption{[FIGURE 4 HERE]}
-% Chroma plots for the simulated datasets.
+\begin{figure}[h]
+\caption{[FIGURE 4 ABOUT HERE] Chroma plots for the simulated datasets.
 Dataset A in top panel and dataset B in bottom panel.
-Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.
+Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.}
 \label{plot.simChroma}
@@
 These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}.
-More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in the supplementray material.
+More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in the supplementray material (S2).
 We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10.
@@
 
-\begin{table}[!ht]
+\begin{table}[ht]
 \centering
+\caption{\textbf{Median values of each of the six dimensions of psychological well-being, stratified by country.} 25 and 75 percentiles are listed in parentheses.
+The scales are constructed such that they all run from 0-10.}
+\label{tableDistr}
+
 \begin{tabular}{lccc}
@@
 \end{tabular} 
-\caption{Median values of each of the six dimensions of psychological well-being, stratified by country.
-25 and 75 percentiles are listed in parentheses.
-Note that the scales are constructed such that they all run from 0-10.}
-\label{tableDistr}
 \end{table}
@@
 
-\begin{figure}[!h]
-\caption{[FIGURE 5 HERE]}
-% CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being.
+\begin{figure}[h]
+\caption{[FIGURE 5 ABOUT HERE] CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being.
 The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
-In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.
+In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
 \label{plotBG.cehair}
@@
 
-\begin{figure}[!h]
-\caption{[FIGURE 6 HERE]}
-% Chroma plot for comparing Danish and Bulgarian data on psychological well-being.
-A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data.
+\begin{figure}[h]
+\caption{[FIGURE 6 ABOUT HERE] Chroma plot for comparing Danish and Bulgarian data on psychological well-being.
+A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgarian- and Danish psychological well-being data.
 The component-bars are annotated with their relative variance contributions.
+}
 \label{plotBG.pancake}
@@
 The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalues) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle- and chroma plots).
-In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, corresponding well with the previous findings mentioned above.
+In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristics play a relatively larger role, corresponding well with the previous findings mentioned above.
 Thus, the datasets are fundamentally different and we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics \cite{ESStopline5}.
@@
 
-\begin{figure}[!h]
-\caption{[FIGURE 7 HERE]}
-% A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being.
-The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.
+\begin{figure}[h]
+\caption{[FIGURE 7 ABOUT HERE] A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being.
+The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
 \label{plotSE.cehair}
@@
 
-\begin{figure}[!h]
-\caption{[FIGURE 8 HERE]}
-% Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples.
-For each component, the bar is annotated with its cumulative variance contribution, that is, how much variance can be explained by having information of this and the preceding components.
+\begin{figure}[h]
+\caption{[FIGURE 8 ABOUT HERE] Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples.
+For each component, the bar is annotated with its cumulative variance contribution, that is, how much variance can be explained by having information of this and the preceding components.}
 \label{plotSE.pancake}
@@
 
-All in all, PCADSC represents a first step towards addressing a known issue that has otherwise been met with unsatisfactory ad-hoc methods whose assumptions rest upon the very hypothesis they are testing.
+All in all, the PCADSC methods represents a first step towards addressing a known issue that has otherwise been met with unsatisfactory ad-hoc methods whose assumptions rest upon the very hypothesis they are testing.
 The PCADSC tools perform quite well in the data example provided here, yielding conclusions that are in accordance with well-established theories and they are simple to use and test further using the PCADSC \texttt{R} package.
 
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for
-% step-by-step instructions.
-%
+\section*{Conclusion}
+
+We have presented three new tools for visual, exploratory data structure comparisons that utilizes covariance matrix decompositions obtained by use of principal component analysis.
+In the provided data example, comparing data on personal happiness across three European countries, the PCADSC tools were able to identify both comparable datasets and non-comparable datasets, as determined according to existing subject-matter literature.
+Moreover, the tools were useful for reaching meaningful interpretations concerning the nature of the identified non-comparability.
+
+\bibliographystyle{apalike}
+\bibliography{bib}
 
@@
 
-% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Appendix.}
+\subsection*{S0 Appendix: A brief introduction to principal component analysis}
+
+\label{appendix.introPCA}
+Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = {(\bar{x}_1, ..., \bar{x}_d)}^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) {(x_i-\bar{x})}^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$.
+We assume in the following that all variables of $X$ have a numerical interpretation, e.g.
+by being continuous or ordinal categorical.
+For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost.
+More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the original $d$-dimensional one.
+Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
+
+\begin{equation}
+\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
+\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
+\end{equation}
+
+The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \cite{HastieEtAl2009}.
+More specifically, the rank-q-reconstruction error is attained when we choose
+
+\begin{equation*}
+\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\},
+\end{equation*}
+where $\eta_1, ..., \eta_q$ are the first $q$ eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$.
+In the PCA framework, we refer to the eigenvectors as \textit{loadings}.
+The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
+\begin{equation*}
+\text{trace}(S) = \sum_{j=1}^d \hat{V}(x_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
+\end{equation*}
+where $x_j \in \RR^n$ denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
+This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
+And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are %different,
+distinct, the representation hereby obtained is a valid object of inference.
+If the covariance matrix does not have full rank, which is the case in particular when $n < d$, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
+
+The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
+\begin{align*}
+\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
+\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2,
+\end{align*}
+where the initial subspace is defined as $\hat{K}_0 = \{0\}$.
+It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
+This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$.
+This implies that the loadings and the variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.
+
+\subsection*{S1 Appendix: Details about the simulated data}
 \label{appendix.simData}
-{\bf Details about the simulated data.}
+This supplementary appendix describes the origin of the simulated data used for presenting the PCADSC plots.
+Dataset A consisted of 1000 independent simulated realizations from the $N(0, \Sigma_1)$-distribution, while dataset B consisted of 500 independent realizations from $N(0, \Sigma_1)$ and 500 independent realizations from $N(0, \Sigma_2)$.
+Dataset A was furthermore appended with a grouping variable, randomly dividing the observations into two groups.
+Dataset B also included a grouping variable, and this variable contained information about which of the two normal distributions each observation was simulated from.
+The covariance matrices, $\Sigma_1$ and $\Sigma_2$, were defined by
 
-\paragraph*{S1 Table.}
+$$\Sigma_1 = \begin{pmatrix}
+    1.0  &  0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
+ 0.0  &  1.0 & 0.0 & 0.0&  0.0 & 0.0 \\
+ 0.0   & 0.0 &  1.0 & 0.7 & 0.0 & 0.0 \\
+ 0.0 &  0.0 & 0.7 & 1.0 & 0.0 & 0.0 \\
+0.0 &    0.0 &  0.0 & 0.0 & 1.0 & 0.4 \\
+ 0.0 & 0.0 &  0.0 & 0.0 & 0.4 & 1.0
+\end{pmatrix}$$
+and
+$$
+\Sigma_2 = \begin{pmatrix}
+ 1.0 & 0.2 & 0.1 & 0.0 & 0.0 & 0.0 \\
+ 0.2 & 1.0 & 0.1 & 0.0 & 0.0 & 0.0 \\
+ 0.1 & 0.1 & 1.0 & 0.0 & 0.0 & 0.0 \\
+ 0.0 & 0.0 & 0.0 & 1.0 & 0.3 & 0.1 \\
+ 0.0 & 0.0 & 0.0 & 0.3 & 1.0 & 0.2 \\
+ 0.0 & 0.0 & 0.0 & 0.1 & 0.2 & 1.0
+\end{pmatrix},
+$$
+respectively.
+
+\subsection*{S2 Table: ESS psychological wellbeing scale definitions}
+\begin{table}[h]
+\centering
+\caption{{\bf ESS psychological wellbeing scale definitions.} Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
+Note that before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
 \label{table:items}
-{\bf ESS psychological wellbeing scale definitions.} Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
-Note that before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.
 
-\newpage
+\scriptsize
+\rowcolors{1}{}{lightgray}
+\bgroup
+\def\arraystretch{1.5}
+\begin{tabular}{ll}
+\hline
+Scale & Items \\
+\hline
+\rowcolor{lightgray} & How satisfied with life as a whole \\
+\rowcolor{lightgray}\multirow{-2}{*}{Evaluative wellbeing }& How happy are you\\
 
-\bibliographystyle{apalike}
-\bibliography{bib}
+\rowcolor{white} & Felt sad, how often in the past week\\
+\rowcolor{white} & Felt depressed, how often in the past week\\
+\rowcolor{white} & Enjoyed life, how often in the past week\\
+\rowcolor{white} & Were happy, how often in the past week\\
+\rowcolor{white} & Felt anxious, how often in the past week\\
+\rowcolor{white}\multirow{-6}{*}{Emotional wellbeing} & Felt calm and peaceful, how often in the past week\\
+
+\rowcolor{lightgray} & Free to decide how to live my life\\
+\rowcolor{lightgray} & Little chance to show how capable I am\\
+\rowcolor{lightgray} & Feel accomplishment from what I do\\
+\rowcolor{lightgray} & Interested in what you are doing\\
+\rowcolor{lightgray} & Absorbed in what you are doing\\
+\rowcolor{lightgray} & Enthusiastic about what you are doing\\
+\rowcolor{lightgray} & Feel what I do in life is valuable and worthwhile\\
+\rowcolor{lightgray} & Have a sense of direction\\
+\rowcolor{lightgray} & Always optimistic about my future\\
+\rowcolor{lightgray} & There are lots of things I feel I am good at\\
+\rowcolor{lightgray} & In general feel very positive about myself\\
+\rowcolor{lightgray} & At times feel as if I am a failure\\
+\rowcolor{lightgray} & When things go wrong in my life it takes a long time to get back to normal\\
+\rowcolor{lightgray}\multirow{-14}{*}{Functioning} & Deal with important problems\\
+
+\rowcolor{white} & Felt everything did as effort, how often in the past week\\
+\rowcolor{white} & Sleep was restless, how often in the past week\\
+\rowcolor{white} & Could not get going, how often in the past week\\
+\rowcolor{white}\multirow{-4}{*}{Vitality} & Had a lot of energy, how often in the past week\\
+
+\rowcolor{lightgray} &  Most people can be trusted\\
+\rowcolor{lightgray} &  People try to take advantage\\
+\rowcolor{lightgray} &  Most of the time people are helpful\\
+\rowcolor{lightgray} &  Feel people in local area help one another\\
+\rowcolor{lightgray}\multirow{-5}{*}{Community wellbeing}& Feel close to the people in local area\\
+
+\rowcolor{white} & How many with whom you can discuss intimate matters\\
+\rowcolor{white} & Feel appreciated by those you are close to\\
+\rowcolor{white} & Receive help and support\\
+\rowcolor{white}\multirow{-4}{*}{Supportive relationships} & Felt lonely, how often in the past week\\
+\hline
+\end{tabular}
+\egroup
+\end{table}
 \end{document}

### 2018-05-21 R/CEPlot.R
Added info about ordering of datasets to the plots

@@ -82,6 +82,7 @@ CEPlot <- function(x, nDraw = NULL) {
   CvM.pvalue <- obj$CvM.pvalue
   B <- obj$B
   d <- obj$d
+  splitLevels <- obj$splitLevels
 
   #Make CE plot
   if(is.null(nDraw)) nDraw <- min(B, 20)
@@ -115,7 +116,9 @@ CEPlot <- function(x, nDraw = NULL) {
                x = -Inf, y = Inf, hjust = "left", vjust = "top",
                label.r = unit(0, "lines"),
                label.padding = unit(1, "lines")) +
-      annotate(geom = "line", x = xVals, y = y.obs, size = 1)
+      annotate(geom = "line", x = xVals, y = y.obs, size = 1) +
+      ggtitle(paste("Cumulative eigenvalue difference:", splitLevels[1],
+                    "-", splitLevels[2]))
 }
 
 

### 2018-05-21 R/doCE.R

@@ -48,6 +48,7 @@ doCE <- function(x, ...) {
 #x: pcaRes
 #' @export
 doCE.pcaRes <- function(x, data, B, ...) {
+  splitLevels <- x$splitLevels
 
   #check whether B is positive
   if (B < 1) {
@@ -91,7 +92,8 @@ doCE.pcaRes <- function(x, data, B, ...) {
   #Pack and return output
   out <- list(d = d, B = B, xVals = xVals, y.obs = y.obs, y.sim = y.sim,
               KS.obs = KS.obs, KS.pvalue = KS.pvalue,
-              CvM.obs = CvM.obs, CvM.pvalue = CvM.pvalue)
+              CvM.obs = CvM.obs, CvM.pvalue = CvM.pvalue,
+              splitLevels = splitLevels)
   class(out) <- "CEInfo"
   out
 }

### 2018-05-21 article/Old versions/MPS/PCADSC_article.tex
Changed manuscript according to reviewer comments (everything but component permutation issue) and styleguide.

@@
 
+\usepackage{tikz}
+\usetikzlibrary{arrows}
+\usetikzlibrary{shapes, positioning, calc}
+
+\usepackage{float}
+
 %\DisableLigatures[f]{encoding = *, family = * }
@@
 
+\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
 \renewcommand{\baselinestretch}{2}
@@
 The Principal Component Analysis-based Data Structure Comparisons (PCADSC) tools are three new non-parametric, visual diagnostic tools for investigating differences in structure for two subsets of a dataset through covariance matrix comparisons by use of principal component analysis.
-The PCADCS tools are demonstrated in a data example using European Social Survey data on psychological well-being in three countries, Denmark, Sweden and Bulgaria.
-The data structures are found to be different in Denmark and Bulgaria, and thus a comparison of e.g.
-mean psychological well-being scores is not meaningful.
+The PCADCS tools are demonstrated in a data example using European Social Survey data on psychological well-being in three countries, Denmark, Sweden, and Bulgaria.
+The data structures are found to be different in Denmark and Bulgaria, and thus a comparison of for example mean psychological well-being scores is not meaningful.
 However, when comparing Denmark and Sweden, very similar data structures, and thus comparable concepts of well-being, are found.
@@
 Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability needs to be assessed empirically.
-This happens e.g.
-when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration or the order of survey questionnaire items are applied.
-All of these practices are essential parts of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions.
+This happens for instance when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration, or the order of survey questionnaire items are applied.
+While a mixture of modes of administration, for example mail and telephone, can improve response rates in surveys, it can also be quite problematic by inducing differences in response behavior that may lead to biased results \cite{Brambilla1987,McHorney1994}.
+Powers, Mishra, and Young report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful \cite{Powers2005}, so the comparability issue is not just a statistical puzzle.
+Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu (2016), who warns against combining online panel data with intercept samples (a pool of respondents obtained through banners, ads, or promotions).\nocite{Liu2016} But mixing several data collection methods is an essential part of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions.
 However, if we do not address the question of comparability among the data subsets arriving from e.g.
 different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.
+
+The comparability issue in these examples can be summarized as follows: Assume that we have two datasets with the same variables, but different observations, often represented as a single dataset with a subset-inducing variable, and that we wish to compare them without specifying a model, or even a variable of interest.
+The central question is then whether the two datasets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be dealt with in later statistical modelling.
 
@@
 It addresses the interplay between the model and the data, not the data alone.
-However, useful tools for initial, exploratory investigations of data comparability that help researchers identify potential problems or challenges early in the study design development are mostly absent.
-What is needed are procedures that compares differences in data structures in two subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
-This paper suggests procedures for linear data structures Simple methods like variable-by-variable tests of distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables, but entry-by-entry comparisons of two empirical correlation matrices quickly become unmanageable as the number of variables increase.
+However, useful tools for initial, exploratory investigations of data comparability %that help researchers identify potential problems or challenges early in the study design development
+are mostly absent.
+What is needed are procedures that compare differences in data structures in two subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.
+Simple methods like variable-by-variable tests of distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables, but entry-by-entry comparisons of two empirical correlation matrices quickly become unmanageable as the number of variables increase.
 
-While the issue is not in any way new, its relative importance is increasing due to new developments in data availability and data collection.
+%While the issue is not in any way new, its relative importance is increasing due to new developments in data availability and data collection.
 Classical statistical methods are generally aimed at analyzing data from designed experiments and historically, statistical analyses have been conducted by researchers who knew the design and the source of the dataset well.
@@
 
-The data structure comparison problem is however not limited to the context of new data sharing platforms.
-Even in single-population survey studies, a mixture of modes of administration is often used, e.g.
-mail and telephone, and while this can improve response rates, it can also be quite problematic \cite{Brambilla1987,McHorney1994}, and differences in response behavior can lead to biased results.
-Powers, Mishra and Young report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful \cite{Powers2005}, so the comparability issue is not just a statistical puzzle.
-Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu (2016), who warns against combining online panel data with intercept samples (a pool of respondents obtained through banners, ads, or promotions).\nocite{Liu2016}
-
-The comparability issue in these examples can be summarized as follows: Assume that we have two datasets with the same variables, but different observations (often represented as a single dataset with a subset-inducing variable) and that we wish to compare them without specifying a model, or even a variable of interest.
-The central question is then whether the two datasets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be dealt with in later statistical modelling.
-
-We propose three visual tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC).
+We propose three visual tools for data structure comparisons, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC).
 These methods use principal component decomposition of the empirical covariance matrix in the two subsets to create intuitive visualizations of data structure differences.
+The use of PCA implies that we can only compare datasets with numerical variables and that we focus on the linear aspect of the  data structures.
 The proposed tools are implemented in the \texttt{R} package \texttt{PCADSC} \cite{PCADSC}.
-
-This manuscript describes the procedures, including a brief introduction to principal component analysis (PCA), and presents a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey.
-More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is completely ignored e.g.
+Below, we describe the procedures and present a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey.
+More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is ignored e.g.
 when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project \cite{WHR2016}.
@@
 
-PCADSC compares the covariance matrices of the two subsets of a dataset.
-If all variables in the datasets are jointly normal with known means, the covariance matrices are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing interrelations between variables.
+PCADSC compares the covariance matrices of two subsets of a dataset.
+If all variables in the subsets are jointly normal with known means, the covariance matrices are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing linear interrelations between variables.
 This makes the empirical covariance matrix a reasonable place to start looking for differences in data structures.
-Computing empirical covariance matrices for the two subsets and comparing these entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets.
+Computing empirical covariance matrices for the two subsets and comparing them entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets.
+We refer to Appendix S1 for a brief introduction to PCA.
 
-At this point a general remark about standardization is in place.
-PCA deconstructs the covariance matrix in components according to the most explained variance.
-This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will always be deemed highly influential, no matter the structure of the data.
-Therefore, the variables are typically standardized prior to performing PCA.
-Note that the covariance matrix of the standardized variables is the same as the correlation matrix of the original variables, so post-standardization PCA simply corresponds to performing data structure comparisons of the correlation matrices rather than the covariance matrices.
-The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation.
-In this paper we will adopt this standardization policy, and in particular, we will allow for differences between scalings of the original, raw variables in the two data subsets.
+We assume that the variables have been standardized before PCA is conducted, which is equivalent to working with the empirical correlation matrices rather than the empirical covariance matrices.
+We also assume that all considered variables have a numerical interpretation, for example by being continuous or ordinally categorical.
+Moreover, we use the following notation:
+$\mathbf{X} \in \RR^{n_x \times d} $ and $\mathbf{Y} \in \RR^{n_y \times d} $ are datasets containing the same number of variables, $d$, but possibly different numbers of observations, $n_x$ and $n_y$.
+We use $X_j$ to refer to the $j$th variables of $\mathbf{X}$ and $x_{ij}$ to refer to the $i$th observation  within that variable, while $x_{i \cdot}$ is the full $i$th observation row.
+Note that $X_j \in \RR^{n_x}$, $x_{ij} \in \RR$, and $x_{i \cdot} \in \RR^d$.
+We let $\bar{X} = (\bar{X}_1, ...
+, \bar{X}_d)^T = \frac{1}{n_x} \sum_{i = 1} x_{i \cdot}$ denote the variable averages and use $S_x =  \frac{1}{n_x - 1} \sum_{i = i}^n (x_{i \cdot} - \bar{X}) - (x_{i \cdot} - \bar{X})^T \in \RR^{d \times d}$ to denote the empirical covariance matrix of $\mathbf{X}$.
 
 \subsection*{Using PCA for data structure comparisons}
+%PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and eigenvalues are meaningful and interesting objects to compare across such different datasets.
 
-PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets.
-We consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ and let $\bar{x} = {(\bar{x}_1, ..., \bar{x}_d)}^\top = \frac{1}{n} \sum_{i=1}^n x_n$ and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) {(x_i-\bar{x})}^\top \in \RR^{d \times d}$ denote their averages and the empirical covariance matrix of the full data matrix, respectively.
-A brief introdution to CFA is included in the supplementary material.
-
-Usually, in PCA, the main interest lies in the \textit{scores}, i.e.
-the coordinates $\eta_j^\top (x_i - \bar{x})$ of the projections observations onto the loadings, but the variation components and the accompanying loadings also describe aspects of the data.
-If two different datasets with the same variables, but different samples of observations, have similar loadings, then the variables can be argued to measure the same underlying quantities in both datasets.
-Similar loadings can occur in situations where the two sets of scores are different, e.g.\ if the two datasets come from two different populations of subjects.
-On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.
-
-The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables.
-We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$.
-%
-Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For these %two
-three datasets, we complete the following steps:
+The tools presented here are all based on comparing the PCA results across two different datasets that contain the same variables, $\mathbf{X}$ and $\mathbf{Y}$. Let 
+$$\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{n_x + n_y \times d}$$
+ be the combined dataset. For each of these three datasets, we complete the following steps (here described for $\mathbf{X}$ only):
 \begin{enumerate}
-\item Within each dataset, standardize each of the variables to have mean zero and unit standard deviation.
-Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ and $\tilde{x}_j \in \RR^d$ for $j=1,\dotsc,n$ be the standardized datasets.
+\item Standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{\mathbf{X}}_{j} \in \RR^{n_x \times d}$ be the standardized dataset.
 \item Form the principal component analyses
 \begin{align*}
-S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} \\
-S &= \frac{1}{n-1} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top,
+S_x &= \frac{1}{n_x-1} \sum_{i=1}^{n_x} \tilde{x}_{i \cdot} \tilde{x}_{i \cdot}^\top = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top 
 \end{align*}
-thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda_k$ for $k=1,\dotsc,d$.
+thereby obtaining loadings $\eta_{j}^x$ and eigenvalues $\lambda^x_j$ for $j=1,\dotsc,d$.
 \end{enumerate}
 The hereby obtained PCA decompositions of the correlation matrices can then be compared.
-We present three diagnostic plots that are designed to shine a light on different types and levels of data structure differences.
-These three plots are:
-\begin{description}
-\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA.
-The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
-\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e.
-loading and variance component pairs) are most similar and most different across the two datasets
-\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
-\end{description}
-For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.
-
-Note that the standardization implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equal 1, and thus also that
+The standardization implies that the diagonal elements of $S_x$, $S_y$, and $S_z$ all equal 1, and thus also that
 \begin{equation*}
-\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d.
+\sum_{j=1}^d \lambda_j^x = \sum_{j=1}^d \lambda_{j}^y = \sum_{j=1}^d \lambda_{j}^z =  d.
 \end{equation*}
 This identity will simplify some expressions below.
+Using a greedy interpretation of PCA, which is often overlooked in the literature, we can note that the sequence of loadings $\eta_j$ and their associated eigenvalues $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$.
+This implies that the loadings and the eigenvalues can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.
 
-While we present the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions.
-These plots are available in Figures \ref{plot.simCE}, \ref{plot.simAngle} and \ref{plot.simChroma} for the CE-, angle- and chroma plots, respectively, and they are based on two simulated datasets:
+We present three diagnostic plots that are designed to shine a light on different types and levels of data structure differences, namely the \textit{cumulative eigenvalue (CE) plot}, the \textit{angle plot}, and the \textit{chroma plot}.
+These plots are dsecribed in turn below.
+For the deepest understanding of the data structure differences in two datasets, we suggest using all of the three plots in the same order as they are presented.
+While we describe the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions.
+These plots are available in Figures \ref{plot.simCE}, \ref{plot.simAngle}, and \ref{plot.simChroma} for the CE-, angle-, and chroma plots, respectively, and they are based on two simulated datasets:
 \begin{description}
@@
 
-In \textit{Model 1} the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable $X_1$.
-Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models (S1).
-Further details about data simulation is also available in the supplementary material.
-In dataset A, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures.
-In dataset B, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.
+In \textit{Model 1} the observed variables $V_3$ and $V_4$ are caused by a common, unobserved variable $L_1$.
+Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models, in Appendix S2.
+In dataset $\mathbf{A}$, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures.
+In dataset $\mathbf{B}$, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.
 
-\begin{figure}[h]
-\caption{[FIGURE 1 ABOUT HERE].
+%\begin{figure}[h]
+%\caption{[FIGURE 1 ABOUT HERE].
 Graphical representations of models used for simulating example data.
-Square nodes represent observed variables, while rounded nodes represent latent variables.
+Square nodes represent observed variables, while %rounded nodes represent latent variables.
 Arrows are used to illustrate causal links.}
-\label{fig:simGraphs}
-\end{figure}
+%\label{fig:simGraphs}
+%\end{figure}
 
-\subsubsection*{The cumulative eigenvalue plot}
-The cumulative eigenvalue (CE) plot compares the variation components, i.e.
-the eigenvalues of the correlation matrix.
+\subsection*{The cumulative eigenvalue plot}
+The cumulative eigenvalue (CE) plot compares the eigenvalues of the correlation matrices.
 These eigenvalues represent how much information is withheld in each component in terms of explained variance.
-Thus, by comparing cumulative sums of variance components, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative.
+Thus, by comparing cumulative sums of eigenvalues, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative.
 In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
@@
 &(0,0), \;
-(\lambda_1,\lambda_{11}-\lambda_{21}), \;
-(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), \;
+(\lambda_1^z,\lambda_{1}^x-\lambda_{1}^y), \;
+(\lambda_1^z + \lambda_2^z,\lambda_{1}^x+\lambda_{2}^x-\lambda_{1}^y-\lambda_{2}^y), \;
 \ldots, \\
-&\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
+&\bigg( \sum_{j=1}^d \lambda_j^z, \sum_{j=1}^d \lambda_{j}^x - \sum_{j=1}^d \lambda_{j}^y \bigg)
 \end{align*}
-Note that due to the standardization, the last point will always be equal to $(d,0)$.
+Due to the standardization, the last point will always be equal to $(d,0)$.
 Thus, this curve will begin and end at the x-axis.
-And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are.
-Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does.
-Two examples of cumulative eigenvalue plots are available in Figure \ref{plot.simCE}.
+And the larger excursions it makes away from the  first axis, the less alike the cumulative eigenvalue sums for the two datasets are.
+Moreover, a positive cumulative difference implies that dataset $\mathbf{X}$ holds more information in the first components than dataset $\mathbf{Y}$ does.
+By using the cumulative eigenvalues of $\mathbf{Z}$ as the first coordinate, we obtain a visualization where a component take up an amount of horizontal space that corresponds to the variance explained by the component (that is, its eigenvalue), thereby letting the more infuential components drive the visual impression.
 
-\begin{figure}[h]
-\caption{[FIGURE 2 ABOUT HERE] CE plots for the simulated datasets.
-Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel.
+%\begin{figure}[h]
+%\caption{[FIGURE 2 ABOUT HERE] CE plots for the simulated datasets.
+Dataset A (with no differences in data structure) in top panel, and dataset B %(where observations stem from different models) in bottom panel.
 Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures.
 The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.}
-\label{plot.simCE} 
-\end{figure}
+%\label{plot.simCE} 
+%\end{figure}
 
+Two examples of cumulative eigenvalue plots are available in Figure \ref{plot.simCE}.
 In the top panel, we see a cumulative eigenvalue curve that falls well within the confidence region, thereby suggesting no difference in eigenvalues (as expected).
@@
 \begin{align*}
-\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
-\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} {\bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)}^2.
+\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{j}^x - \sum_{j=1}^k \lambda_{j}^y \bigg\rvert, &
+\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k^z + \lambda_{k+1}^z}{2} {\bigg( \sum_{j=1}^k \lambda_{j}^x - \sum_{j=1}^k \lambda_{j}^y \bigg)}^2.
 \end{align*}
-We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_1$ and $n_2$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics.
+We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_x$ and $n_y$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics.
 This should be done a large (e.g.
@@
 
-\subsubsection*{The angle plot}
-This plot simultaneously compares the variation components and the loadings.
-Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets.
-Then the empirical correlation matrix for the first dataset, $S_1$, has the following orthogonal decomposition in the coordinate system of the second dataset
+\subsection*{The angle plot}
+The angle plot compares both loadings and eigenvalues at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e.
+loading and eigenvalue pairs) are most similar and most different across the two datasets.
+Let $\lambda_{\max} = \max\{ \lambda_{1}^x, \lambda_{1}^y \}$ be the largest eigenvalue for the two datasets.
+Then the empirical correlation matrix for the $\mathbf{X}$ dataset, $S_x$, has the following orthogonal decomposition in the coordinate system of the $\mathbf{Y}$ dataset
 \begin{equation*}
-S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
+S_x = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top
 = \lambda_{\max} \sum_{k=1}^d
-\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
-{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)}^\top,
+\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)
+{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)}^\top,
 \end{equation*}
-and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset.
+and we have a similar decomposition of $S_y$ in the coordinate system of the $\mathbf{X}$ dataset.
 We propose to visualize these two decompositions in a $d \times d$ grid display.
@@
 To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows.
-The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
+The lengths $\mu_{jk}$ and $\nu_{jk}$, and the angle $\theta_{jk}$, are given by
 \begin{align*}
-\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
-\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
-\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
+\mu_{jk} &= \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert, &
+\nu_{jk} &= \sqrt{\frac{\lambda_{j}^y}{\lambda_{\max}}} \lvert {\eta_{j}^y}^\top \eta_{k}^x \rvert, &
+\theta_{jk} &= \acos(\lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert).
 \end{align*}
-Note that for two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the sample correlation.
-Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance component contributions.
-The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors.
-This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal.
+For two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the empirical correlation function.
+Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance contributions.
+The absolute value of the projection $\eta_{k}^x {\eta_{j}^y}^\top$ is inserted due to the indeterminacy of the direction of loading vectors.
+This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal.
 Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.
@@
 
-\begin{figure}[h]
-\caption{[FIGURE 3 ABOUT HERE] Angle plots for the simulated datasets.
+%\begin{figure}[h]
+%\caption{[FIGURE 3 ABOUT HERE] Angle plots for the simulated datasets.
 Dataset A in top panel and dataset B in bottom panel.
 Blue arrows show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.}
-\label{plot.simAngle}
-\end{figure}
+%\label{plot.simAngle}
+%\end{figure}
 
-For dataset A, we find very short off-diagonal arrows, suggesting that the two PCA decompositions agree on the relative placement of the information in  the data among the 6 PCs.
-For dataset B, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.
+\hl{For dataset $\mathbf{A}$, we find very short off-diagonal arrows, suggesting that the two PCA decompositions agree on the relative placement of the information in  the data among the 6 PCs.
+For dataset $\mathbf{B}$, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}
 
-In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the first dataset in the coordinate system of the second dataset.
-Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the second dataset in the coordinate system of the first dataset.
-If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
-Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
-And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.
+In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the $\mathbf{A}$ dataset in the coordinate system of the $\mathbf{B}$ dataset.
+Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the $\mathbf{B}$ dataset in the coordinate system of the $\mathbf{A}$ dataset.
+\hl{If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
+Differences in the eigenvalues are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
+And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.}
 
-\subsubsection*{The chroma plot}
-This plot compares the loadings of the two datasets.
+\subsection*{The chroma plot}
+The chroma plot is primarily an illustration of differences in loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
 The chroma plot consists of two panels, one for each dataset, made up of colored bars.
@@
 \begin{align*}
-\omega_{1ij} &= \frac{\lvert\eta_{1ij}\rvert}{\sum_{k=1}^d \lvert\eta_{1ik}\rvert}, &
-\omega_{2ij} &= \frac{\lvert\eta_{2ij}\rvert}{\sum_{k=1}^d \lvert\eta_{2ik}\rvert},
+\omega_{ij}^x &= \frac{\lvert\eta_{ij}^x \rvert}{\sum_{k=1}^d \lvert\eta_{ik}^x\rvert}, &
+\omega_{ij}^y &= \frac{\lvert\eta_{ij}^y\rvert}{\sum_{k=1}^d \lvert\eta_{ik}^y\rvert},
 \end{align*}
-where $\eta_{1ij}$ and $\eta_{2ij}$ denotes the $j$th entry of $\eta_{1i}$ and $\eta_{2i}$, respectively.
-Due to the indeterminacy of the sign, all the signs are removed from the coefficients in the loadings.
-The bars are ordered according to the variation components and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance component contributions,
+where $\eta_{ij}^x$ and $\eta_{ij}^y$ denotes the $j$th entry of $\eta_{i}^x$ and $\eta_{i}^y$, respectively.
+Due to the indeterminacy of the loading signs, all the signs are removed from the coefficients in the loadings.
+The bars are ordered according to the eigenvalues and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance contributions,
 \begin{align*}
-\tilde\sigma^c_{1i} &= \frac{\sum_{j=1}^i \lambda_{1j}}{\sum_{k=1}^d \lambda_{1k}} = \frac{\sum_{j=1}^i \lambda_{1j}}{d}, &
-\tilde\sigma^c_{2i} &= \frac{\sum_{j=1}^i \lambda_{2j}}{\sum_{k=1}^d \lambda_{2k}} = \frac{\sum_{j=1}^i \lambda_{2j}}{d}.
+\tilde\Sigma^x_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^x}{\sum_{k=1}^d \lambda_{k}^x} = \frac{\sum_{j=1}^i \lambda_{j}^x}{d}, &
+\tilde\Sigma^y_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^y}{\sum_{k=1}^d \lambda_{k}^y} = \frac{\sum_{j=1}^i \lambda_{j}^y}{d}.
 \end{align*}
-Especially when $d$ is large, we recommend plotting only a select set of interesting principal components, e.g.
-as identified by use of the angle plot.
-In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$.
+Especially when $d$ is large, we recommend plotting only a select set of interesting principal components (for instance identified by use of the angle plot).
+In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{i}^x = \frac{\lambda_{i}^x}{d}$ and $\tilde\sigma_{i}^y = \frac{\lambda_{i}^y}{d}$.
+
+%\begin{figure}[h]
+%\caption{[FIGURE 4 ABOUT HERE] Chroma plots for the simulated datasets.
+Dataset $\mathbf{A}$ in top panel and dataset $\mathbf{B}$ in bottom panel.
+Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.}
+%\label{plot.simChroma}
+%\end{figure}
+
 In Figure  \ref{plot.simChroma}, two examples of chroma plots are available.
-
-\begin{figure}[h]
-\caption{[FIGURE 4 ABOUT HERE] Chroma plots for the simulated datasets.
-Dataset A in top panel and dataset B in bottom panel.
-Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.}
-\label{plot.simChroma}
-\end{figure}
-
-For dataset A, we see very similar visual patterns across the two panels, corresponding well with the fact that both groups are drawn from the same underlying model.
-For dataset B, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups.
+For dataset $\mathbf{A}$, we see very similar visual patterns across the two panels, corresponding well with the fact that both groups are drawn from the same underlying model.
+For dataset $\mathbf{B}$, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups.
 The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.
 For each component, the loadings describe how influential the different variables are on that component.
-Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such as \textit{"variable $x$ is generally more influential in subset 1 than it is in subset 2"}, thereby helping us to understand where and why potential data structure differences are found.
+Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such as \textit{"variable 1 is generally more influential in subset A than it is in subset B"}, thereby helping us to understand where and why potential data structure differences are found.
+
+\subsection*{A data structure comparison workflow}
+%\begin{figure}
+%\center
+%\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=5cm, align=center,
+%	minimum height=2cm,
+%  thick,main node/.style={rectangle, rounded corners,draw,font=\small\normalfont},
+%  sq node/.style={rectangle,draw,font=\small\normalfont}]
+%  \node[sq node] (1) {Input: Two datasets $\mathbf{X} \in \RR^{n_x \times d}$ \\ and $\mathbf{Y} \in \RR^{n_y \times d}$};
+%  \node[sq node, below = 2 cm of 1] (2) {Make CE plot};
+%  \node[sq node, below left of = 2] (3) {Make angle plot};
+%  \node[sq node, below right of = 2] (4) {Stop and conclude that \\ the datasets are \textit{not} \\ similar in structure};
+%  \node[sq node, below left of = 3] (5) {Stop and conclude that  \\ the datasets \textit{are} \\ similar in structure};
+%  \node[sq node, below right of = 3] (6) {Make chroma plot: \\  Explain dissimilarities};
+%  \path[every node/.style={sloped, anchor = south, font=\small}]
+%    (1) edge node  {} (2)
+%    (2) edge node {Similarity} (3)
+%    (2) edge node {Dissimilarity} (4)
+%    (3) edge node {Similarity} (5)
+%    (3) edge node {Dissimilarity} (6);
+% \end{tikzpicture}
+% \caption{The PCADSC workflow, starting with two datasets and then successively applying the three visual tools for obtaining a deeper and deeper %understanding of similarities and dissimilarities in the data structures.}
+% \label{fig:flowchart}
+% \end{figure}
+
+Finally, we turn to a few comments about how the three plots presented thus far can be used together.
+We suggest that the plots are applied in the  order in which they were presented, thereby moving from overall assesments of similarity to explanations about the nature of potential data structure differences.
+This workflow is summarized in Figure \ref{fig:flowchart} and should follow three steps:
+\begin{enumerate}
+\item Make a CE plot.
+If this plots suggests similarity in eigenvalues then it is appropriate to move on to the angle plot to assess similarity in loadings.
+Otherwise, it can be concluded that the structures of the two datasets are not similar.
+\item Make an angle plot.
+If this plots suggests similarities in both eigenvalues and loadings, then the data structure comparison exploration is concluded.
+If the angle plots suggests differences in data structures, identify which components are responsible for the differences.
+\item Make a chroma plot of the components selected to be different by the angle plot. Use this information to discuss \textit{how} and \textit{why} the data structures might be different. 
+\end{enumerate}
+We will now turn to a data example, showing how the PCADSC tools can be used in practice.
 
@@
 
-If two countries differ e.g.
-in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one.
-More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being.
+If for instance two countries differ in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one.
+More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital, and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being.
 The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures.
@@
 Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
-These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}.
-More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in the supplementray material (S2).
+These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing}, and \textit{Supportive relationships}.
+More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in Appendix S3.
 We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10.
@@
 This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero.
+The CE curve generally lies above zero, suggesting that the Bulgarian dataset has larger eigenvalues for all but the last component, thereby explaining more variance at lower dimensionalities.
 
-\begin{figure}[h]
-\caption{[FIGURE 5 ABOUT HERE] CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being.
-The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
+%\begin{figure}[h]
+%\caption{[FIGURE 6 ABOUT HERE] CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being.
+%The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
 In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
-\label{plotBG.cehair}
-\end{figure}
+%\label{plotBG.cehair}
+%\end{figure}
 
-Moving on to the angle plot, we find that the differences are primarily to be found in the second, third and fourth principal components (PCs).
+Moving on to the angle plot, we find that the differences are primarily to be found in the second, third, and fourth principal components (PCs).
 The blue arrows visualize the decomposition of the principal components for the Bulgarian dataset in the coordinate system of the Danish dataset.
@@
 
-\begin{figure}[h]
-\caption{[FIGURE 6 ABOUT HERE] Chroma plot for comparing Danish and Bulgarian data on psychological well-being.
-A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgarian- and Danish psychological well-being data.
+%\begin{figure}[h]
+%\caption{[FIGURE 7 ABOUT HERE] Chroma plot for comparing Danish and Bulgarian data on psychological well-being.
+A chroma plot comparing the 2nd, 3rd, and 4th principal components of the Bulgarian- and Danish psychological well-being data.
 The component-bars are annotated with their relative variance contributions.
 }
-\label{plotBG.pancake}
-\end{figure}
+%\label{plotBG.pancake}
+%\end{figure}
 
@@
 This is also reflected by the two tests, which now produce $p$-values of $p_\text{KS} = 0.17$ and $p_\text{CvM} = 0.11$, respectively.
-Thus, it is not unreasonable to assume equal variance components in the two datasets.
+Thus, it is not unreasonable to assume equal eigenvalues in the two datasets.
 
-\begin{figure}[h]
-\caption{[FIGURE 7 ABOUT HERE] A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being.
+%\begin{figure}[h]
+%\caption{[FIGURE 8 ABOUT HERE] A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being.
 The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
-\label{plotSE.cehair}
-\end{figure}
+%\label{plotSE.cehair}
+%\end{figure}
 
 The angle plot in Figure \ref{plotSE.cehair} shows that the two datasets agree very strongly about the relative importance of the six scales in the six PCs, as almost all off-diagonal arrows are practically non-existent.
-This implies that if one already has e.g.
-the information held in the first PC from the Danish data, this information is in itself mostly sufficient to describe the first PC of the Swedish data.
+This implies that if one already has the information held in the first PC from the Danish data, this information is in itself mostly sufficient to describe the first PC of the Swedish data.
 
@@
 We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25\% of the variance in both datasets.
-In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g.
-conclude using Table~\ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally happier than the least happy people in Sweden.
+In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and for instance conclude using Table~\ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally happier than the least happy people in Sweden.
 A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.
 
-\begin{figure}[h]
-\caption{[FIGURE 8 ABOUT HERE] Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples.
+%\begin{figure}[h]
+%\caption{[FIGURE 9 ABOUT HERE] Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples.
 For each component, the bar is annotated with its cumulative variance contribution, that is, how much variance can be explained by having information of this and the preceding components.}
-\label{plotSE.pancake}
-\end{figure}
+%\label{plotSE.pancake}
+%\end{figure}
 
@@
 
-Three new tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if two datasets can be combined for analysis were proposed and discussed in this paper.
+Three new tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if two datasets can be combined for analysis were proposed and discussed.
 They all employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
@@
 This highlights lack of data comparability as an empirically relevant and concerning issue that there is no way around: When conducting rankings, the issue of data structure heterogeneity should always be addressed.
+
+Usually, when using PCA, the main interest lies in the \textit{scores}, i.e.
+the coordinates $\eta_j^\top (x_i - \bar{x})$ of the projections observations onto the loadings, but the eigenvalues and the accompanying loadings are in fact more suitable objects for assessing data structure similarity.
+If two dataset subsets have similar loadings, then they can be argued to measure the same underlying quantities.
+Similar loadings can occur in situations where the two sets of scores are different, for example if the two datasets come from two different populations of subjects.
+On the other hand, if the loading patterns are different in the two subsets, then this indicates that their respective variable interplays differ, and hence it would be criticizable to use these variables for comparisons across the two subsets.
+The PCADSC tools are novel in their emphasis on the information held in the loadings rather than the scores.
 
@@
 
-All in all, the PCADSC methods represents a first step towards addressing a known issue that has otherwise been met with unsatisfactory ad-hoc methods whose assumptions rest upon the very hypothesis they are testing.
+A different aspect of the data size, namely the number of variables $d$, may also limit the useability of the PCADSC tools.
+While the CE plot scales well with large $d$, the angle plot and the chroma plot can become difficult to read if there are a lot of variables and thus a large number of principal components.
+However, because the angle plot lets the arrow length depend on the size of the correpsonding eigenvalue, its visual impression should help guide a user towards influential \textit{and} dissimilar components even though $d$ is large.
+And when these components have been identified, the chroma plot can focus only on them, thereby avoiding plotting the loading patterns of too many components.
+
+All in all, the PCADSC methods represent a first step towards addressing a known issue that has otherwise been met with unsatisfactory ad-hoc methods whose assumptions rest upon the very hypothesis they are testing.
 The PCADSC tools perform quite well in the data example provided here, yielding conclusions that are in accordance with well-established theories and they are simple to use and test further using the PCADSC \texttt{R} package.
@@
 We have presented three new tools for visual, exploratory data structure comparisons that utilizes covariance matrix decompositions obtained by use of principal component analysis.
+The tools compare the linear structures of numerical datasets.
 In the provided data example, comparing data on personal happiness across three European countries, the PCADSC tools were able to identify both comparable datasets and non-comparable datasets, as determined according to existing subject-matter literature.
@@
 
+\newpage
+
+\section*{Figures}
+
+\begin{figure}[H]
+\center
+Model 1: \medskip \\
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=2.5cm,
+  thick,main node/.style={rectangle, rounded corners,draw,font=\small\normalfont},
+  sq node/.style={rectangle,draw,font=\small\normalfont}]
+  \node[sq node] (1) {$V_1$};
+  \node[sq node, right of = 1] (2) {$V_2$};
+  \node[sq node, right of = 2] (3) {$V_3$};
+  \node[sq node, right of = 3] (4) {$V_4$};
+  \node[sq node, right of = 4] (5) {$V_5$};
+  \node[sq node, right of = 5] (6) {$V_6$};
+  \node[main node, above right=1cm of 3] (7) {$L_1$};
+  \node[main node, above right=1cm of 5] (8) {$L_2$};
+  \path[every node/.style={font=\sffamily\small}]
+    (7) edge node  {} (3)
+    (7) edge node {} (4)
+    (8) edge node {} (5)
+    (8) edge node {} (6);
+ \end{tikzpicture}
+\\
+\bigskip
+Model 2: \medskip \\
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=2.5cm,
+  thick,main node/.style={rectangle, rounded corners,draw,font=\small\normalfont},
+  sq node/.style={rectangle,draw,font=\small\normalfont}]
+  \node[sq node] (1) {$V_1$};
+  \node[sq node, right of = 1] (2) {$V_2$};
+  \node[sq node, right of = 2] (3) {$V_3$};
+  \node[sq node, right of = 3] (4) {$V_4$};
+  \node[sq node, right of = 4] (5) {$V_5$};
+  \node[sq node, right of = 5] (6) {$V_6$};
+  \node[main node, above=1cm of 2] (7) {$K_1$};
+  \node[main node, above=1cm of 5] (8) {$K_2$};
+  \path[every node/.style={font=\small}]
+    (7) edge node  {} (1)
+    (7) edge node {} (2)
+    (7) edge node {} (3)
+    (8) edge node {} (4)
+    (8) edge node {} (5)
+    (8) edge node {} (6);
+ \end{tikzpicture}
+ \caption{Graphical representations of models used for simulating example data.
+Square nodes represent observed variables, while rounded nodes represent latent variables.
+Arrows are used to illustrate causal links.}
+\label{fig:simGraphs}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\includegraphics[scale=0.8]{Figure2_v2.pdf}
+\caption{CE plots for the simulated datasets.
+Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel.
+Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures.
+The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.}
+\label{plot.simCE} 
+\end{figure}
+
+\begin{figure}[H]
+\center
+\includegraphics[scale=0.8]{Figure3_v2.pdf}
+\caption{Angle plots for the simulated datasets.
+Dataset A in top panel and dataset B in bottom panel.
+Blue arrows show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.}
+\label{plot.simAngle}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\includegraphics[scale=0.8]{Figure4_v2.pdf}
+\caption{Chroma plots for the simulated datasets.
+Dataset $\mathbf{A}$ in top panel and dataset $\mathbf{B}$ in bottom panel.
+Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.}
+\label{plot.simChroma}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=5cm, align=center,
+	minimum height=2cm,
+  thick,main node/.style={rectangle, rounded corners,draw,font=\small\normalfont},
+  sq node/.style={rectangle,draw,font=\small\normalfont}]
+  \node[sq node] (1) {Input: Two datasets $\mathbf{X} \in \RR^{n_x \times d}$ \\ and $\mathbf{Y} \in \RR^{n_y \times d}$};
+  \node[sq node, below = 2 cm of 1] (2) {Make CE plot};
+  \node[sq node, below left of = 2] (3) {Make angle plot};
+  \node[sq node, below right of = 2] (4) {Stop and conclude that \\ the datasets are \textit{not} \\ similar in structure};
+  \node[sq node, below left of = 3] (5) {Stop and conclude that  \\ the datasets \textit{are} \\ similar in structure};
+  \node[sq node, below right of = 3] (6) {Make chroma plot: \\  Explain dissimilarities};
+  \path[every node/.style={sloped, anchor = south, font=\small}]
+    (1) edge node  {} (2)
+    (2) edge node {Similarity} (3)
+    (2) edge node {Dissimilarity} (4)
+    (3) edge node {Similarity} (5)
+    (3) edge node {Dissimilarity} (6);
+ \end{tikzpicture}
+ \caption{The PCADSC workflow, starting with two datasets and then successively applying the three visual tools for obtaining a deeper and deeper understanding of similarities and dissimilarities in the data structures.}
+ \label{fig:flowchart}
+ \end{figure}
+
+\begin{figure}[H]
+\center
+\includegraphics[scale=0.8]{Figure6_v2.pdf}
+\caption{CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being.
+The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
+In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
+\label{plotBG.cehair}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\includegraphics[scale=0.8]{Figure7_v2.pdf}
+\caption{Chroma plot for comparing Danish and Bulgarian data on psychological well-being.
+A chroma plot comparing the 2nd, 3rd, and 4th principal components of the Bulgarian- and Danish psychological well-being data.
+The component-bars are annotated with their relative variance contributions.
+}
+\label{plotBG.pancake}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\includegraphics[scale=0.8]{Figure8_v2.pdf}
+\caption{A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being.
+The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
+\label{plotSE.cehair}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\includegraphics[scale=0.8]{Figure9_v2.pdf}
+\caption{Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples.
+For each component, the bar is annotated with its cumulative variance contribution, that is, how much variance can be explained by having information of this and the preceding components.}
+\label{plotSE.pancake}
+\end{figure}
+
+\newpage
+
 \section*{Supplementary material}
 
-\subsection*{S0 Appendix: A brief introduction to principal component analysis}
+\subsection*{S1 Appendix: A brief introduction to principal component analysis}
+\label{appendix.introPCA}
 
-\label{appendix.introPCA}
-Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = {(\bar{x}_1, ..., \bar{x}_d)}^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) {(x_i-\bar{x})}^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$.
-We assume in the following that all variables of $X$ have a numerical interpretation, e.g.
-by being continuous or ordinal categorical.
+In this appendix we introduce PCA. We suppress the subscripts and superscripts used in the article because we only consider a single dataset here, denoted by $\mathbf{X} \in \RR^{d \times n}$.
+
 For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost.
-More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the original $d$-dimensional one.
+More specifically, we wish to minimize the loss when looking at a projection of $\mathbf{X}$ onto a $q$-dimensional space, rather than the original $d$-dimensional one.
 Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
@@
 \begin{equation}
-\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
-\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
+\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_{i \cdot} - \bar{X} - z \rVert^2 =
+\min_{K_q} \sum_{i=1}^n \lVert x_{i \cdot} - \bar{X} - \text{proj}_{K_q}(x_{i \cdot} - \bar{X}) \rVert^2.
 \end{equation}
@@
 \begin{equation*}
-\text{trace}(S) = \sum_{j=1}^d \hat{V}(x_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
+\text{trace}(S) = \sum_{j=1}^d \hat{V}(X_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top \mathbf{X}^\top) = \sum_{j=1}^d \lambda_j,
 \end{equation*}
-where $x_j \in \RR^n$ denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function.
+where $X_j \in \RR^n$ denotes the $j$th variable of $\mathbf{X}$ and $\hat{V}$ is the empirical variance function.
 This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
-And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are %different,
-distinct, the representation hereby obtained is a valid object of inference.
-If the covariance matrix does not have full rank, which is the case in particular when $n < d$, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.
+And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are distinct, the representation hereby obtained is a valid object of inference.
 
@@
 \begin{align*}
-\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
-\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2,
+\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_{i \cdot} - \bar{X}) \rVert^2, &
+\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_{i \cdot} - \bar{X}) \rVert^2,
 \end{align*}
@@
 It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
-This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$.
-This implies that the loadings and the variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.
 
-\subsection*{S1 Appendix: Details about the simulated data}
+At this point a general remark about standardization is in place.
+PCA deconstructs the covariance matrix in components according to the most explained variance.
+This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will always be deemed highly influential, no matter the structure of the data.
+Therefore, the variables are typically standardized prior to performing PCA.
+The covariance matrix of the standardized variables is the same as the correlation matrix of the original variables, so post-standardization PCA simply corresponds to performing data structure comparisons of the correlation matrices rather than the covariance matrices.
+The standardization makes the variables comparable on the same scale, namely units of standard deviation.
+We have adopted this standardization policy.
+
+\subsection*{S2 Appendix: Details about the simulated data}
 \label{appendix.simData}
 This supplementary appendix describes the origin of the simulated data used for presenting the PCADSC plots.
-Dataset A consisted of 1000 independent simulated realizations from the $N(0, \Sigma_1)$-distribution, while dataset B consisted of 500 independent realizations from $N(0, \Sigma_1)$ and 500 independent realizations from $N(0, \Sigma_2)$.
-Dataset A was furthermore appended with a grouping variable, randomly dividing the observations into two groups.
-Dataset B also included a grouping variable, and this variable contained information about which of the two normal distributions each observation was simulated from.
-The covariance matrices, $\Sigma_1$ and $\Sigma_2$, were defined by
+Dataset $\mathbf{A}$ consisted of 1000 independent simulated realizations from the $N(0, \Sigma_1)$-distribution, while dataset $\mathbf{B}$ consisted of 500 independent realizations from $N(0, \Sigma_1)$ and 500 independent realizations from $N(0, \Sigma_2)$.
+Dataset $\mathbf{A}$ was furthermore appended with a grouping variable, randomly dividing the observations into two groups.
+Dataset $\mathbf{B}$ also included a grouping variable, and this variable contained information about which of the two normal distributions each observation was simulated from.
+The covariance matrices, $\Sigma_A$ and $\Sigma_B$, were defined by
 
-$$\Sigma_1 = \begin{pmatrix}
+$$\Sigma_A = \begin{pmatrix}
     1.0  &  0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
@@
 $$
-\Sigma_2 = \begin{pmatrix}
+\Sigma_B = \begin{pmatrix}
  1.0 & 0.2 & 0.1 & 0.0 & 0.0 & 0.0 \\
@@
 
-\subsection*{S2 Table: ESS psychological wellbeing scale definitions}
+\subsection*{S3 Table: ESS psychological wellbeing scale definitions}
 \begin{table}[h]
@@
 \caption{{\bf ESS psychological wellbeing scale definitions.} Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
-Note that before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
+Before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
 \label{table:items}

### 2018-05-21 ESS/sim_notes.R

@@ -38,51 +38,82 @@ po2 <- PCADSC(dF2, "group")
 
 CE1 <- CEPlot(po1) +
   scale_y_continuous(breaks = round(seq(-0.6, 0.6, 0.1),1), limits = c(-0.5, 0.3)) +
-  ggtitle("Dataset B") +  theme(plot.title = element_text(hjust = 0.5))
+  ggtitle("Dataset B", subtitle = "Cumulative difference in eigenvalues: Group 1 - Group 2") +
+  theme(panel.border = element_blank(),
+        axis.line=element_line(),
+        plot.subtitle = element_text(size = 11, hjust = -0.11),
+        plot.title = element_text(hjust = 0.5)) +
+  ylab("")
 
 CE2 <- CEPlot(po2) +
   scale_y_continuous(breaks = round(seq(-0.6, 0.6, 0.1),1), limits = c(-0.3, 0.3)) +
-  ggtitle("Dataset A") +  theme(plot.title = element_text(hjust = 0.5))
+  ggtitle("Dataset A", subtitle = "Cumulative difference in eigenvalues: Group 1 - Group 2") +
+  theme(panel.border = element_blank(),
+        axis.line=element_line(),
+        plot.subtitle = element_text(size = 11, hjust = -0.11),
+        plot.title = element_text(hjust = 0.5)) +
+  ylab("")
 
 angle1 <- anglePlot(po1) +
-  ggtitle("Dataset B") +  theme(plot.title = element_text(hjust = 0.5))
+  theme(panel.border = element_blank(),
+        axis.line=element_line(),
+        plot.subtitle = element_text(size = 11, hjust = -0.03),
+        plot.title = element_text(hjust = 0.5)) +
+  ggtitle("Dataset B", subtitle = "PCs for Group 2") +
+  ylab("")
+
 angle2 <- anglePlot(po2) +
-  ggtitle("Dataset A") +  theme(plot.title = element_text(hjust = 0.5))
+  theme(panel.border = element_blank(),
+        axis.line=element_line(),
+        plot.subtitle = element_text(size = 11, hjust = -0.03),
+        plot.title = element_text(hjust = 0.5)) +
+  ggtitle("Dataset A", subtitle = "PCs for Group 2") +
+  ylab("")
 
 chroma1 <- chromaPlot(po1) +
[…]

### 2018-05-21 ESS/ess_notes.R
Jeg har forsøgt at tilføje 95 pct konfidensintervaller på anglePlot. Disse er kun baseret på vinklerne, og længde korrektionen stammende fra egenværdierne er sadt til 1. Implementeringen i objekterne er måske heller ikke helt ren.

@@ -190,6 +190,7 @@ xtable(m, digits = 2)
 library(PCADSC)
 library(ggplot2)
 library(gridExtra)
+#library(grid)
 
 #DK/BG
 set.seed(123424)
@@ -200,23 +201,49 @@ a2 <- PCADSC(data_BGDK, "cntry",  c("EvaluativeWellbeing",
                                     "CommunityWellbeing",
                                     "SupportiveRelationships"))
 
-fig5_CE <- CEPlot(a2) +
-  scale_y_continuous(breaks = round(seq(-0.3, 0.5, 0.1),1), limits = c(-0.2, 0.5))
-fig5_angle <- anglePlot(a2)
+fig6_CE <- CEPlot(a2) +
+  scale_y_continuous(breaks = round(seq(-0.3, 0.5, 0.1),1), limits = c(-0.2, 0.5)) +
+  theme(panel.border = element_blank(),
+        axis.line=element_line(),
+        plot.title = element_text(size = 11, hjust = -0.09)) +
+  ggtitle("Cumulative difference in eigenvalues: BG - DK") +
+  ylab("")
+fig6_angle <- anglePlot(a2) +
+  theme(panel.border = element_blank(),
+        axis.line=element_line(),
+        plot.title = element_text(size = 11, hjust = -0.03)) +
+  ggtitle("PCs for DK") +
+  ylab("")
 
-fig6_chroma <- chromaPlot(a2, useComps = 2:4) #figure 6
+fig7_chroma <- chromaPlot(a2, useComps = 2:4) +
+  theme(#panel.border = element_blank(),
+        axis.line=element_line(),
+       # plot.title = element_text(size = 11, hjust = -0.02),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank()) #+
+ # ggtitle("PCs for DK") +
+#  ylab("")
 
 
-ggsave(plot = grid.arrange(fig5_CE, fig5_angle, nrow = 2), file = "./article/PlosOne/Fig5.eps",
-       device = "eps", dpi = 400, width = 7, height = 8)
+#ggsave(plot = grid.arrange(fig5_CE, fig5_angle, nrow = 2), file = "./article/PlosOne/Fig5.eps",
+#       device = "eps", dpi = 400, width = 7, height = 8)
 
[…]

### 2018-05-21 R/PCADSC.R

@@ -146,7 +146,7 @@ PCADSC <- function(data, splitBy, vars=NULL, doCE = TRUE,
   chromaInfo <- NULL
 
   if (doCE) CEInfo <- doCE(pcaRes, data, B)
-  if (doAngle) angleInfo <- doAngle(pcaRes)
+  if (doAngle) angleInfo <- doAngle(pcaRes, data, B)                  # CHANGED to include simulations
   if (doChroma) chromaInfo <- doChroma(pcaRes)
 
   out <- list(pcaRes = pcaRes, CEInfo = CEInfo, angleInfo = angleInfo,

### 2018-05-21 R/anglePlot.R

@@ -67,6 +67,25 @@ anglePlot <- function(x) {
   d <- obj$d
   splitLevels <- obj$splitLevels
 
+  # -------------------------------------------------------
+  # -------------------------------------------------------
+  # Make data frame for adding 95pct confidence intervals
+  mydata <- data.frame(g=NULL,x=NULL,y=NULL)
+  for (i in 1:d) for (j in 1:d) {
+    theta <- quantile(obj$angles.sim[i,j,],c(0.025,0.975))
+    theta <- seq(theta[1],theta[2],length.out = 100)
+    mydata <- rbind(mydata,
+                    data.frame(g=paste(c(i,j,"1st"),collapse="."),
+                               x=c(i,i+cos(theta)*cos(pi/4+theta/2)),
+                               y=c(j,j+cos(theta)*sin(pi/4+theta/2))),
+                    data.frame(g=paste(c(i,j,"2nd"),collapse="."),
+                               x=c(i,i+cos(theta)*cos(pi/4-theta/2)),
+                               y=c(j,j+cos(theta)*sin(pi/4-theta/2))))
+  }
+  # -------------------------------------------------------
+  # -------------------------------------------------------
+
+
   #Make anglePlot
   ggplot(aF, aes_string(x = "x", y = "y", col = "type",
                         xend = "xend", yend = "yend")) +
@@ -77,5 +96,6 @@ anglePlot <- function(x) {
     xlab(paste("PCs for", splitLevels[1])) +
     ylab(paste("PCs for", splitLevels[2])) +
     geom_segment(arrow = arrow(length = unit(arrow.len, "inch"))) +
-    scale_color_manual(values = c("blue", "red"), guide = FALSE)
+    scale_color_manual(values = c("blue", "red"), guide = FALSE) +
+    geom_polygon(aes(group=g,x=x,y=y),data=mydata,inherit.aes = FALSE,fill="gray85",alpha=1)   # Add confidence intervals
 }

### 2018-05-21 R/doAngle.R

@@ -38,13 +38,13 @@
 #' @seealso \code{\link{anglePlot}}, \code{\link{PCADSC}}
 #'
 #' @export
-doAngle <- function(x) {
+doAngle <- function(x, ...) {
   UseMethod("doAngle")
 }
 
 #x: pcaRes
 #' @export
-doAngle.pcaRes <- function(x) {
+doAngle.pcaRes <- function(x, data, B, ...) {
   load1 <- x$load1
   load2 <- x$load2
   eigen1 <- x$eigen1
@@ -80,14 +80,44 @@ doAngle.pcaRes <- function(x) {
                             rep(1:d, each = d) + c(len2)*sin(c(angles)/2)),
                    type = rep(c("1st", "2nd"), each = d*d))
 
+  # -------------------------------------------------------------------
+  # -------------------------------------------------------------------
+  # The following is change from corresponding simulation part of doCE
+  splitLevels <- x$splitLevels
+
+  #check whether B is positive
+  if (B < 1) {
+    stop("B must be positive.")
+  }
+
+  #Unpack x
+  n1 <- x$n1
+  nBoth <- x$nBoth
+  d <- x$d
+  vars <- x$vars
+
+  #Calculate cumulative sums for the randomly partitioned data
+  angles.sim   <- array(0, dim=c(d,d,B))
+
+  for (i in 1:B) {
+    splitVar <- rep("2", nBoth)
+    ii <- sample(1:nBoth, n1)
+    splitVar[ii] <- "1"
+    data$splitVar <- factor(splitVar)
+    myPCA <- doPCA(data, "splitVar", c("1", "2"), vars, doBoth = FALSE)
+    angles.sim[,,i] <- acos(abs(t(myPCA$load1)%*%myPCA$load2))
+  }
+  # -------------------------------------------------------------------
+  # -------------------------------------------------------------------
+
   #pack and return output
-  out <- list(aF = aF, splitLevels = splitLevels, d = d)
+  out <- list(aF = aF, splitLevels = splitLevels, angles.sim = angles.sim, d = d)
   class(out) <- "angleInfo"
   out
 }
 
 #' @export
-doAngle.PCADSC <- function(x) {
+doAngle.PCADSC <- function(x, ...) {
   x$angleInfo <- doAngle(x$pcaRes)
   x
 }

### 2018-06-11 R/anglePlot.R
Fixed minor stuff in angle-functions so that the package can be installed

@@ -40,7 +40,7 @@
 #'
 #' @importFrom ggplot2 ggplot aes_string scale_x_continuous scale_y_continuous
 #' theme_bw theme element_blank xlab ylab geom_segment unit scale_color_manual
-#' geom_polygon
+#' geom_polygon scale_fill_grey
 #' @importFrom grid arrow
 #' @export
 anglePlot <- function(x) {

### 2018-06-11 R/doAngle.R

@@ -148,7 +148,6 @@ doAngle.pcaRes <- function(x, data, B, ...) {
 
 #' @export
 doAngle.PCADSC <- function(x, ...) {
-  x$angleInfo <- doAngle(x$pcaRes)
   if ("B" %in% names(list(...))) b <- list(...)$B
   else b <- x$B
 

### 2018-06-13 R/anglePlot.R
cirkulære konfidensområder

@@ -71,7 +71,7 @@ anglePlot <- function(x) {
 
   #Make anglePlot
   ggplot(cR,aes_string(group = "g:quantile", x = "x", y = "y", fill="quantile")) +
-    geom_polygon() + scale_fill_grey(start=0.7,end=1,guide=FALSE) +
+    geom_polygon() + scale_fill_grey(start=0.7,end=0.9,guide=FALSE) +
     geom_segment(aes_string(x = "x", y = "y", col = "type", xend = "xend", yend = "yend"),
                  aF,arrow = arrow(length = unit(arrow.len, "inch")),inherit.aes = FALSE) +
     scale_x_continuous(limits = c(0.5,d+0.5), breaks = 1:d) +

### 2018-06-13 R/doAngle.R

@@ -119,20 +119,26 @@ doAngle.pcaRes <- function(x, data, B, ...) {
   for (i in 1:d) for (j in 1:d) {
     if (i==j) {
       pval[i,j] <- mean(angles[i,j]<=angles.sim[i,j,])
+      my.q <- c(0,quantile(angles.sim[i,j,],seq(0.05,1,0.05)))
     } else {
       pval[i,j] <- mean(angles[i,j]>=angles.sim[i,j,])
+      my.q <- c(quantile(angles.sim[i,j,],seq(0,0.95,0.05)),pi/2)
     }
-    my.q <- quantile(angles.sim[i,j,],seq(0,1,0.05))
     for (k in 1:20) {
       cR <- rbind(cR,
                   data.frame(g=paste(c(i,j,"1st"),collapse="."),
                              quantile=ifelse(i==j,k,21-k),
-                             x=c(i,i+len[i]*cos(my.q[k+0:1])*cos(pi/4+my.q[k+0:1]/2)),
-                             y=c(j,j+len[i]*cos(my.q[k+0:1])*sin(pi/4+my.q[k+0:1]/2))),
-                  data.frame(g=paste(c(i,j,"2nd"),collapse="."),
+                             x=c(i,i+len1[i,j]*cos(pi/4+my.q[k+0:1]/2)),
+                             y=c(j,j+len1[i,j]*sin(pi/4+my.q[k+0:1]/2))),
+      #            x=c(i,i+sqrt(eigen1[i]/max.eigen)*cos(pi/4+my.q[k+0:1]/2)),
+      #            y=c(j,j+sqrt(eigen1[j]/max.eigen)*sin(pi/4+my.q[k+0:1]/2))),
+      data.frame(g=paste(c(i,j,"2nd"),collapse="."),
                              quantile=ifelse(i==j,k,21-k),
-                             x=c(i,i+len[j]*cos(my.q[k+0:1])*cos(pi/4-my.q[k+0:1]/2)),
-                             y=c(j,j+len[j]*cos(my.q[k+0:1])*sin(pi/4-my.q[k+0:1]/2))))
+                             x=c(i,i+len2[i,j]*cos(pi/4-my.q[k+0:1]/2)),
+                             y=c(j,j+len2[i,j]*sin(pi/4-my.q[k+0:1]/2))))
[…]

### 2018-06-14 R/anglePlot.R
Sections of same angular size used in plotting of confidence regions

@@ -70,13 +70,13 @@ anglePlot <- function(x) {
   splitLevels <- obj$splitLevels
 
   #Make anglePlot
-  ggplot(cR,aes_string(group = "g:quantile", x = "x", y = "y", fill="quantile")) +
+  ggplot(cR,aes_string(group = "g", x = "x", y = "y", fill = "percent")) +
+    scale_color_manual(values = c("blue", "red"), guide = FALSE) +
     geom_polygon() + scale_fill_grey(start=0.7,end=0.9,guide=FALSE) +
     geom_segment(aes_string(x = "x", y = "y", col = "type", xend = "xend", yend = "yend"),
                  aF,arrow = arrow(length = unit(arrow.len, "inch")),inherit.aes = FALSE) +
-    scale_x_continuous(limits = c(0.5,d+0.5), breaks = 1:d) +
-    scale_y_continuous(limits = c(0.5,d+0.5), breaks = 1:d) +
-    scale_color_manual(values = c("blue", "red"), guide = FALSE) +
+    scale_x_continuous(limits = c(0.5,d+0.71), breaks = 1:d) +
+    scale_y_continuous(limits = c(0.5,d+0.71), breaks = 1:d) +
     theme_bw() +
     theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
     xlab(paste("PCs for", splitLevels[1])) +

### 2018-06-14 R/doAngle.R

@@ -119,35 +119,43 @@ doAngle.pcaRes <- function(x, data, B, ...) {
   for (i in 1:d) for (j in 1:d) {
     if (i==j) {
       pval[i,j] <- mean(angles[i,j]<=angles.sim[i,j,])
-      my.q <- c(0,quantile(angles.sim[i,j,],seq(0.05,1,0.05)))
+      my.angles <- seq(0,max(angles.sim[i,j,]),length.out = max(2,1+ceiling(max(angles.sim[i,j,])/0.05)))
+      percent <- my.angles[-1]
+      for (k in 2:length(my.angles)) {
+        percent[k-1] <- max(1,ceiling(mean(angles.sim[i,j,]<=mean(my.angles[(k-1):k]))*100))
+      }
     } else {
       pval[i,j] <- mean(angles[i,j]>=angles.sim[i,j,])
-      my.q <- c(quantile(angles.sim[i,j,],seq(0,0.95,0.05)),pi/2)
+      my.angles <- seq(min(angles.sim[i,j,]),pi/2,length.out = max(2,1+ceiling((pi/2-min(angles.sim[i,j,]))/0.05)))
+      percent <- my.angles[-1]
+      for (k in 2:length(my.angles)) {
+        percent[k-1] <- max(1,ceiling(mean(angles.sim[i,j,]>=mean(my.angles[(k-1):k]))*100))
+      }
     }
-    for (k in 1:20) {
+    for (k in 2:length(my.angles)) {
       cR <- rbind(cR,
-                  data.frame(g=paste(c(i,j,"1st"),collapse="."),
-                             quantile=ifelse(i==j,k,21-k),
-                             x=c(i,i+len1[i,j]*cos(pi/4+my.q[k+0:1]/2)),
-                             y=c(j,j+len1[i,j]*sin(pi/4+my.q[k+0:1]/2))),
-      #            x=c(i,i+sqrt(eigen1[i]/max.eigen)*cos(pi/4+my.q[k+0:1]/2)),
-      #            y=c(j,j+sqrt(eigen1[j]/max.eigen)*sin(pi/4+my.q[k+0:1]/2))),
-      data.frame(g=paste(c(i,j,"2nd"),collapse="."),
-                             quantile=ifelse(i==j,k,21-k),
-                             x=c(i,i+len2[i,j]*cos(pi/4-my.q[k+0:1]/2)),
-                             y=c(j,j+len2[i,j]*sin(pi/4-my.q[k+0:1]/2))))
[…]

### 2018-06-14 ESS/ess_notes.R
New versions of all figures after update of anglePlot

@@ -213,7 +213,8 @@ fig6_angle <- anglePlot(a2) +
         axis.line=element_line(),
         plot.title = element_text(size = 11, hjust = -0.03)) +
   ggtitle("PCs for DK") +
-  ylab("")
+  ylab("") +
+  scale_fill_manual(values = c("lightgreen", rep("grey", 18), "lightgreen"), guide = FALSE)
 
 fig7_chroma <- chromaPlot(a2, useComps = 2:4) +
   theme(#panel.border = element_blank(),
@@ -237,11 +238,11 @@ fig7_chroma <- chromaPlot(a2, useComps = 2:4) +
 #ggsave(plot = fig6_chroma, file = "P:/PCADSC/CSP/chromaBGDK.pdf", width = 7, height = 5)
 
 ggsave(plot = grid.arrange(fig6_CE, fig6_angle, nrow = 2),
-       file = "P:/PCADSC/R/article/Old versions/MPS/Figure6_v2.pdf",
+       file = "P:/PCADSC/R/article/Old versions/MPS/Figure6_v3.pdf",
              dpi = 400, width = 7, height = 8)
 
 ggsave(plot = fig7_chroma,
-       file = "P:/PCADSC/R/article/Old versions/MPS/Figure7_v2.pdf",
+       file = "P:/PCADSC/R/article/Old versions/MPS/Figure7_v3.pdf",
        dpi = 400, width = 7, height = 4)
 
 
@@ -266,7 +267,8 @@ fig8_angle <- anglePlot(a1) +
         axis.line=element_line(),
         plot.title = element_text(size = 11, hjust = -0.03)) +
   ggtitle("PCs for SE") +
-  ylab("")
+  ylab("")  +
+  scale_fill_manual(values = c("lightgreen", rep("grey", 18), "lightgreen"), guide = FALSE)
 
 fig9_chroma <- chromaPlot(a1) +
   theme(#panel.border = element_blank(),
@@ -290,10 +292,10 @@ fig9_chroma <- chromaPlot(a1) +
 #ggsave(plot = fig8_chroma, file = "P:/PCADSC/CSP/chromaSEDK.pdf", width = 7, height = 5)
 
 ggsave(plot = grid.arrange(fig8_CE, fig8_angle, nrow = 2),
-       file = "P:/PCADSC/R/article/Old versions/MPS/Figure8_v2.pdf",
[…]

### 2018-06-14 ESS/sim_notes.R

@@ -1,6 +1,7 @@
 library(MASS)
 library(PCADSC)
 library(ggplot2)
+library(gridExtra)
 
 #Simulate data from covariance matrices
 #Note: Names (S1 and S2) are opposite of that in the article.
@@ -129,15 +130,15 @@ chroma2 <- chromaPlot(po2) +
 
 
 ggsave(plot = grid.arrange(CE2, CE1, nrow = 2),
-        file = "P:/PCADSC/R/article/Old versions/MPS/Figure2_v2.pdf",
+        file = "P:/PCADSC/R/article/Old versions/MPS/Figure2_v3.pdf",
        dpi = 400, width = 7, height = 8)
 
 
 ggsave(plot = grid.arrange(angle2, angle1, nrow = 2),
-       file = "P:/PCADSC/R/article/Old versions/MPS/Figure3_v2.pdf",
+       file = "P:/PCADSC/R/article/Old versions/MPS/Figure3_v3.pdf",
        dpi = 400, width = 7, height = 8)
 
 ggsave(plot = grid.arrange(chroma2, chroma1, nrow = 2),
-       file = "P:/PCADSC/R/article/Old versions/MPS/Figure4_v2.pdf",
+       file = "P:/PCADSC/R/article/Old versions/MPS/Figure4_v3.pdf",
        dpi = 400, width = 7, height = 8)
 

### 2018-06-14 R/doAngle.R
Fixed bug concerning column names in cR

@@ -115,7 +115,7 @@ doAngle.pcaRes <- function(x, data, B, ...) {
 
   # Make data frame for visualizing confidence region
   pval <- matrix(0,d,d); colnames(pval) <- colnames(angles); rownames(pval) <- rownames(angles)
-  cR <- data.frame(g=NULL,quantile=NULL,x=NULL,y=NULL)
+  cR <- data.frame(g=NULL,percent=NULL,x=NULL,y=NULL)
   for (i in 1:d) for (j in 1:d) {
     if (i==j) {
       pval[i,j] <- mean(angles[i,j]<=angles.sim[i,j,])

### 2018-06-14 ESS/ess_notes.R
new version of ESS figures

@@ -213,8 +213,7 @@ fig6_angle <- anglePlot(a2) +
         axis.line=element_line(),
         plot.title = element_text(size = 11, hjust = -0.03)) +
   ggtitle("PCs for DK") +
-  ylab("") +
-  scale_fill_manual(values = c("lightgreen", rep("grey", 18), "lightgreen"), guide = FALSE)
+  ylab("")
 
 fig7_chroma <- chromaPlot(a2, useComps = 2:4) +
   theme(#panel.border = element_blank(),
@@ -267,8 +266,7 @@ fig8_angle <- anglePlot(a1) +
         axis.line=element_line(),
         plot.title = element_text(size = 11, hjust = -0.03)) +
   ggtitle("PCs for SE") +
-  ylab("")  +
-  scale_fill_manual(values = c("lightgreen", rep("grey", 18), "lightgreen"), guide = FALSE)
+  ylab("")
 
 fig9_chroma <- chromaPlot(a1) +
   theme(#panel.border = element_blank(),

### 2018-06-14 R/doAngle.R
longer arrows

@@ -69,8 +69,8 @@ doAngle.pcaRes <- function(x, data, B, ...) {
   len1 <- len2 <- matrix(NA, d, d)
 
   for (i in 1:d) for (j in 1:d) {
-    len1[i,j] <- sqrt(eigen1[i]/max.eigen)*abs(sum(load1[,i]*load2[,j]))
-    len2[i,j] <- sqrt(eigen2[j]/max.eigen)*abs(sum(load1[,i]*load2[,j]))
+    len1[i,j] <- sqrt(eigen1[i]/max.eigen*abs(sum(load1[,i]*load2[,j])))
+    len2[i,j] <- sqrt(eigen2[j]/max.eigen*abs(sum(load1[,i]*load2[,j])))
   }
 
   #Store in data.frame ready for plotting

### 2018-06-15 ESS/ess_notes.R
Updated y-axis labels on chroma plots to fit with MPS requirements (not in code, only for article)

@@ -220,7 +220,8 @@ fig7_chroma <- chromaPlot(a2, useComps = 2:4) +
         axis.line=element_line(),
        # plot.title = element_text(size = 11, hjust = -0.02),
         panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank()) #+
+        panel.grid.minor = element_blank()) +
+  scale_x_discrete("", limits = rev(c("", "PC2", "PC3", "PC4", "")))
  # ggtitle("PCs for DK") +
 #  ylab("")
 
@@ -268,12 +269,16 @@ fig8_angle <- anglePlot(a1) +
   ggtitle("PCs for SE") +
   ylab("")
 
-fig9_chroma <- chromaPlot(a1) +
+fig9_chroma <-
+  chromaPlot(a1) +
   theme(#panel.border = element_blank(),
     axis.line=element_line(),
     # plot.title = element_text(size = 11, hjust = -0.02),
     panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank()) #+
+    panel.grid.minor = element_blank()) +
+  scale_x_reverse("", breaks = 1:6, labels = paste("PC", 1:6, sep = ""))
+
+#+
 # ggtitle("PCs for DK") +
 #  ylab("")
 

### 2018-06-16 R/doAngle.R
Redid angle-plot with previous lengths , which is mathematically justified

@@ -69,8 +69,8 @@ doAngle.pcaRes <- function(x, data, B, ...) {
   len1 <- len2 <- matrix(NA, d, d)
 
   for (i in 1:d) for (j in 1:d) {
-    len1[i,j] <- sqrt(eigen1[i]/max.eigen*abs(sum(load1[,i]*load2[,j])))
-    len2[i,j] <- sqrt(eigen2[j]/max.eigen*abs(sum(load1[,i]*load2[,j])))
+    len1[i,j] <- sqrt(eigen1[i]/max.eigen)*abs(sum(load1[,i]*load2[,j]))
+    len2[i,j] <- sqrt(eigen2[j]/max.eigen)*abs(sum(load1[,i]*load2[,j]))
   }
 
   #Store in data.frame ready for plotting

### 2018-06-17 article/Old versions/MPS/PCADSC_article.tex
New versions of all figures after update of anglePlot
Started editing of manuscript
CE-plot section updated

@@
 \usepackage{apalike}
-\usepackage[utf8x]{inputenc}
+%\usepackage[utf8x]{inputenc}
+\usepackage[utf8]{inputenc}
 %\usepackage{nameref,hyperref}
@@
 The tools presented here are all based on comparing the PCA results across two different datasets that contain the same variables, $\mathbf{X}$ and $\mathbf{Y}$. Let 
-$$\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{n_x + n_y \times d}$$
+$$\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{(n_x + n_y) \times d}$$
  be the combined dataset. For each of these three datasets, we complete the following steps (here described for $\mathbf{X}$ only):
@@
 This identity will simplify some expressions below.
-Using a greedy interpretation of PCA, which is often overlooked in the literature, we can note that the sequence of loadings $\eta_j$ and their associated eigenvalues $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$.
+We note that the sequence of loadings $\eta_j$ and their associated eigenvalues $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$.
 This implies that the loadings and the eigenvalues can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.
@@
 
-In \textit{Model 1} the observed variables $V_3$ and $V_4$ are caused by a common, unobserved variable $L_1$.
+In \textit{model 1} the observed variables $V_3$ and $V_4$ are caused by a common, unobserved variable $L_1$.
 Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models, in Appendix S2.
@@
 Moreover, a positive cumulative difference implies that dataset $\mathbf{X}$ holds more information in the first components than dataset $\mathbf{Y}$ does.
-By using the cumulative eigenvalues of $\mathbf{Z}$ as the first coordinate, we obtain a visualization where a component take up an amount of horizontal space that corresponds to the variance explained by the component (that is, its eigenvalue), thereby letting the more infuential components drive the visual impression.
+By using the cumulative eigenvalues of $\mathbf{Z}$ as the first coordinate, we obtain a visualization where a component take up an amount of horizontal space that corresponds to the variance explained by the component (that is, its eigenvalue), thereby letting the more influential components drive the visual impression.
 
@@
 
-Two examples of cumulative eigenvalue plots are available in Figure \ref{plot.simCE}.
-In the top panel, we see a cumulative eigenvalue curve that falls well within the confidence region, thereby suggesting no difference in eigenvalues (as expected).
-This also agrees with the large $p$-values and the fact that only a single underlying model was used for generating this data.
-In the bottom panel, the conclusion is reversed, which is also in accordance with the nature of the simulated data, which contains two distinct groups.
 In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
@@
 If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicates differences between the two datasets.
+
+Two examples of cumulative eigenvalue plots are available in Figure \ref{plot.simCE}.
+In the top panel, we see a cumulative eigenvalue curve that falls well within the confidence region, thereby suggesting no difference in eigenvalues (as expected).
+This also agrees with the large $p$-values and the fact that only a single underlying model was used for generating this data.
+In the bottom panel, the conclusion is reversed, which is also in accordance with the nature of the simulated data, which contains two distinct groups.
 
@@
 Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance contributions.
-The absolute value of the projection $\eta_{k}^x {\eta_{j}^y}^\top$ is inserted due to the indeterminacy of the direction of loading vectors.
+The absolute value of the projection ${\eta_{k}^x}^\top \eta_{j}^y$ is inserted due to the indeterminacy of the direction of loading vectors.
 This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal.
@@
 \center
-\includegraphics[scale=0.8]{Figure2_v2.pdf}
+\includegraphics[scale=0.8]{Figure2_v3.pdf}
 \caption{CE plots for the simulated datasets.
@@
 \center
-\includegraphics[scale=0.8]{Figure3_v2.pdf}
+\includegraphics[scale=0.8]{Figure3_v3.pdf}
 \caption{Angle plots for the simulated datasets.
@@
 \center
-\includegraphics[scale=0.8]{Figure4_v2.pdf}
+\includegraphics[scale=0.8]{Figure4_v3.pdf}
 \caption{Chroma plots for the simulated datasets.
@@
 \center
-\includegraphics[scale=0.8]{Figure6_v2.pdf}
+\includegraphics[scale=0.8]{Figure6_v3.pdf}
 \caption{CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being.
@@
 \center
-\includegraphics[scale=0.8]{Figure7_v2.pdf}
+\includegraphics[scale=0.8]{Figure7_v3.pdf}
 \caption{Chroma plot for comparing Danish and Bulgarian data on psychological well-being.
@@
 \center
-\includegraphics[scale=0.8]{Figure8_v2.pdf}
+\includegraphics[scale=0.8]{Figure8_v3.pdf}
 \caption{A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being.
@@
 \center
-\includegraphics[scale=0.8]{Figure9_v2.pdf}
+\includegraphics[scale=0.8]{Figure9_v3.pdf}
 \caption{Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples.

### 2018-06-17 article/Old versions/MPS/PCADSC_article.tex
angle-plot updated!

@@
 Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.
-Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data.
+
+In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the $\mathbf{A}$ dataset in the coordinate system of the $\mathbf{B}$ dataset.
+Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the $\mathbf{B}$ dataset in the coordinate system of the $\mathbf{A}$ dataset.
+\hl{If we have coinciding blue and red arrows along the diagonal in the grid display, and arrows of vanishing lengths in the off-diagonal cells, then the structures of the two datasets are identical.
+In contrast, differences in the eigenvalues are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
+And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.}
+
+\hl{However, there is a generic situation where the angles between the arrows lack discriminative power, namely when two or more neighbouring eigenvalues are identical or close.
+For instance, multiple eigenvalues close to 1 arise if more than one of the variables are uncorrelated, or close to uncorrelated, of all the other variables.
+In this situation only the corresponding eigenspace is well estimates, but the rotation of the associated eigenvectors will be subject to sampling variability.
+To investigate if we are in such a situation of identical data structures with neighbouring eigenvalues being close, where comparably long arrows may occur in the corresponding blocks of off-diagonal cells, we have implemented a permutation test similar as done for the CE-plot.
+For each random reallocation we recompute the angles $\theta_{jk}$.
+The resampled angles may both be used for quantitative tests, where large angles are critical for the diagonal cells and small angles are critical for the off-diagonal cells, as well as for the visualisation of the sampled angle regions.
+In the visualisation we propose to present the entire sampled region of angles for each cell, but grey shaded according to percentiles such that more critical angles are lighter grey.
+In our implementation we do not present the p-values for the quantitative tests, but the p-values are available in the R-object to support the visualization if necessary.
+However, it is important to remember that there are $d \times d$ p-values, which means that attention should be given to a possible multiple testing issue.}
 
@@
 
-\hl{For dataset $\mathbf{A}$, we find very short off-diagonal arrows, suggesting that the two PCA decompositions agree on the relative placement of the information in  the data among the 6 PCs.
+\hl{Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data.
+For dataset $\mathbf{A}$, we find very short off-diagonal arrows except for the off-diagonal cells corresponding to the 3'rd and 4'th eigenvalues.
+However, we see that the arrows in cell (3,4) and (4,3) lie well within the shaded angle regions (with associated p-values 0.9119 and 0.9103).
+This indicates that we are in a situation with identical data structure, and with 3'rd and 4'th eigenvalues being close (this corresponds to the two independent variables in \emph{model 1}).
 For dataset $\mathbf{B}$, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}
-
-In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the $\mathbf{A}$ dataset in the coordinate system of the $\mathbf{B}$ dataset.
-Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the $\mathbf{B}$ dataset in the coordinate system of the $\mathbf{A}$ dataset.
-\hl{If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length.
-Differences in the eigenvalues are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
-And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.}
 

### 2018-06-17 ESS/sim_notes.R

@@ -6,7 +6,8 @@ library(gridExtra)
 #Simulate data from covariance matrices
 #Note: Names (S1 and S2) are opposite of that in the article.
 
-set.seed(11234)
+#set.seed(11234)
+set.seed(170618)
 
 S1 <- diag(6)
 S1[c(1, 2, 3), 1] <- S1[1, c(1, 2, 3)] <- c(1, 0.2, 0.1)
@@ -31,7 +32,6 @@ dF1$group <- rep(c("Group 1", "Group 2"), each = 500)
 
 dF2 <- as.data.frame(mvrnorm(1000, rep(0,6), S2))
 dF2$group <- rep(c("Group 1", "Group 2"), each = 500)
-
 
 #Do PCADSC
 po1 <- PCADSC(dF1, "group")
@@ -128,17 +128,16 @@ chroma2 <- chromaPlot(po2) +
 #ggsave(plot = chroma2, file = "./article/simChroma2.pdf",
 #       width = 8, height = 5)
 
-
+#P:/PCADSC/R/
 ggsave(plot = grid.arrange(CE2, CE1, nrow = 2),
-        file = "P:/PCADSC/R/article/Old versions/MPS/Figure2_v3.pdf",
+       file = "article/Old versions/MPS/Figure2_v3.pdf",
        dpi = 400, width = 7, height = 8)
 
-
 ggsave(plot = grid.arrange(angle2, angle1, nrow = 2),
-       file = "P:/PCADSC/R/article/Old versions/MPS/Figure3_v3.pdf",
+       file = "article/Old versions/MPS/Figure3_v3.pdf",
        dpi = 400, width = 7, height = 8)
 
 ggsave(plot = grid.arrange(chroma2, chroma1, nrow = 2),
-       file = "P:/PCADSC/R/article/Old versions/MPS/Figure4_v3.pdf",
+       file = "article/Old versions/MPS/Figure4_v3.pdf",
        dpi = 400, width = 7, height = 8)
 

### 2018-06-18 article/Old versions/MPS/PCADSC_article.tex
Data analysis updated

@@
 In our implementation we do not present the p-values for the quantitative tests, but the p-values are available in the R-object to support the visualization if necessary.
-However, it is important to remember that there are $d \times d$ p-values, which means that attention should be given to a possible multiple testing issue.}
+It is important to remember that there are $d \times d$ p-values, which means that attention should be given to a possible multiple testing issue.}
 
@@
 For dataset $\mathbf{A}$, we find very short off-diagonal arrows except for the off-diagonal cells corresponding to the 3'rd and 4'th eigenvalues.
-However, we see that the arrows in cell (3,4) and (4,3) lie well within the shaded angle regions (with associated p-values 0.9119 and 0.9103).
-This indicates that we are in a situation with identical data structure, and with 3'rd and 4'th eigenvalues being close (this corresponds to the two independent variables in \emph{model 1}).
+But since the arrows in cells (3,4) and (4,3) lie well within the shaded angle regions (supported by p-values 0.9119 and 0.9103) this is in agreement with identical data structures, where the 3'rd and 4'th eigenvalues are close (in this situation corresponding to the two independent variables in \emph{model 1}).
 For dataset $\mathbf{B}$, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}
@@
 
-In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though Veenhoven presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability \cite{Veenhoven2012}, while Lolle \& Andersen show highly potent translation issues for the term \textit{happiness} \cite{Lolle2016}.
+In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparability, though Veenhoven presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability \cite{Veenhoven2012}, while Lolle \& Andersen show highly potent translation issues for the term \textit{happiness} \cite{Lolle2016}.
 
@@
 The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures.
-And therefore, comparing the two countries in these measures is not a meaningful endeavor.
+And therefore, comparing the two countries in these measures is not a meaningful endeavour.
 
@@
 Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3.
-Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.
+\hl{These statements are supported by p-values for the hypotheses of no loading that all are less than $0.002$.} Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.
 
@@
 This general approach to model-fit evaluation is very helpful in combination with permutation tests, as inspired by Lin and colleagues \cite{LinEtAl2002}.
-Similarly, the angle plot could also be annotated with $p$-values from a permutation test based on appropriate test statistics.
+
+%Similarly, the angle plot could also be annotated with $p$-values from a permutation test based on appropriate test statistics.
 

### 2018-06-23 article/Old versions/MPS/PCADSC_article.tex
Minor changes to manuscript + figure code

@@
 We let $\bar{X} = (\bar{X}_1, ...
-, \bar{X}_d)^T = \frac{1}{n_x} \sum_{i = 1} x_{i \cdot}$ denote the variable averages and use $S_x =  \frac{1}{n_x - 1} \sum_{i = i}^n (x_{i \cdot} - \bar{X}) - (x_{i \cdot} - \bar{X})^T \in \RR^{d \times d}$ to denote the empirical covariance matrix of $\mathbf{X}$.
+, \bar{X}_d)^T = \frac{1}{n_x} \sum_{i = 1} x_{i \cdot}$ denote the variable averages and use $S_x =  \frac{1}{n_x - 1} \sum_{i = i}^n (x_{i \cdot} - \bar{X})(x_{i \cdot} - \bar{X})^T \in \RR^{d \times d}$ to denote the empirical covariance matrix of $\mathbf{X}$.
 
@@
 The absolute value of the projection ${\eta_{k}^x}^\top \eta_{j}^y$ is inserted due to the indeterminacy of the direction of loading vectors.
-This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal.
+This indeterminacy implies that the angle between loadings from the two datasets can always be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal.
 Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.
@@
 Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the $\mathbf{B}$ dataset in the coordinate system of the $\mathbf{A}$ dataset.
-\hl{If we have coinciding blue and red arrows along the diagonal in the grid display, and arrows of vanishing lengths in the off-diagonal cells, then the structures of the two datasets are identical.
+If we have coinciding blue and red arrows along the diagonal in the grid display, and arrows of vanishing lengths in the off-diagonal cells, then the structures of the two datasets are identical.
 In contrast, differences in the eigenvalues are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal.
-And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.}
+And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.
 
-\hl{However, there is a generic situation where the angles between the arrows lack discriminative power, namely when two or more neighbouring eigenvalues are identical or close.
+However, there is a generic situation where the angles between the arrows lack discriminative power, namely when two or more neighboring eigenvalues are identical or close.
+We refer to this situation as \textit{eigenvalue indeterminancy}.
 For instance, multiple eigenvalues close to 1 arise if more than one of the variables are uncorrelated, or close to uncorrelated, of all the other variables.
-In this situation only the corresponding eigenspace is well estimates, but the rotation of the associated eigenvectors will be subject to sampling variability.
+In this situation only the corresponding eigenspace is well estimated, but the rotation of the associated eigenvectors will be subject to sampling variability.
 To investigate if we are in such a situation of identical data structures with neighbouring eigenvalues being close, where comparably long arrows may occur in the corresponding blocks of off-diagonal cells, we have implemented a permutation test similar as done for the CE-plot.
@@
 The resampled angles may both be used for quantitative tests, where large angles are critical for the diagonal cells and small angles are critical for the off-diagonal cells, as well as for the visualisation of the sampled angle regions.
-In the visualisation we propose to present the entire sampled region of angles for each cell, but grey shaded according to percentiles such that more critical angles are lighter grey.
-In our implementation we do not present the p-values for the quantitative tests, but the p-values are available in the R-object to support the visualization if necessary.
-It is important to remember that there are $d \times d$ p-values, which means that attention should be given to a possible multiple testing issue.}
+In the visualization we propose to present the entire sampled region of angles for each cell, but gray shaded according to percentiles such that more critical angles are lighter gray.
+%In our implementation, we do not present the $p$-values for the quantitative tests, but the $p$-values are available in the R-object to support the visualization if necessary.
+It is important to remember that there are $d \times d$ $p$-values, which means that attention should be given to a possible multiple testing issue.
 
@@
 
-\hl{Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data.
-For dataset $\mathbf{A}$, we find very short off-diagonal arrows except for the off-diagonal cells corresponding to the 3'rd and 4'th eigenvalues.
-But since the arrows in cells (3,4) and (4,3) lie well within the shaded angle regions (supported by p-values 0.9119 and 0.9103) this is in agreement with identical data structures, where the 3'rd and 4'th eigenvalues are close (in this situation corresponding to the two independent variables in \emph{model 1}).
-For dataset $\mathbf{B}$, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}
+Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data.
+For dataset $\mathbf{A}$, we find very short off-diagonal arrows except for the off-diagonal cells corresponding to the 3rd and 4th eigenvalues.
+But since the arrows in cells (3,4) and (4,3) lie well within the shaded angle regions (supported by p-values 0.9119 and 0.9103) this is in agreement with identical data structures, where the 3rd and 4th eigenvalues suffer from indeterminancy (in this situation due to the two independent variables in \emph{model 1}).
+For dataset $\mathbf{B}$, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.
 
@@
 In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{i}^x = \frac{\lambda_{i}^x}{d}$ and $\tilde\sigma_{i}^y = \frac{\lambda_{i}^y}{d}$.
+If there is eigenvalue indeterminancy for neighboring components, as discussed above, the loading patterns for these components will not have a meaningful interpretation.
+Therefore, one should not compare the chroma bars for component that were identified to suffer from eigenvalue indeterminancy by the angle plot.
 
@@
 In Figure  \ref{plot.simChroma}, two examples of chroma plots are available.
-For dataset $\mathbf{A}$, we see very similar visual patterns across the two panels, corresponding well with the fact that both groups are drawn from the same underlying model.
+For dataset $\mathbf{A}$, we see very similar visual patterns across the two panels for components 1, 2, 5 and 6, corresponding well with the fact that both groups are drawn from the same underlying model.
+Components 3 and 4 do not have similar loading patterns across the two subsets of the data, but this is expected since we identified these components to suffer from eigenvalue indeterminancy in the angle plot.
 For dataset $\mathbf{B}$, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups.
@@
 If the angle plots suggests differences in data structures, identify which components are responsible for the differences.
-\item Make a chroma plot of the components selected to be different by the angle plot. Use this information to discuss \textit{how} and \textit{why} the data structures might be different. 
+\item Make a chroma plot of the components selected to be different by the angle plot, but whose differences are not due to eigenvalue indeterminancy among neighboring components.
+Use this information to discuss \textit{how} and \textit{why} the data structures might be different.
 \end{enumerate}
@@
 Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3.
-\hl{These statements are supported by p-values for the hypotheses of no loading that all are less than $0.002$.} Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.
+%\hl{These statements are supported by p-values for the hypotheses of no loading that all are less than $0.002$.} 
+Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.
 

### 2018-06-23 ESS/sim_notes.R

@@ -38,7 +38,7 @@ po1 <- PCADSC(dF1, "group")
 po2 <- PCADSC(dF2, "group")
 
 CE1 <- CEPlot(po1) +
-  scale_y_continuous(breaks = round(seq(-0.6, 0.6, 0.1),1), limits = c(-0.5, 0.3)) +
+  scale_y_continuous(breaks = round(seq(-0.6, 0.6, 0.1),1), limits = c(-0.6, 0.3)) +
   ggtitle("Dataset B", subtitle = "Cumulative difference in eigenvalues: Group 1 - Group 2") +
   theme(panel.border = element_blank(),
         axis.line=element_line(),
@@ -47,7 +47,7 @@ CE1 <- CEPlot(po1) +
   ylab("")
 
 CE2 <- CEPlot(po2) +
-  scale_y_continuous(breaks = round(seq(-0.6, 0.6, 0.1),1), limits = c(-0.3, 0.3)) +
+  scale_y_continuous(breaks = round(seq(-0.6, 0.6, 0.1),1), limits = c(-0.2, 0.2)) +
   ggtitle("Dataset A", subtitle = "Cumulative difference in eigenvalues: Group 1 - Group 2") +
   theme(panel.border = element_blank(),
         axis.line=element_line(),

### 2019-12-18 article/Old versions/JAS/PCADSC_article_JAS_resub_jan2020.tex
old changes to meta files

--- 2018-06-23 article/Old versions/MPS/PCADSC_article.tex
@@
-\documentclass[a4paper,12pt]{article}
+% interacttfssample.tex
+% v1.05 - August 2017
 
-\usepackage{geometry}
-\geometry{margin=1in}
+\documentclass[]{interact}
 
-\usepackage[english]{babel}
-\usepackage{authblk}
-\usepackage{geometry}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{mathrsfs}
-%\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
-\usepackage{apalike}
-%\usepackage[utf8x]{inputenc}
-\usepackage[utf8]{inputenc}
-%\usepackage{nameref,hyperref}
-% line numbers
-%\usepackage[right]{lineno}
-% ligatures disabled
+\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
+\usepackage[caption=false]{subfig}% Support for small, `sub' figures and tables
+%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
+
+%\usepackage[doublespacing]{setspace}% To produce a `double spaced' document if required
+%\setlength\parindent{24pt}% To increase paragraph indentation when line spacing is doubled
+%\setlength\bibindent{2em}% To increase hanging indent in bibliography when line spacing is doubled
+
+\usepackage[numbers,sort&compress]{natbib}% Citation support using natbib.sty
+\bibpunct[, ]{[}{]}{,}{n}{,}{,}% Citation support using natbib.sty
+\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
+
+\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{proposition}[theorem]{Proposition}
+
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{example}[theorem]{Example}
+
+\theoremstyle{remark}
+\newtheorem{remark}{Remark}
+\newtheorem{notation}{Notation}
+
 \usepackage{microtype}
@@
 \definecolor{lightgray}{HTML}{e6e6e6}
+
+%\usepackage{caption}
+%\captionsetup{font={stretch=2}}
 
@@
 
-\usepackage{setspace}
-\doublespacing
+%\usepackage{setspace}
+%\doublespacing
 
@@
 
-\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
-\renewcommand{\baselinestretch}{2}
+\makeatletter
+\newcounter{savesection}
+\newcounter{apdxsection}
+\renewcommand\appendix{\par
+  \setcounter{savesection}{\value{section}}%
+  \setcounter{section}{\value{apdxsection}}%
+  \setcounter{subsection}{0}%
+  \gdef\thesection{\@Alph\c@section}}
+\newcommand\unappendix{\par
+  \setcounter{apdxsection}{\value{section}}%
+  \setcounter{section}{\value{savesection}}%
+  \setcounter{subsection}{0}%
+  \gdef\thesection{\@arabic\c@section}}
+\makeatother
 
-\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
-T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-
-\def\volumeyear{2018}
+%\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
+%\renewcommand{\baselinestretch}{2}
 
@@
 
-%\runninghead{PCADSC}
+\articletype{Application Note}% Specify the article type or omit as appropriate
 
@@
 
-\author[1]{}
-%\author[1]{Anne Helby Petersen}
-%\author[1]{Bo Markussen}
-%\author[3]{Karl Bang Christensen}
-%\affil[1]{Department of Public Health, University of Copenhagen, Copenhagen, Denmark}
-%\affil[2]{Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}
-\date{}
+%\author{
+%\name{Anne Helby Petersen\textsuperscript{a}\thanks{CONTACT Anne Helby Petersen. Email: ahpe@sund.ku.dk}, Bo Markussen\textsuperscript{b} and Karl Bang Christensen\textsuperscript{a}}
+%\affil{\textsuperscript{a}Department of Public Health, University of Copenhagen, Copenhagen, Denmark; \textsuperscript{b}Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}
+%}
+
 \maketitle
-
-%\corrauth{Anne Helby Petersen}
-%\email{ahpe@sund.ku.dk}
-
-\renewcommand{\abstractname}{Abstract}
-\renewcommand{\figurename}{Figure}
-\renewcommand{\tablename}{Table}
 
@@
 
-%% Include all macros below
+\begin{keywords}
+Principal component analysis, exploratory data analysis, covariance matrix,
+data structure 
+\end{keywords}
+\title{}
 
-\section*{Introduction}\label{sec:introduction}
+\section{Introduction}\label{sec:introduction}
 
-Data comparability is a classical topic in statistics and a recurring one in psychometrics.
-Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability needs to be assessed empirically.
-This happens for instance when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration, or the order of survey questionnaire items are applied.
+Data comparability is a recurring topic in a wide array of statistical applications.
+Often, data will be collected in such a way that it is essentially divided into several subsets whose comparability needs to be assessed empirically.
+This happens for instance when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration of a survey, or the order of survey questionnaire items are applied.
 While a mixture of modes of administration, for example mail and telephone, can improve response rates in surveys, it can also be quite problematic by inducing differences in response behavior that may lead to biased results \cite{Brambilla1987,McHorney1994}.
 Powers, Mishra, and Young report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful \cite{Powers2005}, so the comparability issue is not just a statistical puzzle.
-Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu (2016), who warns against combining online panel data with intercept samples (a pool of respondents obtained through banners, ads, or promotions).\nocite{Liu2016} But mixing several data collection methods is an essential part of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions.
+Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu, who warns against combining online panel data with intercept samples (a pool of respondents obtained through banners, ads, or promotions)\cite{Liu2016}.
+But mixing several data collection methods is essential for acheiving further advances in surveying- and data collection methodology, as it allows us to simultaneously build on existing methods and answer new empirical questions.
 However, if we do not address the question of comparability among the data subsets arriving from e.g.
@@
 The proposed tools are implemented in the \texttt{R} package \texttt{PCADSC} \cite{PCADSC}.
-Below, we describe the procedures and present a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey.
+%Below, we describe the procedures and present a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey.
 More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is ignored e.g.
@@
 
-\section*{PCA-based tools for data structure comparisons}\label{sec:pcadscintro}
+\subsection{Comparing psychological well-being}
+An example of where the data structure comparability issue must be considered is when conducting international rankings of countries according to e.g.
+educational quality (e.g.
+the PISA project) and citizen happiness (e.g.
+the UN \textit{World Happiness Report} project).
+From a methodological point of view, such international rankings are problematic, as they rely on the fundamental assumption that the measured concepts are inherently the same across countries.
+
+We use data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
+In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparability, though Veenhoven presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability \cite{Veenhoven2012}, while Lolle \& Andersen show highly potent translation issues for the term \textit{happiness} \cite{Lolle2016}.
+If for instance two countries differ in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one.
+More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital, and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being.
+The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures.
+And therefore, comparing the two countries in these measures is not a meaningful endeavour.
+
+Our starting point is Denmark, a small, Northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \cite{WHR2016}, and we wish to investigate if this title is really meaningful at all.
+In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria.
+Though both countries are European and thus neither very far apart geographically nor culturally, these two countries have previously been highlighted to be very different in terms of what defines happiness \cite{ESStopline5}.
+Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \cite{Rodriguez2014}.
+When comparing Northern European countries to other European countries, a much less pronounced relationship between the two concepts is found.
+In particular, interpersonal relations play a less important role in Denmark, compared to Bulgaria.
+Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.
+We also compare the Danish data with Swedish data in order to investigate if the PCADSC tools actually do hold discriminatory power.
+Denmark and Sweden are both Northern European countries and are often deemed very similar in terms of culture and history.
+Therefore, we expect fundamental concepts, such as psychological well-being, to be similar across these two countries.
+
+\section{Methodology}\label{sec:pcadscintro}
 
@@
 Computing empirical covariance matrices for the two subsets and comparing them entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets.
-We refer to Appendix S1 for a brief introduction to PCA.
+%We refer to Appendix \ref{appendix.introPCA} for a brief introduction to PCA.
 
@@
 
-\subsection*{Using PCA for data structure comparisons}
+\subsection{Using PCA for data structure comparisons}
 %PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and eigenvalues are meaningful and interesting objects to compare across such different datasets.
@@
 The tools presented here are all based on comparing the PCA results across two different datasets that contain the same variables, $\mathbf{X}$ and $\mathbf{Y}$. Let 
-$$\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{(n_x + n_y) \times d}$$
+\begin{equation*}
+\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{(n_x + n_y) \times d}
+\end{equation*}
  be the combined dataset. For each of these three datasets, we complete the following steps (here described for $\mathbf{X}$ only):
@@
 In \textit{model 1} the observed variables $V_3$ and $V_4$ are caused by a common, unobserved variable $L_1$.
-Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models, in Appendix S2.
+Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models, in Appendix \ref{appendix.simData}.
 In dataset $\mathbf{A}$, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures.
@@
 
-\subsection*{The cumulative eigenvalue plot}
+\subsection{The cumulative eigenvalue plot}
 The cumulative eigenvalue (CE) plot compares the eigenvalues of the correlation matrices.
@@
 
-\subsection*{The angle plot}
+\subsection{The angle plot}
 The angle plot compares both loadings and eigenvalues at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e.
@@
 
-\subsection*{The chroma plot}
+\subsection{The chroma plot}
 The chroma plot is primarily an illustration of differences in loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
@@
 
-\subsection*{A data structure comparison workflow}
+\subsection{A data structure comparison workflow}
 %\begin{figure}
@@
 
-\section*{Comparing psychological well-being: A data example}\label{sec:dataexample}
+\section{Application}\label{sec:dataexample}
 
-We illustrate the three tools using data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
-The topic of our investigation is motivated by an increasingly popular tendency to publish rankings of countries in fields as different as educational quality (e.g.
-the PISA project) and citizen happiness (e.g.
-the UN \textit{World Happiness Report} project).
-From a methodological point of view, such international rankings are problematic, as they rely on the fundamental assumption that the measured concepts are inherently the same across countries.
-The PCADSC tools qualify as a suite of methods for exploring the validity of this assumption empirically.
-
-In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparability, though Veenhoven presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability \cite{Veenhoven2012}, while Lolle \& Andersen show highly potent translation issues for the term \textit{happiness} \cite{Lolle2016}.
-
-If for instance two countries differ in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one.
-More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital, and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being.
-The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures.
-And therefore, comparing the two countries in these measures is not a meaningful endeavour.
-
-In this section, we use the PCADSC tools to examine international differences in one aspect of happiness, namely psychological well-being.
-Our starting point is Denmark, a small, Northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \cite{WHR2016}, and we wish to investigate if this title is really meaningful at all.
-In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria.
-Though both countries are European and thus neither very far apart geographically nor culturally, these two countries have previously been highlighted to be very different in terms of what defines happiness \cite{ESStopline5}.
-Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \cite{Rodriguez2014}.
-When comparing Northern European countries to other European countries, a much less pronounced relationship between the two concepts is found.
-In particular, interpersonal relations play a less important role in Denmark, compared to Bulgaria.
-Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.
-
-We also compare the Danish data with Swedish data in order to investigate if the PCADSC tools actually do hold discriminatory power.
-Denmark and Sweden are both Northern European countries and are often deemed very similar in terms of culture and history.
-Therefore, we expect fundamental concepts, such as psychological well-being, to be similar across these two countries.
-
+We now illustrate the three PCADSC tools using data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
 All computations and figures presented in this section were created using the \R{R} package \R{PCADSC} \cite{PCADSC}.
 
-\subsection*{Data}
+\subsection{Data}
 
@@
 Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
-These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing}, and \textit{Supportive relationships}.
-More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in Appendix S3.
+These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative well-being}, \textit{Emotional well-being}, \textit{Functioning}, \textit{Vitality}, \textit{Community well-being}, and \textit{Supportive relationships}.
+More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in Appendix \ref{appendix.essscales}.
 We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10.
@@
 All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations.
-Table~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
+Figure~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
 
-\begin{table}[ht]
-\centering
-\caption{\textbf{Median values of each of the six dimensions of psychological well-being, stratified by country.} 25 and 75 percentiles are listed in parentheses.
+%\begin{figure}[ht]
+%\centering
+%\caption{\textbf{Median values of each of the six dimensions of psychological well-being, stratified by country.} 25 and %75 percentiles are listed in parentheses.
 The scales are constructed such that they all run from 0-10.}
-\label{tableDistr}
+%\label{tableDistr}
+%
+%\begin{tabular}{lccc}
+%\hline
+%  & Denmark & Bulgaria & Sweden \\
+%\hline
+%Evaluative well-being    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
+%Emotional well-being     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
+%Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
+%Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
+%Community well-being     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
+%Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
+%\hline
+%\end{tabular} 
+%\end{figure}
 
-\begin{tabular}{lccc}
-\hline
-  & Denmark & Bulgaria & Sweden \\
-\hline
-Evaluative wellbeing    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
-Emotional wellbeing     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
-Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
-Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
-Community wellbeing     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
-Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
-\hline
-\end{tabular} 
-\end{table}
-
-\subsection*{Comparing Denmark and Bulgaria}
+\subsection{Comparing Denmark and Bulgaria}
 Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales.
@@
 The chroma plot in Figure \ref{plotBG.pancake} allows us to look closer into these components.
-Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish.
+Here, we find that the relative importance of the \textit{Community well-being} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish.
 In the Danish data, on the other hand, we find that \textit{Vitality} and \textit{Emotional well-being} seem to play bigger roles, as they appear with larger loadings in more high-ranking components in this sample, relative to the Bulgarian.
@@
 
-\subsection*{Comparing Denmark and Sweden}
+\subsection{Comparing Denmark and Sweden}
 We now turn to the comparison of Denmark and Sweden in terms of psychological well-being.
@@
 We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25\% of the variance in both datasets.
-In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and for instance conclude using Table~\ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally happier than the least happy people in Sweden.
+In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and for instance conclude using Figure~\ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally happier than the least happy people in Sweden.
 A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.
@@
 
-\section*{Discussion}\label{sec.Discussion}
+\section{Conclusion and discussion}\label{sec.Discussion}
 
@@
 
-\section*{Conclusion}
+%\section{Conclusion}
 
-We have presented three new tools for visual, exploratory data structure comparisons that utilizes covariance matrix decompositions obtained by use of principal component analysis.
+%We have presented three new tools for visual, exploratory data structure comparisons that utilizes covariance matrix decompositions obtained by use of principal component analysis.
 The tools compare the linear structures of numerical datasets.
@@
 
-\bibliographystyle{apalike}
+\bibliographystyle{tfs}
 \bibliography{bib}
+
+\appendix
+
+\section{Details about the simulated data}
+\label{appendix.simData}
+This supplementary appendix describes the origin of the simulated data used for presenting the PCADSC plots.
+Dataset $\mathbf{A}$ consisted of 1000 independent simulated realizations from the $N(0, \Sigma_1)$-distribution, while dataset $\mathbf{B}$ consisted of 500 independent realizations from $N(0, \Sigma_1)$ and 500 independent realizations from $N(0, \Sigma_2)$.
+Dataset $\mathbf{A}$ was furthermore appended with a grouping variable, randomly dividing the observations into two groups.
+Dataset $\mathbf{B}$ also included a grouping variable, and this variable contained information about which of the two normal distributions each observation was simulated from.
+The covariance matrices, $\Sigma_A$ and $\Sigma_B$, were defined by
+
+\begin{equation*}
+\Sigma_A = \begin{pmatrix}
+    1.0  &  0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
+ 0.0  &  1.0 & 0.0 & 0.0&  0.0 & 0.0 \\
+ 0.0   & 0.0 &  1.0 & 0.7 & 0.0 & 0.0 \\
+ 0.0 &  0.0 & 0.7 & 1.0 & 0.0 & 0.0 \\
+0.0 &    0.0 &  0.0 & 0.0 & 1.0 & 0.4 \\
+ 0.0 & 0.0 &  0.0 & 0.0 & 0.4 & 1.0
+\end{pmatrix}
+\end{equation*}
+and
+\begin{equation*}
+\Sigma_B = \begin{pmatrix}
+ 1.0 & 0.2 & 0.1 & 0.0 & 0.0 & 0.0 \\
+ 0.2 & 1.0 & 0.1 & 0.0 & 0.0 & 0.0 \\
+ 0.1 & 0.1 & 1.0 & 0.0 & 0.0 & 0.0 \\
+ 0.0 & 0.0 & 0.0 & 1.0 & 0.3 & 0.1 \\
+ 0.0 & 0.0 & 0.0 & 0.3 & 1.0 & 0.2 \\
+ 0.0 & 0.0 & 0.0 & 0.1 & 0.2 & 1.0
+\end{pmatrix},
+\end{equation*}
+respectively.
+
+\section{ESS psychological well-being scale definitions}
+\label{appendix.essscales}
+\begin{table}[H]
+\renewcommand\thetable{C1}
+\centering
+\caption{ESS psychological well-being scale definitions.
+Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
+Before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
+\label{table:items}
+\scriptsize
+\rowcolors{1}{}{lightgray}
+\bgroup
+\def\arraystretch{1.5}
+\begin{tabular}{ll}
+\hline
+Scale & Items \\
+\hline
+\rowcolor{lightgray} & How satisfied with life as a whole \\
+\rowcolor{lightgray}\multirow{-2}{*}{Evaluative well-being }& How happy are you\\
+
+\rowcolor{white} & Felt sad, how often in the past week\\
+\rowcolor{white} & Felt depressed, how often in the past week\\
+\rowcolor{white} & Enjoyed life, how often in the past week\\
+\rowcolor{white} & Were happy, how often in the past week\\
+\rowcolor{white} & Felt anxious, how often in the past week\\
+\rowcolor{white}\multirow{-6}{*}{Emotional well-being} & Felt calm and peaceful, how often in the past week\\
+
+\rowcolor{lightgray} & Free to decide how to live my life\\
+\rowcolor{lightgray} & Little chance to show how capable I am\\
+\rowcolor{lightgray} & Feel accomplishment from what I do\\
+\rowcolor{lightgray} & Interested in what you are doing\\
+\rowcolor{lightgray} & Absorbed in what you are doing\\
+\rowcolor{lightgray} & Enthusiastic about what you are doing\\
+\rowcolor{lightgray} & Feel what I do in life is valuable and worthwhile\\
+\rowcolor{lightgray} & Have a sense of direction\\
+\rowcolor{lightgray} & Always optimistic about my future\\
+\rowcolor{lightgray} & There are lots of things I feel I am good at\\
+\rowcolor{lightgray} & In general feel very positive about myself\\
+\rowcolor{lightgray} & At times feel as if I am a failure\\
+\rowcolor{lightgray} & When things go wrong in my life it takes a long time to get back to normal\\
+\rowcolor{lightgray}\multirow{-14}{*}{Functioning} & Deal with important problems\\
+
+\rowcolor{white} & Felt everything did as effort, how often in the past week\\
+\rowcolor{white} & Sleep was restless, how often in the past week\\
+\rowcolor{white} & Could not get going, how often in the past week\\
+\rowcolor{white}\multirow{-4}{*}{Vitality} & Had a lot of energy, how often in the past week\\
+
+\rowcolor{lightgray} &  Most people can be trusted\\
+\rowcolor{lightgray} &  People try to take advantage\\
+\rowcolor{lightgray} &  Most of the time people are helpful\\
+\rowcolor{lightgray} &  Feel people in local area help one another\\
+\rowcolor{lightgray}\multirow{-5}{*}{Community well-being}& Feel close to the people in local area\\
+
+\rowcolor{white} & How many with whom you can discuss intimate matters\\
+\rowcolor{white} & Feel appreciated by those you are close to\\
+\rowcolor{white} & Receive help and support\\
+\rowcolor{white}\multirow{-4}{*}{Supportive relationships} & Felt lonely, how often in the past week\\
+\hline
+\end{tabular}
+\egroup
+\end{table}
+
+\unappendix
 
@@
 
-\section*{Figures}
+\setcounter{table}{0}
+
+\begin{table}[H]
+\centering
+\caption{Median values of each of the six dimensions of psychological well-being, stratified by country.
+25 and 75 percentiles are listed in parentheses.
+The scales are constructed such that they all run from 0-10.}
+\label{tableDistr}
+
+\begin{tabular}{lccc}
+\hline
+  & Denmark & Bulgaria & Sweden \\
+\hline
+Evaluative well-being    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
+Emotional well-being     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
+Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
+Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
+Community well-being     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
+Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
+\hline
+\end{tabular} 
+\end{table}
+
+\newpage
 
@@
  \end{tikzpicture}
- \caption{Graphical representations of models used for simulating example data.
+ \caption{}%Graphical representations of models used for simulating example data.
 Square nodes represent observed variables, while rounded nodes represent latent variables.
@@
 \center
-\includegraphics[scale=0.8]{Figure2_v3.pdf}
-\caption{CE plots for the simulated datasets.
+\includegraphics[scale=0.8]{Figure2_PCADSC_v3.pdf}
+\caption{}%CE plots for the simulated datasets.
 Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel.
@@
 \center
-\includegraphics[scale=0.8]{Figure3_v3.pdf}
-\caption{Angle plots for the simulated datasets.
+\includegraphics[scale=0.8]{Figure3_PCADSC_v3.pdf}
+\caption{}%Angle plots for the simulated datasets.
 Dataset A in top panel and dataset B in bottom panel.
@@
 \center
-\includegraphics[scale=0.8]{Figure4_v3.pdf}
-\caption{Chroma plots for the simulated datasets.
+\includegraphics[scale=0.8]{Figure4_PCADSC_v3.pdf}
+\caption{}%Chroma plots for the simulated datasets.
 Dataset $\mathbf{A}$ in top panel and dataset $\mathbf{B}$ in bottom panel.
@@
  \end{tikzpicture}
- \caption{The PCADSC workflow, starting with two datasets and then successively applying the three visual tools for obtaining a deeper and deeper understanding of similarities and dissimilarities in the data structures.}
+ \caption{}%The PCADSC workflow, starting with two datasets and then successively applying the three visual tools for obtaining a deeper and deeper understanding of similarities and dissimilarities in the data structures.}
  \label{fig:flowchart}
@@
 
+\newpage
+
 \begin{figure}[H]
 \center
-\includegraphics[scale=0.8]{Figure6_v3.pdf}
-\caption{CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being.
+\includegraphics[scale=0.8]{Figure7_PCADSC_v3.pdf}
+\caption{}%CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being.
 The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
@@
 \center
-\includegraphics[scale=0.8]{Figure7_v3.pdf}
-\caption{Chroma plot for comparing Danish and Bulgarian data on psychological well-being.
+\includegraphics[scale=0.8]{Figure8_PCADSC_v3.pdf}
+\caption{}%Chroma plot for comparing Danish and Bulgarian data on psychological well-being.
 A chroma plot comparing the 2nd, 3rd, and 4th principal components of the Bulgarian- and Danish psychological well-being data.
@@
 \center
-\includegraphics[scale=0.8]{Figure8_v3.pdf}
-\caption{A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being.
+\includegraphics[scale=0.8]{Figure9_PCADSC_v3.pdf}
+\caption{}%A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being.
 The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
@@
 \center
-\includegraphics[scale=0.8]{Figure9_v3.pdf}
-\caption{Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples.
+\includegraphics[scale=0.8]{Figure10_PCADSC_v3.pdf}
+\caption{}%Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples.
 For each component, the bar is annotated with its cumulative variance contribution, that is, how much variance can be explained by having information of this and the preceding components.}
@@
 
-\section*{Supplementary material}
+\begin{description}
+\item[Figure 1] Graphical representations of models used for simulating example data.
+Square nodes represent observed variables, while rounded nodes represent latent variables.
+Arrows are used to illustrate causal links.
+\item[Figure 2] CE plots for the simulated datasets.
+Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel.
+Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures.
+The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.
+\item[Figure 3] Angle plots for the simulated datasets.
+Dataset A in top panel and dataset B in bottom panel.
+Blue arrows show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.
+\item[Figure 4] Chroma plots for the simulated datasets.
+Dataset $\mathbf{A}$ in top panel and dataset $\mathbf{B}$ in bottom panel.
+Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.
+\item[Figure 5] The PCADSC workflow, starting with two datasets and then successively applying the three visual tools for obtaining a deeper and deeper understanding of similarities and dissimilarities in the data structures.
+\item[Figure 6] CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being.
+The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures.
+In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.
+\item[Figure 7] Chroma plot for comparing Danish and Bulgarian data on psychological well-being.
+A chroma plot comparing the 2nd, 3rd, and 4th principal components of the Bulgarian- and Danish psychological well-being data.
+The component-bars are annotated with their relative variance contributions.
+\item[Figure 8] A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being.
+The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.
+\item[Figure 9] Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples.
+For each component, the bar is annotated with its cumulative variance contribution, that is, how much variance can be explained by having information of this and the preceding components.
+\end{description}
 
-\subsection*{S1 Appendix: A brief introduction to principal component analysis}
-\label{appendix.introPCA}
-
-In this appendix we introduce PCA. We suppress the subscripts and superscripts used in the article because we only consider a single dataset here, denoted by $\mathbf{X} \in \RR^{d \times n}$.
-
-For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost.
-More specifically, we wish to minimize the loss when looking at a projection of $\mathbf{X}$ onto a $q$-dimensional space, rather than the original $d$-dimensional one.
-Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
-
-\begin{equation}
-\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_{i \cdot} - \bar{X} - z \rVert^2 =
-\min_{K_q} \sum_{i=1}^n \lVert x_{i \cdot} - \bar{X} - \text{proj}_{K_q}(x_{i \cdot} - \bar{X}) \rVert^2.
-\end{equation}
-
-The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \cite{HastieEtAl2009}.
-More specifically, the rank-q-reconstruction error is attained when we choose
-
-\begin{equation*}
-\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\},
-\end{equation*}
-where $\eta_1, ..., \eta_q$ are the first $q$ eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$.
-In the PCA framework, we refer to the eigenvectors as \textit{loadings}.
-The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
-\begin{equation*}
-\text{trace}(S) = \sum_{j=1}^d \hat{V}(X_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top \mathbf{X}^\top) = \sum_{j=1}^d \lambda_j,
-\end{equation*}
-where $X_j \in \RR^n$ denotes the $j$th variable of $\mathbf{X}$ and $\hat{V}$ is the empirical variance function.
-This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
-And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are distinct, the representation hereby obtained is a valid object of inference.
-
-The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
-\begin{align*}
-\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_{i \cdot} - \bar{X}) \rVert^2, &
-\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_{i \cdot} - \bar{X}) \rVert^2,
-\end{align*}
-where the initial subspace is defined as $\hat{K}_0 = \{0\}$.
-It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
-
-At this point a general remark about standardization is in place.
-PCA deconstructs the covariance matrix in components according to the most explained variance.
-This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will always be deemed highly influential, no matter the structure of the data.
-Therefore, the variables are typically standardized prior to performing PCA.
-The covariance matrix of the standardized variables is the same as the correlation matrix of the original variables, so post-standardization PCA simply corresponds to performing data structure comparisons of the correlation matrices rather than the covariance matrices.
-The standardization makes the variables comparable on the same scale, namely units of standard deviation.
-We have adopted this standardization policy.
-
-\subsection*{S2 Appendix: Details about the simulated data}
-\label{appendix.simData}
-This supplementary appendix describes the origin of the simulated data used for presenting the PCADSC plots.
-Dataset $\mathbf{A}$ consisted of 1000 independent simulated realizations from the $N(0, \Sigma_1)$-distribution, while dataset $\mathbf{B}$ consisted of 500 independent realizations from $N(0, \Sigma_1)$ and 500 independent realizations from $N(0, \Sigma_2)$.
-Dataset $\mathbf{A}$ was furthermore appended with a grouping variable, randomly dividing the observations into two groups.
-Dataset $\mathbf{B}$ also included a grouping variable, and this variable contained information about which of the two normal distributions each observation was simulated from.
-The covariance matrices, $\Sigma_A$ and $\Sigma_B$, were defined by
-
-$$\Sigma_A = \begin{pmatrix}
-    1.0  &  0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
- 0.0  &  1.0 & 0.0 & 0.0&  0.0 & 0.0 \\
- 0.0   & 0.0 &  1.0 & 0.7 & 0.0 & 0.0 \\
- 0.0 &  0.0 & 0.7 & 1.0 & 0.0 & 0.0 \\
-0.0 &    0.0 &  0.0 & 0.0 & 1.0 & 0.4 \\
- 0.0 & 0.0 &  0.0 & 0.0 & 0.4 & 1.0
-\end{pmatrix}$$
-and
-$$
-\Sigma_B = \begin{pmatrix}
- 1.0 & 0.2 & 0.1 & 0.0 & 0.0 & 0.0 \\
- 0.2 & 1.0 & 0.1 & 0.0 & 0.0 & 0.0 \\
- 0.1 & 0.1 & 1.0 & 0.0 & 0.0 & 0.0 \\
- 0.0 & 0.0 & 0.0 & 1.0 & 0.3 & 0.1 \\
- 0.0 & 0.0 & 0.0 & 0.3 & 1.0 & 0.2 \\
- 0.0 & 0.0 & 0.0 & 0.1 & 0.2 & 1.0
-\end{pmatrix},
-$$
-respectively.
-
-\subsection*{S3 Table: ESS psychological wellbeing scale definitions}
-\begin{table}[h]
-\centering
-\caption{{\bf ESS psychological wellbeing scale definitions.} Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
-Before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
-\label{table:items}
-
-\scriptsize
-\rowcolors{1}{}{lightgray}
-\bgroup
-\def\arraystretch{1.5}
-\begin{tabular}{ll}
-\hline
-Scale & Items \\
-\hline
-\rowcolor{lightgray} & How satisfied with life as a whole \\
-\rowcolor{lightgray}\multirow{-2}{*}{Evaluative wellbeing }& How happy are you\\
-
-\rowcolor{white} & Felt sad, how often in the past week\\
-\rowcolor{white} & Felt depressed, how often in the past week\\
-\rowcolor{white} & Enjoyed life, how often in the past week\\
-\rowcolor{white} & Were happy, how often in the past week\\
-\rowcolor{white} & Felt anxious, how often in the past week\\
-\rowcolor{white}\multirow{-6}{*}{Emotional wellbeing} & Felt calm and peaceful, how often in the past week\\
-
-\rowcolor{lightgray} & Free to decide how to live my life\\
-\rowcolor{lightgray} & Little chance to show how capable I am\\
-\rowcolor{lightgray} & Feel accomplishment from what I do\\
-\rowcolor{lightgray} & Interested in what you are doing\\
-\rowcolor{lightgray} & Absorbed in what you are doing\\
-\rowcolor{lightgray} & Enthusiastic about what you are doing\\
-\rowcolor{lightgray} & Feel what I do in life is valuable and worthwhile\\
-\rowcolor{lightgray} & Have a sense of direction\\
-\rowcolor{lightgray} & Always optimistic about my future\\
-\rowcolor{lightgray} & There are lots of things I feel I am good at\\
-\rowcolor{lightgray} & In general feel very positive about myself\\
-\rowcolor{lightgray} & At times feel as if I am a failure\\
-\rowcolor{lightgray} & When things go wrong in my life it takes a long time to get back to normal\\
-\rowcolor{lightgray}\multirow{-14}{*}{Functioning} & Deal with important problems\\
-
-\rowcolor{white} & Felt everything did as effort, how often in the past week\\
-\rowcolor{white} & Sleep was restless, how often in the past week\\
-\rowcolor{white} & Could not get going, how often in the past week\\
-\rowcolor{white}\multirow{-4}{*}{Vitality} & Had a lot of energy, how often in the past week\\
-
-\rowcolor{lightgray} &  Most people can be trusted\\
-\rowcolor{lightgray} &  People try to take advantage\\
-\rowcolor{lightgray} &  Most of the time people are helpful\\
-\rowcolor{lightgray} &  Feel people in local area help one another\\
-\rowcolor{lightgray}\multirow{-5}{*}{Community wellbeing}& Feel close to the people in local area\\
-
-\rowcolor{white} & How many with whom you can discuss intimate matters\\
-\rowcolor{white} & Feel appreciated by those you are close to\\
-\rowcolor{white} & Receive help and support\\
-\rowcolor{white}\multirow{-4}{*}{Supportive relationships} & Felt lonely, how often in the past week\\
-\hline
-\end{tabular}
-\egroup
-\end{table}
 \end{document}

### 2019-12-18 article/Old versions/JAS/PCADSC_article_JAS_v1.tex
old changes to meta files

--- 2019-12-18 article/Old versions/JAS/PCADSC_article_JAS_resub_jan2020.tex
@@
 
-\articletype{Application Note}% Specify the article type or omit as appropriate
+\articletype{Original Research Article}% Specify the article type or omit as appropriate
 
@@
 Computing empirical covariance matrices for the two subsets and comparing them entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets.
-%We refer to Appendix \ref{appendix.introPCA} for a brief introduction to PCA.
+We refer to Appendix \ref{appendix.introPCA} for a brief introduction to PCA.
 
@@
 In \textit{model 1} the observed variables $V_3$ and $V_4$ are caused by a common, unobserved variable $L_1$.
-Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models, in Appendix \ref{appendix.simData}.
+Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models, in Appendix \ref{appendix.introPCA}.
 In dataset $\mathbf{A}$, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures.
@@
 
+\section{A brief introduction to principal component analysis}
+\label{appendix.introPCA}
+
+In this appendix we introduce PCA. We suppress the subscripts and superscripts used in the article because we only consider a single dataset here, denoted by $\mathbf{X} \in \RR^{d \times n}$.
+
+For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost.
+More specifically, we wish to minimize the loss when looking at a projection of $\mathbf{X}$ onto a $q$-dimensional space, rather than the original $d$-dimensional one.
+Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
+
+\begin{equation*}
+\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_{i \cdot} - \bar{X} - z \rVert^2 =
+\min_{K_q} \sum_{i=1}^n \lVert x_{i \cdot} - \bar{X} - \text{proj}_{K_q}(x_{i \cdot} - \bar{X}) \rVert^2.
+\end{equation*}
+
+The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \cite{HastieEtAl2009}.
+More specifically, the rank-q-reconstruction error is attained when we choose
+
+\begin{equation*}
+\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\},
+\end{equation*}
+where $\eta_1, ..., \eta_q$ are the first $q$ eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$.
+In the PCA framework, we refer to the eigenvectors as \textit{loadings}.
+The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
+\begin{equation*}
+\text{trace}(S) = \sum_{j=1}^d \hat{V}(X_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top \mathbf{X}^\top) = \sum_{j=1}^d \lambda_j,
+\end{equation*}
+where $X_j \in \RR^n$ denotes the $j$th variable of $\mathbf{X}$ and $\hat{V}$ is the empirical variance function.
+This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe.
+And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are distinct, the representation hereby obtained is a valid object of inference.
+
+The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
+\begin{align*}
+\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_{i \cdot} - \bar{X}) \rVert^2, &
+\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_{i \cdot} - \bar{X}) \rVert^2,
+\end{align*}
+where the initial subspace is defined as $\hat{K}_0 = \{0\}$.
+It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.
+
+At this point a general remark about standardization is in place.
+PCA deconstructs the covariance matrix in components according to the most explained variance.
+This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will always be deemed highly influential, no matter the structure of the data.
+Therefore, the variables are typically standardized prior to performing PCA.
+The covariance matrix of the standardized variables is the same as the correlation matrix of the original variables, so post-standardization PCA simply corresponds to performing data structure comparisons of the correlation matrices rather than the covariance matrices.
+The standardization makes the variables comparable on the same scale, namely units of standard deviation.
+We have adopted this standardization policy.
+
 \section{Details about the simulated data}

### 2019-12-18 supplementary_code/future extensions/summary.R

@@ -1,0 +1,22 @@
+#'@import pander pander
+#'@export
+summary.CEInfo <- function(x) {
+  if (is.null(x)) {
+    cat("No cumulative eigenvalue (CE) information available")
+  } else {
+  y.min <- apply(x$y.sim, 1, quantile, probs=0.025)
+  y.max <- apply(x$y.sim, 1, quantile, probs=0.975)
+  outsideCI <- any(x$y.obs < x$y.min | x$y.obs > x$y.max)
+
+  restab <- matrix(c("Number of variables:",
+                     "Number of permutation iterations:",
+                     "Pointwise 95% CI exceeded:",
+                     "Kolmogorov-Smirnov test p-value:",
+                     "Cramer-von Mises test p-value:",
+                     x$d, x$B, outsideCI, x$KS.pvalue,
+                     x$CvM.pvalue), 5, 2)
+
+  cat("Cumulative eigenvalue (CE) information")
+  cat(pander(restab, justify = "lr"))
+  }
+}

### 2019-12-19 article/Old versions/JAS/PCADSC_article_JAS_resub_jan2020.tex

@@
 \newcommand{\acos}{\text{arccos}}
+\newcommand{\blue}[1]{\textcolor{blue}{#1}}
 
@@
 \section{Introduction}\label{sec:introduction}
+
+\let\thefootnote\relax\footnote{Note: An early draft of this article was included in the conference proceedings \textit{Symposium i Anvendt Statistik 2018.
+Copenhagen: The National Research Center for the Working Environment in Denmark \& University of Copenhagen (2018)}.}
 
@@
 We present three diagnostic plots that are designed to shine a light on different types and levels of data structure differences, namely the \textit{cumulative eigenvalue (CE) plot}, the \textit{angle plot}, and the \textit{chroma plot}.
-These plots are dsecribed in turn below.
+These plots are described in turn below.
 For the deepest understanding of the data structure differences in two datasets, we suggest using all of the three plots in the same order as they are presented.
@@
 Three new tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if two datasets can be combined for analysis were proposed and discussed.
-They all employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.
+They all employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create three intuitive visualizations of data structure differences.
+\blue{The first visualization, the cumulative eigenvalue plot, compares the eigenvalues from the two datasets.
+The second visualization, the angle plot, uses the interpretation of eigenvalues as explained variance to focus a comparison of loadings (eigenvectors).
+Finally, the third visulization, the chroma plot, allows for qualitative explanations of differences in loading patterns in selected principal components.}
 
@@
 \begin{table}[H]
-\renewcommand\thetable{C1}
+\renewcommand\thetable{B1}
 \centering
@@
 \end{table}
+
+\section{ESS data structure comparisons using pairwise complete cases}
+\label{appendix.pairwisecc}
+
+This appendix presents the results of the PCADSC procedures described in Section \ref{sec:dataexample} when allowing for pairwise complete cases to contribute to the computations.
+The numbers of pairwise complete cases in the Danish, Bulgarian and Swedish samples, respectively, are presented in Tables \ref{tableNPairDK}, \ref{tableNPairBG} and \ref{tableNPairSE} below.
+We use the following abbreviations for variable names:
+\begin{description}
+\item[Eval.:] Evaluative well-being
+\item[Emot.:] Emotional well-being
+\item[Func.:] Functioning
+\item[Vita.:] Vitality
+\item[Comm.:] Community well-being
+\item[Supp.:] Supportive relationships
+\end{description}
+Figures \ref{plotBG.cehair.pairwiseCC} and \ref{plotBG.pancake.pairwiseCC} show the resulting angle plot, CE plot and chroma plot for the comparison of the Danish and Bulgarian data.
+We see virtually no differences relative to the figures presented in Section \ref{sec:dataexample} that used complete cases only.
+Figures \ref{plotSE.cehair.pairwiseCC} and \ref{plotSE.pancake.pairwiseCC}  show the PCADSC plots for the comparison of Danish and Swedish data based on pairwise complete cases, and again, we find only very minor differences from the plots in Section \ref{sec:dataexample}.
+In conclusion, in this specific data example, the results are robust towards whether only complete cases or pairwise complete cases are used.
+
+\begin{table}[ht]
+\renewcommand\thetable{C1}
+\centering
+\caption{Number of observations in the Danish sample that are complete for each pairwise combination of variables.
+The total number of observations, including incomplete observations, is $n_\text{DK, total} = 1650$}
+\label{tableNPairDK}
+\begin{tabular}{rrrrrrr}
+  \hline
+ & Eval. & Emot. & Func. & Vita. & Comm. & Supp. \\ 
+  \hline
+Eval. & 1644 & 1608 & 1572 & 1629 & 1589 & 1613 \\ 
+  Emot. & 1608 & 1613 & 1553 & 1606 & 1564 & 1588 \\ 
+  Func. & 1572 & 1553 & 1573 & 1568 & 1535 & 1556 \\ 
+  Vita. & 1629 & 1606 & 1568 & 1635 & 1583 & 1608 \\ 
+  Comm. & 1589 & 1564 & 1535 & 1583 & 1592 & 1565 \\ 
+  Supp. & 1613 & 1588 & 1556 & 1608 & 1565 & 1617 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+\begin{table}[ht]
+\renewcommand\thetable{C2}
+\centering
+\caption{Number of observations in the Bulgarian sample that are complete for each pairwise combination of variables.
+The total number of observations, including incomplete observations, is $n_\text{BG, total} = 2260$}
+\label{tableNPairBG}
+\begin{tabular}{rrrrrrr}
+  \hline
+ & Eval. & Emot. & Func. & Vita. & Comm. & Supp. \\ 
+  \hline
+Eval. & 2223 & 2144 & 1975 & 2094 & 2124 & 2127 \\ 
+  Emot. & 2144 & 2169 & 1960 & 2079 & 2080 & 2094 \\ 
+  Func. & 1975 & 1960 & 1996 & 1921 & 1938 & 1941 \\ 
+  Vita. & 2094 & 2079 & 1921 & 2119 & 2034 & 2052 \\ 
+  Comm. & 2124 & 2080 & 1938 & 2034 & 2151 & 2073 \\ 
+  Supp. & 2127 & 2094 & 1941 & 2052 & 2073 & 2155 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+\begin{table}[ht]
+\renewcommand\thetable{C3}
+\centering
+\caption{Number of observations in the Swedish sample that are complete for each pairwise combination of variables.
+The total number of observations, including incomplete observations, is $n_\text{SE, total} = 1847$}
+\label{tableNPairSE}
+\begin{tabular}{rrrrrrr}
+  \hline
+ & Eval. & Emot. & Func. & Vita. & Comm. & Supp. \\ 
+  \hline
+Eval. & 1843 & 1832 & 1784 & 1837 & 1801 & 1831 \\ 
+  Emot. & 1832 & 1835 & 1782 & 1830 & 1795 & 1826 \\ 
+  Func. & 1784 & 1782 & 1788 & 1784 & 1752 & 1780 \\ 
+  Vita. & 1837 & 1830 & 1784 & 1841 & 1800 & 1830 \\ 
+  Comm. & 1801 & 1795 & 1752 & 1800 & 1804 & 1794 \\ 
+  Supp. & 1831 & 1826 & 1780 & 1830 & 1794 & 1835 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+\begin{figure}[H]
+\renewcommand\thefigure{C1}
+\center
+\includegraphics[scale=0.8]{Figure7_PCADSC_v3_partialcomplete.pdf}
+\caption{CE plot (top panel) and angle plot (bottom panel) comparing Danish and Bulgarian data using pairwise complete observations.}
+\label{plotBG.cehair.pairwiseCC}
+\end{figure}
+
+\begin{figure}[H]
+\renewcommand\thefigure{C2}
+\center
+\includegraphics[scale=0.8]{Figure8_PCADSC_v3_partialcomplete.pdf}
+\caption{Chroma plot for comparing Danish and Bulgarian data  data using pairwise complete observations.}
+\label{plotBG.pancake.pairwiseCC}
+\end{figure}
+
+\begin{figure}[H]
+\renewcommand\thefigure{C3}
+\center
+\includegraphics[scale=0.8]{Figure9_PCADSC_v3_partialcomplete.pdf}
+\caption{CE plot (top panel) and angle plot (bottom panel) comparing Danish and Swedish data using pairwise complete observations.}
+\label{plotSE.cehair.pairwiseCC}
+\end{figure}
+
+\begin{figure}[H]
+\renewcommand\thefigure{C4}
+\center
+\includegraphics[scale=0.8]{Figure10_PCADSC_v3_partialcomplete.pdf}
+\caption{Chroma plot for comparing Danish and Swedish  data using pairwise complete observations.}
+\label{plotSE.pancake.pairwiseCC}
+\end{figure}
 

### 2019-12-28 dataMaid_test.Rmd

---
dataMaid: yes
title: test
subtitle: "Autogenerated data summary from dataMaid"
date: 2018-05-18 12:11:11
output: pdf_document
documentclass: report
header-includes:
  - \renewcommand{\chaptername}{Part}
  - \newcommand{\fullline}{\noindent\makebox[\linewidth]{\rule{\textwidth}{0.4pt}}}
  - \newcommand{\bminione}{\begin{minipage}{0.75 \textwidth}}
  - \newcommand{\bminitwo}{\begin{minipage}{0.25 \textwidth}}
  - \newcommand{\emini}{\end{minipage}}
---
```{r , echo=FALSE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
library(ggplot2)
library(pander)
```
```{r visualFunctions, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
ggAggHist <- getFromNamespace("ggAggHist", "dataMaid")
ggAggBarplot <- getFromNamespace("ggAggBarplot", "dataMaid")
```
# Data report overview
The dataset examined has the following dimensions:

---------------------------------
Feature                    Result
------------------------ --------
Number of observations        500

Number of variables             1
---------------------------------

### Checks performed
The following variable checks were performed, depending on the data type of each variable:

----------------------------------------------------------------------------------------------------------------------------------
&nbsp;                                                 character    factor    labelled   numeric    integer    logical     Date   
----------------------------------------------------- ----------- ---------- ---------- ---------- ---------- --------- ----------
Identify miscoded missing values                       $\times$    $\times$   $\times$   $\times$   $\times$             $\times$

Identify prefixed and suffixed whitespace              $\times$    $\times$   $\times$

Identify levels with < 6 obs.                          $\times$    $\times$   $\times$

Identify case issues                                   $\times$    $\times$   $\times$

Identify misclassified numeric or integer variables    $\times$    $\times$   $\times$

Identify outliers (Turkish Boxplot style)                                                $\times$   $\times$

Identify outliers                                                                                                        $\times$ 
----------------------------------------------------------------------------------------------------------------------------------

Please note that all numerical values in the following have been rounded to 2 decimals.

# Summary table

----------------------------------------------------------------------------------
&nbsp;   Variable class     # unique values   Missing observations  Any problems? 
-------- ---------------- ----------------- ---------------------- ---------------
[x]      numeric                        415                 0.20 %    $\times$    
----------------------------------------------------------------------------------

# Variable list
## x

\bminione

---------------------------------------
Feature                          Result
------------------------- -------------
Variable type                   numeric

Number of missing obs.        1 (0.2 %)

Number of unique values             414

Median                            10.22

1st and 3rd quartiles       9.91; 10.47

Min. and max.                  -Inf; 14
---------------------------------------

\emini
\bminitwo
```{r Var-1-x, echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(factorV = structure(1:20, .Label = c("[3.17,3.71]", 
"(3.71,4.25]", "(4.25,4.79]", "(4.79,5.34]", "(5.34,5.88]", "(5.88,6.42]", 
"(6.42,6.96]", "(6.96,7.5]", "(7.5,8.04]", "(8.04,8.58]", "(8.58,9.12]", 
"(9.12,9.67]", "(9.67,10.2]", "(10.2,10.7]", "(10.7,11.3]", "(11.3,11.8]", 
"(11.8,12.4]", "(12.4,12.9]", "(12.9,13.5]", "(13.5,14]"), class = "factor"), 
    Freq = c(1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 4L, 58L, 
    184L, 190L, 35L, 16L, 7L, 2L, 0L, 1L), xmin = c(3.16992500144231, 
    3.71127890213383, 4.25263280282536, 4.79398670351688, 5.33534060420841, 
    5.87669450489993, 6.41804840559146, 6.95940230628298, 7.50075620697451, 
    8.04211010766603, 8.58346400835755, 9.12481790904908, 9.6661718097406, 
    10.2075257104321, 10.7488796111237, 11.2902335118152, 11.8315874125067, 
    12.3729413131982, 12.9142952138897, 13.4556491145813), xmax = c(3.71127890213383, 
    4.25263280282536, 4.79398670351688, 5.33534060420841, 5.87669450489993, 
    6.41804840559146, 6.95940230628298, 7.50075620697451, 8.04211010766603, 
    8.58346400835755, 9.12481790904908, 9.6661718097406, 10.2075257104321, 
    10.7488796111237, 11.2902335118152, 11.8315874125067, 12.3729413131982, 
    12.9142952138897, 13.4556491145813, 13.9970030152728), ymin = c(0, 
    0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), 
    ymax = c(1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 4L, 58L, 
    184L, 190L, 35L, 16L, 7L, 2L, 0L, 1L)), .Names = c("factorV", 
"Freq", "xmin", "xmax", "ymin", "ymax"), row.names = c(NA, -20L
), class = "data.frame"), vnam = "x")
```
\emini

- The following suspected missing value codes enter as regular values: \"-Inf\".

- Note that a check function found the following problematic values: \"-Inf\", \"3.17\", \"8.84\", \"8.94\", \"9.02\", \"11.35\", \"11.4\", \"11.41\", \"11.41\", \"11.47\" (21 additional values omitted).

\fullline

Report generation information:

 *  Created by Anne Helby Petersen (username: `zms499`).

 *  Report creation time: fr maj 18 2018 12:11:15

 *  Report Was run from directory: `P:/PCADSC/R`

 *  dataMaid v1.1.2 [Pkg: 2018-05-03 from CRAN (R 3.4.4)]

 *  R version 3.4.2 (2017-09-28).

 *  Platform: x86_64-w64-mingw32/x64 (64-bit)(Windows 7 x64 (build 7601) Service Pack 1).

 *  Function call: `makeDataReport(data = test, checks = setChecks(numeric = defaultNumericChecks(add = "identifyOutliersTBStyle", 
    remove = "identifyOutliers"), integer = defaultIntegerChecks(add = "identifyOutliersTBStyle", 
    remove = "identifyOutliers")))`

### 2019-12-28 article/Old versions/JAS/PCADSC_article_MPS_final.tex

--- 2018-06-23 article/Old versions/MPS/PCADSC_article.tex
@@
 \definecolor{lightgray}{HTML}{e6e6e6}
+
+\usepackage{caption}
+\captionsetup{font={stretch=2}}
 
@@
 The tools presented here are all based on comparing the PCA results across two different datasets that contain the same variables, $\mathbf{X}$ and $\mathbf{Y}$. Let 
-$$\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{(n_x + n_y) \times d}$$
+\begin{equation}
+\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{(n_x + n_y) \times d}
+\end{equation}
  be the combined dataset. For each of these three datasets, we complete the following steps (here described for $\mathbf{X}$ only):
@@
 \item Form the principal component analyses
-\begin{align*}
+\begin{align}
 S_x &= \frac{1}{n_x-1} \sum_{i=1}^{n_x} \tilde{x}_{i \cdot} \tilde{x}_{i \cdot}^\top = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top 
-\end{align*}
+\end{align}
 thereby obtaining loadings $\eta_{j}^x$ and eigenvalues $\lambda^x_j$ for $j=1,\dotsc,d$.
@@
 The standardization implies that the diagonal elements of $S_x$, $S_y$, and $S_z$ all equal 1, and thus also that
-\begin{equation*}
+\begin{equation}
 \sum_{j=1}^d \lambda_j^x = \sum_{j=1}^d \lambda_{j}^y = \sum_{j=1}^d \lambda_{j}^z =  d.
-\end{equation*}
+\end{equation}
 This identity will simplify some expressions below.
@@
 In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
-\begin{align*}
+\begin{align}
 &(0,0), \;
@@
 &\bigg( \sum_{j=1}^d \lambda_j^z, \sum_{j=1}^d \lambda_{j}^x - \sum_{j=1}^d \lambda_{j}^y \bigg)
-\end{align*}
+\end{align}
 Due to the standardization, the last point will always be equal to $(d,0)$.
@@
 In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
-\begin{align*}
+\begin{align}
 \text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{j}^x - \sum_{j=1}^k \lambda_{j}^y \bigg\rvert, &
 \text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k^z + \lambda_{k+1}^z}{2} {\bigg( \sum_{j=1}^k \lambda_{j}^x - \sum_{j=1}^k \lambda_{j}^y \bigg)}^2.
-\end{align*}
+\end{align}
 We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_x$ and $n_y$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics.
@@
 Then the empirical correlation matrix for the $\mathbf{X}$ dataset, $S_x$, has the following orthogonal decomposition in the coordinate system of the $\mathbf{Y}$ dataset
-\begin{equation*}
+\begin{equation}
 S_x = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top
@@
 {\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)}^\top,
-\end{equation*}
+\end{equation}
 and we have a similar decomposition of $S_y$ in the coordinate system of the $\mathbf{X}$ dataset.
@@
 The lengths $\mu_{jk}$ and $\nu_{jk}$, and the angle $\theta_{jk}$, are given by
-\begin{align*}
+\begin{align}
 \mu_{jk} &= \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert, &
@@
 \theta_{jk} &= \acos(\lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert).
-\end{align*}
+\end{align}
 For two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the empirical correlation function.
@@
 The width of the $j$th colored segment is given by
-\begin{align*}
+\begin{align}
 \omega_{ij}^x &= \frac{\lvert\eta_{ij}^x \rvert}{\sum_{k=1}^d \lvert\eta_{ik}^x\rvert}, &
 \omega_{ij}^y &= \frac{\lvert\eta_{ij}^y\rvert}{\sum_{k=1}^d \lvert\eta_{ik}^y\rvert},
-\end{align*}
+\end{align}
 where $\eta_{ij}^x$ and $\eta_{ij}^y$ denotes the $j$th entry of $\eta_{i}^x$ and $\eta_{i}^y$, respectively.
@@
 The bars are ordered according to the eigenvalues and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance contributions,
-\begin{align*}
+\begin{align}
 \tilde\Sigma^x_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^x}{\sum_{k=1}^d \lambda_{k}^x} = \frac{\sum_{j=1}^i \lambda_{j}^x}{d}, &
 \tilde\Sigma^y_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^y}{\sum_{k=1}^d \lambda_{k}^y} = \frac{\sum_{j=1}^i \lambda_{j}^y}{d}.
-\end{align*}
+\end{align}
 Especially when $d$ is large, we recommend plotting only a select set of interesting principal components (for instance identified by use of the angle plot).
@@
 Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being.
-These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing}, and \textit{Supportive relationships}.
+These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative well-being}, \textit{Emotional well-being}, \textit{Functioning}, \textit{Vitality}, \textit{Community well-being}, and \textit{Supportive relationships}.
 More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in Appendix S3.
@@
 All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations.
-Table~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
+Figure~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.
 
-\begin{table}[ht]
-\centering
-\caption{\textbf{Median values of each of the six dimensions of psychological well-being, stratified by country.} 25 and 75 percentiles are listed in parentheses.
+%\begin{figure}[ht]
+%\centering
+%\caption{\textbf{Median values of each of the six dimensions of psychological well-being, stratified by country.} 25 and %75 percentiles are listed in parentheses.
 The scales are constructed such that they all run from 0-10.}
-\label{tableDistr}
-
-\begin{tabular}{lccc}
-\hline
-  & Denmark & Bulgaria & Sweden \\
-\hline
-Evaluative wellbeing    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
-Emotional wellbeing     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
-Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
-Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
-Community wellbeing     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
-Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
-\hline
-\end{tabular} 
-\end{table}
+%\label{tableDistr}
+%
+%\begin{tabular}{lccc}
+%\hline
+%  & Denmark & Bulgaria & Sweden \\
+%\hline
+%Evaluative well-being    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
+%Emotional well-being     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
+%Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
+%Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
+%Community well-being     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
+%Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
+%\hline
+%\end{tabular} 
+%\end{figure}
 
@@
 \center
-\includegraphics[scale=0.8]{Figure2_v3.pdf}
+\includegraphics[scale=0.8]{Figure2_AnneHelbyPetersen_v3.pdf}
 \caption{CE plots for the simulated datasets.
@@
 \center
-\includegraphics[scale=0.8]{Figure3_v3.pdf}
+\includegraphics[scale=0.8]{Figure3_AnneHelbyPetersen_v3.pdf}
 \caption{Angle plots for the simulated datasets.
@@
 \center
-\includegraphics[scale=0.8]{Figure4_v3.pdf}
+\includegraphics[scale=0.8]{Figure4_AnneHelbyPetersen_v3.pdf}
 \caption{Chroma plots for the simulated datasets.
@@
 
+\newpage
+
+\begin{figure}[H]
+\centering
+\caption{Median values of each of the six dimensions of psychological well-being, stratified by country.
+25 and 75 percentiles are listed in parentheses.
+The scales are constructed such that they all run from 0-10.}
+\label{tableDistr}
+
+\begin{tabular}{lccc}
+\hline
+  & Denmark & Bulgaria & Sweden \\
+\hline
+Evaluative well-being    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
+Emotional well-being     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
+Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
+Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
+Community well-being     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
+Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
+\hline
+\end{tabular} 
+\end{figure}
+
 \begin{figure}[H]
 \center
-\includegraphics[scale=0.8]{Figure6_v3.pdf}
+\includegraphics[scale=0.8]{Figure7_AnneHelbyPetersen_v3.pdf}
 \caption{CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being.
@@
 \center
-\includegraphics[scale=0.8]{Figure7_v3.pdf}
+\includegraphics[scale=0.8]{Figure8_AnneHelbyPetersen_v3.pdf}
 \caption{Chroma plot for comparing Danish and Bulgarian data on psychological well-being.
@@
 \center
-\includegraphics[scale=0.8]{Figure8_v3.pdf}
+\includegraphics[scale=0.8]{Figure9_AnneHelbyPetersen_v3.pdf}
 \caption{A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being.
@@
 \center
-\includegraphics[scale=0.8]{Figure9_v3.pdf}
+\includegraphics[scale=0.8]{Figure10_AnneHelbyPetersen_v3.pdf}
 \caption{Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples.
@@
 
-\begin{equation*}
+\begin{equation}
 \hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\},
-\end{equation*}
+\end{equation}
 where $\eta_1, ..., \eta_q$ are the first $q$ eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$.
@@
 The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
-\begin{equation*}
+\begin{equation}
 \text{trace}(S) = \sum_{j=1}^d \hat{V}(X_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top \mathbf{X}^\top) = \sum_{j=1}^d \lambda_j,
-\end{equation*}
+\end{equation}
 where $X_j \in \RR^n$ denotes the $j$th variable of $\mathbf{X}$ and $\hat{V}$ is the empirical variance function.
@@
 The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
-\begin{align*}
+\begin{align}
 \eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_{i \cdot} - \bar{X}) \rVert^2, &
 \lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_{i \cdot} - \bar{X}) \rVert^2,
-\end{align*}
+\end{align}
 where the initial subspace is defined as $\hat{K}_0 = \{0\}$.
@@
 
-$$\Sigma_A = \begin{pmatrix}
+\begin{equation}
+\Sigma_A = \begin{pmatrix}
     1.0  &  0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
@@
  0.0 & 0.0 &  0.0 & 0.0 & 0.4 & 1.0
-\end{pmatrix}$$
+\end{pmatrix}
+\end{equation}
 and
-$$
+\begin{equation}
 \Sigma_B = \begin{pmatrix}
@@
 \end{pmatrix},
-$$
+\end{equation}
 respectively.
 
-\subsection*{S3 Table: ESS psychological wellbeing scale definitions}
-\begin{table}[h]
+\subsection*{S3 Figure: ESS psychological well-being scale definitions}
+\begin{figure}[h]
 \centering
-\caption{{\bf ESS psychological wellbeing scale definitions.} Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
+\caption*{Figure S3.1: ESS psychological well-being scale definitions.
+Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}.
 Before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
@@
 \rowcolor{lightgray} & How satisfied with life as a whole \\
-\rowcolor{lightgray}\multirow{-2}{*}{Evaluative wellbeing }& How happy are you\\
+\rowcolor{lightgray}\multirow{-2}{*}{Evaluative well-being }& How happy are you\\
 
@@
 \rowcolor{white} & Felt anxious, how often in the past week\\
-\rowcolor{white}\multirow{-6}{*}{Emotional wellbeing} & Felt calm and peaceful, how often in the past week\\
+\rowcolor{white}\multirow{-6}{*}{Emotional well-being} & Felt calm and peaceful, how often in the past week\\
 
@@
 \rowcolor{lightgray} &  Feel people in local area help one another\\
-\rowcolor{lightgray}\multirow{-5}{*}{Community wellbeing}& Feel close to the people in local area\\
+\rowcolor{lightgray}\multirow{-5}{*}{Community well-being}& Feel close to the people in local area\\
 
@@
 \egroup
-\end{table}
+\end{figure}
 \end{document}

### 2019-12-29 R/QQanglePlot.R
And again

@@ -65,5 +65,3 @@ QQanglePlot <- function(x) {
              label.padding = unit(1, "lines")) +
     theme_light()
 }
-
-# end
